### 2018-09-05 string.tex
add latex source file

\documentclass[prb,preprint]{revtex4-1} 
% The line above defines the type of LaTeX document.
% Note that AJP uses the same style as Phys. Rev. B (prb).

% The % character begins a comment, which continues to the end of the line.

\usepackage{amsmath}  % needed for \tfrac, \bmatrix, etc.
\usepackage{amsfonts} % needed for bold Greek, Fraktur, and blackboard bold
\usepackage{graphicx} % needed for figures

\begin{document}

% Be sure to use the \title, \author, \affiliation, and \abstract macros
% to format your title page.  Don't use lower-level macros to  manually
% adjust the fonts and centering.

\title{Light inextensible strings under tension in the Schwarzschild geometry}
% In a long title you can use \\ to force a line break at a certain location.

\author{Robin K. S. Hankin}
\email{hankin.robin@gmail.com} % optional
\altaffiliation[permanent address: ]{2-14 Wakefield Street, Auckland, New Zealand} % optional second address
% If there were a second author at the same address, we would put another 
% \author{} statement here.  Don't combine multiple authors in a single
% \author statement.
\affiliation{AUT University, 2-14 Wakefield Street, Auckland, New Zealand}
% Please provide a full mailing address here.

% See the REVTeX documentation for more examples of author and affiliation lists.

\date{\today}

\begin{abstract}
Here I analyze the behaviour of light inextensible strings under
tension in the vicinity of a black hole.  Proper length (given by the
spatial component of the Schwarzshild metric) is minimized by the
Euler-Lagrange equations.  Light inextensible taut strings may assume
a range of distinct configurations with an arbitrarily high winding
number around the central singularity, and here I give a range of
numerical solutions and theoretical observations.
\end{abstract}

% AJP requires an abstract for all regular article submissions.
% Abstracts are optional for submissions to the "Notes and Discussions" section.

\maketitle % title page is now complete

\section{Introduction}

In this short paper I consider the behaviour of a light inextensible
string in the exterior Schwarzschild metric.  A light string is not
subject to gravity, and being under tension means that its
configuration is an extremum of proper length.  Of course, no real
string is perfectly weightless or inextensible.  With tension~$T$,
gravitational acceleration $g$ and (proper) mass per (proper) unit
length~$\mu$, then ideal behaviour will be appropriate for
lengthscales $L\lesssim\frac{T}{\mu g}$.  Using Newtonian expressions
for the strength of gravity at the event horizon gives
$L\ll\frac{TGM}{\mu c^2}$.  For the string to behave itself near the
event horizon on lengthscales comparable with the Schwarzschild radius
would require $T\gtrsim\mu c^2$.  Exceeding this criterion would
violate the dominant energy condition (and in this context we note
that transverse waves would propagate at speed~$\sqrt{T/\mu}\sim c$).

Nevertheless, the idealization of a light inextensible string is an
interesting and informative limit: its analysis interprets the concept
``straight line'' and suggests a possible way of manipulating a black
hole.

The Schwarzschild metric

\begin{equation}\label{schwarzschild}
ds^2= -dt^2\left(1-2M/r\right) +\frac{dr^2}{1-2M/r} + r^2\left(d\theta^2 + \sin^2\theta d\phi^2\right)
\end{equation}

where units in which~$2GM=c=1$ are used, has been known for over a
century and makes a number of counter-intuitive predictions.  Here I
consider the spatial component

\begin{equation}\label{spatial_schwarzschild}
  ds^2= \frac{dr^2}{1-2M/r} + r^2d\phi^2
\end{equation}

confined to equatorial points~($\phi=\pi/2$); we
understand~$1<r<\infty$ throughout.  A string of constant~$r$ will,
trivially, have proper length~$2\pi r$.

\subsection*{Radial strings}

Considering first the radial component we may calculate the
proper length $L\left(r_1,r_2\right)$ of a string stretching
from~$r=r_1\geq 1$ to~$r_2\geq r_1$ at constant $\theta,\phi$:

\begin{equation}\label{radial_string_length}
  \int_{r=r_1}^{r_2}\frac{dr}{\sqrt{1-2M/r}}=
  \left.
  \sqrt{r(r-1)} +\frac{1}{2}\log\left(
  \frac{r+\sqrt{r-1}}{r-\sqrt{r-1}}\right)
  \right|_{r_1}^{r_2}
  \end{equation}

Observe that integration may be performed with a lower limit
of~$r_1=1$ with no difficulty, even though the radial component of the
Schwarzschild metric approaches infinity there.  The proper string
length from~$r=1$ to~$r=1+r_0$, say, is given by elementary calculus
and we can say:

\begin{equation}\label{proper_string_length_large_r}
  L\left(1,1+r_0\right)=
  r_0 + \frac{\log(r_0)}{2}
  + \log(2) -\frac{1}{2} 
  +o(1)\qquad\mbox{as $r_0\longrightarrow\infty$}
\end{equation}

%In[23]:=  p = Assuming[r0>1,Integrate[1/Sqrt[1-1/r],{r,1,1+r0}]] 
%
%                                            r0
%Out[23]= Sqrt[r0 (1 + r0)] + ArcTanh[Sqrt[------]]
%                                          1 + r0
%
%In[24]:= Limit[p -(r0 + Log[2] + 1/2 +Log[r0]/2) , {r0 -> Infinity}] 
%
%Out[24]= {0}

For small values of~$r_0$ we have

\begin{equation}\label{proper_string_length_small_r}
L\left(1,1+r_0\right) =   \sqrt{r_0}\left(2 - \frac{1}{6}r_0+ o\left(r_0^2\right)\right)\qquad\mbox{as $r_0\longrightarrow 0^+$}
\end{equation}

%var('r,r0')
%def g(r):  return sqrt(r*(r-1)) + log(    (r+sqrt(r-1))/(r-sqrt(r-1))  )/2
% (   (g(1+r0)-g(1))/sqrt(r0)    ).taylor(r0,0,2)

% 3/40*r0^2 - 1/6*r0 + 2

\subsection{Radial and angular components}

Taut strings with a metric corresponding
to~\ref{spatial_schwarzschild} may be visualized using the device of
{\em Flamm's paraboloid\,}~\cite{flamm1916}
(figure~\ref{flamm_nostring}), in which a 2D surface corresponding
to~$\theta=0$ is embedded into~$\mathbb{R}^3$ in such a way that
distances on the surface correspond to the spatial components of the
Schwarzschild metric.  Note the absence of points within the event
horizon.

Integrating equation~\ref{spatial_schwarzschild} along a path between
two points~$p_1=\left(r_1,\phi_1\right)$
and~$p_2=\left(r_2,\phi_2\right)$ gives the {\em proper} length of the
path, which is the length of a string joining~$p_1$ and~$p_2$.  A taut
but light string from~$p_1$ to~$p_2$ will adopt an extremal-length
path, as measured by equation~\ref{spatial_schwarzschild}.  Such paths
may be found by the calculus of variations; parametrizing a curve in
terms of~$r=r\left(\phi\right)$ gives us a path length of

\begin{equation}
%  \int_{p_1}^{p_2}\sqrt{\frac{\left(r'\right)^2}{1-2M/r} + r^2}d\phi=
  \int_{p_1}^{p_2}\left(\frac{\left(r'\right)^2}{1-2M/r} + r^2\right)^\frac{1}{2}d\phi=
  \int_{p_1}^{p_2}F\left(r,r'\right)d\phi.
\end{equation}

where dashes denote differentiation with respect to~$\phi$.  The
Euler-Lagrange equations for this system are

\begin{equation}
  \frac{d}{d\phi}\frac{\partial F}{\partial r'}-\frac{\partial F}{\partial r}=0.
\end{equation}

This gives, after simplification, an expression for the second
derivative of~$r$:

\begin{equation}\label{rdashdash}
  r''\left(\phi\right) =
  (r-1) + \frac{(4r-3)\left(r'\right)^2}{2r\left(r-1\right)},\qquad r>1
\end{equation}

This differential equation does not appear to have a simple analytical
solution (I present some numerical results below) but nevertheless we
may make several observations.  Equation~\ref{rdashdash} is a
second-order nonlinear ordinary differential equation and thus has two
constants of integration.  Consider a solution where there exists a
point~$\left(r,\phi\right), r>1$ at which~$r'(\phi)=0$.  By Rolle's
theorem, this point must be unique, as~$r''(\phi)>0$, and be the
closest approach of the string to the event horizon.  Further, because
the system is symmetrical under~$\phi\longrightarrow-\phi$, the string
must be symmetrical about this point; and we can see that~$r$ can take
arbitrarily large values as~$\phi\longrightarrow\pm\infty$: the string
has two ``free'' ends which continue to asymptotically flat space.

Solutions for which~$r'(\phi)>0$ whenever~$r>1$ correspond to strings
that have only one free end; the other is tangential to the event
horizon, except in the exceptional case of constant~$\phi$ when the
string is radial for~$r>1$.  The (signed) radius of curvature~$R$ is
readily evaluated:

\begin{equation}\label{roc}
  R = \frac{
    \left(r^2 + \left(r'\right)^2\right)^\frac{3}{2}
  }{
%    \left|
    r^2 + 2\left(r'\right)^2-rr''
%    \right|
  }
  =
  \frac{
    \left(r^2 + \left(r'\right)^2\right)^\frac{3}{2}
  }{
    r-\frac{\left(r'\right)^2}{2\left(r-1\right)}
  }\qquad\mbox{along a taut string}
\end{equation}

which would give~$R$ as a function of~$\left(r,r'\right)$.
Considering~$r'(\phi)=0$ shows that if a string is tangential at a
point with~$r_0>1$, then it has a radius of curvature of~$r_0^2$ at
that point.  Any circular path, including a circular orbit at~$r=r_0$,
has a radius of curvature of $r_0<r_0^2$, showing that circular orbits
cannot follow a taut string.  Also observe that null geodesics have a
radius of curvature of $2r_0^2/3 < r_0^2$, showing that light cannot
follow a taut string either.

% \subsection{Particle paths}
% 
% Free particle paths are given by
% 
% \begin{equation}
% \frac{d^2u}{d\phi^2} + u = \frac{2}{\ell^2} + \frac{3}{2}
% \end{equation}
% 
% (Rindler 11.45) where~$u=r^{-1}<1$ is the inverse radius and $\ell$ is a constant.
% This, in conjunction with equation~\ref{roc}, shows that tangential
% particle paths have a radius of curvature
% of~$\frac{2\ell^2}{1+3\ell^2/r^2}$.  This cannot equal the radius of
% curvature of a tangential string, which is~$r^2$ for any value
% of~$\ell$; thus no tangentially moving particle can follow a string
% path, no matter how fast it moves.

Equation~\ref{roc} also shows that~$R$ becomes infinite, and the
string has a point of inflection,
if~$\left(r'\right)^2=2r\left(r-1\right)$.

\begin{figure}[h!]
\centering
\includegraphics{flamm_nostring.pdf}
\caption{Flamm's paraboloid}
\label{flamm_nostring}
\end{figure}

\begin{figure}[h!]
\centering
\includegraphics{flamm_string.pdf}
\caption{Flamm's paraboloid with a string}
\label{flamm_withstring}
\end{figure}

\section{Results}
Light inextensible strings fall into two distinct classes: those with
two free ends, and those with one free end (the other ``end'' is
tangential to the event horizon).  Results for strings with two free
ends are presented first.

\subsection{Strings with two free ends}

The canonical result is shown in Figure~\ref{flamm_withstring}:
Flamm's paraboloid with a taut string.  Note that an actual 3D model
would require the string to be constrained above and below.

Figure~\ref{closest_approach} shows strings indexed by distance of
approach to the singularity, which occurs at $\phi=0$.  Strings with a
sufficiently close passage to the event horizon self-intersect;
numerical methods show that the critical value is~$\simeq
1.076109317$, with an uncertainty of 1 in the last digit: if a string
with two free ends passes closer than this, it forms a closed loop.

\begin{figure}[h!]
\centering
\includegraphics{closest_approach.pdf}
\caption{Strings arranged by closest approach to the event horizon}
\label{closest_approach}
\end{figure}

It is possible to parametrize the string in a number of different
ways.  Figure~\ref{angle_at_r_equals_2} shows the same physical
system, but the strings are rotated so that they pass through the
point~$\left(2,0\right)$; the strings are indexed by the
angle~$\alpha$ made with a radius at that point.

\begin{figure}[h!]
\centering
\includegraphics{angle_at_r_equals_2.pdf}
\caption{Strings \label{angle_at_r_equals_2} passing through $\left(2,0\right)$, two free ends}
\end{figure}

\subsection{Strings with one free end}

\begin{figure}[h!]
\centering
\includegraphics{one_free_end_r_equals_2.pdf}
\caption{Strings arranged by angle made at $r=2$, one free end}
\label{y}
\end{figure}

\begin{figure}[h!]
\centering
\includegraphics{one_free_end_fixed_EH_intersection.pdf}
\caption{Strings arranged by QFGASDSF}
\label{fixed_EH_intersection}
\end{figure}

\begin{figure}[h!]
\centering
\includegraphics{curvature_switch.pdf}
\caption{Strings arranged by QFGASDSF}
\label{curvature_switch}
\end{figure}

\section{Discussion}

The analysis above was restricted to the case~$r\geq 1$, where it is
possible to hold objects steady.  Of course, once within the event
horizon, even a light inextensible string cannot hold itself steady
(at least in Schwarzschild coordinates) and one would have to account
for the dynamics of thes system.

Although the concept of light inextensible string is not entirely
consistent---tugging one end of a string cannot produce an
instantaneous response at the other end---it is possible to speculate
about what might happen if~$r<1$.

We might consider two spaceships, or perhaps more appropriately
spiders, on identical freefall radial plunge trajectories separated by
a constant angle~$\theta$.  The spiders might indeed pass a light
inextensible string between themselves and reel it in so as to
maintain a tension (but not so tight as to perturb their radial fall);
what happens to the string?  The spiders will lose causal contact with
one another after a finite proper time but before encountering the
singularity, and it is not clear what happens to their string.

Considering Figure~\ref{closest_approach} in which strings under
tension are near the black hole.  One's classical intuition would
suggest that the black hole feels a force to the right, and should
start accelerating.  However, it is not clear by what mechanism such a
force would be transmitted; the situation is steady, so no
gravitational waves are produced; the string is inextensible, so
stores no energy.  Likewise, Figure~\ref{fixed_EH_intersection} would
suggest that angular momentum is transmitted to the black hole, but no
mechanism is apparent.

This work considered only the Schwarzschild metric and generalizing it
to the Kerr metric~\cite{asdfsadfasfd} might be interesting.

\appendix*   % Omit the * if there's more than one appendix.

\section{Uninteresting stuff}

Appendices are for material that is needed for completeness but
not sufficiently interesting to include in the main body of the paper.  Most
articles don't need any appendices, but feel free to use them when
appropriate.  This sample article needs an appendix only to illustrate how 
to create an appendix.

\section{Conclusions and further work}

\begin{acknowledgments}

We gratefully acknowledge 
\end{acknowledgments}

\begin{thebibliography}{99}
% The numeral (here 99) in curly braces is nominally the number of entries in
% the bibliography. It's supposed to affect the amount of space around the
% numerical labels, so only the number of digits should matter--and even that
% seems to make no discernible difference.

\bibitem{flamm1916} Flamm, L.: Beitr\"{a}ge zur Einsteinschen
  Gravitationstheorie.  Physikalische Zeitschrift XVII, 448–454 (1916)

\bibitem{noBIBTeX} Many \LaTeX\ users manage their bibliographic data with 
a tool called BIB\TeX.  Unfortunately, AJP cannot accept BIB\TeX\ files; all 
bibliographic references must be incorporated into the manuscript file
as shown here, at least when you send an editable file for production.

\bibitem{dyson} Freeman J. Dyson, ``Feynman's proof of the Maxwell equations,''
Am. J. Phys. \textbf{58} (3), 209--211.  
% The issue number (3) in this citation is optional, because AJP's pagination 
% is by volume.

\end{thebibliography}

% If your manuscript is conditionally accepted, the editors will ask you to
% submit your editable LaTeX source file.  Before doing so, you should move
% all tables and figure captions to the end, as shown below.  Tables come 
% first, followed by figure captions (with figure inclusions commented-out).
% Figures should be submitted as separate files, collected with the
% LaTeX file into a single .zip archive.

%\newpage   % Start a new page for tables

%\begin{table}[h!]
%\centering
%\caption{Elementary bosons}
%\begin{ruledtabular}
%\begin{tabular}{l c c c c p{5cm}}
%Name & Symbol & Mass (GeV/$c^2$) & Spin & Discovered & Interacts with \\
%\hline
%Photon & $\gamma$ & \ \ 0 & 1 & 1905 & Electrically charged particles \\
%Gluons & $g$ & \ \ 0 & 1 & 1978 & Strongly interacting particles (quarks and gluons) \\
%Weak charged bosons & $W^\pm$ & \ 82 & 1 & 1983 & Quarks, leptons, $W^\pm$, $Z^0$, $\gamma$ \\
%Weak neutral boson & $Z^0$ & \ 91 & 1 & 1983 & Quarks, leptons, $W^\pm$, $Z^0$ \\
%Higgs boson & $H$ & 126 & 0 & 2012 & Massive particles (according to theory) \\
%\end{tabular}
%\end{ruledtabular}
%\label{bosons}
%\end{table}

%\newpage   % Start a new page for figure captions

%\section*{Figure captions}

%\begin{figure}[h!]
%\centering
%\includegraphics[width=5in]{ThreeSunsets.jpg}   % This line stays commented-out
%\caption{Three overlaid sequences of photos of the setting sun, taken
%near the December solstice (left), September equinox (center), and
%June solstice (right), all from the same location at 41$^\circ$ north
%latitude. The time interval between images in each sequence is approximately
%four minutes.}
%\label{sunsets}
%\end{figure}

\end{document}

### 2018-09-05 closest_approach2.R
another R script but for very close approaches

@@ -1,0 +1,67 @@
+## lots of blue strings with varying closest approach distances.  The
+## emphasis here is on self-intersecting strings.
+
+## This script runs standalone but is designed to be called from
+## maker.R [cf Makefile], which creates the pdf file
+## "closest_approach2.pdf"
+
+source("usefulfuncs.R") # defines stringpoints()
+
+
+mask <- TRUE
+pars <- c(eel=1)  # dummy
+
+## plot setup:
+jj <- 6
+par(xpd=TRUE)
+plot(NULL,asp=1,xlim=c(-jj,jj),ylim=c(-jj,jj),type='l',axes=FALSE,,xlab='',ylab='')
+
+polargrid(rlab=6.5)
+## setup ends
+
+dist <- exp(seq(from=log(1.01), to=log(1.076109317),len=300))
+
+
+thetamax <- 0.1 + dist*0  # default value, (almost) always do-able
+
+cutoffmatrix <- matrix(c(
+    1.000, pi + 0.00  ,
+    1.011, pi + 1.20  ,
+    1.013, pi + 1.15  ,
+    1.015, pi + 1.05  ,
+    1.017, pi + 0.91  ,
+    1.020, pi + 0.83  ,
+    1.025, pi + 0.72  ,
+    1.030, pi + 0.57  ,
+    1.035, pi + 0.44  ,
+    1.040, pi + 0.40  ,
+    1.045, pi + 0.30  ,
+    1.050, pi + 0.20  ,
+    1.060, pi + 0.11  ,
+    1.070, pi + 0.05  ,
+    1.080, pi + 0.00
+    ),byrow=TRUE,ncol=2)
+f <- fun(cutoffmatrix[,1],cutoffmatrix[,2])
+
+
+## Thus f(1.03) = pi, f(1.12) = 2.5
+
+for(i in seq_along(dist)){
+  tseq <- seq(from=0,to=f(dist[i]),len=100)
+  xy <- stringpoints(y_start=dist[i],initial_string_angle = 0,theta=tseq)
+  points(xy,type='l',col=rainbow(length(dist))[i])
+  
+#  xy <- stringpoints(y_start=dist[i],initial_string_angle = 0, theta=-tseq)
+#  points(xy,type='l',col=rainbow(length(dist))[i])
+}
+
+## mask strings too far from the black hole:
+if(mask){
[…]

### 2018-09-05 closest_approach2.R
finer distinctions in cutoffmatrix

@@ -20,18 +20,26 @@ polargrid(rlab=6.5)
 ## setup ends
 
 dist <- exp(seq(from=log(1.01), to=log(1.076109317),len=300))
-
+dist <- exp(seq(from=log(1.005), to=log(1.076109317),len=3000))[1:200]
 
 thetamax <- 0.1 + dist*0  # default value, (almost) always do-able
 
 cutoffmatrix <- matrix(c(
     1.000, pi + 0.00  ,
+    1.004, pi + 1.70  ,
+    1.005, pi + 1.70  ,
+    1.006, pi + 1.69  ,
+    1.007, pi + 1.55  ,
+    1.008, pi + 1.45  ,
+    1.009, pi + 1.35  ,
+    1.010, pi + 1.27  ,
     1.011, pi + 1.20  ,
+    1.012, pi + 1.20  ,
     1.013, pi + 1.15  ,
     1.015, pi + 1.05  ,
     1.017, pi + 0.91  ,
     1.020, pi + 0.83  ,
-    1.025, pi + 0.72  ,
+    1.025, pi + 0.71  ,
     1.030, pi + 0.57  ,
     1.035, pi + 0.44  ,
     1.040, pi + 0.40  ,

### 2018-09-06 string.tex
tiny changes

@@
 Figure~\ref{closest_approach} shows strings indexed by distance of
-approach to the singularity, which occurs at $\phi=0$.  Strings with a
+approach to the singularity, which occurs at~$\phi=0$.  Strings with a
 sufficiently close passage to the event horizon self-intersect;
@@
 ways.  Figure~\ref{angle_at_r_equals_2} shows the same physical
-system, but the strings are rotated so that they pass through the
+system, but the strings are arranged so that they pass through the
 point~$\left(2,0\right)$; the strings are indexed by the
@@
 \includegraphics{angle_at_r_equals_2.pdf}
-\caption{Strings \label{angle_at_r_equals_2} passing through $\left(2,0\right)$, two free ends}
+\caption{Strings \label{angle_at_r_equals_2} passing through
+  $\left(2,0\right)$, two free ends}
 \end{figure}

### 2018-09-06 usefulfuncs.R
include a dseq() function, a distorted seq()

@@ -24,7 +24,7 @@ library("deSolve")   # needed for ode()
   ## pars <- c(eel=1) # [dummy]
   ## stringpoints(2,0,theta=seq(from=0,to=pi/2,len=100))
 
-  print(initial_string_angle)
+  print(c(y_start,initial_string_angle))
   yini <- c(y=y_start, ydash=y_start*tan(initial_string_angle))
 
   bh <- ode(yini,theta,taut_string,pars,rtol=1e-6)
@@ -93,3 +93,6 @@ rotmat <- function(theta){
     
 
 
+dseq <- function(from,to,length.out,power=1){
+  from + (to-from)*seq(from=0,to=1,length.out=length.out)^power
+}

### 2018-09-06 closest_approach2.R
refine plotting region

@@ -19,19 +19,52 @@ plot(NULL,asp=1,xlim=c(-jj,jj),ylim=c(-jj,jj),type='l',axes=FALSE,,xlab='',ylab=
 polargrid(rlab=6.5)
 ## setup ends
 
-dist <- exp(seq(from=log(1.01), to=log(1.076109317),len=300))
-dist <- exp(seq(from=log(1.005), to=log(1.076109317),len=3000))[1:200]
+
+dist <- exp(seq(from=log(1.0001), to=log(1.076109317),len=80))
+dist <- seq(from=(1.0001), to=1.076109317,len=80)
+dist <- dseq(from=(1.00012), to=1.076109317,len=100,power=2.3)[1:30]
 
 thetamax <- 0.1 + dist*0  # default value, (almost) always do-able
 
 cutoffmatrix <- matrix(c(
-    1.000, pi + 0.00  ,
-    1.004, pi + 1.70  ,
-    1.005, pi + 1.70  ,
-    1.006, pi + 1.69  ,
-    1.007, pi + 1.55  ,
-    1.008, pi + 1.45  ,
-    1.009, pi + 1.35  ,
+    1.0000, pi + 3.30  ,
+    1.00011, pi + 4.50  ,
+    1.00013, pi + 4.35  ,
+    1.00015, pi + 4.30  ,
+    1.00017, pi + 4.20  ,
+    1.00020, pi + 4.06  ,
+    1.00023, pi + 3.95  ,
+    1.00025, pi + 3.90 ,
+    1.00027, pi + 3.80 ,
+    1.00030, pi + 3.80  ,
+    1.00035, pi + 3.70 ,
+    1.00040, pi + 3.60 ,
+    1.00045, pi + 3.50 ,
+    1.0005, pi + 3.445 ,
+    1.0006, pi + 3.30  ,
+    1.0007, pi + 3.20  ,
+    1.0008, pi + 3.10  ,
+    1.0009, pi + 3.00  ,
+    1.0010, pi + 2.90  ,
+    1.0011, pi + 2.85  ,
+    1.0012, pi + 2.75  ,
+    1.0014, pi + 2.68  ,
+    1.0016, pi + 2.56  ,
+    1.0018, pi + 2.47  ,
+    1.0020, pi + 2.45  ,
+    1.0022, pi + 2.32  ,
+    1.0025, pi + 2.30  ,
+    1.0028, pi + 2.20  ,
[…]

### 2018-09-06 maker.R
plot closest_approach2.pdf

@@ -2,6 +2,10 @@ pdf(file="closest_approach.pdf")
 source("closest_approach.R")
 dev.off()
 
+pdf(file="closest_approach2.pdf")
+source("closest_approach2.R")
+dev.off()
+
 pdf(file="angle_at_r_equals_2.pdf")
 source("angle_at_r_equals_2.R")
 dev.off()

### 2018-09-06 closest_approach2.R
documentation

@@ -1,11 +1,13 @@
 ## lots of blue strings with varying closest approach distances.  The
-## emphasis here is on self-intersecting strings.
+## emphasis here is on self-intersecting strings with two free ends.
+## For a self-intersecting string, the closest approach to the black
+## hole must be less than about 1.076109317.
 
 ## This script runs standalone but is designed to be called from
 ## maker.R [cf Makefile], which creates the pdf file
 ## "closest_approach2.pdf"
 
-source("usefulfuncs.R") # defines stringpoints()
+source("usefulfuncs.R") # defines stringpoints() and dseq()
 
 
 mask <- TRUE
@@ -22,7 +24,7 @@ polargrid(rlab=6.5)
 
 dist <- exp(seq(from=log(1.0001), to=log(1.076109317),len=80))
 dist <- seq(from=(1.0001), to=1.076109317,len=80)
-dist <- dseq(from=(1.00012), to=1.076109317,len=100,power=2.3)[1:30]
+dist <- dseq(from=(1.00012), to=1.076109317,len=100,power=2.3)
 
 thetamax <- 0.1 + dist*0  # default value, (almost) always do-able
 

### 2018-09-06 string.tex
include self-intersecting strings

@@
 
-Figure~\ref{closest_approach} shows strings indexed by distance of
-approach to the singularity, which occurs at~$\phi=0$.  Strings with a
-sufficiently close passage to the event horizon self-intersect;
-numerical methods show that the critical value is~$\simeq
-1.076109317$, with an uncertainty of 1 in the last digit: if a string
-with two free ends passes closer than this, it forms a closed loop.
+Figure~\ref{closest_approach_non_self_intersecting} shows
+nonselfintersecting strings indexed by distance of approach to the
+singularity, which occurs at~$\phi=0$.  Strings with a sufficiently
+close passage to the event horizon self-intersect; numerical methods
+show that the critical value is~$\simeq 1.076109317$, with an
+uncertainty of 1 in the last digit: if a string with two free ends
+passes closer than this, it forms a closed loop.
 
@@
 \centering
-\includegraphics{closest_approach.pdf}
-\caption{Strings arranged by closest approach to the event horizon}
-\label{closest_approach}
+\includegraphics{closest_approach_nonselfintersecting.pdf}
+\caption{Strings arranged by closest approach to the event horizon:
+  non-self-intersecting strings}
+\label{closest_approach_non_self_intersecting}
+\end{figure}
+
+\begin{figure}[h!]
+\centering
+\includegraphics{closest_approach_selfintersecting.pdf}
+\caption{Strings arranged by closest approach to the event horizon:
+ self-intersecting strings}
+\label{closest_approach_self_intersecting}
 \end{figure}

### 2018-09-06 maker.R
automatically generate self-intersecting strings diagram

@@ -1,8 +1,8 @@
-pdf(file="closest_approach.pdf")
+pdf(file="closest_approach_nonselfintersecting.pdf")
 source("closest_approach.R")
 dev.off()
 
-pdf(file="closest_approach2.pdf")
+pdf(file="closest_approach_selfintersecting.pdf")
 source("closest_approach2.R")
 dev.off()
 

### 2018-09-06 closest_approach2.R
remove redundant dist setting

@@ -22,8 +22,6 @@ polargrid(rlab=6.5)
 ## setup ends
 
 
-dist <- exp(seq(from=log(1.0001), to=log(1.076109317),len=80))
-dist <- seq(from=(1.0001), to=1.076109317,len=80)
 dist <- dseq(from=(1.00012), to=1.076109317,len=100,power=2.3)
 
 thetamax <- 0.1 + dist*0  # default value, (almost) always do-able

### 2018-09-06 closest_approach.R
tidyup and use cutoffmatrix for ease of modification

@@ -19,17 +19,45 @@ plot(NULL,asp=1,xlim=c(-jj,jj),ylim=c(-jj,jj),type='l',axes=FALSE,,xlab='',ylab=
 polargrid(rlab=6.5)
 ## setup ends
 
-dist <- exp(seq(from=log(1.01),to=log(6.5),len=40))
+
+dist <- dseq(from=1.076109317, to=7.5, len=100, power=2)
 
 thetamax <- 0.1 + dist*0  # default value, (almost) always do-able
 
+cutoffmatrix <- matrix(c(
+ 1.060, pi  ,
+ 1.076, 3.1 ,
+ 1.077, 3.1 ,
+ 1.078, 3.1 ,
+ 1.079, 3.1 ,
+ 1.080, 3.0 ,
+ 1.090, 3.0 ,
+ 1.10, 2.90 ,
+ 1.11, 2.87 ,
+ 1.13, 2.79 ,
+ 1.15, 2.71 ,
+ 1.17, 2.61 ,
+ 1.19, 2.51 ,
+ 1.21, 2.45 ,
+ 1.25, 2.42 ,
+ 1.31, 2.37 ,
+ 1.32, 2.30 ,
+ 1.33, 2.30 ,
+ 1.35, 2.30 ,
+ 1.41, 2.20 ,
+ 1.51, 2.10 ,
+ 1.70, 2.00 ,
+ 1.80, 1.90 ,
+ 1.90, 1.90 ,
+ 2.00, 1.90 ,
+ 2.10, 1.90 ,
+ 2.20, 1.60 ,
+ 2.40, 1.60 ,
+ 5.00, 1.60 ,
+ 9.00, 1.60 ),
+ ncol=2,byrow=TRUE)
 
-cuts <- c(        1,   1.06,  1.07, 1.08,   1.11,    1.13,    1.15,     1.17,  1.19,    1.21,    1.31,   1.41,  1.51, 1.7, 2.01,  5.00)
-vals <- c(  NA,      pi,    pi,   3.1,   2.87,   2.79,    2.71,     2.61,   2.51,  2.4,      2.4,     2.2,   2.1,  2.0, 1.9,    1.6,     1.0)
-f <- fun(cuts,vals)
-
-
-## Thus f(1.03) = pi, f(1.12) = 2.5
+f <- fun(cutoffmatrix[,1],cutoffmatrix[,2])
 
 for(i in seq_along(dist)){
   tseq <- seq(from=0,to=f(dist[i]),len=100)

### 2018-09-06 angle_at_r_equals_2.R
put grid plot at the end so the strings do not obscure the angles

@@ -18,7 +18,7 @@ mask <- TRUE    # set to FALSE to see entire string
 jj <- 6.5
 par(pty='s')
 plot(NULL,asp=1,xlim=c(-jj,jj),ylim=c(-jj,jj),type='l',xlab='',ylab='',axes=FALSE)
-polargrid(rlab=6.5)
+
 
 
 dist_from_hole <- 2
@@ -117,5 +117,5 @@ if(mask){
 }
 
 
-
+polargrid(rlab=6.5)
 event_horizon()

### 2018-09-06 closest_approach.R

@@ -13,16 +13,10 @@ pars <- c(eel=1)  # dummy
 jj <- 6
 par(xpd=TRUE)
 plot(NULL,asp=1,xlim=c(-jj,jj),ylim=c(-jj,jj),type='l',axes=FALSE,,xlab='',ylab='')
-
-
-
-polargrid(rlab=6.5)
 ## setup ends
 
 
 dist <- dseq(from=1.076109317, to=7.5, len=100, power=2)
-
-thetamax <- 0.1 + dist*0  # default value, (almost) always do-able
 
 cutoffmatrix <- matrix(c(
  1.060, pi  ,
@@ -75,4 +69,6 @@ for (theta in seq(from=0,to=2*pi,len=100)){
   polygon(jjxy[,1],jjxy[,2],col='white',border=NA)
 }
 
+
+polargrid(rlab=6.5)
 event_horizon()

### 2018-09-06 closest_approach2.R

@@ -17,14 +17,11 @@ pars <- c(eel=1)  # dummy
 jj <- 6
 par(xpd=TRUE)
 plot(NULL,asp=1,xlim=c(-jj,jj),ylim=c(-jj,jj),type='l',axes=FALSE,,xlab='',ylab='')
-
-polargrid(rlab=6.5)
 ## setup ends
 
 
 dist <- dseq(from=(1.00012), to=1.076109317,len=100,power=2.3)
 
-thetamax <- 0.1 + dist*0  # default value, (almost) always do-able
 
 cutoffmatrix <- matrix(c(
     1.0000, pi + 3.30  ,
@@ -105,4 +102,6 @@ if(mask){
   }
 }
 
+
+polargrid(rlab=6.5)
 event_horizon()

### 2018-09-09 string.tex
rewording

@@
 passes closer than this, it forms a closed loop.
+Figure~\ref{closest_approach_non_self_intersecting} shows some
+non-self-intersecting strings and
+figure~\ref{closest_approach_self_intersecting} shows some strings
+that self-intersect.
 
@@
 \includegraphics{closest_approach_nonselfintersecting.pdf}
-\caption{Strings arranged by closest approach to the event horizon:
-  non-self-intersecting strings}
+\caption{Strings arranged by distance~$d$ of closest approach to the
+  event horizon, occurring at~$\phi=0$; non-self-intersecting strings}
 \label{closest_approach_non_self_intersecting}
@@
 
-\begin{acknowledgments}
-
-We gratefully acknowledge 
-\end{acknowledgments}
+Try the Kerr metric too.
 

### 2018-09-30 string_microarticle.tex
Add this in Results In Physics (microarticle) format

--- 2018-09-09 string.tex
@@
-\documentclass[prb,preprint]{revtex4-1} 
-% The line above defines the type of LaTeX document.
-% Note that AJP uses the same style as Phys. Rev. B (prb).
+%% This is file `rinp-template.tex',
+%% 
+%% Copyright 2016 Elsevier Ltd
+%% 
+%% This file is part of the 'Elsarticle Bundle'.
+%% ---------------------------------------------
+%% 
+%% It may be distributed under the conditions of the LaTeX Project Public
+%% License, either version 1.2 of this license or (at your option) any
+%% later version.  The latest version of this license is in
+%%    http://www.latex-project.org/lppl.txt
+%% and version 1.2 or later is part of all distributions of LaTeX
+%% version 1999/12/01 or later.
+%% 
+%% The list of all files belonging to the 'Elsarticle Bundle' is
+%% given in the file `manifest.txt'.
+%% 
+%% Template article for Elsevier's document class `elsarticle'
+%% with harvard style bibliographic references
+%%
+%% $Id: $
+%%
+%% Use the option review to obtain double line spacing
+%\documentclass[times,review,preprint,authoryear]{elsarticle}
 
-% The % character begins a comment, which continues to the end of the line.
+%% Use the options `twocolumn,final' to obtain the final layout
+%% Use longtitle option to break abstract to multiple pages if overfull.
+%% For Review pdf (With double line spacing)
+%\documentclass[times,twocolumn,review]{elsarticle}
+%% For abstracts longer than one page.
+%\documentclass[times,twocolumn,review,longtitle]{elsarticle}
+%% For Review pdf without preprint line
+%\documentclass[times,twocolumn,review,nopreprintline]{elsarticle}
+%% Final pdf
+\documentclass[times,twocolumn,final]{elsarticle}
+%%
+%\documentclass[times,twocolumn,final,longtitle]{elsarticle}
+%%
 
-\usepackage{amsmath}  % needed for \tfrac, \bmatrix, etc.
-\usepackage{amsfonts} % needed for bold Greek, Fraktur, and blackboard bold
-\usepackage{graphicx} % needed for figures
+%% Stylefile to load RINP template
+\usepackage{rinp}
+\usepackage{framed,multirow}
+
+%% The amssymb package provides various useful mathematical symbols
+\usepackage{amssymb}
+\usepackage{latexsym}
+
+% Following three lines are needed for this document.
+% If you are not loading colors or url, then these are
+% not required.
+\usepackage{url}
+\usepackage{xcolor}
+\definecolor{newcolor}{rgb}{.8,.349,.1}
+
+\journal{Results in Physics}
 
@@
 
-% Be sure to use the \title, \author, \affiliation, and \abstract macros
-% to format your title page.  Don't use lower-level macros to  manually
-% adjust the fonts and centering.
+\verso{Robin Hankin}
+
+\begin{frontmatter}
 
 \title{Light inextensible strings under tension in the Schwarzschild geometry}
-% In a long title you can use \\ to force a line break at a certain location.
 
-\author{Robin K. S. Hankin}
-\email{hankin.robin@gmail.com} % optional
-\altaffiliation[permanent address: ]{2-14 Wakefield Street, Auckland, New Zealand} % optional second address
-% If there were a second author at the same address, we would put another 
-% \author{} statement here.  Don't combine multiple authors in a single
-% \author statement.
-\affiliation{AUT University, 2-14 Wakefield Street, Auckland, New Zealand}
-% Please provide a full mailing address here.
+\author[1]{Robin \snm{Hankin}\corref{cor1}}
+\cortext[cor1]{Corresponding author: 
+  Tel.: +64-09-921-9999;  
+  fax: +64-09-921-9998;}
 
-% See the REVTeX documentation for more examples of author and affiliation lists.
+\address[1]{Auckland University of Technology, 2-14 Wakefield Street,
+  Auckland 1142, New Zealand}
 
-\date{\today}
+\received{1 May 2013}
+\finalform{10 May 2013}
+\accepted{13 May 2013}
+\availableonline{15 May 2013}
+\communicated{R. Hankin}
 
 \begin{abstract}
+%%%
 Here I analyze the behaviour of light inextensible strings under
@@
 numerical solutions and theoretical observations.
+%%%%
 \end{abstract}
 
-% AJP requires an abstract for all regular article submissions.
-% Abstracts are optional for submissions to the "Notes and Discussions" section.
+\begin{keyword}
+%% MSC codes here, in the form: \MSC code \sep code
+%% or \MSC[2008] code \sep code (2000 is the default)
+\MSC 83C57\sep 35Q75
+%% Keywords
+\KWD Black Holes\sep Light inextensible string
+\end{keyword}
 
-\maketitle % title page is now complete
+\end{frontmatter}
 
+%\linenumbers
+
+%% main text
 \section{Introduction}
-
+\label{intro}
 In this short paper I consider the behaviour of a light inextensible
@@
 ``straight line'' and suggests a possible way of manipulating a black
-hole.
-
-The Schwarzschild metric
+hole.  The Schwarzschild metric
 
 \begin{equation}\label{schwarzschild}
-ds^2= -dt^2\left(1-2M/r\right) +\frac{dr^2}{1-2M/r} + r^2\left(d\theta^2 + \sin^2\theta d\phi^2\right)
+ds^2= -dt^2\left(1-1/r\right) +\frac{dr^2}{1-1/r} + r^2\left(d\theta^2 + \sin^2\theta d\phi^2\right)
 \end{equation}
@@
 century and makes a number of counter-intuitive predictions.  Here I
-consider the spatial component
-
-\begin{equation}\label{spatial_schwarzschild}
-  ds^2= \frac{dr^2}{1-2M/r} + r^2d\phi^2
-\end{equation}
-
-confined to equatorial points~($\phi=\pi/2$); we
+consider the spatial component~$ds^2= \frac{dr^2}{1-1/r} +
+r^2d\phi^2$, confined to equatorial points~($\phi=\pi/2$); we
 understand~$1<r<\infty$ throughout.  A string of constant~$r$ will,
 trivially, have proper length~$2\pi r$.
-
-\subsection*{Radial strings}
 
@@
 of~$r_1=1$ with no difficulty, even though the radial component of the
-Schwarzschild metric approaches infinity there.  The proper string
-length from~$r=1$ to~$r=1+r_0$, say, is given by elementary calculus
-and we can say:
-
-\begin{equation}\label{proper_string_length_large_r}
-  L\left(1,1+r_0\right)=
-  r_0 + \frac{\log(r_0)}{2}
-  + \log(2) -\frac{1}{2} 
-  +o(1)\qquad\mbox{as $r_0\longrightarrow\infty$}
-\end{equation}
-
-%In[23]:=  p = Assuming[r0>1,Integrate[1/Sqrt[1-1/r],{r,1,1+r0}]] 
-%
-%                                            r0
-%Out[23]= Sqrt[r0 (1 + r0)] + ArcTanh[Sqrt[------]]
-%                                          1 + r0
-%
-%In[24]:= Limit[p -(r0 + Log[2] + 1/2 +Log[r0]/2) , {r0 -> Infinity}] 
-%
-%Out[24]= {0}
-
-For small values of~$r_0$ we have
-
-\begin{equation}\label{proper_string_length_small_r}
-L\left(1,1+r_0\right) =   \sqrt{r_0}\left(2 - \frac{1}{6}r_0+ o\left(r_0^2\right)\right)\qquad\mbox{as $r_0\longrightarrow 0^+$}
-\end{equation}
-
-%var('r,r0')
-%def g(r):  return sqrt(r*(r-1)) + log(    (r+sqrt(r-1))/(r-sqrt(r-1))  )/2
-% (   (g(1+r0)-g(1))/sqrt(r0)    ).taylor(r0,0,2)
-
-% 3/40*r0^2 - 1/6*r0 + 2
+Schwarzschild metric approaches infinity there.
 
@@
 
-Taut strings with a metric corresponding
-to~\ref{spatial_schwarzschild} may be visualized using the device of
-{\em Flamm's paraboloid\,}~\cite{flamm1916}
-(figure~\ref{flamm_nostring}), in which a 2D surface corresponding
-to~$\theta=0$ is embedded into~$\mathbb{R}^3$ in such a way that
-distances on the surface correspond to the spatial components of the
-Schwarzschild metric.  Note the absence of points within the event
-horizon.
-
-Integrating equation~\ref{spatial_schwarzschild} along a path between
-two points~$p_1=\left(r_1,\phi_1\right)$
-and~$p_2=\left(r_2,\phi_2\right)$ gives the {\em proper} length of the
-path, which is the length of a string joining~$p_1$ and~$p_2$.  A taut
-but light string from~$p_1$ to~$p_2$ will adopt an extremal-length
-path, as measured by equation~\ref{spatial_schwarzschild}.  Such paths
-may be found by the calculus of variations; parametrizing a curve in
-terms of~$r=r\left(\phi\right)$ gives us a path length of
+Integrating along a path between two
+points~$p_1=\left(r_1,\phi_1\right)$ and~$p_2=\left(r_2,\phi_2\right)$
+gives the {\em proper} length of the path, which is the length of a
+string joining~$p_1$ and~$p_2$.  A taut but light string from~$p_1$
+to~$p_2$ will adopt an extremal-length path.  Such paths may be found
+by the calculus of variations; parametrizing a curve in terms
+of~$r=r\left(\phi\right)$ gives us a path length of
 
@@
 where dashes denote differentiation with respect to~$\phi$.  The
-Euler-Lagrange equations for this system are
-
-\begin{equation}
-  \frac{d}{d\phi}\frac{\partial F}{\partial r'}-\frac{\partial F}{\partial r}=0.
-\end{equation}
-
-This gives, after simplification, an expression for the second
-derivative of~$r$:
+Euler-Lagrange equations for this system,
+$\frac{d}{d\phi}\frac{\partial F}{\partial r'}-\frac{\partial
+  F}{\partial r}=0$ give, after simplification, an expression for the
+second derivative of~$r$:
 
@@
 This differential equation does not appear to have a simple analytical
-solution (I present some numerical results below) but nevertheless we
-may make several observations.  Equation~\ref{rdashdash} is a
-second-order nonlinear ordinary differential equation and thus has two
-constants of integration.  Consider a solution where there exists a
-point~$\left(r,\phi\right), r>1$ at which~$r'(\phi)=0$.  By Rolle's
-theorem, this point must be unique, as~$r''(\phi)>0$, and be the
-closest approach of the string to the event horizon.  Further, because
-the system is symmetrical under~$\phi\longrightarrow-\phi$, the string
-must be symmetrical about this point; and we can see that~$r$ can take
-arbitrarily large values as~$\phi\longrightarrow\pm\infty$: the string
-has two ``free'' ends which continue to asymptotically flat space.
-
-Solutions for which~$r'(\phi)>0$ whenever~$r>1$ correspond to strings
-that have only one free end; the other is tangential to the event
-horizon, except in the exceptional case of constant~$\phi$ when the
-string is radial for~$r>1$.  The (signed) radius of curvature~$R$ is
-readily evaluated:
-
-\begin{equation}\label{roc}
-  R = \frac{
-    \left(r^2 + \left(r'\right)^2\right)^\frac{3}{2}
-  }{
-%    \left|
-    r^2 + 2\left(r'\right)^2-rr''
-%    \right|
-  }
-  =
-  \frac{
-    \left(r^2 + \left(r'\right)^2\right)^\frac{3}{2}
-  }{
-    r-\frac{\left(r'\right)^2}{2\left(r-1\right)}
-  }\qquad\mbox{along a taut string}
-\end{equation}
-
-which would give~$R$ as a function of~$\left(r,r'\right)$.
-Considering~$r'(\phi)=0$ shows that if a string is tangential at a
-point with~$r_0>1$, then it has a radius of curvature of~$r_0^2$ at
-that point.  Any circular path, including a circular orbit at~$r=r_0$,
-has a radius of curvature of $r_0<r_0^2$, showing that circular orbits
-cannot follow a taut string.  Also observe that null geodesics have a
-radius of curvature of $2r_0^2/3 < r_0^2$, showing that light cannot
-follow a taut string either.
-
-% \subsection{Particle paths}
-% 
-% Free particle paths are given by
-% 
-% \begin{equation}
-% \frac{d^2u}{d\phi^2} + u = \frac{2}{\ell^2} + \frac{3}{2}
-% \end{equation}
-% 
-% (Rindler 11.45) where~$u=r^{-1}<1$ is the inverse radius and $\ell$ is a constant.
-% This, in conjunction with equation~\ref{roc}, shows that tangential
-% particle paths have a radius of curvature
-% of~$\frac{2\ell^2}{1+3\ell^2/r^2}$.  This cannot equal the radius of
-% curvature of a tangential string, which is~$r^2$ for any value
-% of~$\ell$; thus no tangentially moving particle can follow a string
-% path, no matter how fast it moves.
-
-Equation~\ref{roc} also shows that~$R$ becomes infinite, and the
-string has a point of inflection,
-if~$\left(r'\right)^2=2r\left(r-1\right)$.
-
-\begin{figure}[h!]
-\centering
-\includegraphics{flamm_nostring.pdf}
-\caption{Flamm's paraboloid}
-\label{flamm_nostring}
-\end{figure}
-
-\begin{figure}[h!]
-\centering
-\includegraphics{flamm_string.pdf}
-\caption{Flamm's paraboloid with a string}
-\label{flamm_withstring}
-\end{figure}
+solution; I present some numerical results below.  Solutions for
+which~$r'(\phi)>0$ whenever~$r>1$ correspond to strings that have only
+one free end; the other is tangential to the event horizon, except in
+the exceptional case of constant~$\phi$ when the string is radial
+for~$r>1$.  It is straightforward to show that the radius of curvature
+of a taut string becomes infinite, and the string has a point of
+inflection, if~$\left(r'\right)^2=2r\left(r-1\right)$.
 
@@
 two free ends, and those with one free end (the other ``end'' is
-tangential to the event horizon).  Results for strings with two free
-ends are presented first.
-
-\subsection{Strings with two free ends}
-
-The canonical result is shown in Figure~\ref{flamm_withstring}:
-Flamm's paraboloid with a taut string.  Note that an actual 3D model
-would require the string to be constrained above and below.
+tangential to the event horizon).
 
@@
 \centering
-\includegraphics{closest_approach_nonselfintersecting.pdf}
-\caption{Strings arranged by distance~$d$ of closest approach to the
-  event horizon, occurring at~$\phi=0$; non-self-intersecting strings}
+\includegraphics[width=70mm]{closest_approach_nonselfintersecting.pdf}
+\caption{Non-self intersecting strings arranged by distance~$d$ of
+  closest approach to the event horizon, occurring at~$\phi=0$}
 \label{closest_approach_non_self_intersecting}
@@
 \centering
-\includegraphics{closest_approach_selfintersecting.pdf}
+\includegraphics[width=70mm]{closest_approach_selfintersecting.pdf}
 \caption{Strings arranged by closest approach to the event horizon:
- self-intersecting strings}
+ self-intersecting strings shown for $\phi\geq 0$}
 \label{closest_approach_self_intersecting}
@@
 
-It is possible to parametrize the string in a number of different
-ways.  Figure~\ref{angle_at_r_equals_2} shows the same physical
-system, but the strings are arranged so that they pass through the
-point~$\left(2,0\right)$; the strings are indexed by the
-angle~$\alpha$ made with a radius at that point.
-
-\begin{figure}[h!]
-\centering
-\includegraphics{angle_at_r_equals_2.pdf}
-\caption{Strings \label{angle_at_r_equals_2} passing through
-  $\left(2,0\right)$, two free ends}
-\end{figure}
+Figure~\ref{one_free_end_r_equals_2} shows some strings that pass
+through~$\left(2,0\right)$ and tangent to the event horizon.
 
@@
 \centering
-\includegraphics{one_free_end_r_equals_2.pdf}
+\includegraphics[width=70mm]{one_free_end_r_equals_2.pdf}
 \caption{Strings arranged by angle made at $r=2$, one free end}
-\label{y}
-\end{figure}
-
-\begin{figure}[h!]
-\centering
-\includegraphics{one_free_end_fixed_EH_intersection.pdf}
-\caption{Strings arranged by QFGASDSF}
-\label{fixed_EH_intersection}
-\end{figure}
-
-\begin{figure}[h!]
-\centering
-\includegraphics{curvature_switch.pdf}
-\caption{Strings arranged by QFGASDSF}
-\label{curvature_switch}
+\label{one_free_end}
 \end{figure}
@@
 
-Although the concept of light inextensible string is not entirely
-consistent---tugging one end of a string cannot produce an
-instantaneous response at the other end---it is possible to speculate
-about what might happen if~$r<1$.
-
-We might consider two spaceships, or perhaps more appropriately
-spiders, on identical freefall radial plunge trajectories separated by
-a constant angle~$\theta$.  The spiders might indeed pass a light
-inextensible string between themselves and reel it in so as to
-maintain a tension (but not so tight as to perturb their radial fall);
-what happens to the string?  The spiders will lose causal contact with
-one another after a finite proper time but before encountering the
-singularity, and it is not clear what happens to their string.
-
-Considering Figure~\ref{closest_approach} in which strings under
-tension are near the black hole.  One's classical intuition would
-suggest that the black hole feels a force to the right, and should
-start accelerating.  However, it is not clear by what mechanism such a
-force would be transmitted; the situation is steady, so no
-gravitational waves are produced; the string is inextensible, so
-stores no energy.  Likewise, Figure~\ref{fixed_EH_intersection} would
-suggest that angular momentum is transmitted to the black hole, but no
-mechanism is apparent.
+In figure~\ref{closest_approach_non_self_intersecting}, one's
+classical intuition would suggest that the black hole feels a force to
+the right, and should start accelerating.  However, it is not clear by
+what mechanism such a force would be transmitted; the situation is
+steady, so no gravitational waves are produced; the string is
+inextensible, so stores no energy.  Likewise,
+Figure~\ref{one_free_end} would suggest that angular momentum is
+transmitted to the black hole, but no mechanism is apparent.
 
@@
 
-\appendix*   % Omit the * if there's more than one appendix.
-
-\section{Uninteresting stuff}
-
-Appendices are for material that is needed for completeness but
-not sufficiently interesting to include in the main body of the paper.  Most
-articles don't need any appendices, but feel free to use them when
-appropriate.  This sample article needs an appendix only to illustrate how 
-to create an appendix.
-
-\section{Conclusions and further work}
-
-Try the Kerr metric too.
-
-\begin{thebibliography}{99}
-% The numeral (here 99) in curly braces is nominally the number of entries in
-% the bibliography. It's supposed to affect the amount of space around the
-% numerical labels, so only the number of digits should matter--and even that
-% seems to make no discernible difference.
-
-\bibitem{flamm1916} Flamm, L.: Beitr\"{a}ge zur Einsteinschen
-  Gravitationstheorie.  Physikalische Zeitschrift XVII, 448–454 (1916)
-
-\bibitem{noBIBTeX} Many \LaTeX\ users manage their bibliographic data with 
-a tool called BIB\TeX.  Unfortunately, AJP cannot accept BIB\TeX\ files; all 
-bibliographic references must be incorporated into the manuscript file
-as shown here, at least when you send an editable file for production.
-
-\bibitem{dyson} Freeman J. Dyson, ``Feynman's proof of the Maxwell equations,''
-Am. J. Phys. \textbf{58} (3), 209--211.  
-% The issue number (3) in this citation is optional, because AJP's pagination 
-% is by volume.
-
-\end{thebibliography}
-
-% If your manuscript is conditionally accepted, the editors will ask you to
-% submit your editable LaTeX source file.  Before doing so, you should move
-% all tables and figure captions to the end, as shown below.  Tables come 
-% first, followed by figure captions (with figure inclusions commented-out).
-% Figures should be submitted as separate files, collected with the
-% LaTeX file into a single .zip archive.
-
-%\newpage   % Start a new page for tables
-
-%\begin{table}[h!]
-%\centering
-%\caption{Elementary bosons}
-%\begin{ruledtabular}
-%\begin{tabular}{l c c c c p{5cm}}
-%Name & Symbol & Mass (GeV/$c^2$) & Spin & Discovered & Interacts with \\
-%\hline
-%Photon & $\gamma$ & \ \ 0 & 1 & 1905 & Electrically charged particles \\
-%Gluons & $g$ & \ \ 0 & 1 & 1978 & Strongly interacting particles (quarks and gluons) \\
-%Weak charged bosons & $W^\pm$ & \ 82 & 1 & 1983 & Quarks, leptons, $W^\pm$, $Z^0$, $\gamma$ \\
-%Weak neutral boson & $Z^0$ & \ 91 & 1 & 1983 & Quarks, leptons, $W^\pm$, $Z^0$ \\
-%Higgs boson & $H$ & 126 & 0 & 2012 & Massive particles (according to theory) \\
-%\end{tabular}
-%\end{ruledtabular}
-%\label{bosons}
-%\end{table}
-
-%\newpage   % Start a new page for figure captions
-
-%\section*{Figure captions}
-
-%\begin{figure}[h!]
-%\centering
-%\includegraphics[width=5in]{ThreeSunsets.jpg}   % This line stays commented-out
-%\caption{Three overlaid sequences of photos of the setting sun, taken
-%near the December solstice (left), September equinox (center), and
-%June solstice (right), all from the same location at 41$^\circ$ north
-%latitude. The time interval between images in each sequence is approximately
-%four minutes.}
-%\label{sunsets}
-%\end{figure}
+%%Vancouver style references.
+\bibliographystyle{model3-num-names}
+\bibliography{refs}
 
 \end{document}
+
+%%

### 2018-10-01 string_microarticle.tex
tidyup and terseification

@@
 for the strength of gravity at the event horizon gives
-$L\ll\frac{TGM}{\mu c^2}$.  For the string to behave itself near the
-event horizon on lengthscales comparable with the Schwarzschild radius
-would require $T\gtrsim\mu c^2$.  Exceeding this criterion would
-violate the dominant energy condition (and in this context we note
-that transverse waves would propagate at speed~$\sqrt{T/\mu}\sim c$).
-
-Nevertheless, the idealization of a light inextensible string is an
-interesting and informative limit: its analysis interprets the concept
-``straight line'' and suggests a possible way of manipulating a black
-hole.  The Schwarzschild metric
+$L\ll\frac{TGM}{\mu c^2}$.  For the string to behave as though light
+on lengthscales comparable with the Schwarzschild radius would require
+$T\gtrsim\mu c^2$.  Exceeding this criterion would violate the
+dominant energy condition (and in this context we note that transverse
+waves would propagate at speed~$\sqrt{T/\mu}\sim c$).  Nevertheless,
+the idealization of a light inextensible string is an interesting and
+informative limit: its analysis interprets the concept ``straight
+line'' and suggests a possible way of manipulating a black hole.  The
+Schwarzschild metric
 
@@
 where units in which~$2GM=c=1$ are used, has been known for over a
-century and makes a number of counter-intuitive predictions.  Here I
-consider the spatial component~$ds^2= \frac{dr^2}{1-1/r} +
-r^2d\phi^2$, confined to equatorial points~($\phi=\pi/2$); we
-understand~$1<r<\infty$ throughout.  A string of constant~$r$ will,
-trivially, have proper length~$2\pi r$.
+century~\cite{schwarzschild1916} and makes a number of
+counter-intuitive predictions.  Here I consider the spatial
+component~$ds^2= \frac{dr^2}{1-1/r} + r^2d\phi^2$, confined to
+equatorial points~($\phi=\pi/2$); we understand~$1<r<\infty$
+throughout.  A string of constant~$r$ will, trivially, have proper
+length~$2\pi r$.
 
@@
 \begin{equation}\label{radial_string_length}
-  \int_{r=r_1}^{r_2}\frac{dr}{\sqrt{1-2M/r}}=
+  \int_{r=r_1}^{r_2}\frac{dr}{\sqrt{1-1/r}}=
   \left.
@@
 Schwarzschild metric approaches infinity there.
-
-\subsection{Radial and angular components}
 
@@
 %  \int_{p_1}^{p_2}\sqrt{\frac{\left(r'\right)^2}{1-2M/r} + r^2}d\phi=
-  \int_{p_1}^{p_2}\left(\frac{\left(r'\right)^2}{1-2M/r} + r^2\right)^\frac{1}{2}d\phi=
+  \int_{p_1}^{p_2}\left(\frac{\left(r'\right)^2}{1-1/r} + r^2\right)^\frac{1}{2}d\phi=
   \int_{p_1}^{p_2}F\left(r,r'\right)d\phi.
@@
 figure~\ref{closest_approach_self_intersecting} shows some strings
-that self-intersect.
+that self-intersect.  Figure~\ref{one_free_end} shows some strings
+that pass through~$\left(2,0\right)$ and tangent to the event horizon.
 
@@
 \end{figure}
-
-Figure~\ref{one_free_end_r_equals_2} shows some strings that pass
-through~$\left(2,0\right)$ and tangent to the event horizon.
-
-\subsection{Strings with one free end}
 
@@
 This work considered only the Schwarzschild metric and generalizing it
-to the Kerr metric~\cite{asdfsadfasfd} might be interesting.
+to the Kerr metric~\cite{kerr1963} might be interesting.
 

### 2018-10-01 string_microarticle.tex
rephrasing

@@
 Here I analyze the behaviour of light inextensible strings under
-tension in the vicinity of a black hole.  Proper length (given by the
-spatial component of the Schwarzshild metric) is minimized by the
-Euler-Lagrange equations.  Light inextensible taut strings may assume
-a range of distinct configurations with an arbitrarily high winding
-number around the central singularity, and here I give a range of
-numerical solutions and theoretical observations.
+tension in the vicinity of a nonrotating black hole.  Proper length
+(given by the spatial component of the Schwarzshild metric) is
+minimized by the Euler-Lagrange equations.  Light inextensible taut
+strings may assume topologically distinct configurations, and here I
+give some numerical solutions and theoretical observations.
 %%%%
@@
 \begin{equation}\label{schwarzschild}
-ds^2= -dt^2\left(1-1/r\right) +\frac{dr^2}{1-1/r} + r^2\left(d\theta^2 + \sin^2\theta d\phi^2\right)
+ds^2= -dt^2\left(1-1/r\right) +\frac{dr^2}{1-1/r} + r^2\left(d\theta^2 + \sin^2\theta d\phi^2\right),
 \end{equation}
 
-where units in which~$2GM=c=1$ are used, has been known for over a
-century~\cite{schwarzschild1916} and makes a number of
+\noindent where units in which~$2GM=c=1$ are used, has been known for
+over a century~\cite{schwarzschild1916} and makes a number of
 counter-intuitive predictions.  Here I consider the spatial
 component~$ds^2= \frac{dr^2}{1-1/r} + r^2d\phi^2$, confined to
-equatorial points~($\phi=\pi/2$); we understand~$1<r<\infty$
+equatorial points~($\theta=\pi/2$); we understand~$1<r<\infty$
 throughout.  A string of constant~$r$ will, trivially, have proper
@@
 
-where dashes denote differentiation with respect to~$\phi$.  The
+\noindent where dashes denote differentiation with respect to~$\phi$.  The
 Euler-Lagrange equations for this system,
@@
   r''\left(\phi\right) =
-  (r-1) + \frac{(4r-3)\left(r'\right)^2}{2r\left(r-1\right)},\qquad r>1
+  (r-1) + \frac{(4r-3)\left(r'\right)^2}{2r\left(r-1\right)},\qquad r>1.
 \end{equation}
@@
 This differential equation does not appear to have a simple analytical
-solution; I present some numerical results below.  Solutions for
-which~$r'(\phi)>0$ whenever~$r>1$ correspond to strings that have only
-one free end; the other is tangential to the event horizon, except in
-the exceptional case of constant~$\phi$ when the string is radial
-for~$r>1$.  It is straightforward to show that the radius of curvature
-of a taut string becomes infinite, and the string has a point of
-inflection, if~$\left(r'\right)^2=2r\left(r-1\right)$.
+solution; I present some numerical results below.  It is
+straightforward to show that the radius of curvature of a taut string
+becomes infinite, and the string has a point of inflection,
+if~$\left(r'\right)^2=2r\left(r-1\right)$.
 
@@
 
-Figure~\ref{closest_approach_non_self_intersecting} shows
-nonselfintersecting strings indexed by distance of approach to the
-singularity, which occurs at~$\phi=0$.  Strings with a sufficiently
-close passage to the event horizon self-intersect; numerical methods
-show that the critical value is~$\simeq 1.076109317$, with an
-uncertainty of 1 in the last digit: if a string with two free ends
-passes closer than this, it forms a closed loop.
-Figure~\ref{closest_approach_non_self_intersecting} shows some
-non-self-intersecting strings and
+Figure~\ref{closest_approach_non_self_intersecting} shows non-self
+intersecting strings indexed by distance of approach to the
+singularity, which is taken to occur at~$\phi=0$.  Strings with a
+sufficiently close passage to the event horizon self-intersect;
+numerical methods show that the critical value is~$\simeq
+1.076109317$, with an uncertainty of 1 in the last digit: if the
+closest approach is less than this value, the string crosses itself
+at~$\phi=\pi/2$.  Figure~\ref{closest_approach_non_self_intersecting}
+shows some non-self-intersecting strings and
 figure~\ref{closest_approach_self_intersecting} shows some strings
 that self-intersect.  Figure~\ref{one_free_end} shows some strings
-that pass through~$\left(2,0\right)$ and tangent to the event horizon.
+that pass through~$\left(2,0\right)$ and are tangent to the event horizon.
 
@@
 (at least in Schwarzschild coordinates) and one would have to account
-for the dynamics of thes system.
+for the dynamics of the system.
 
@@
 the right, and should start accelerating.  However, it is not clear by
-what mechanism such a force would be transmitted; the situation is
+what mechanism such a force would be transmitted: the situation is
 steady, so no gravitational waves are produced; the string is
@@
 
+\section*{Supplementary Material}
+
+Calculations are performed using the R programming language; source
+code for the figures is available at
+
+\url{https://github.com/RobinHankin/string.git}
+
 \end{document}

### 2018-10-01 string_longer_document.tex
add the longer paper

--- 2018-09-09 string.tex
@@
 \begin{abstract}
-Here I analyze the behaviour of light inextensible strings under
-tension in the vicinity of a black hole.  Proper length (given by the
-spatial component of the Schwarzshild metric) is minimized by the
-Euler-Lagrange equations.  Light inextensible taut strings may assume
-a range of distinct configurations with an arbitrarily high winding
-number around the central singularity, and here I give a range of
-numerical solutions and theoretical observations.
+
 \end{abstract}
@@
 uncertainty of 1 in the last digit: if a string with two free ends
-passes closer than this, it forms a closed loop.
+passes closer than this to the event horizon,
 Figure~\ref{closest_approach_non_self_intersecting} shows some

### 2018-10-01 string_microarticle.tex
more copyediting

@@
 
-\received{1 May 2013}
-\finalform{10 May 2013}
-\accepted{13 May 2013}
-\availableonline{15 May 2013}
+\received{day month year}
+\finalform{day month year}
+\accepted{day month year}
+\availableonline{day month year}
 \communicated{R. Hankin}
@@
 %%%
-Here I analyze the behaviour of light inextensible strings under
-tension in the vicinity of a nonrotating black hole.  Proper length
-(given by the spatial component of the Schwarzshild metric) is
+Here I analyze the behaviour of stationary, light inextensible strings
+under tension in the vicinity of a nonrotating black hole.  Proper
+length (given by the spatial component of the Schwarzshild metric) is
 minimized by the Euler-Lagrange equations.  Light inextensible taut
@@
 \label{intro}
-In this short paper I consider the behaviour of a light inextensible
-string in the exterior Schwarzschild metric.  A light string is not
-subject to gravity, and being under tension means that its
-configuration is an extremum of proper length.  Of course, no real
-string is perfectly weightless or inextensible.  With tension~$T$,
-gravitational acceleration $g$ and (proper) mass per (proper) unit
-length~$\mu$, then ideal behaviour will be appropriate for
-lengthscales $L\lesssim\frac{T}{\mu g}$.  Using Newtonian expressions
-for the strength of gravity at the event horizon gives
-$L\ll\frac{TGM}{\mu c^2}$.  For the string to behave as though light
-on lengthscales comparable with the Schwarzschild radius would require
-$T\gtrsim\mu c^2$.  Exceeding this criterion would violate the
-dominant energy condition (and in this context we note that transverse
-waves would propagate at speed~$\sqrt{T/\mu}\sim c$).  Nevertheless,
-the idealization of a light inextensible string is an interesting and
-informative limit: its analysis interprets the concept ``straight
-line'' and suggests a possible way of manipulating a black hole.  The
-Schwarzschild metric
+Light inextensible strings are a stalwart feature of elementary
+physics.  A light string is not subject to gravity, and being under
+tension means that its configuration is an extremum of proper length.
+Of course, no real string is perfectly weightless or inextensible.
+With tension~$T$, gravitational acceleration $g$ and (proper) mass per
+(proper) unit length~$\mu$, then a string will behave as though light
+on lengthscales $L\lesssim\frac{T}{\mu g}$.  Using Newtonian
+expressions for the strength of gravity at the event horizon of a
+black hole gives $L\lesssim\frac{TGM}{\mu c^2}$.  For the string to
+behave as though light on lengthscales comparable with the
+Schwarzschild radius would require $T\gtrsim\mu c^2$.  Exceeding this
+criterion would violate the dominant energy condition (and in this
+context we note that transverse waves would propagate at
+speed~$\sqrt{T/\mu}\sim c$).  Nevertheless, the idealization of a
+light inextensible string is an interesting and informative limit: its
+analysis interprets the concept ``straight line'' and suggests a
+possible way of manipulating a black hole.  The Schwarzschild metric
 
@@
   \int_{p_1}^{p_2}\left(\frac{\left(r'\right)^2}{1-1/r} + r^2\right)^\frac{1}{2}d\phi=
-  \int_{p_1}^{p_2}F\left(r,r'\right)d\phi.
+  \int_{p_1}^{p_2}F\left(r,r'\right)d\phi
 \end{equation}
@@
 \section{Results}
-Light inextensible strings fall into two distinct classes: those with
-two free ends, and those with one free end (the other ``end'' is
-tangential to the event horizon).
+Solutions to equation~\ref{rdashdash} fall into two distinct classes:
+those with two free ends, and those with one free end (the other
+``end'' is tangential to the event horizon, in general).
 
 Figure~\ref{closest_approach_non_self_intersecting} shows non-self
-intersecting strings indexed by distance of approach to the
+intersecting strings indexed by minimum distance of approach to the
 singularity, which is taken to occur at~$\phi=0$.  Strings with a
-sufficiently close passage to the event horizon self-intersect;
-numerical methods show that the critical value is~$\simeq
-1.076109317$, with an uncertainty of 1 in the last digit: if the
-closest approach is less than this value, the string crosses itself
-at~$\phi=\pi/2$.  Figure~\ref{closest_approach_non_self_intersecting}
-shows some non-self-intersecting strings and
-figure~\ref{closest_approach_self_intersecting} shows some strings
-that self-intersect.  Figure~\ref{one_free_end} shows some strings
-that pass through~$\left(2,0\right)$ and are tangent to the event horizon.
+sufficiently close passage to the event horizon self-intersect
+(Figure~\ref{closest_approach_self_intersecting}).  Numerical methods
+show that the critical value is~$\simeq 1.076109317$, with an
+uncertainty of~1 in the last digit: if the closest approach is less
+than this value, the string crosses itself at~$\phi=\pi/2$.
+Figure~\ref{one_free_end} shows some strings that are tangent to the
+event horizon and pass through the point~$\left(2,0\right)$
 
@@
 \caption{Non-self intersecting strings arranged by distance~$d$ of
-  closest approach to the event horizon, occurring at~$\phi=0$}
+  closest approach to the event horizon, occurring tangentially
+  at~$\phi=0$}
 \label{closest_approach_non_self_intersecting}
@@
 
-The analysis above was restricted to the case~$r\geq 1$, where it is
-possible to hold objects steady.  Of course, once within the event
-horizon, even a light inextensible string cannot hold itself steady
-(at least in Schwarzschild coordinates) and one would have to account
-for the dynamics of the system.
+Light inextensible strings under tension are one interpretation of
+``straight line'' in general relativity, the others being null
+geodesics and world lines of massive objects.  Here it is shown that
+such strings adopt non-trivial configurations including
+self-intersecting paths.
 
-In figure~\ref{closest_approach_non_self_intersecting}, one's
+In Figure~\ref{closest_approach_non_self_intersecting}, one's
 classical intuition would suggest that the black hole feels a force to
-the right, and should start accelerating.  However, it is not clear by
+the left, and should start accelerating.  However, it is not clear by
 what mechanism such a force would be transmitted: the situation is

### 2018-10-01 string_microarticle.tex
replace diagram with the fixed_EH_intersection work

@@
 than this value, the string crosses itself at~$\phi=\pi/2$.
-Figure~\ref{one_free_end} shows some strings that are tangent to the
-event horizon and pass through the point~$\left(2,0\right)$
+Figure~\ref{fixed_EH_intersection} shows some strings that are tangent
+to the event horizon at~$\phi=0$.
 
@@
 \centering
-\includegraphics[width=70mm]{one_free_end_r_equals_2.pdf}
-\caption{Strings arranged by angle made at $r=2$, one free end}
-\label{one_free_end}
+\includegraphics[width=70mm]{one_free_end_fixed_EH_intersection.pdf}
+\caption{Strings tangential to event horizon}
+\label{fixed_EH_intersection}
 \end{figure}
@@
 inextensible, so stores no energy.  Likewise,
-Figure~\ref{one_free_end} would suggest that angular momentum is
+Figure~\ref{fixed_EH_intersection} would suggest that angular momentum is
 transmitted to the black hole, but no mechanism is apparent.

### 2018-10-01 one_free_end_fixed_EH_intersection.R
tweaks to graphical plotting to make better diagram

@@ -7,10 +7,9 @@ pars <- c(eel=1)  # dummy
 source("usefulfuncs.R")
 
 ## setup:
-jj <- 4.1
+jj <- 3.1
 par(pty='s')
 plot(NULL,asp=1,xlim=c(-jj,jj),ylim=c(-jj,jj),type='n',axes=F,xlab='',ylab='')
-polargrid(1:4,rlab=3.5)
 
 
 
@@ -81,13 +80,19 @@ for(i in seq_along(start_angles)){
   
   rm <- rotmat(angle_of_closest_approach)
   
-  points(xy_inward  %*% rm ,type='l',col=cols[i],lwd=2)
+  jj <- xy_inward %*% rm 
+  points(jj,type='l',col=cols[i],lwd=2)
+
+  jj[,2] <- -jj[,2] 
+  points(jj,type='l',col=cols[i],lwd=2)
+  
   if(FALSE){
     points(xy_outward %*% rm ,type='l',col=cols[i],lwd=5)
   }
   
 }
-event_horizon()
-points(1,0,pch=16,cex=2,col='gray')
 
 
+
+polargrid(1:3,rlab=2.5)
+event_horizon()  # event horizon last: the black hole should be truly black

### 2018-10-01 one_free_end_fixed_EH_intersection.R
tweaks for plotting

@@ -12,17 +12,16 @@ par(pty='s')
 plot(NULL,asp=1,xlim=c(-jj,jj),ylim=c(-jj,jj),type='n',axes=F,xlab='',ylab='')
 
 
-
 dist_from_hole <- 3
 
 ## One string per start angle, each one a different colour:
-start_angles <- c(seq(from=1.18,by=0.01,to=1.5))
+start_angles <- c(seq(from=1.18,by=0.03,to=1.56))
 
 
 start_angles <- rev(start_angles)
 
-n <- length(start_angles)
-cols <- rainbow(n+round(n/7))
+n <- length(start_angles)+1
+cols <- rainbow(n+round(n/7))[-1]  # "[-1]" so the radial string is correctly coloured
 
 theta_start <- 0
 
@@ -41,10 +40,25 @@ theta_end2[cont(start_angles, c(1.34,1.35))] <-  pi*0.62
 theta_end2[cont(start_angles, c(1.35,1.36))] <-  pi*0.62
 theta_end2[cont(start_angles, c(1.36,1.42))] <-  pi*0.4
 theta_end2[cont(start_angles, c(1.42,1.44))] <-  pi*0.3
-theta_end2[cont(start_angles, c(1.44,1.51))] <-  pi*0.2
-
+theta_end2[cont(start_angles, c(1.44,1.45))] <-  pi*0.2
+theta_end2[cont(start_angles, c(1.45,1.46))] <-  pi*0.2
+theta_end2[cont(start_angles, c(1.46,1.47))] <-  pi*0.2
+theta_end2[cont(start_angles, c(1.47,1.48))] <-  pi*0.2
+theta_end2[cont(start_angles, c(1.48,1.49))] <-  pi*0.2
+theta_end2[cont(start_angles, c(1.49,1.50))] <-  pi*0.2
+theta_end2[cont(start_angles, c(1.50,1.51))] <-  pi*0.15
+theta_end2[cont(start_angles, c(1.51,1.52))] <-  pi*0.1
+theta_end2[cont(start_angles, c(1.52,1.53))] <-  pi*0.1
+theta_end2[cont(start_angles, c(1.53,1.54))] <-  pi*0.07
+theta_end2[cont(start_angles, c(1.54,1.55))] <-  pi*0.05
+theta_end2[cont(start_angles, c(1.55,1.56))] <-  pi*0.02
+theta_end2[cont(start_angles, c(1.56,1.57))] <-  pi*0.02
 
 closest_approach <- start_angles + NA
+
+
[…]

### 2018-10-01 string_microarticle.tex
sharpen up the wording

@@
 Solutions to equation~\ref{rdashdash} fall into two distinct classes:
-those with two free ends, and those with one free end (the other
-``end'' is tangential to the event horizon, in general).
-
-Figure~\ref{closest_approach_non_self_intersecting} shows non-self
-intersecting strings indexed by minimum distance of approach to the
-singularity, which is taken to occur at~$\phi=0$.  Strings with a
-sufficiently close passage to the event horizon self-intersect
-(Figure~\ref{closest_approach_self_intersecting}).  Numerical methods
-show that the critical value is~$\simeq 1.076109317$, with an
-uncertainty of~1 in the last digit: if the closest approach is less
-than this value, the string crosses itself at~$\phi=\pi/2$.
-Figure~\ref{fixed_EH_intersection} shows some strings that are tangent
-to the event horizon at~$\phi=0$.
+those with $r>1$ at all points, and those which attain $r=1$, in which
+case the string is tangential to the event horizon, in general.
+Considering first cases with $r>1$, numerical methods show that if the
+closest approach is greater than a critical value of~$\simeq
+1.076109317(1)$, the string does not self-intersect
+(Figure~\ref{closest_approach_non_self_intersecting}); if the closest
+approach is less than this value, the then string crosses itself
+at~$\phi=\pi/2$ (Figure~\ref{closest_approach_self_intersecting}).
+The case where the string is tangent to the event horizon
+at~$r=1,\phi=0$ is shown in Figure~\ref{fixed_EH_intersection}.
 
@@
 steady, so no gravitational waves are produced; the string is
-inextensible, so stores no energy.  Likewise,
-Figure~\ref{fixed_EH_intersection} would suggest that angular momentum is
-transmitted to the black hole, but no mechanism is apparent.
-
-This work considered only the Schwarzschild metric and generalizing it
-to the Kerr metric~\cite{kerr1963} might be interesting.
+inextensible, so stores no energy.  This work considered only the
+Schwarzschild metric and generalizing it to the Kerr
+metric~\cite{kerr1963} might be interesting.
 

### 2018-10-01 closest_approach.R
fewer strings, easier to see in a small diagram

@@ -16,7 +16,7 @@ plot(NULL,asp=1,xlim=c(-jj,jj),ylim=c(-jj,jj),type='l',axes=FALSE,,xlab='',ylab=
 ## setup ends
 
 
-dist <- dseq(from=1.076109317, to=7.5, len=100, power=2)
+dist <- dseq(from=1.076109317, to=7.5, len=30, power=2)
 
 cutoffmatrix <- matrix(c(
  1.060, pi  ,

### 2018-10-03 closest_approach2.R
expand F to FALSE

@@ -16,7 +16,7 @@ pars <- c(eel=1)  # dummy
 ## plot setup:
 jj <- 6
 par(xpd=TRUE)
-plot(NULL,asp=1,xlim=c(-jj,jj),ylim=c(-jj,jj),type='l',axes=FALSE,,xlab='',ylab='')
+plot(NULL,asp=1,xlim=c(-jj,jj),ylim=c(-jj,jj),type='l',axes=F,xlab='',ylab='')
 ## setup ends
 
 

### 2018-10-03 one_free_end_fixed_EH_intersection.R
tweak for appearance

@@ -7,8 +7,9 @@ pars <- c(eel=1)  # dummy
 source("usefulfuncs.R")
 
 ## setup:
-jj <- 3.1
-par(pty='s')
+jj <- 2.5
+par(xpd=T)
+par(pty='m')
 plot(NULL,asp=1,xlim=c(-jj,jj),ylim=c(-jj,jj),type='n',axes=F,xlab='',ylab='')
 
 

### 2018-10-03 string_microarticle.tex
various rewriting tweaks

@@
 Here I analyze the behaviour of stationary, light inextensible strings
-under tension in the vicinity of a nonrotating black hole.  Proper
-length (given by the spatial component of the Schwarzshild metric) is
-minimized by the Euler-Lagrange equations.  Light inextensible taut
-strings may assume topologically distinct configurations, and here I
-give some numerical solutions and theoretical observations.  
+under tension in the vicinity of a nonrotating black hole.  Such
+strings minimize their proper length, given by integrating the spatial
+equatorial component of the Schwarzshild metric.  The string path is
+calculated by numerical solution to the Euler-Lagrange equations.  If
+the string is entirely outside the event horizon, its closest approach
+is to the singularity is tangential; if the distance of closest
+approach is less than about $1.076109317\cdot(2GM)$, the string
+self-intersects.  Light taut strings interpret the concept ``straight
+line'' and their behaviour suggests a possible way of manipulating a
+black hole.
 %%%%
@@
 %% Keywords
-\KWD Black Holes\sep Light inextensible string
+\KWD Black holes\sep Light inextensible string
 \end{keyword}
@@
 (proper) unit length~$\mu$, then a string will behave as though light
-on lengthscales $L\lesssim\frac{T}{\mu g}$.  Using Newtonian
+on lengthscales $L\lesssim T/\left(\mu g\right)$.  Using Newtonian
 expressions for the strength of gravity at the event horizon of a
-black hole gives $L\lesssim\frac{TGM}{\mu c^2}$.  For the string to
-behave as though light on lengthscales comparable with the
+black hole gives $L\lesssim TGM/\left(\mu c^2\right)$.  For the string
+to behave as though light on lengthscales comparable with the
 Schwarzschild radius would require $T\gtrsim\mu c^2$.  Exceeding this
@@
 counter-intuitive predictions.  Here I consider the spatial
-component~$ds^2= \frac{dr^2}{1-1/r} + r^2d\phi^2$, confined to
-equatorial points~($\theta=\pi/2$); we understand~$1<r<\infty$
-throughout.  A string of constant~$r$ will, trivially, have proper
-length~$2\pi r$.
+component~$ds^2= dr^2/\left(1-1/r\right) + r^2d\phi^2$, confined to
+equatorial points~($\theta=\pi/2$); we understand~$1\leq r<\infty$
+throughout.  A circular string at constant~$r$ will, trivially, have
+proper length~$2\pi r$.
 
@@
 
-Light inextensible strings under tension are one interpretation of
-``straight line'' in general relativity, the others being null
-geodesics and world lines of massive objects.  Here it is shown that
-such strings adopt non-trivial configurations including
-self-intersecting paths.
+Light inextensible strings under tension adopt non-trivial
+configurations including self-intersecting paths.  The path of such
+strings furnishes an interpretation of the concept ``straight line''
+in general relativity, others being null geodesics and world lines of
+massive objects.  Comparing a world line with that of nearby taut
+strings thus furnishes another insight into path curvature,
+complementing the analysis of Abramowicz~\cite{abramowicz1992}.
 
@@
 what mechanism such a force would be transmitted: the situation is
-steady, so no gravitational waves are produced; the string is
-inextensible, so stores no energy.  This work considered only the
-Schwarzschild metric and generalizing it to the Kerr
-metric~\cite{kerr1963} might be interesting.
+steady, so no gravitational waves are produced, and the string is
+inextensible so stores no energy.
 
@@
 code for the figures is available at
-
+\\
+\\
 \url{https://github.com/RobinHankin/string.git}

### 2018-10-03 usefulfuncs.R
whitespace plus docs

@@ -83,16 +83,14 @@ library("deSolve")   # needed for ode()
   ## setup ends
 }
 
-fun <- function(cuts,vals){
+`fun` <- function(cuts,vals){
   return(function(x){vals[findInterval(x, c(-Inf, cuts))]})
 }
 
-rotmat <- function(theta){
+`rotmat` <- function(theta){
     matrix(c(cos(theta),sin(theta),-sin(theta),cos(theta)),2,2)
 }
-    
 
-
-dseq <- function(from,to,length.out,power=1){
+`dseq` <- function(from,to,length.out,power=1){  # dseq == distorted seq
   from + (to-from)*seq(from=0,to=1,length.out=length.out)^power
 }

### 2018-10-03 string_microarticle.tex
minor wordsmithing to the abstract

@@
 strings minimize their proper length, given by integrating the spatial
-equatorial component of the Schwarzshild metric.  The string path is
+component of the Schwarzshild metric along the string.  The path is
 calculated by numerical solution to the Euler-Lagrange equations.  If
@@
 is to the singularity is tangential; if the distance of closest
-approach is less than about $1.076109317\cdot(2GM)$, the string
-self-intersects.  Light taut strings interpret the concept ``straight
-line'' and their behaviour suggests a possible way of manipulating a
-black hole.
+approach is less than about $1.076109317$ Schwarzschild radii, the
+string self-intersects.  If the string attains $r=2GM$, then it will
+in general be tangential to the event horizon at that point.  Light
+taut strings interpret the concept ``straight line'' and their
+behaviour suggests a possible way of manipulating a black hole.
 %%%%

### 2018-10-03 closest_approach.R
tweak line width and number of strings

@@ -56,10 +56,10 @@ f <- fun(cutoffmatrix[,1],cutoffmatrix[,2])
 for(i in seq_along(dist)){
   tseq <- seq(from=0,to=f(dist[i]),len=100)
   xy <- stringpoints(y_start=dist[i],initial_string_angle = 0,theta=tseq)
-  points(xy,type='l',col=rainbow(length(dist))[i])
+  points(xy,type='l',col=rainbow(length(dist))[i],lwd=2)
   
   xy <- stringpoints(y_start=dist[i],initial_string_angle = 0, theta=-tseq)
-  points(xy,type='l',col=rainbow(length(dist))[i])
+  points(xy,type='l',col=rainbow(length(dist))[i],lwd=2)
 }
 
 # mask strings too far from the black hole:

### 2018-10-03 closest_approach2.R

@@ -20,7 +20,7 @@ plot(NULL,asp=1,xlim=c(-jj,jj),ylim=c(-jj,jj),type='l',axes=F,xlab='',ylab='')
 ## setup ends
 
 
-dist <- dseq(from=(1.00012), to=1.076109317,len=100,power=2.3)
+dist <- dseq(from=(1.00012), to=1.076109317,len=50,power=3.3)
 
 
 cutoffmatrix <- matrix(c(
@@ -87,7 +87,7 @@ f <- fun(cutoffmatrix[,1],cutoffmatrix[,2])
 for(i in seq_along(dist)){
   tseq <- seq(from=0,to=f(dist[i]),len=100)
   xy <- stringpoints(y_start=dist[i],initial_string_angle = 0,theta=tseq)
-  points(xy,type='l',col=rainbow(length(dist))[i])
+  points(xy,type='l',col=rainbow(length(dist))[i],lwd=2)
   
 #  xy <- stringpoints(y_start=dist[i],initial_string_angle = 0, theta=-tseq)
 #  points(xy,type='l',col=rainbow(length(dist))[i])

### 2018-10-03 string_microarticle.tex
say explicitly where the PDF files are created
typo
add Flamm's paraboloid reference plus slight rewording to re-attain two page maximum

@@
 \noindent where units in which~$2GM=c=1$ are used, has been known for
-over a century~\cite{schwarzschild1916} and makes a number of
-counter-intuitive predictions.  Here I consider the spatial
+over a century~\cite{schwarzschild1916}.  Here I consider the spatial
 component~$ds^2= dr^2/\left(1-1/r\right) + r^2d\phi^2$, confined to
@@
 This differential equation does not appear to have a simple analytical
-solution; I present some numerical results below.  It is
-straightforward to show that the radius of curvature of a taut string
-becomes infinite, and the string has a point of inflection,
+solution; solutions correspond to a taut string confined to Flamm's
+paraboloid~\cite{flamm1916}.  It is straightforward to show that the
+string has a point of inflection
 if~$\left(r'\right)^2=2r\left(r-1\right)$.
@@
 case the string is tangential to the event horizon, in general.
-Considering first cases with $r>1$, numerical methods show that if the
-closest approach is greater than a critical value of~$\simeq
-1.076109317(1)$, the string does not self-intersect
-(Figure~\ref{closest_approach_non_self_intersecting}); if the closest
-approach is less than this value, the then string crosses itself
-at~$\phi=\pi/2$ (Figure~\ref{closest_approach_self_intersecting}).
-The case where the string is tangent to the event horizon
-at~$r=1,\phi=0$ is shown in Figure~\ref{fixed_EH_intersection}.
+Numerical methods show that if the closest approach is greater than a
+critical value of~$\simeq 1.076109317(1)$, the string does not
+self-intersect (Figure~\ref{closest_approach_non_self_intersecting});
+if the closest approach is less than this value, the then string
+crosses itself at~$\phi=\pi/2$
+(Figure~\ref{closest_approach_self_intersecting}).  The case where the
+string is tangent to the event horizon at~$r=1,\phi=0$ is shown in
+Figure~\ref{fixed_EH_intersection}.
 
-\begin{figure}[h!]
+\begin{figure}[h!] % Created by "closest_approach.R"
 \centering
@@
 
-\begin{figure}[h!]
+\begin{figure}[h!] % created by "closest_approach2.R"
 \centering
@@
 
-\begin{figure}[h!]
+\begin{figure}[h!] % created by "one_free_end_fixed_EH_intersection.R"
 \centering
@@
 \bibliographystyle{model3-num-names}
-\bibliography{refs}
+\bibliography{stringrefs}
 

### 2019-02-05 string_ajp.tex
start manuscript for American Journal of Physics

--- 2018-10-03 string_microarticle.tex
@@
-%% This is file `rinp-template.tex',
-%% 
-%% Copyright 2016 Elsevier Ltd
-%% 
-%% This file is part of the 'Elsarticle Bundle'.
-%% ---------------------------------------------
-%% 
-%% It may be distributed under the conditions of the LaTeX Project Public
-%% License, either version 1.2 of this license or (at your option) any
-%% later version.  The latest version of this license is in
-%%    http://www.latex-project.org/lppl.txt
-%% and version 1.2 or later is part of all distributions of LaTeX
-%% version 1999/12/01 or later.
-%% 
-%% The list of all files belonging to the 'Elsarticle Bundle' is
-%% given in the file `manifest.txt'.
-%% 
-%% Template article for Elsevier's document class `elsarticle'
-%% with harvard style bibliographic references
-%%
-%% $Id: $
-%%
-%% Use the option review to obtain double line spacing
-%\documentclass[times,review,preprint,authoryear]{elsarticle}
+\documentclass[prb,preprint]{revtex4-1} 
+% The line above defines the type of LaTeX document.
+% Note that AJP uses the same style as Phys. Rev. B (prb).
 
-%% Use the options `twocolumn,final' to obtain the final layout
-%% Use longtitle option to break abstract to multiple pages if overfull.
-%% For Review pdf (With double line spacing)
-%\documentclass[times,twocolumn,review]{elsarticle}
-%% For abstracts longer than one page.
-%\documentclass[times,twocolumn,review,longtitle]{elsarticle}
-%% For Review pdf without preprint line
-%\documentclass[times,twocolumn,review,nopreprintline]{elsarticle}
-%% Final pdf
-\documentclass[times,twocolumn,final]{elsarticle}
-%%
-%\documentclass[times,twocolumn,final,longtitle]{elsarticle}
-%%
+% The % character begins a comment, which continues to the end of the line.
 
-%% Stylefile to load RINP template
-\usepackage{rinp}
-\usepackage{framed,multirow}
-
-%% The amssymb package provides various useful mathematical symbols
-\usepackage{amssymb}
-\usepackage{latexsym}
-
-% Following three lines are needed for this document.
-% If you are not loading colors or url, then these are
-% not required.
-\usepackage{url}
-\usepackage{xcolor}
-\definecolor{newcolor}{rgb}{.8,.349,.1}
-
-\journal{Results in Physics}
+\usepackage{amsmath}  % needed for \tfrac, \bmatrix, etc.
+\usepackage{amsfonts} % needed for bold Greek, Fraktur, and blackboard bold
+\usepackage{graphicx} % needed for figures
 
@@
 
-\verso{Robin Hankin}
-
-\begin{frontmatter}
+% Be sure to use the \title, \author, \affiliation, and \abstract macros
+% to format your title page.  Don't use lower-level macros to  manually
+% adjust the fonts and centering.
 
 \title{Light inextensible strings under tension in the Schwarzschild geometry}
+% In a long title you can use \\ to force a line break at a certain location.
 
-\author[1]{Robin \snm{Hankin}\corref{cor1}}
-\cortext[cor1]{Corresponding author: 
-  Tel.: +64-09-921-9999;  
-  fax: +64-09-921-9998;}
+\author{Robin K. S. Hankin}
+\email{hankin.robin@gmail.com} % optional
+\altaffiliation[permanent address: ]{55 Wellesley Street East,
+Auckland, New Zealand}
+% If there were a second author at the same address, we would put another 
+% \author{} statement here.  Don't combine multiple authors in a single
+% \author statement.
+\affiliation{School of Engineering, Computer and Mathematical Sciences, Auckland University of Technology}
+% Please provide a full mailing address here.
 
-\address[1]{Auckland University of Technology, 2-14 Wakefield Street,
-  Auckland 1142, New Zealand}
+% See the REVTeX documentation for more examples of author and affiliation lists.
 
-\received{day month year}
-\finalform{day month year}
-\accepted{day month year}
-\availableonline{day month year}
-\communicated{R. Hankin}
+\date{\today}
 
 \begin{abstract}
-%%%
 Here I analyze the behaviour of stationary, light inextensible strings
@@
 approach is less than about $1.076109317$ Schwarzschild radii, the
-string self-intersects.  If the string attains $r=2GM$, then it will
-in general be tangential to the event horizon at that point.  Light
-taut strings interpret the concept ``straight line'' and their
-behaviour suggests a possible way of manipulating a black hole.
-%%%%
+string self-intersects.  If the string attains $r=2GM/c^2$, then it
+will in general be tangential to the event horizon.  Light taut
+strings interpret the concept ``straight line'' and their behaviour
+suggests a possible way of manipulating a black hole.
 \end{abstract}
+% AJP requires an abstract for all regular article submissions.
+% Abstracts are optional for submissions to the "Notes and Discussions" section.
 
-\begin{keyword}
-%% MSC codes here, in the form: \MSC code \sep code
-%% or \MSC[2008] code \sep code (2000 is the default)
-\MSC 83C57\sep 35Q75
-%% Keywords
-\KWD Black holes\sep Light inextensible string
-\end{keyword}
+\maketitle % title page is now complete
 
-\end{frontmatter}
+\section{Introduction} % Section titles are automatically converted to all-caps.
+% Section numbering is automatic.
 
-%\linenumbers
-
-%% main text
 \section{Introduction}

### 2019-02-05 string_ajp.tex
more work for AJP submission

@@
 
+Elementary methods show that the proper length $\ell$ of radial strings has
+some peculiar features.  Firstly, consider the proper length from
+$r=1$ to $r=1+w$ where $w>0, \left|w\right|\ll 1$.  Then
+
+\begin{equation}
+  \ell=\int_{r=1}^{r=1+w}ds
+  =2\sqrt{w} + \frac{1}{3}w^{3/2} + O(w^{5/2})
+\end{equation}
+
+This would indicate that, close to the Schwarzschild radius $r=1$,
+large (but finite) amounts of string are required to connect two
+points with slightly differing radial coordinates.  Passing to the
+case $w\gg 1$ we have
+
+\begin{equation}
+  \ell  =\frac{1}{2} + \log(2) + w + \frac{1}{8}w^{-1} + O(w^{-2}),
+\end{equation}
+
+\noindent suggesting that there is ``extra radial space'', of length
+over $1.19$ Schwarzschild radii, sensible at large distances from the
+event horizon.  This might suggest that radial strings have to be
+longer than expected.
+
+\subsection{Nonradial string}
 Integrating along a path between two
 points~$p_1=\left(r_1,\phi_1\right)$ and~$p_2=\left(r_2,\phi_2\right)$
-gives the {\em proper} length of the path, which is the length of a
-string joining~$p_1$ and~$p_2$.  A taut but light string from~$p_1$
-to~$p_2$ will adopt an extremal-length path.  Such paths may be found
-by the calculus of variations; parametrizing a curve in terms
-of~$r=r\left(\phi\right)$ gives us a path length of
+gives the {\em proper} length of the path, which is the length of an
+inextensible string joining~$p_1$ and~$p_2$.  A taut but light string
+from~$p_1$ to~$p_2$ will adopt an extremal-length path.  Such paths
+may be found by the calculus of variations; parametrizing a curve in
+terms of~$r=r\left(\phi\right)$ gives us a path length of
 
@@
 
-This differential equation does not appear to have a simple analytical
-solution; solutions correspond to a taut string confined to Flamm's
-paraboloid~\cite{flamm1916}.  It is straightforward to show that the
-string has a point of inflection
+This differential equation does not appear to be soluble analytically;
+solutions correspond to a taut string confined to Flamm's
+paraboloid~\cite{flamm1916}; see figure~\ref{flamm}.  It is
+straightforward to show that the string has a point of inflection
 if~$\left(r'\right)^2=2r\left(r-1\right)$.
+
+\begin{figure}[h!] % Created by "flamm.R"
+\centering
+\includegraphics[width=70mm]{flamm_string.pdf}
+\caption{Perspective fiew of Flamm's paraboloid with superimposed minimal-length path corresponding to a taut, light string}
+\label{flamm}
+\end{figure}
 

### 2019-02-07 string_ajp.tex
more editing for AJP

@@
 points with slightly differing radial coordinates.  Passing to the
-case $w\gg 1$ we have
+case $w\gg 1$ we have an asymptotic expansion that begins
 
@@
 over $1.19$ Schwarzschild radii, sensible at large distances from the
-event horizon.  This might suggest that radial strings have to be
-longer than expected.
+event horizon.  This might suggest that radial strings have longer
+proper length than expected.
 
@@
 
-This differential equation does not appear to be soluble analytically;
-solutions correspond to a taut string confined to Flamm's
-paraboloid~\cite{flamm1916}; see figure~\ref{flamm}.  It is
-straightforward to show that the string has a point of inflection
-if~$\left(r'\right)^2=2r\left(r-1\right)$.
-
 \begin{figure}[h!] % Created by "flamm.R"
@@
 \includegraphics[width=70mm]{flamm_string.pdf}
-\caption{Perspective fiew of Flamm's paraboloid with superimposed minimal-length path corresponding to a taut, light string}
+\caption{Perspective fiew of Flamm's paraboloid with superimposed minimal-length
+  path corresponding to a taut, light string}
 \label{flamm}
 \end{figure}
+
+Solutions to this differential equation correspond to a taut string
+confined to Flamm's paraboloid~\cite{flamm1916}; see
+figure~\ref{flamm}.
+
+There does not appear to be a simple analytical solution but
+nevertheless we may make several observations.
+Equation~\ref{rdashdash} is a second-order nonlinear ordinary
+differential equation and thus has two constants of integration.
+There are two qualitatively different types of solutions: those with a
+tangential point ($r'=0,r>1$), and those without.  Consider first
+solutions with such a point: there exists a pair~$\left(r,\phi\right),
+r>1$ at which~$r'(\phi)=0$.  By Rolle's theorem, this point must be
+unique, as~$r''(\phi)>0$, and be the closest approach of the string to
+the event horizon.
+
+Further, because the system is symmetrical
+under~$\phi\longrightarrow-\phi$, the string must be symmetrical about
+at this point.  The (signed) radius of curvature~$R$ is readily
+evaluated:
+
+\begin{equation}\label{roc}
+  R = \frac{
+    \left(r^2 + \left(r'\right)^2\right)^\frac{3}{2}
+  }{
+%    \left|
+    r^2 + 2\left(r'\right)^2-rr''
+%    \right|
+  }
+  =
+  \frac{
+    \left(r^2 + \left(r'\right)^2\right)^\frac{3}{2}
+  }{
+    r-\frac{\left(r'\right)^2}{2\left(r-1\right)}
+  }\qquad\mbox{along a taut string}
+\end{equation}
+
+\noindent which would give~$R$ as a function of~$\left(r,r'\right)$.
+Considering~$r'(\phi)=0$ shows that if a string is tangential at a
+point with~$r_0>1$, then it has a radius of curvature of~$r_0^2$ at
+that point.  Any circular path, including a circular orbit at~$r=r_0$,
+has a radius of curvature of $r_0<r_0^2$, showing that circular orbits
+cannot follow a taut string at closest approach.  Also observe that
+null geodesics have a radius of curvature of $2r_0^2/3 < r_0^2$,
+showing that light cannot follow a taut string at closest approach
+either.
+
+Compare the above analysis with a taut {\em heavy} string which
+adopts, at least locally, a catenary configuration at closest
+approach~\cite{nguyen2007}.  Such a configuration is manifestly convex
+to the singularity, in contrast to light strings which are concave.
+
+If, conversely, $r'(\phi)>0$ whenever~$r>1$, then the string has only
+one free end; the other is tangential to the event horizon, except in
+the exceptional case of constant~$\phi$ when the string is radial
+for~$r>1$.
 
@@
 \centering
-\includegraphics[width=70mm]{closest_approach_nonselfintersecting.pdf}
+\includegraphics[width=170mm]{closest_approach_nonselfintersecting.pdf}
 \caption{Non-self intersecting strings arranged by distance~$d$ of
@@
 \centering
-\includegraphics[width=70mm]{closest_approach_selfintersecting.pdf}
+\includegraphics[width=170mm]{closest_approach_selfintersecting.pdf}
 \caption{Strings arranged by closest approach to the event horizon:
@@
 \centering
-\includegraphics[width=70mm]{one_free_end_fixed_EH_intersection.pdf}
+\includegraphics[width=170mm]{one_free_end_fixed_EH_intersection.pdf}
 \caption{Strings tangential to event horizon}

### 2019-02-07 flamm.R
functionality to optionally include the legend

@@ -12,19 +12,24 @@ source("usefulfuncs.R")
 pars <- c(eel=1) # dummy
 
 
-flamm <- function(string=TRUE, ...){
+flamm <- function(string=TRUE, drawlegend=FALSE, ...){
+
+  ## string, drawlengend: Boolean specifying whether or not to draw
+  ## the string and/or the legend.  "...", further arguments passed to
+  ## scatter3D().
 
   ## values of r and theta to plot: use longer vectors to make the plot
   ## look better but this takes longer
+
 
   thetavals <- seq(from=0,to=2*pi,len=100)
   rvals <-  seq(from=1,to=10,len=30)
 
   ## set up the axis:
   scatter3D(x=1:2, y=1:2, z=1:2,   # dummy points
-            bty=NULL,type='n',lwd=1,pch=NA,
+            type='n',lwd=1,pch=NA,
             r=100000,box=FALSE,scale=FALSE,colkey=FALSE,
-            xlim=c(-10,10),ylim=c(-10,10),zlim=c(-32,0),
+            xlim=c(-5,5),ylim=c(-5,5),zlim=c(-16,0),
             ...
             )
 
@@ -87,25 +92,29 @@ flamm <- function(string=TRUE, ...){
             lwd=2, col='black',
             add=TRUE,ticktype='detailed'
             )
-  legend("topright",
-         legend = c(
-             expression(paste("lines of constant ",r)),
-             expression(paste("lines of constant ",phi)),
-             "taut string"),
-         col = c("blue","red","black"),
-         lty = 1,
-         lwd = c(0.1,0.1,2)
-         )
+
+    if(drawlegend){
+      legend("topright",
+             legend = c(
+                 expression(paste("lines of constant ",r)),
+                 expression(paste("lines of constant ",phi)),
+                 "taut string"),
+             col = c("blue","red","black"),
+             lty = 1,
+             lwd = c(0.1,0.1,2)
+             )
+    }
   } else {
-  legend("topright",
-         legend = c(
-             expression(paste("lines of constant ",r)),
-             expression(paste("lines of constant ",phi))),
-         col = c("blue","red","black"),
-         lty = 1,
-         lwd = c(0.1,0.1)
-         )
+    if(drawlegend){
+      legend("topright",
+             legend = c(
+                 expression(paste("lines of constant ",r)),
+                 expression(paste("lines of constant ",phi))),
+             col = c("blue","red","black"),
[…]

### 2019-02-07 maker.R
do not include the legend

@@ -28,7 +28,7 @@ flamm(string=FALSE,phi=50,theta=40)
 dev.off()
 
 pdf(file="flamm_string.pdf")
-flamm(string=TRUE,phi=50,theta=50)
+flamm(string=TRUE,addlegend=FALSE,phi=50,theta=50)
 dev.off()
 
 pdf(file="light_start_at_r_equals_2.pdf")

### 2019-02-07 radius_of_curvature_switch.R
whitespace

@@ -1,9 +1,6 @@
-## lots of blue strings with varying closest approach distances.
-
 ## This script runs standalone but is designed to be called from
 ## maker.R [cf Makefile], which creates the pdf file
 ## "curvature_switch.pdf"
-
 
 source("usefulfuncs.R") # defines stringpoints()
 
@@ -46,13 +43,11 @@ for(i in seq_along(dist)){
   wm <-  which.min(rr)
   theta1 <- theta-theta[wm]
   xy1 <- cbind(r*cos(theta1),r*sin(theta1))
-##  points(xy1[ thing,],type='l',col=rainbow(length(dist))[i])  # nice rainbow colours
   points(xy1[ thing,],type='l',col='red')   # red
   points(xy1[!thing,],type='l',col='black')
  
   theta2 <- -theta-theta[wm]
   xy2 <- cbind(r*cos(theta2),r*sin(theta2))
-#  points(xy2[ thing,],type='l',col=rainbow(length(dist))[i])  # nice rainbow colours
   points(xy2[ thing,],type='l',col='red')
   points(xy2[!thing,],type='l',col='black')
 

### 2019-02-07 string_ajp.tex
include radius of curvature switch figure

@@
 \includegraphics[width=70mm]{flamm_string.pdf}
-\caption{Perspective fiew of Flamm's paraboloid with superimposed minimal-length
+\caption{Perspective view of Flamm's paraboloid with superimposed minimal-length
   path corresponding to a taut, light string}
@@
 showing that light cannot follow a taut string at closest approach
-either.
+either.  The string has a point of inflection
+if~$\left(r'\right)^2=2r(r-1)$.
 
@@
 
-\section{Results}
+\section{Numerical results}
+
 Solutions to equation~\ref{rdashdash} fall into two distinct classes:
@@
 \includegraphics[width=170mm]{closest_approach_selfintersecting.pdf}
-\caption{Strings arranged by closest approach to the event horizon:
- self-intersecting strings shown for $\phi\geq 0$}
+\caption{Self-intersecting strings arranged by closest approach to the
+  event horizon which occurs at $\phi=0$.  
+  shown for $\phi\geq 0$}
 \label{closest_approach_self_intersecting}
@@
 \label{fixed_EH_intersection}
+\end{figure}
+
+\begin{figure}[h!]
+\centering
+\includegraphics{curvature_switch.pdf} % created by "radius_of_curvature_switch.R"
+\caption{Strings arranged by point of curvature switch, which occurs
+  at~$\phi=0$.  Colour changes from black to red at points of
+  infinite~$R$; black indicates convex to the origin and red indicates
+  concave regions}
+\label{curvature_switch}
 \end{figure}

### 2019-02-13 string_ajp.tex
typo
more rewording

@@
 
-\section{Introduction}
 \label{intro}
@@
 
-Solutions to equation~\ref{rdashdash} fall into two distinct classes:
-those with $r>1$ at all points, and those which attain $r=1$, in which
-case the string is tangential to the event horizon, in general.
-Numerical methods show that if the closest approach is greater than a
+Here I solve equation~\ref{rdashdash} using the \verb+deSolve+ package
+for ordinary differential equations~\cite{soetart2010}.  Solutions
+fall into two distinct classes: those with $r>1$ at all points, and
+those which attain $r=1$, in which case the string is tangential to
+the event horizon, unless it is radial.
+
+Taking strings for which $r>1$ first,
+figure~\ref{closest_approach_non_self_intersecting} shows a sequence
+of non-self-intersecting numerical solutions to
+equation~\ref{rdashdash}, rotated so that the tangential point of
+closest approach occurs at~$\phi=0$.
+Figure~\ref{closest_approach_self_intersecting} shows the
+corresponding diagram for self-interesecting strings.  A bisection
+technique reveals that if the closest approach is greater than a
 critical value of~$\simeq 1.076109317(1)$, the string does not
-self-intersect (Figure~\ref{closest_approach_non_self_intersecting});
-if the closest approach is less than this value, the then string
-crosses itself at~$\phi=\pi/2$
-(Figure~\ref{closest_approach_self_intersecting}).  The case where the
-string is tangent to the event horizon at~$r=1,\phi=0$ is shown in
+self-intersect; if the closest approach is less than this value, then
+the string crosses itself at~$\phi=\pi/2$.  As discussed above, if
+$\left(r'\right)^2=2r(r-1)$, the string has a point of inflection at
+which point the radius of curvature becomes infinite.
+Figure~\ref{curvature_switch} shows non-self intersecting strings
+rotated so that the point of inflection occurs at~$\phi=0$; we see
+that the strings have a finite length in which convex to the origin
+and two semi-infinite stretches in which the curvature goes the other
+way.
+
+The case where the string is tangent to the event horizon at~$\phi=0$,
+including a radial string, is shown in
 Figure~\ref{fixed_EH_intersection}.
 
-\begin{figure}[h!] % Created by "closest_approach.R"
+\begin{figure}[p] % Created by "closest_approach.R"
 \centering
@@
 
-\begin{figure}[h!] % created by "closest_approach2.R"
+\begin{figure}[p] % created by "closest_approach2.R"
 \centering
@@
 
-\begin{figure}[h!] % created by "one_free_end_fixed_EH_intersection.R"
-\centering
-\includegraphics[width=170mm]{one_free_end_fixed_EH_intersection.pdf}
-\caption{Strings tangential to event horizon}
-\label{fixed_EH_intersection}
-\end{figure}
-
-\begin{figure}[h!]
+\begin{figure}[p]
 \centering
@@
 \label{curvature_switch}
+\end{figure}
+
+\begin{figure}[p] % created by "one_free_end_fixed_EH_intersection.R"
+\centering
+\includegraphics[width=170mm]{one_free_end_fixed_EH_intersection.pdf}
+\caption{Strings tangential to event horizon}
+\label{fixed_EH_intersection}
 \end{figure}
@@
 
-Calculations are performed using the R programming language; source
-code for the figures is available at
+Calculations are performed using the R programming language~\citep{rcore2018};
+source code for the figures is available at
 \\

### 2019-02-19 string_ajp.tex
rewording and add Abramowicz as a ref

@@
 speed~$\sqrt{T/\mu}\sim c$).  Nevertheless, the idealization of a
-light inextensible string is an interesting and informative limit: its
-analysis interprets the concept ``straight line'' and suggests a
-possible way of manipulating a black hole.  The Schwarzschild metric
+string as light and inextensible is an interesting and informative
+limit: its analysis interprets the concept ``straight line'' and
+suggests a possible way of manipulating a black hole.  Peculiar
+counterintuitive behaviour such as inward-directed centrifugal force
+in the vicinity of a black hole is well
+known~\citep{abramowicz1992,abramowicz1993}, and the analysis of taut
+strings furnishes another everyday window into general relativity.
+Note in particular that Abramowicz states ``feels with his hands''
+
+The Schwarzschild metric
 
@@
 steady, so no gravitational waves are produced, and the string is
-inextensible so stores no energy.
+inextensible so stores no energy.  Further work might include analysis
+of taut strings near a spinning black hole.
 

### 2019-02-27 light_closest_approach.R
tweaks to plotsize, more docs, legend

@@ -1,4 +1,10 @@
-## Null geodesics.  Here, the light starts tangentially from phi=0.
+## Null geodesics in the Schwarzschild geometry.
+
+## This script runs standalone but is designed to be called from
+## maker.R [cf Makefile], which creates the pdf file
+## "light_closest_approach.pdf"
+
+## Here, the light starts tangentially from phi=0.
 ## It is like closest_approach.R and closest_approach2.R but for light
 ## beams rather than string.  Compare light_start_at_r_equals_2.R,
 ## which has light beams starting from (2,0) going in all directions.
@@ -20,7 +26,7 @@ source("usefullightfuncs.R") # defines lightpoints()
 
 r_start <- 2 # starting radius for light ray
 mask <- TRUE    # set to FALSE to see entire geodesic (TRUE is time-consuming but needed for publication quality)
-size_of_plot <- 6
+size_of_plot <- 3.5
 
 ## plot setup:
 par(xpd=TRUE)
@@ -140,4 +146,5 @@ circ(1.5,lwd=3,lty=1,col="red")   # ergosphere
 
 event_horizon(fill=FALSE)
 
-
+legend("bottomleft",bg="white",legend=c("null geodesic","closed photon orbit","event horizon"),
+       lty=1,col=c("red","red","black"),lwd=c(1,3,1))

### 2019-02-27 maker.R
add null geodesic calculator

@@ -35,4 +35,8 @@ pdf(file="light_start_at_r_equals_2.pdf")
 source("light_start_at_r_equals_2.R")
 dev.off()
 
+pdf(file="light_closest_approach.pdf")
+source("light_closest_approach.R")
+dev.off()
+
 

### 2019-02-27 string_ajp.tex
discuss null geodesics

@@
 Figure~\ref{closest_approach_self_intersecting} shows the
-corresponding diagram for self-interesecting strings.  A bisection
-technique reveals that if the closest approach is greater than a
-critical value of~$\simeq 1.076109317(1)$, the string does not
-self-intersect; if the closest approach is less than this value, then
-the string crosses itself at~$\phi=\pi/2$.  As discussed above, if
-$\left(r'\right)^2=2r(r-1)$, the string has a point of inflection at
-which point the radius of curvature becomes infinite.
-Figure~\ref{curvature_switch} shows non-self intersecting strings
-rotated so that the point of inflection occurs at~$\phi=0$; we see
-that the strings have a finite length in which convex to the origin
-and two semi-infinite stretches in which the curvature goes the other
-way.
+corresponding diagram for self-interesecting strings
+(figure~\ref{light_closest_approach} shows comparable null geodesic
+diagram for comparison).  A bisection technique reveals that if the
+closest approach is greater than a critical value of~$\simeq
+1.076109317(1)$, the string does not self-intersect; if the closest
+approach is less than this value, then the string crosses itself
+at~$\phi=\pi/2$.  As discussed above, if $\left(r'\right)^2=2r(r-1)$,
+the string has a point of inflection at which point the radius of
+curvature becomes infinite.  Figure~\ref{curvature_switch} shows
+non-self intersecting strings rotated so that the point of inflection
+occurs at~$\phi=0$; we see that the strings have a finite length in
+which convex to the origin and two semi-infinite stretches in which
+the curvature goes the other way.
 
@@
 \label{closest_approach_self_intersecting}
+\end{figure}
+
+\begin{figure}[p] % Created by "light_closest_approach.R"
+\centering
+\includegraphics[width=170mm]{light_closest_approach.pdf}
+\caption{Null geodesics in the Schwarzschild geometry starting~$(r,0)$
+  and initially tangential.  Note the differences between these curves
+  and the taut strings shown elsewhere: unlike taug strings, null
+  geodesics have positive curvature everywhere, may cross the event
+  horizon inwards, and are never tangential to the event horizon}
+\label{light_closest_approach}
 \end{figure}

### 2019-02-27 closest_approach.R
modification for publication, remove rainbow colour and standardize strings to be blue.  Option can be reversed by setting userainbow to TRUE

@@ -10,9 +10,10 @@ source("usefulfuncs.R") # defines stringpoints()
 pars <- c(eel=1)  # dummy
 
 ## plot setup:
-jj <- 6
+jj <- c(-6,6)
 par(xpd=TRUE)
-plot(NULL,asp=1,xlim=c(-jj,jj),ylim=c(-jj,jj),type='l',axes=FALSE,,xlab='',ylab='')
+plot(NULL,asp=1,xlim=jj,ylim=jj,type='l',axes=FALSE,,xlab='',ylab='')
+userainbow <- FALSE   # TRUE = rainbow, FALSE = all strings blue
 ## setup ends
 
 
@@ -54,12 +55,19 @@ cutoffmatrix <- matrix(c(
 f <- fun(cutoffmatrix[,1],cutoffmatrix[,2])
 
 for(i in seq_along(dist)){
+  if(userainbow){
+    stringcol <- rainbow(length(dist))[ii]
+  } else {
+    stringcol <- "blue"
+  }
+
   tseq <- seq(from=0,to=f(dist[i]),len=100)
   xy <- stringpoints(y_start=dist[i],initial_string_angle = 0,theta=tseq)
-  points(xy,type='l',col=rainbow(length(dist))[i],lwd=2)
+  points(xy,type='l',col=stringcol,lwd=2)
   
   xy <- stringpoints(y_start=dist[i],initial_string_angle = 0, theta=-tseq)
-  points(xy,type='l',col=rainbow(length(dist))[i],lwd=2)
+  points(xy,type='l',col=stringcol,lwd=2)
+
 }
 
 # mask strings too far from the black hole:

### 2019-02-27 closest_approach2.R

@@ -17,6 +17,7 @@ pars <- c(eel=1)  # dummy
 jj <- 6
 par(xpd=TRUE)
 plot(NULL,asp=1,xlim=c(-jj,jj),ylim=c(-jj,jj),type='l',axes=F,xlab='',ylab='')
+userainbow <- FALSE   # colour of strings (FALSE = all strings blue)
 ## setup ends
 
 
@@ -85,9 +86,16 @@ f <- fun(cutoffmatrix[,1],cutoffmatrix[,2])
 ## Thus f(1.03) = pi, f(1.12) = 2.5
 
 for(i in seq_along(dist)){
+
+  if(userainbow){
+    stringcol <- rainbow(length(dist))[ii]
+  } else {
+    stringcol <- "blue"
+  }
+
   tseq <- seq(from=0,to=f(dist[i]),len=100)
   xy <- stringpoints(y_start=dist[i],initial_string_angle = 0,theta=tseq)
-  points(xy,type='l',col=rainbow(length(dist))[i],lwd=2)
+  points(xy,type='l',col=stringcol,lwd=2)
   
 #  xy <- stringpoints(y_start=dist[i],initial_string_angle = 0, theta=-tseq)
 #  points(xy,type='l',col=rainbow(length(dist))[i])

### 2019-02-27 one_free_end_fixed_EH_intersection.R

@@ -7,13 +7,15 @@ pars <- c(eel=1)  # dummy
 source("usefulfuncs.R")
 
 ## setup:
-jj <- 2.5
+jj <- c(-2.5,2.5)
 par(xpd=T)
 par(pty='m')
-plot(NULL,asp=1,xlim=c(-jj,jj),ylim=c(-jj,jj),type='n',axes=F,xlab='',ylab='')
-
-
+plot(NULL,asp=1,xlim=jj,ylim=jj,type='n',axes=F,xlab='',ylab='')
 dist_from_hole <- 3
+userainbow <- FALSE  # TRUE = rainbow, FALSE = all strings blue
+
+## setup ends
+
 
 ## One string per start angle, each one a different colour:
 start_angles <- c(seq(from=1.18,by=0.03,to=1.56))
@@ -23,6 +25,7 @@ start_angles <- rev(start_angles)
 
 n <- length(start_angles)+1
 cols <- rainbow(n+round(n/7))[-1]  # "[-1]" so the radial string is correctly coloured
+if(!userainbow){  cols[] <- "blue"}
 
 theta_start <- 0
 

### 2019-02-27 radius_of_curvature_switch.R

@@ -43,12 +43,12 @@ for(i in seq_along(dist)){
   wm <-  which.min(rr)
   theta1 <- theta-theta[wm]
   xy1 <- cbind(r*cos(theta1),r*sin(theta1))
-  points(xy1[ thing,],type='l',col='red')   # red
+  points(xy1[ thing,],type='l',col='blue')   # blue
   points(xy1[!thing,],type='l',col='black')
  
   theta2 <- -theta-theta[wm]
   xy2 <- cbind(r*cos(theta2),r*sin(theta2))
-  points(xy2[ thing,],type='l',col='red')
+  points(xy2[ thing,],type='l',col='blue')
   points(xy2[!thing,],type='l',col='black')
 
   print(paste("done ",dist[i],sep = ""))

### 2019-02-28 angle_at_r_equals_2.R
make strings blue by default, not rainbow

@@ -18,10 +18,9 @@ mask <- TRUE    # set to FALSE to see entire string
 jj <- 6.5
 par(pty='s')
 plot(NULL,asp=1,xlim=c(-jj,jj),ylim=c(-jj,jj),type='l',xlab='',ylab='',axes=FALSE)
-
-
-
+userainbow <- FALSE   # colour of strings (FALSE = all strings blue)
 dist_from_hole <- 2
+## setup ends
 
 ## Define start_angles, specifying the angle that the string makes at
 ## (2,0) from a tangent:
@@ -33,6 +32,10 @@ start_angles <- seq(from=-0.8,to=0.0,by=0.015)
 
 n <- length(start_angles)
 cols <- rainbow(n+round(n/7))
+
+  if(!userainbow){
+    cols[] <- "blue"
+  }
 
 theta_start <- 0
 

### 2019-02-28 string_ajp.tex
include strings passing through (2,0)
try to include all the graphs
another diagram

@@
 
+\begin{figure}[p]
+\centering
+\includegraphics{angle_at_r_equals_2.pdf} % created by "angle_at_r_equals_2.R"
+\caption{Strings passing through $(2,0)$}
+\label{strings_r_equals_2}
+\end{figure}
+
+\begin{figure}[p]
+\centering
+\includegraphics{light_start_at_r_equals_2.pdf} % created by "angle_start_at_r_equals_2.R"
+\caption{Null geodesics passing through $(2,0)$, shown for $0\leq\phi\leq 2\pi$}
+\label{light_r_equals_2}
+\end{figure}
+
+\begin{figure}[p] % created by "one_free_end_r_equals_2.R"
+\centering
+\includegraphics[width=170mm]{one_free_end_r_equals_2.pdf}
+\caption{Strings passing through $(2,0)$ tangential to event horizon}
+\label{one_free_end_r_equals_2}
+\end{figure}
+
 \begin{figure}[p] % created by "one_free_end_fixed_EH_intersection.R"

### 2019-03-04 closest_approach.R
slight improvement in appearance

@@ -17,7 +17,7 @@ userainbow <- FALSE   # TRUE = rainbow, FALSE = all strings blue
 ## setup ends
 
 
-dist <- dseq(from=1.076109317, to=7.5, len=30, power=2)
+dist <- dseq(from=1.076109317, to=7.5, len=30, power=1.7)
 
 cutoffmatrix <- matrix(c(
  1.060, pi  ,

### 2019-03-04 light_start_at_r_equals_2.R
docs

@@ -140,7 +140,7 @@ for(initialangletotangent in start_angle){
   points(xy1,col='red',type='l')
 }
 
-## radial ray
+## radial ray (inward and outward):
 segments(x0=0,x1=4,y0=0,col='red')
 
 ## grid:

### 2019-03-04 angle_at_r_equals_2.R
change so that the final (radial) string looks right, as before the delta of the last string was incorrect

@@ -28,7 +28,7 @@ dist_from_hole <- 2
 
 ## One string per start angle, each one a different colour:
 delta_angle <- 0.02
-start_angles <- seq(from=-0.8,to=pi/2,by=delta_angle)
+start_angles <- seq(from=-0.8,to=pi/2,length=100)
 start_angles <- start_angles[-length(start_angles)]
 
 
@@ -104,7 +104,7 @@ for(i in seq_along(start_angles)){
   points(xy,type='l',col=cols[i],lwd=1)
 }
 
-
+## Radial string:
 points(rbind(c(2,0),c(100,0)),col=cols[i+1],type='l')
 
 ## Now the downward strings:

### 2019-03-06 light_start_at_r_equals_1.R
longer strings

@@ -42,41 +42,57 @@ plot(NULL,asp=1,xlim=jj,ylim=jj,type='l',axes=FALSE,xlab='',ylab='')
 
 cutoffmatrix  <- matrix(c(
     -pi/2,0.001,
-    -1.57,0.001,
-    -1.55, 0.001,
-    -1.53, 0.02,
-    -1.51, 0.03,
-    -1.49, 0.1,
-    -1.47, 0.07,
+    -1.57,0.003,
+    -1.56, 0.004,
+    -1.55, 0.02,
+    -1.54, 0.03,
+    -1.53, 0.04,
+    -1.52, 0.045,
+    -1.51, 0.05,
+    -1.50, 0.06,
+    -1.49, 0.08,
+    -1.47, 0.09,
+    -1.46, 0.11,
     -1.45, 0.12,
-    -1.4, 0.1,
-    -1.3, 0.2,
+    -1.4, 0.14,
+    -1.3, 0.24,
     -1.2, 0.3,
     -1.1, 0.5,
-    -1.0, 0.5,
-    -0.9, 0.6,
+    -1.0, 0.6,
+    -0.9, 0.8,
     -0.8, 1.0,
-    -0.7, 1.1,
-    -0.6, 0.8,
-    -0.5, 0.9,
-    -0.4, 1.1,
-    -0.3, 1.2,
-    -0.2, 1.4,
-    -0.1, pi/2 + 0.2,
-    0.0, pi/2 + 0.4,
-    0.1, pi/2 +0.3  ,
-    0.15, pi/2 +0.3 ,
-    0.2, pi/2   ,
-    0.30, pi/2  ,
-    0.4, pi/2  ,
-    0.5, pi/2  ,
-    0.6, pi/2  ,
-    0.7, 1.2  ,
-    0.8, 1.0  ,
-    0.9, 1.0  ,
-    1.0, 1.0  ,
-    1.1, 0.7  ,
-    1.2, 0.6  ,
+    -0.7, 1.2,
+    -0.6, 1.5,
+    -0.55, 1.8,
+    -0.5, 2.1,
+    -0.45, 2.3,
+    -0.4, 3.3,
+    -0.38, 4.5,
+    -0.371, 5.5,
+    -0.37, 2*pi,
+    -0.36, 2*pi,
+    -0.35, 5.5,
+    -0.3, 4.2,
+    -0.2, 3.6,
+    -0.15, 3.2,
[…]

### 2019-03-06 string_ajp.tex
put the figures last
minor tweaks

@@
 strings furnishes another everyday window into general relativity.
-Note in particular that Abramowicz states ``feels with his hands''
+Another possibility, presented by \citet{damour1978} and discussed by
+\citet{thorne1986}, is to pass electric current through such taut
+strings, an option not pursued here.
 
@@
 
+\section{Discussion}
+
+Light inextensible strings under tension adopt non-trivial
+configurations including self-intersecting paths.  The path of such
+strings furnishes an interpretation of the concept ``straight line''
+in general relativity, others being null geodesics and world lines of
+massive objects.  Comparing a world line with that of nearby taut
+strings thus furnishes another insight into path curvature,
+complementing the analysis of Abramowicz~\cite{abramowicz1992}.
+
+In Figure~\ref{closest_approach_non_self_intersecting}, one's
+classical intuition would suggest that the black hole feels a force to
+the left, and should start accelerating.  However, it is not clear by
+what mechanism such a force would be transmitted: the situation is
+steady, so no gravitational waves are produced, and the string is
+inextensible so stores no energy.  Further work might include analysis
+of taut strings near a spinning black hole.
+
+%%Vancouver style references.
+\bibliographystyle{model3-num-names}
+\bibliography{stringrefs}
+
+\section*{Supplementary Material}
+
+Calculations are performed using the R programming language~\citep{rcore2018};
+source code for the figures is available at
+\\
+\\
+\url{https://github.com/RobinHankin/string.git}
+
+\section*{Figures}
+
 \begin{figure}[p] % Created by "closest_approach.R"
@@
 \caption{Self-intersecting strings arranged by closest approach to the
-  event horizon which occurs at $\phi=0$.  
-  shown for $\phi\geq 0$}
+  event horizon which occurs at $\phi=0$.  Strings shown for $\phi\geq
+  0$}
 \label{closest_approach_self_intersecting}
@@
 \includegraphics[width=170mm]{light_closest_approach.pdf}
-\caption{Null geodesics in the Schwarzschild geometry starting~$(r,0)$
-  and initially tangential.  Note the differences between these curves
-  and the taut strings shown elsewhere: unlike taug strings, null
-  geodesics have positive curvature everywhere, may cross the event
-  horizon inwards, and are never tangential to the event horizon}
+\caption{Null geodesics in the Schwarzschild geometry passing
+  through~$(r,0)$, at which point they are tangential.  Note the
+  differences between these curves and the taut strings shown
+  elsewhere: unlike taug strings, null geodesics have positive
+  curvature everywhere, may cross the event horizon inwards, and are
+  never tangential to the event horizon}
 \label{light_closest_approach}
@@
 
-\section{Discussion}
-
-Light inextensible strings under tension adopt non-trivial
-configurations including self-intersecting paths.  The path of such
-strings furnishes an interpretation of the concept ``straight line''
-in general relativity, others being null geodesics and world lines of
-massive objects.  Comparing a world line with that of nearby taut
-strings thus furnishes another insight into path curvature,
-complementing the analysis of Abramowicz~\cite{abramowicz1992}.
-
-In Figure~\ref{closest_approach_non_self_intersecting}, one's
-classical intuition would suggest that the black hole feels a force to
-the left, and should start accelerating.  However, it is not clear by
-what mechanism such a force would be transmitted: the situation is
-steady, so no gravitational waves are produced, and the string is
-inextensible so stores no energy.  Further work might include analysis
-of taut strings near a spinning black hole.
-
-%%Vancouver style references.
-\bibliographystyle{model3-num-names}
-\bibliography{stringrefs}
-
-\section*{Supplementary Material}
-
-Calculations are performed using the R programming language~\citep{rcore2018};
-source code for the figures is available at
-\\
-\\
-\url{https://github.com/RobinHankin/string.git}
-
 \end{document}
-
-%%

### 2019-03-13 light_start_at_r_equals_1.R
minor visual tweaks

@@ -7,7 +7,8 @@ source("usefulfuncs.R") # defines polargrid() etc
 source("usefullightfuncs.R") # defines lightpoints()
 
 r_start <- 1 # starting radius for light ray
-mask <- FALSE    # set to FALSE to see entire geodesic
+mask <- TRUE    # set to TRUE to mask the diagram to a circle; set to
+                # FALSE for debugging (masking is time-consuming)
 size_of_plot <- 3
 
 ## plot setup:
@@ -75,24 +76,37 @@ cutoffmatrix  <- matrix(c(
     -0.3, 4.2,
     -0.2, 3.6,
     -0.15, 3.2,
-    -0.1, 2.9,
-    0.0, pi/2 + 0.7,
-    0.1, pi/2 +0.7  ,
-    0.15, pi/2 +0.4 ,
-    0.2, pi/2+0.3   ,
-    0.30, 2  ,
-    0.4, 2  ,
+    -0.1, 3.1,
+    -0.05, 2.9,
+    0.0, 2.9,
+    0.1,2.9,
+    0.15,2.9,
+    0.2,2.9,
+    
+    0.5, 2.5,
+    0.1, 2.4  ,
+    0.15, 2.4,
+    0.2, 2.5   ,
+    0.25, 2.4  ,
+    0.30, 2.4  ,
+    0.4, 2.2  ,
     0.5, 2.2  ,
-    0.6, 1.8  ,
-    0.7, 1.9  ,
-    0.8, 1.6  ,
-    0.9, 1.4  ,
+
+    0.6, 2.1  ,
+    0.7, 2.1  ,
+    0.8, 2.0  ,
+
+    0.9, 1.8  ,
     1.0, 1.7  ,
-    1.1, 1.6  ,
-    1.2, 1.2  ,
-    1.3, 1.1  , 
-    1.4, 1.0  , 
-    1.5, 1.0  , 
+    1.1, 1.7  ,
+    1.2, 1.6  ,
+    1.3, 1.4  , 
+    1.4, 1.2  , 
+    1.5, 1.0  ,
+    1.51,0.8,
+    1.53,0.7,
+    1.54,0.6,
+    1.55,0.6,
     pi/2 , 0.1   
     )
    ,ncol=2,byrow=TRUE)
@@ -101,12 +115,13 @@ cutoffmatrix  <- matrix(c(
 colnames(cutoffmatrix) <- c("cuts","vals")
[…]

### 2019-03-17 maker.R
add light starting at r=1

@@ -31,6 +31,10 @@ pdf(file="flamm_string.pdf")
 flamm(string=TRUE,addlegend=FALSE,phi=50,theta=50)
 dev.off()
 
+pdf(file="light_start_at_r_equals_1.pdf")
+source("light_start_at_r_equals_1.R")
+dev.off()
+
 pdf(file="light_start_at_r_equals_2.pdf")
 source("light_start_at_r_equals_2.R")
 dev.off()

### 2019-03-17 one_free_end_r_equals_2.R
change strings to blue

@@ -9,13 +9,13 @@ source("usefulfuncs.R")
 ## setup:
 jj <- 2
 par(xpd=TRUE)
-
 plot(NULL,asp=1,xlim=c(-jj,jj),ylim=c(-jj,jj),axes=FALSE,xlab='',ylab='')  # close-up
 
 
 closest_approach <- matrix(0,0,2)  # used for debugging theta_end2 values
 colnames(closest_approach) <- c("start_angle","closest_approach")
 dist_from_hole <- 2
+userainbow <- FALSE
 
 ## one string per start angle, each a different colour; start_angle =
 ## pi/2 is a radial string:
@@ -31,6 +31,7 @@ start_angles <- sort(start_angles,decreasing=TRUE)
 
 n <- length(start_angles)
 cols <- rainbow(n+round(n/7))
+if(!userainbow){  cols[] <- "blue"}
 
 theta_start <- 0
 
@@ -85,6 +86,6 @@ for(i in seq_along(start_angles)){
 }
 
 polargrid(rlab=1.8)
-segments(1,0,2,0,col='red')  # radial string
+segments(1,0,2,0,col=cols[1])  # radial string
 points(dist_from_hole,0,pch=16)
 event_horizon(fill=TRUE)

### 2019-03-18 one_free_end_r_equals_2.R
better integration limits

@@ -40,7 +40,8 @@ theta_end2 <- pi*0.05 + 0*start_angles  # downward/inward
 cutoffmatrix <- matrix(c(
 0.10 ,   pi,
 0.50 ,   pi,
-1.00 ,   pi/2 + 0.1,
+0.70 ,   pi + 0.4,
+1.00 ,   pi/2 + 0.4,
 1.10 ,   pi/2,
 1.19 ,   pi/2 ,
 1.21 ,   pi/3,

### 2019-03-18 string_ajp.tex
move the Flamm diagram to the figures, also make it bigger
reword

@@
 
-\begin{figure}[h!] % Created by "flamm.R"
-\centering
-\includegraphics[width=70mm]{flamm_string.pdf}
-\caption{Perspective view of Flamm's paraboloid with superimposed minimal-length
-  path corresponding to a taut, light string}
-\label{flamm}
-\end{figure}
-
 Solutions to this differential equation correspond to a taut string
@@
 non-self intersecting strings rotated so that the point of inflection
-occurs at~$\phi=0$; we see that the strings have a finite length in
-which convex to the origin and two semi-infinite stretches in which
-the curvature goes the other way.
+occurs at~$\phi=0$; we see that each string has a finite length in
+which it is convex to the origin and two semi-infinite stretches in
+which the curvature is negative.
 
@@
 \section*{Figures}
+
+\begin{figure}[h!] % Created by "flamm.R"
+\centering
+\includegraphics[width=170mm]{flamm_string.pdf}
+\caption{Perspective view of Flamm's paraboloid with superimposed minimal-length
+  path corresponding to a taut, light string}
+\label{flamm}
+\end{figure}
 

### 2019-03-20 string_ajp.tex
rearrangement of figures plus typos

@@
 confined to Flamm's paraboloid~\cite{flamm1916}; see
-figure~\ref{flamm}.
+Figure~\ref{flamm}.
 
@@
 Taking strings for which $r>1$ first,
-figure~\ref{closest_approach_non_self_intersecting} shows a sequence
+Figure~\ref{closest_approach_non_self_intersecting} shows a sequence
 of non-self-intersecting numerical solutions to
@@
 corresponding diagram for self-interesecting strings
-(figure~\ref{light_closest_approach} shows comparable null geodesic
+(Figure~\ref{light_closest_approach} shows comparable null geodesic
 diagram for comparison).  A bisection technique reveals that if the
@@
 
-The case where the string is tangent to the event horizon at~$\phi=0$,
-including a radial string, is shown in
-Figure~\ref{fixed_EH_intersection}.
+The cases where the string is tangent to the event horizon
+at~$\phi=0$, including a radial string, are shown in
+Figures~\ref{one_free_end_r_equals_2} and~\ref{fixed_EH_intersection}
+for strings tangential to or missing the Schwarzschild radius
+respectively.  There are other ways to parametrize the strings; Figure
+\ref{strings_r_equals_2}, for example, shows strings passing
+through~$(2,0)$; Figure~\ref{light_r_equals_2} shows light paths for
+comparison.
 
@@
 \includegraphics[width=170mm]{light_closest_approach.pdf}
-\caption{Null geodesics in the Schwarzschild geometry passing
-  through~$(r,0)$, at which point they are tangential.  Note the
-  differences between these curves and the taut strings shown
-  elsewhere: unlike taug strings, null geodesics have positive
-  curvature everywhere, may cross the event horizon inwards, and are
-  never tangential to the event horizon}
+\caption{Null geodesics in the Schwarzschild geometry, tangential
+  at~$\phi=0$.  Note the differences between these curves and the taut
+  strings shown elsewhere: unlike taug strings, null geodesics have
+  positive curvature everywhere, may cross the event horizon inwards,
+  and are never tangential to the event horizon}
 \label{light_closest_approach}
@@
 \includegraphics{curvature_switch.pdf} % created by "radius_of_curvature_switch.R"
-\caption{Strings arranged by point of curvature switch, which occurs
-  at~$\phi=0$.  Colour changes from black to red at points of
-  infinite~$R$; black indicates convex to the origin and red indicates
-  concave regions}
+\caption{Strings arranged by point at which the radius of curvature
+  switches, which occurs at~$\phi=0$.  Colour changes from black to
+  blue at points of infinite~$R$; black indicates convex to the origin
+  and blue indicates concave regions}
 \label{curvature_switch}
+\end{figure}
+
+\begin{figure}[p] % created by "one_free_end_r_equals_2.R"
+\centering
+\includegraphics[width=170mm]{one_free_end_r_equals_2.pdf}
+\caption{Strings tangential to the event horizon passing through $(2,0)$}
+\label{one_free_end_r_equals_2}
+\end{figure}
+
+\begin{figure}[p] % created by "one_free_end_fixed_EH_intersection.R"
+\centering
+\includegraphics[width=170mm]{one_free_end_fixed_EH_intersection.pdf}
+\caption{Strings tangential to event horizon at $(1,0)$}
+\label{fixed_EH_intersection}
 \end{figure}
@@
 
-\begin{figure}[p] % created by "one_free_end_r_equals_2.R"
-\centering
-\includegraphics[width=170mm]{one_free_end_r_equals_2.pdf}
-\caption{Strings passing through $(2,0)$ tangential to event horizon}
-\label{one_free_end_r_equals_2}
-\end{figure}
-
-\begin{figure}[p] % created by "one_free_end_fixed_EH_intersection.R"
-\centering
-\includegraphics[width=170mm]{one_free_end_fixed_EH_intersection.pdf}
-\caption{Strings tangential to event horizon}
-\label{fixed_EH_intersection}
-\end{figure}
-
 \end{document}

### 2019-03-20 angle_at_r_equals_2.R
minor visual changes to bring into line with the light_start_at_r_equal_2 diagram

@@ -1,6 +1,7 @@
-## Lots of rainbow coloured strings all passing through the same point
-## at different angles.  For production PDF image, source file
-## "maker.R", which creates the best PDF.
+## Lots of strings [blue for the paper but can be rainbow colours],
+## all passing through the same point at different angles.  For
+## production PDF image, source file "maker.R", which creates the best
+## PDF.
 
 ## This script runs standalone but is designed to be called from
 ## maker.R [cf Makefile], which creates the pdf file
@@ -15,9 +16,12 @@ source("usefulfuncs.R")
 mask <- TRUE    # set to FALSE to see entire string
 
 ## setup:
-jj <- 6.5
+
 par(pty='s')
-plot(NULL,asp=1,xlim=c(-jj,jj),ylim=c(-jj,jj),type='l',xlab='',ylab='',axes=FALSE)
+par(xpd=TRUE)
+size_of_plot <- 3
+jj <- c(-size_of_plot,size_of_plot)
+plot(NULL,asp=1,xlim=jj,ylim=jj,type='l',xlab='',ylab='',axes=FALSE)
 userainbow <- FALSE   # colour of strings (FALSE = all strings blue)
 dist_from_hole <- 2
 ## setup ends
@@ -126,7 +130,7 @@ points(dist_from_hole,0,pch=16)
 
 ## Mask strings too far from the black hole:
 if(mask){
-  xy <- cbind(c(7,30,30,7),c(-0.5,-0.5,0.5,0.5))
+  xy <- cbind(c(4,30,30,4),c(-0.5,-0.5,0.5,0.5))
   howmany <- 100  # 100 for production, 10 for testing
   for (theta in seq(from=0,to=2*pi,len=howmany)){
     jjxy <- xy %*% rotmat(theta)

### 2019-03-20 usefulfuncs.R
change default to FALSE for displaying the angles, it looks sooo much better without the labels

@@ -38,7 +38,7 @@ library("deSolve")   # needed for ode()
 ## function cont() returns TRUE if x is in the interval specified
 `cont` <- function(x,interval){ (x-interval[1])*(x-interval[2]) <=0}
 
-`polargrid` <- function(r=1:7,n=12,rlab=max(r)/2, labels=TRUE,...){
+`polargrid` <- function(r=1:7,n=12,rlab=max(r)/2, labels=FALSE,...){
   fish <- c(
       expression(pi/6),    # Note off-by-one error
       expression(pi/3),

### 2019-03-20 angle_at_r_equals_2.R
rationalise masking by defining a function mask() in usefulfuncs.R and using it

@@ -13,7 +13,7 @@ pars <- c(eel=1)  # dummy
 source("usefulfuncs.R")
 
 
-mask <- TRUE    # set to FALSE to see entire string
+domask <- TRUE    # set to FALSE to see entire string
 
 ## setup:
 
@@ -129,14 +129,7 @@ points(dist_from_hole,0,pch=16)
 
 
 ## Mask strings too far from the black hole:
-if(mask){
-  xy <- cbind(c(4,30,30,4),c(-0.5,-0.5,0.5,0.5))
-  howmany <- 100  # 100 for production, 10 for testing
-  for (theta in seq(from=0,to=2*pi,len=howmany)){
-    jjxy <- xy %*% rotmat(theta)
-    polygon(jjxy[,1],jjxy[,2],col='white',border=NA) # white for production
-  }
-}
+if(domask){ mask(4) }
 
 
 polargrid(rlab=6.5)

### 2019-03-20 closest_approach.R

@@ -10,6 +10,7 @@ source("usefulfuncs.R") # defines stringpoints()
 pars <- c(eel=1)  # dummy
 
 ## plot setup:
+domask <- TRUE
 jj <- c(-6,6)
 par(xpd=TRUE)
 plot(NULL,asp=1,xlim=jj,ylim=jj,type='l',axes=FALSE,,xlab='',ylab='')
@@ -70,13 +71,8 @@ for(i in seq_along(dist)){
 
 }
 
-# mask strings too far from the black hole:
-xy <- cbind(c(7,30,30,7),c(-0.5,-0.5,0.5,0.5))
-for (theta in seq(from=0,to=2*pi,len=100)){
-  jjxy <- xy %*% rotmat(theta)
-  polygon(jjxy[,1],jjxy[,2],col='white',border=NA)
-}
-
+ ## mask strings too far from the black hole:
+if(domask){mask(7)}
 
 polargrid(rlab=6.5)
 event_horizon()

### 2019-03-20 closest_approach2.R

@@ -10,7 +10,7 @@
 source("usefulfuncs.R") # defines stringpoints() and dseq()
 
 
-mask <- TRUE
+domask <- TRUE
 pars <- c(eel=1)  # dummy
 
 ## plot setup:
@@ -104,14 +104,7 @@ for(i in seq_along(dist)){
 }
 
 ## mask strings too far from the black hole:
-if(mask){
-  xy <- cbind(c(7,30,30,7),c(-0.5,-0.5,0.5,0.5))
-  for (theta in seq(from=0,to=2*pi,len=100)){
-    jjxy <- xy %*% rotmat(theta)
-    polygon(jjxy[,1],jjxy[,2],col='white',border=NA)
-  }
-}
-
+if(domask){mask(7)}
 
 polargrid(rlab=6.5)
 event_horizon()

### 2019-03-20 light_closest_approach.R

@@ -25,7 +25,7 @@ source("usefulfuncs.R") # defines polargrid() etc
 source("usefullightfuncs.R") # defines lightpoints()
 
 r_start <- 2 # starting radius for light ray
-mask <- TRUE    # set to FALSE to see entire geodesic (TRUE is time-consuming but needed for publication quality)
+domask <- TRUE    # set to FALSE to see entire geodesic (TRUE is time-consuming but needed for publication quality)
 size_of_plot <- 3.5
 
 ## plot setup:
@@ -127,14 +127,7 @@ polargrid()
 
 
 ## Mask strings too far from the black hole:
-if(mask){
-  xy <- cbind(c(4,30,30,4),c(-0.5,-0.5,0.5,0.5))
-  howmany <- 100  # 100 for production, 10 for testing
-  for (theta in seq(from=0,to=2*pi,len=howmany)){
-    jjxy <- xy %*% rotmat(theta)
-    polygon(jjxy[,1],jjxy[,2],col="white",border=NA) # white for production
-  }
-}
+if(domask){ mask(4) }
 
 circ <- function(r,...){
   th <- seq(from=0,to=2*pi,len=300)

### 2019-03-20 light_start_at_r_equals_2.R

@@ -6,7 +6,7 @@ source("usefulfuncs.R") # defines polargrid() etc
 source("usefullightfuncs.R") # defines lightpoints()
 
 r_start <- 2 # starting radius for light ray
-mask <- TRUE    # set to FALSE to see entire geodesic
+domask <- TRUE    # set to FALSE to see entire geodesic
 size_of_plot <- 3
 
 ## plot setup:
@@ -148,15 +148,7 @@ polargrid()
 
 
 ## Mask strings too far from the black hole:
-if(mask){
-  xy <- cbind(c(4,30,30,4),c(-0.5,-0.5,0.5,0.5))
-  howmany <- 100  # 100 for production, 10 for testing
-  for (theta in seq(from=0,to=2*pi,len=howmany)){
-    jjxy <- xy %*% rotmat(theta)
-    polygon(jjxy[,1],jjxy[,2],col="white",border=NA) # white for production
-  }
-}
-
+if(domask){mask(4)}
 
 event_horizon(fill=FALSE)
 points(2,0,pch=16)

### 2019-03-20 radius_of_curvature_switch.R

@@ -5,7 +5,7 @@
 source("usefulfuncs.R") # defines stringpoints()
 
 pars <- c(eel=1)  # dummy
-
+domask <- TRUE
 ## plot setup:
 jj <- 5.5
 par(xpd=TRUE)
@@ -55,11 +55,5 @@ for(i in seq_along(dist)){
 
 }
 
-# mask strings too far from the black hole:
-xy <- cbind(c(7,30,30,7),c(-0.5,-0.5,0.5,0.5))
-for (theta in seq(from=0,to=2*pi,len=100)){
-  jjxy <- xy %*% rotmat(theta)
-  polygon(jjxy[,1],jjxy[,2],col='white',border=NA)
-}
-
+if(domask){mask(7)}
 event_horizon()

### 2019-03-20 usefulfuncs.R

@@ -102,3 +102,12 @@ library("deSolve")   # needed for ode()
   f <- function(r){2*r^2/3}  # radius of curvature of a ray at radius r moving tangentially
   circ(c(x-f(r),0),f(r),...)  # shows an osculating circle
 }
+
+`mask` <- function(radius,n=100,...){  # n=100 for production, 10 for testing
+  BIG <- 30
+  xy <- cbind(c(radius,BIG,BIG,radius),c(-0.5,-0.5,0.5,0.5))
+  for (theta in seq(from=0,to=2*pi,len=n)){
+    jjxy <- xy %*% rotmat(theta)
+    polygon(jjxy[,1],jjxy[,2],col='white',border=NA,...) # white for production
+  }
+}

### 2019-03-20 angle_at_r_equals_2.R
remove unused pars variable

@@ -8,11 +8,7 @@
 ## "angle_at_r_equals_2.pdf".
 
 
-pars <- c(eel=1)  # dummy
-
 source("usefulfuncs.R")
-
-
 domask <- TRUE    # set to FALSE to see entire string
 
 ## setup:

### 2019-03-20 closest_approach.R

@@ -7,8 +7,6 @@
 
 source("usefulfuncs.R") # defines stringpoints()
 
-pars <- c(eel=1)  # dummy
-
 ## plot setup:
 domask <- TRUE
 jj <- c(-6,6)

### 2019-03-20 closest_approach2.R

@@ -11,7 +11,7 @@ source("usefulfuncs.R") # defines stringpoints() and dseq()
 
 
 domask <- TRUE
-pars <- c(eel=1)  # dummy
+
 
 ## plot setup:
 jj <- 6

### 2019-03-20 flamm.R

@@ -9,9 +9,6 @@
 library("plot3D")
 source("usefulfuncs.R")
 
-pars <- c(eel=1) # dummy
-
-
 flamm <- function(string=TRUE, drawlegend=FALSE, ...){
 
   ## string, drawlengend: Boolean specifying whether or not to draw

### 2019-03-20 one_free_end_r_equals_2.R

@@ -2,8 +2,6 @@
 ## maker.R [cf Makefile], which creates the pdf file
 ## "one_free_end_r_equals_2.pdf"
 
-pars <- c(eel=1)  # dummy for the ODE solver
-
 source("usefulfuncs.R")
 
 ## setup:

### 2019-03-20 radius_of_curvature_switch.R

@@ -4,7 +4,6 @@
 
 source("usefulfuncs.R") # defines stringpoints()
 
-pars <- c(eel=1)  # dummy
 domask <- TRUE
 ## plot setup:
 jj <- 5.5

### 2019-03-20 usefulfuncs.R
more tidyup (fun() is now replaced with builtin approxfun())

@@ -58,7 +58,7 @@ library("deSolve")   # needed for ode()
   jj <- seq(from=0,to=2*pi,len=100)
   xy <- cbind(cos(jj),sin(jj))
   for(i in r){ points(xy*i,type='l',lty=3,lwd=0.5, col='gray', ...) }
-  angs <- seq_len(n)*2*pi/n
+  angs <- seq_len(n)*2*pi/n  # last element is 0 (mod 2*pi)
   for(a in seq_along(angs)){
       segments(
           x0=0, y0=0,
@@ -80,9 +80,6 @@ library("deSolve")   # needed for ode()
   ## setup ends
 }
 
-`fun` <- function(cuts,vals){
-  return(function(x){vals[findInterval(x, c(-Inf, cuts))]})
-}
 
 `rotmat` <- function(theta){
     matrix(c(cos(theta),sin(theta),-sin(theta),cos(theta)),2,2)

### 2019-03-21 string_ajp.tex
slight rearrangement of diagrams plus some typos

@@
 Here I solve equation~\ref{rdashdash} using the \verb+deSolve+ package
-for ordinary differential equations~\cite{soetart2010}.  Solutions
-fall into two distinct classes: those with $r>1$ at all points, and
-those which attain $r=1$, in which case the string is tangential to
-the event horizon, unless it is radial.
+for ordinary differential equations~\cite{soetart2010}.  Non-radial
+solutions fall into two distinct classes: those with $r>1$ at all
+points, and those which attain $r=1$, in which case the string is
+tangential to the event horizon.
 
-Taking strings for which $r>1$ first,
+We first consider strings for which $r>1$ everywhere (that is,
+completely exterior strings).
 Figure~\ref{closest_approach_non_self_intersecting} shows a sequence
@@
 corresponding diagram for self-interesecting strings
-(Figure~\ref{light_closest_approach} shows comparable null geodesic
-diagram for comparison).  A bisection technique reveals that if the
-closest approach is greater than a critical value of~$\simeq
-1.076109317(1)$, the string does not self-intersect; if the closest
-approach is less than this value, then the string crosses itself
-at~$\phi=\pi/2$.  As discussed above, if $\left(r'\right)^2=2r(r-1)$,
-the string has a point of inflection at which point the radius of
-curvature becomes infinite.  Figure~\ref{curvature_switch} shows
-non-self intersecting strings rotated so that the point of inflection
-occurs at~$\phi=0$; we see that each string has a finite length in
-which it is convex to the origin and two semi-infinite stretches in
-which the curvature is negative.
+(Figure~\ref{light_closest_approach} shows equivalent null geodesics
+for comparison).  A bisection technique reveals that if the closest
+approach is greater than a critical value of~$\simeq 1.076109317(1)$,
+the string does not self-intersect; if the closest approach is less
+than this value, then the string crosses itself at~$\phi=\pi/2$.  As
+discussed above, if $\left(r'\right)^2=2r(r-1)$, the string has a
+point of inflection at which point the radius of curvature becomes
+infinite.  Figure~\ref{curvature_switch} shows non-self intersecting
+strings rotated so that the point of inflection occurs at~$\phi=0$; we
+see that each string has a finite length in which it is convex to the
+origin and two semi-infinite stretches in which the curvature is
+negative.
 
-The cases where the string is tangent to the event horizon
-at~$\phi=0$, including a radial string, are shown in
-Figures~\ref{one_free_end_r_equals_2} and~\ref{fixed_EH_intersection}
-for strings tangential to or missing the Schwarzschild radius
-respectively.  There are other ways to parametrize the strings; Figure
-\ref{strings_r_equals_2}, for example, shows strings passing
-through~$(2,0)$; Figure~\ref{light_r_equals_2} shows light paths for
+There are other ways to parametrize completely exterior strings;
+Figure \ref{strings_r_equals_2}, for example, shows strings passing
+through~$(2,0)$.  Figure~\ref{light_r_equals_2} shows light paths for
 comparison.
+
+The cases where the string attains $r=1$ are tangent to the event
+horizon (or, exceptionally, purely radial).
+Figure~\ref{fixed_EH_intersection} shows strings tangential to the
+event horizon at~$\phi=0$ and Figure~\ref{one_free_end_r_equals_2}
+shows tangential strings passing through $(2,0)$.
 
@@
 
-\begin{figure}[p] % created by "one_free_end_r_equals_2.R"
-\centering
-\includegraphics[width=170mm]{one_free_end_r_equals_2.pdf}
-\caption{Strings tangential to the event horizon passing through $(2,0)$}
-\label{one_free_end_r_equals_2}
-\end{figure}
-
-\begin{figure}[p] % created by "one_free_end_fixed_EH_intersection.R"
-\centering
-\includegraphics[width=170mm]{one_free_end_fixed_EH_intersection.pdf}
-\caption{Strings tangential to event horizon at $(1,0)$}
-\label{fixed_EH_intersection}
-\end{figure}
-
 \begin{figure}[p]
@@
 
+\begin{figure}[p] % created by "one_free_end_fixed_EH_intersection.R"
+\centering
+\includegraphics[width=170mm]{one_free_end_fixed_EH_intersection.pdf}
+\caption{Strings tangential to event horizon at $(1,0)$}
+\label{fixed_EH_intersection}
+\end{figure}
+
+\begin{figure}[p] % created by "one_free_end_r_equals_2.R"
+\centering
+\includegraphics[width=170mm]{one_free_end_r_equals_2.pdf}
+\caption{Strings tangential to the event horizon passing through $(2,0)$}
+\label{one_free_end_r_equals_2}
+\end{figure}
+
 \end{document}

### 2019-03-21 string_ajp.tex
typos

@@
 strings minimize their proper length, given by integrating the spatial
-component of the Schwarzshild metric along the string.  The path is
+component of the Schwarzschild metric along the string.  The path is
 calculated by numerical solution to the Euler-Lagrange equations.  If
@@
 
-\subsection{Nonradial string}
+\section{Nonradial string}
 Integrating along a path between two
@@
 configurations including self-intersecting paths.  The path of such
-strings furnishes an interpretation of the concept ``straight line''
-in general relativity, others being null geodesics and world lines of
-massive objects.  Comparing a world line with that of nearby taut
-strings thus furnishes another insight into path curvature,
-complementing the analysis of Abramowicz~\cite{abramowicz1992}.
+strings interprets the concept ``straight line'' in general
+relativity, others being null geodesics and world lines of massive
+objects.  Comparing a world line with that of nearby taut strings thus
+furnishes another insight into path curvature, complementing the
+analysis of Abramowicz~\cite{abramowicz1992}.
 

### 2019-05-21 light_start_at_r_equals_1.R
bug where cuts of the cutoffmatrix were out-of-sequence, giving a repeated values error

@@ -78,15 +78,12 @@ cutoffmatrix  <- matrix(c(
     -0.15, 3.2,
     -0.1, 3.1,
     -0.05, 2.9,
+
     0.0, 2.9,
-    0.1,2.9,
+    0.1, 2.9,
     0.15,2.9,
-    0.2,2.9,
+    0.2, 2.0,
     
-    0.5, 2.5,
-    0.1, 2.4  ,
-    0.15, 2.4,
-    0.2, 2.5   ,
     0.25, 2.4  ,
     0.30, 2.4  ,
     0.4, 2.2  ,

### 2019-05-21 string_jpam.tex
fixes #5
fixes #7
fixes #8
fixes #11

--- 2019-03-21 string_ajp.tex
@@
   \sqrt{r(r-1)} +\frac{1}{2}\log\left(
-  \frac{r+\sqrt{r-1}}{r-\sqrt{r-1}}\right)
+  \frac{\sqrt{r}+\sqrt{r-1}}{\sqrt{r}-\sqrt{r-1}}\right)
   \right|_{r_1}^{r_2}
@@
 \begin{equation}
-  \ell  =\frac{1}{2} + \log(2) + w + \frac{1}{8}w^{-1} + O(w^{-2}),
+  \ell = w +  \frac{1}{2}\log w + \frac{1}{2} + \log(2)  +  \frac{1}{8}w^{-1} + O(w^{-2}),
 \end{equation}
@@
 over $1.19$ Schwarzschild radii, sensible at large distances from the
-event horizon.  This might suggest that radial strings have longer
-proper length than expected.
+event horizon.  Radial strings have proper length longer than the
+Schwarzschild radial coordinate difference $w$.
 
@@
 solutions with such a point: there exists a pair~$\left(r,\phi\right),
-r>1$ at which~$r'(\phi)=0$.  By Rolle's theorem, this point must be
-unique, as~$r''(\phi)>0$, and be the closest approach of the string to
-the event horizon.
+r>1$ at which~$r'(\phi)=0$.  This point must be unique,
+as~$r''(\phi)>0$, and be the closest approach of the string to the
+event horizon.
 

### 2019-05-21 string_jpam.tex
fixes #14

@@
 
-Solutions to this differential equation correspond to a taut string
+\noindent 
+Solutions to equation~\ref{rdashdash} correspond to a taut string
 confined to Flamm's paraboloid~\cite{flamm1916}; see
-Figure~\ref{flamm}.
+Figure~\ref{flamm}.  It is interesting to compare
+equation~\ref{rdashdash} with the corresponding equation for null
+geodesics, which is usually given~\cite{wald} in terms of $u=r^{-1}$
+as $u''=3u^2/2-u$ but is equivalent
+to~$r''\left(\phi\right)=r-\frac{3}{2}+\frac{2r'}{r}$.
 
-There does not appear to be a simple analytical solution but
-nevertheless we may make several observations.
-Equation~\ref{rdashdash} is a second-order nonlinear ordinary
-differential equation and thus has two constants of integration.
-There are two qualitatively different types of solutions: those with a
-tangential point ($r'=0,r>1$), and those without.  Consider first
-solutions with such a point: there exists a pair~$\left(r,\phi\right),
-r>1$ at which~$r'(\phi)=0$.  This point must be unique,
-as~$r''(\phi)>0$, and be the closest approach of the string to the
-event horizon.
+There does not appear to be a simple analytical solution to
+equation~\ref{rdashdash} but nevertheless we may make several
+observations.  Equation~\ref{rdashdash} is a second-order nonlinear
+ordinary differential equation and thus has two constants of
+integration.  There are two qualitatively different types of
+solutions: those with a tangential point ($r'=0,r>1$), and those
+without.  Consider first solutions with such a point: there exists a
+pair~$\left(r,\phi\right), r>1$ at which~$r'(\phi)=0$.  This point
+must be unique, as~$r''(\phi)>0$, and be the closest approach of the
+string to the event horizon.
 

### 2019-05-21 string_jpam.tex
fixes #15

@@
 is to the singularity is tangential; if the distance of closest
-approach is less than about $1.076109317$ Schwarzschild radii, the
+approach is less than about $1.0761$ Schwarzschild radii, the
 string self-intersects.  If the string attains $r=2GM/c^2$, then it
@@
 for comparison).  A bisection technique reveals that if the closest
-approach is greater than a critical value of~$\simeq 1.076109317(1)$,
+approach is greater than a critical value of about~$\simeq 1.0761$,
 the string does not self-intersect; if the closest approach is less

### 2019-05-21 string_jpam.tex
fixes #16
fixes #13

@@
 If, conversely, $r'(\phi)>0$ whenever~$r>1$, then the string has only
-one free end; the other is tangential to the event horizon, except in
-the exceptional case of constant~$\phi$ when the string is radial
+one free end; the other is tangential to the event horizon (in the
+sense that ${\displaystyle \lim_{r\longrightarrow 1^+}r'=0}$), except
+in the exceptional case of constant~$\phi$ when the string is radial
 for~$r>1$.
@@
 \includegraphics[width=170mm]{closest_approach_selfintersecting.pdf}
-\caption{Self-intersecting strings arranged by closest approach to the
-  event horizon which occurs at $\phi=0$.  Strings shown for $\phi\geq
-  0$}
+\caption{Self-intersecting strings shown for $\phi\geq 0$.  Closest
+  approach to the event horizon occurs at $\phi=0$ and
+  $r\in\left(1,1.0761\right)$; strings are symmetrical about $\phi=0$}
 \label{closest_approach_self_intersecting}

### 2019-05-21 string_jpam.tex
clarification

@@
 
-If, conversely, $r'(\phi)>0$ whenever~$r>1$, then the string has only
-one free end; the other is tangential to the event horizon (in the
-sense that ${\displaystyle \lim_{r\longrightarrow 1^+}r'=0}$), except
-in the exceptional case of constant~$\phi$ when the string is radial
-for~$r>1$.
+If, conversely, $r'(\phi)\neq 0$ whenever~$r>1$, then the string has
+only one free end; the other is tangential to the event horizon (in
+the sense that ${\displaystyle \lim_{r\longrightarrow 1^+}r'=0}$),
+except in the degenerate case of constant~$\phi$ when the string is
+radial for~$r>1$.
 

### 2019-05-22 string_jpam.tex
fixes #12
fixes #10

@@
 adopts, at least locally, a catenary configuration at closest
-approach~\cite{nguyen2007}.  Such a configuration is manifestly convex
-to the singularity, in contrast to light strings which are concave.
+approach~\cite{nguyen2007}.  A heavy string is manifestly convex to
+the singularity in the sense that $R<0$ if $\phi'=0$, in contrast to
+light strings which have $R>0$.
 
@@
 In Figure~\ref{closest_approach_non_self_intersecting}, one's
-classical intuition would suggest that the black hole feels a force to
-the left, and should start accelerating.  However, it is not clear by
-what mechanism such a force would be transmitted: the situation is
-steady, so no gravitational waves are produced, and the string is
-inextensible so stores no energy.  Further work might include analysis
-of taut strings near a spinning black hole.
+classical intuition, and d'Alembert's principle, would suggest that
+the black hole should accelerate to the left.  Note that the strings
+are already ``straight'' in sense that they are minimal length; so
+altering the tension would not change their path.  However, it is not
+clear by what mechanism the required force would be transmitted: the
+situation is steady, so no gravitational waves are produced, and the
+string is inextensible so stores no energy.  Further work might
+include analysis of taut strings near a spinning black hole.
 

### 2019-05-22 string_jpam.tex
further clarification pursuant to #16

@@
 \caption{Self-intersecting strings shown for $\phi\geq 0$.  Closest
-  approach to the event horizon occurs at $\phi=0$ and
-  $r\in\left(1,1.0761\right)$; strings are symmetrical about $\phi=0$}
+  approach to the event horizon occurs at $\phi=0$ (3 o'clock) and
+  $r\in\left(1,1.0761\right)$; strings are symmetrical about $\phi=0$
+  but for clarity only $\phi\geq 0$ is shown}
 \label{closest_approach_self_intersecting}

### 2019-05-22 light_start_at_r_equals_1.R
comment out printing statements

@@ -123,7 +123,7 @@ start_angle <- start_angle[-c(1,length(start_angle))]
 for(initialangletotangent in start_angle){
 
   final_phi <- min(fmax(initialangletotangent),2*pi)
-  print(c(initialangletotangent,final_phi))
+#  print(c(initialangletotangent,final_phi))
   xy1 <-
     stringu(
         r_start=r_start,  # r=3/2 is a circular (but unstable) orbit

### 2019-05-22 light_start_at_r_equals_2.R

@@ -129,7 +129,7 @@ start_angle <- start_angle[-c(1,length(start_angle))]
 for(initialangletotangent in start_angle){
 
   final_phi <- min(fmax(initialangletotangent),2*pi)
-  print(c(initialangletotangent,final_phi))
+#  print(c(initialangletotangent,final_phi))
   xy1 <-
     stringu(
         r_start=r_start,  # r=3/2 is a circular (but unstable) orbit

### 2019-05-22 one_free_end_fixed_EH_intersection.R

@@ -90,7 +90,7 @@ for(i in seq_along(start_angles)){
           point_of_closest_approach[2]/
           point_of_closest_approach[1]
       )
-  print(angle_of_closest_approach)
+#  print(angle_of_closest_approach)
   if(angle_of_closest_approach>0){
       angle_of_closest_approach <-
           angle_of_closest_approach -pi

### 2019-05-22 radius_of_curvature_switch.R

@@ -27,7 +27,7 @@ f <- approxfun(cuts,vals)
 ## Thus f(1.03) = pi, f(1.12) = 2.5
 
 for(i in seq_along(dist)){
-  print(paste("starting ",dist[i],sep = ""))
+#  print(paste("starting ",dist[i],sep = ""))
   tseq <- seq(from=0,to=f(dist[i]),len=1000)
   bh <- stringpoints(y_start=dist[i],initial_string_angle = 0,theta=tseq,give=TRUE)
   r <- bh[,2]
@@ -50,7 +50,7 @@ for(i in seq_along(dist)){
   points(xy2[ thing,],type='l',col='blue')
   points(xy2[!thing,],type='l',col='black')
 
-  print(paste("done ",dist[i],sep = ""))
+#  print(paste("done ",dist[i],sep = ""))
 
 }
 

### 2019-05-22 usefulfuncs.R

@@ -21,7 +21,7 @@ library("deSolve")   # needed for ode()
   pars <- c(eel=1) # [dummy]
   ## stringpoints(2,0,theta=seq(from=0,to=pi/2,len=100))
 
-  print(c(y_start,initial_string_angle))
+#  print(c(y_start,initial_string_angle))
   yini <- c(y=y_start, ydash=y_start*tan(initial_string_angle))
 
   bh <- ode(yini,theta,taut_string,pars,rtol=1e-6)

### 2019-05-22 string_jpam.tex
units
Branch to address issue #6
put both sets of units side-by-side

--- 2019-05-22 string_jpam.tex
@@
 \begin{equation}\label{schwarzschild}
-ds^2= -dt^2\left(1-1/r\right) +\frac{dr^2}{1-1/r} + r^2\left(d\theta^2 + \sin^2\theta d\phi^2\right),
+ds^2= -dt^2\left(1-2GM/r\right) +\frac{dr^2}{1-2GM/r} + r^2\left(d\theta^2 + \sin^2\theta d\phi^2\right),
 \end{equation}
 
-\noindent where units in which~$2GM=c=1$ are used, has been known for
+\noindent where units in which~$c=1$ are used, has been known for
 over a century~\cite{schwarzschild1916}.  Here I consider the spatial
-component~$ds^2= dr^2/\left(1-1/r\right) + r^2d\phi^2$, confined to
-equatorial points~($\theta=\pi/2$); we understand~$1\leq r<\infty$
+component~$ds^2= dr^2/\left(1-2GM/r\right) + r^2d\phi^2$, confined to
+equatorial points~($\theta=\pi/2$); we understand~$2GM\leq r<\infty$
 throughout.  A circular string at constant~$r$ will, trivially, have
@@
 proper length $L\left(r_1,r_2\right)$ of a string stretching
-from~$r=r_1\geq 1$ to~$r_2\geq r_1$ at constant $\theta,\phi$:
+from~$r=r_1\geq 2GM$ to~$r_2\geq r_1$ at constant $\theta,\phi$:
 
-\begin{equation}\label{radial_string_length}
+\begin{equation}\label{radial_string_length_properunits}
   \int_{r=r_1}^{r_2}\frac{dr}{\sqrt{1-1/r}}=
@@
 
+\begin{equation}\label{radial_string_length}
+  \int_{r=r_1}^{r_2}\frac{dr}{\sqrt{1-2GM/r}}=
+  \left.
+  \sqrt{r(r-2GM)} +\frac{1}{2}\log\left(
+  \frac{\sqrt{r}+\sqrt{r-2GM}}{\sqrt{r}-\sqrt{r-2GM}}\right)
+  \right|_{r_1}^{r_2}
+  \end{equation}
+
 Observe that integration may be performed with a lower limit
-of~$r_1=1$ with no difficulty, even though the radial component of the
+of~$r_1=2GM$ with no difficulty, even though the radial component of the
 Schwarzschild metric approaches infinity there.
@@
 some peculiar features.  Firstly, consider the proper length from
-$r=1$ to $r=1+w$ where $w>0, \left|w\right|\ll 1$.  Then
+$r=2GM$ to $r=2GM+w$ where $w>0, \left|w\right|\ll 1$.  Then
 
@@
 
-This would indicate that, close to the Schwarzschild radius $r=1$,
+\begin{equation}
+  \ell=\int_{r=2GM}^{r=2GM+w}ds
+  =2\sqrt{2GM}w^{1/2} + \frac{w^{3/2}}{3\sqrt{2GM}} + O\left({w}^{5/2}\right)
+\end{equation}
+
+This would indicate that, close to the Schwarzschild radius $r=2GM$,
 large (but finite) amounts of string are required to connect two
 points with slightly differing radial coordinates.  Passing to the
-case $w\gg 1$ we have an asymptotic expansion that begins
+case $w\gg 2GM$ we have an asymptotic expansion that begins
 
@@
 
+\begin{equation}
+  \ell = w +  GM\log\left(\frac{w}{2GM}\right) + GM(1+ 2\log 2)+\frac{GM}{4}w^{-1} + O(w^{-2}),
+\end{equation}
+
 \noindent suggesting that there is ``extra radial space'', of length
-over $1.19$ Schwarzschild radii, sensible at large distances from the
-event horizon.  Radial strings have proper length longer than the
+over $1.19\cdot 2GM$, sensible at large distances from the event
+horizon.  Radial strings have proper length longer than the
 Schwarzschild radial coordinate difference $w$.
@@
 
+\begin{equation}
+%  \int_{p_1}^{p_2}\sqrt{\frac{\left(r'\right)^2}{1-2M/r} + r^2}d\phi=
+  \int_{p_1}^{p_2}\left(\frac{\left(r'\right)^2}{1-2GM/r} + r^2\right)^\frac{1}{2}d\phi=
+  \int_{p_1}^{p_2}F\left(r,r'\right)d\phi
+\end{equation}
+
 \noindent where dashes denote differentiation with respect to~$\phi$.  The
@@
 
+\begin{equation}\label{rdashdash_properunits}
+  r''\left(\phi\right) =
+  (r-1) + \frac{(4r-3)\left(r'\right)^2}{2r\left(r-1\right)},\qquad r>1.
+\end{equation}
 \begin{equation}\label{rdashdash}
   r''\left(\phi\right) =
-  (r-1) + \frac{(4r-3)\left(r'\right)^2}{2r\left(r-1\right)},\qquad r>1.
+  (r-2GM) + \frac{(4r-6GM)\left(r'\right)^2}{2r\left(r-2GM\right)},\qquad r>2GM.
 \end{equation}
@@
 equation~\ref{rdashdash} with the corresponding equation for null
-geodesics, which is usually given~\cite{wald} in terms of $u=r^{-1}$
-as $u''=3u^2/2-u$ but is equivalent
-to~$r''\left(\phi\right)=r-\frac{3}{2}+\frac{2r'}{r}$.
+geodesics, which is usually given~\cite{wald} in terms of $u=2GM/r$ as
+$u''=3u^2/2-u$, but is equivalent
+to~$r''\left(\phi\right)=r-3GM+4GMr'/r$.
 
@@
 integration.  There are two qualitatively different types of
-solutions: those with a tangential point ($r'=0,r>1$), and those
+solutions: those with a tangential point ($r'=0,r>2GM$), and those
 without.  Consider first solutions with such a point: there exists a
-pair~$\left(r,\phi\right), r>1$ at which~$r'(\phi)=0$.  This point
+pair~$\left(r,\phi\right), r>2GM$ at which~$r'(\phi)=0$.  This point
 must be unique, as~$r''(\phi)>0$, and be the closest approach of the
@@
 
-\begin{equation}\label{roc}
+\begin{equation}\label{roc_properunits}
   R = \frac{
@@
 
+\begin{equation}\label{roc}
+  R = \frac{
+    \left(r^2 + \left(r'\right)^2\right)^\frac{3}{2}
+  }{
+%    \left|
+    r^2 + 2\left(r'\right)^2-rr''
+%    \right|
+  }
+  =
+  \frac{
+    \left(r^2 + \left(r'\right)^2\right)^\frac{3}{2}
+  }{
+    r-\frac{\left(r'\right)^2}{2\left(r-2GM\right)}
+  }\qquad\mbox{along a taut string}
+\end{equation}
+
 \noindent which would give~$R$ as a function of~$\left(r,r'\right)$.
 Considering~$r'(\phi)=0$ shows that if a string is tangential at a
-point with~$r_0>1$, then it has a radius of curvature of~$r_0^2$ at
-that point.  Any circular path, including a circular orbit at~$r=r_0$,
-has a radius of curvature of $r_0<r_0^2$, showing that circular orbits
-cannot follow a taut string at closest approach.  Also observe that
-null geodesics have a radius of curvature of $2r_0^2/3 < r_0^2$,
-showing that light cannot follow a taut string at closest approach
-either.  The string has a point of inflection
-if~$\left(r'\right)^2=2r(r-1)$.
+point with~$r_0>2GM$, then it has a radius of curvature of~$r_0^2/2GM$
+at that point.  Any circular path, including a circular orbit
+at~$r=r_0$, has a radius of curvature of $r_0<r_0^2/2GM$, showing that
+circular orbits cannot follow a taut string at closest approach.  Also
+observe that null geodesics have a radius of curvature of $r_0^2/3GM <
+r_0^2/2GM$, showing that light cannot follow a taut string at closest
+approach either.  The string has a point of inflection
+if~$\left(r'\right)^2=2r(r-2GM)$.
 
@@
 
-If, conversely, $r'(\phi)\neq 0$ whenever~$r>1$, then the string has
+If, conversely, $r'(\phi)\neq 0$ whenever~$r>2GM$, then the string has
 only one free end; the other is tangential to the event horizon (in
@@
 except in the degenerate case of constant~$\phi$ when the string is
-radial for~$r>1$.
+radial for~$r>2GM$.
 
@@
 for ordinary differential equations~\cite{soetart2010}.  Non-radial
-solutions fall into two distinct classes: those with $r>1$ at all
-points, and those which attain $r=1$, in which case the string is
+solutions fall into two distinct classes: those with $r>2GM$ at all
+points, and those which attain $r=2GM$, in which case the string is
 tangential to the event horizon.
 
-We first consider strings for which $r>1$ everywhere (that is,
+We first consider strings for which $r>2GM$ everywhere (that is,
 completely exterior strings).
@@
 for comparison).  A bisection technique reveals that if the closest
-approach is greater than a critical value of about~$\simeq 1.0761$,
-the string does not self-intersect; if the closest approach is less
-than this value, then the string crosses itself at~$\phi=\pi/2$.  As
-discussed above, if $\left(r'\right)^2=2r(r-1)$, the string has a
+approach is greater than a critical value of about~$\simeq 1.0761\cdot
+2GM$, the string does not self-intersect; if the closest approach is
+less than this value, then the string crosses itself at~$\phi=\pi/2$.
+As discussed above, if $\left(r'\right)^2=2r(r-2GM)$, the string has a
 point of inflection at which point the radius of curvature becomes
@@
 Figure \ref{strings_r_equals_2}, for example, shows strings passing
-through~$(2,0)$.  Figure~\ref{light_r_equals_2} shows light paths for
+through~$(4GM,0)$.  Figure~\ref{light_r_equals_2} shows light paths for
 comparison.
 
-The cases where the string attains $r=1$ are tangent to the event
+The cases where the string attains $r=2GM$ are tangent to the event
 horizon (or, exceptionally, purely radial).
@@
 event horizon at~$\phi=0$ and Figure~\ref{one_free_end_r_equals_2}
-shows tangential strings passing through $(2,0)$.
+shows tangential strings passing through $(4GM,0)$.
 
@@
 \caption{Self-intersecting strings shown for $\phi\geq 0$.  Closest
-  approach to the event horizon occurs at $\phi=0$ and
-  $r\in\left(1,1.0761\right)$; strings are symmetrical about $\phi=0$}
+  approach to the event horizon occurs at $\phi=0$ (3 o'clock) and
+  $r/2GM\in\left(1,1.0761\right)$; strings are symmetrical about $\phi=0$
+  but for clarity only $\phi\geq 0$ is shown}
 \label{closest_approach_self_intersecting}
@@
 \includegraphics{angle_at_r_equals_2.pdf} % created by "angle_at_r_equals_2.R"
-\caption{Strings passing through $(2,0)$}
+\caption{Strings passing through $(4GM,0)$}
 \label{strings_r_equals_2}
@@
 \includegraphics{light_start_at_r_equals_2.pdf} % created by "angle_start_at_r_equals_2.R"
-\caption{Null geodesics passing through $(2,0)$, shown for $0\leq\phi\leq 2\pi$}
+\caption{Null geodesics passing through $(4GM,0)$, shown for $0\leq\phi\leq 2\pi$}
 \label{light_r_equals_2}
@@
 \includegraphics[width=170mm]{one_free_end_r_equals_2.pdf}
-\caption{Strings tangential to the event horizon passing through $(2,0)$}
+\caption{Strings tangential to the event horizon passing through $(4GM,0)$}
 \label{one_free_end_r_equals_2}

### 2019-05-22 string_jpam.tex
make w dimensionless

@@
 the string is entirely outside the event horizon, its closest approach
-is to the singularity is tangential; if the distance of closest
-approach is less than about $1.0761$ Schwarzschild radii, the
-string self-intersects.  If the string attains $r=2GM/c^2$, then it
+to the singularity is tangential; if the distance of closest
+approach is less than about $1.0761\cdot 2GM$, the
+string self-intersects.  If the string attains $r=2GM$, then it
 will in general be tangential to the event horizon.  Light taut
@@
 
-\begin{equation}\label{radial_string_length_properunits}
-  \int_{r=r_1}^{r_2}\frac{dr}{\sqrt{1-1/r}}=
-  \left.
-  \sqrt{r(r-1)} +\frac{1}{2}\log\left(
-  \frac{\sqrt{r}+\sqrt{r-1}}{\sqrt{r}-\sqrt{r-1}}\right)
-  \right|_{r_1}^{r_2}
-  \end{equation}
+% \begin{equation}\label{radial_string_length_properunits}
+%   \int_{r=r_1}^{r_2}\frac{dr}{\sqrt{1-1/r}}=
+%   \left.
+%   \sqrt{r(r-1)} +\frac{1}{2}\log\left(
+%   \frac{\sqrt{r}+\sqrt{r-1}}{\sqrt{r}-\sqrt{r-1}}\right)
+%   \right|_{r_1}^{r_2}
+%   \end{equation}
 
@@
 some peculiar features.  Firstly, consider the proper length from
-$r=2GM$ to $r=2GM+w$ where $w>0, \left|w\right|\ll 1$.  Then
+$r=2GM$ to $r=2GM(1+w)$ where $w>0, \left|w\right|\ll 1$.  Then
 
-\begin{equation}
+ \begin{equation}\label{ell_properunits}
    \ell=\int_{r=1}^{r=1+w}ds
@@
 
-\begin{equation}
-  \ell=\int_{r=2GM}^{r=2GM+w}ds
-  =2\sqrt{2GM}w^{1/2} + \frac{w^{3/2}}{3\sqrt{2GM}} + O\left({w}^{5/2}\right)
+ \begin{equation}\label{ell}
+   \ell =
+   \int_{r=2GM}^{r=2GM(1+w)}ds
+   =2GM\left(2\sqrt{w} + \frac{1}{3}w^{3/2} + O(w^{5/2})\right).
  \end{equation}
@@
 points with slightly differing radial coordinates.  Passing to the
-case $w\gg 2GM$ we have an asymptotic expansion that begins
+case $w\gg 1$ we have an asymptotic expansion that begins
 
-\begin{equation}
+\begin{equation}\label{asymptotic_ell_properunits}
   \ell = w +  \frac{1}{2}\log w + \frac{1}{2} + \log(2)  +  \frac{1}{8}w^{-1} + O(w^{-2}),
@@
 
-\begin{equation}
-  \ell = w +  GM\log\left(\frac{w}{2GM}\right) + GM(1+ 2\log 2)+\frac{GM}{4}w^{-1} + O(w^{-2}),
+\begin{equation}\label{asymptotic_ell}
+  \ell = 2GM\left(w +  \frac{1}{2}\log w + \frac{1}{2} + \log(2)  +  \frac{1}{8}w^{-1} + O(w^{-2})\right),
 \end{equation}
@@
 
-\begin{equation}
-%  \int_{p_1}^{p_2}\sqrt{\frac{\left(r'\right)^2}{1-2M/r} + r^2}d\phi=
-  \int_{p_1}^{p_2}\left(\frac{\left(r'\right)^2}{1-1/r} + r^2\right)^\frac{1}{2}d\phi=
-  \int_{p_1}^{p_2}F\left(r,r'\right)d\phi
-\end{equation}
+% \begin{equation}
+% %  \int_{p_1}^{p_2}\sqrt{\frac{\left(r'\right)^2}{1-2M/r} + r^2}d\phi=
+%   \int_{p_1}^{p_2}\left(\frac{\left(r'\right)^2}{1-1/r} + r^2\right)^\frac{1}{2}d\phi=
+%   \int_{p_1}^{p_2}F\left(r,r'\right)d\phi
+% \end{equation}
 
@@
 
-\begin{equation}\label{rdashdash_properunits}
-  r''\left(\phi\right) =
-  (r-1) + \frac{(4r-3)\left(r'\right)^2}{2r\left(r-1\right)},\qquad r>1.
-\end{equation}
+% \begin{equation}\label{rdashdash_properunits}
+%   r''\left(\phi\right) =
+%   (r-1) + \frac{(4r-3)\left(r'\right)^2}{2r\left(r-1\right)},\qquad r>1.
+% \end{equation}
 \begin{equation}\label{rdashdash}
@@
 
-\begin{equation}\label{roc_properunits}
-  R = \frac{
-    \left(r^2 + \left(r'\right)^2\right)^\frac{3}{2}
-  }{
-%    \left|
-    r^2 + 2\left(r'\right)^2-rr''
-%    \right|
-  }
-  =
-  \frac{
-    \left(r^2 + \left(r'\right)^2\right)^\frac{3}{2}
-  }{
-    r-\frac{\left(r'\right)^2}{2\left(r-1\right)}
-  }\qquad\mbox{along a taut string}
-\end{equation}
+% \begin{equation}\label{roc_properunits}
+%   R = \frac{
+%     \left(r^2 + \left(r'\right)^2\right)^\frac{3}{2}
+%   }{
+% %    \left|
+%     r^2 + 2\left(r'\right)^2-rr''
+% %    \right|
+%   }
+%   =
+%   \frac{
+%     \left(r^2 + \left(r'\right)^2\right)^\frac{3}{2}
+%   }{
+%     r-\frac{\left(r'\right)^2}{2\left(r-1\right)}
+%   }\qquad\mbox{along a taut string}
+% \end{equation}
 

### 2019-05-23 string_jpam.tex
comment out the properunits versions of the asymptotics
typo

@@
 
- \begin{equation}\label{ell_properunits}
-   \ell=\int_{r=1}^{r=1+w}ds
-   =2\sqrt{w} + \frac{1}{3}w^{3/2} + O(w^{5/2})
- \end{equation}
+%  \begin{equation}\label{ell_properunits}
+%    \ell=\int_{r=1}^{r=1+w}ds
+%    =2\sqrt{w} + \frac{1}{3}w^{3/2} + O(w^{5/2})
+%  \end{equation}
 
@@
 
-\begin{equation}\label{asymptotic_ell_properunits}
-  \ell = w +  \frac{1}{2}\log w + \frac{1}{2} + \log(2)  +  \frac{1}{8}w^{-1} + O(w^{-2}),
-\end{equation}
-
+% \begin{equation}\label{asymptotic_ell_properunits}
+%   \ell = w +  \frac{1}{2}\log w + \frac{1}{2} + \log(2)  +  \frac{1}{8}w^{-1} + O(w^{-2}),
+% \end{equation}
+% 
 \begin{equation}\label{asymptotic_ell}
@@
   at~$\phi=0$.  Note the differences between these curves and the taut
-  strings shown elsewhere: unlike taug strings, null geodesics have
+  strings shown elsewhere: unlike taut strings, null geodesics have
   positive curvature everywhere, may cross the event horizon inwards,

### 2019-05-28 string_jpam.tex
use units in which G=1, basically by replacing GM with M throughout

@@
 to the singularity is tangential; if the distance of closest
-approach is less than about $1.0761\cdot 2GM$, the
-string self-intersects.  If the string attains $r=2GM$, then it
+approach is less than about $1.0761\cdot 2M$, the
+string self-intersects.  If the string attains $r=2M$, then it
 will in general be tangential to the event horizon.  Light taut
@@
 expressions for the strength of gravity at the event horizon of a
-black hole gives $L\lesssim TGM/\left(\mu c^2\right)$.  For the string
+black hole gives $L\lesssim TM/\left(\mu c^2\right)$.  For the string
 to behave as though light on lengthscales comparable with the
@@
 \begin{equation}\label{schwarzschild}
-ds^2= -dt^2\left(1-2GM/r\right) +\frac{dr^2}{1-2GM/r} + r^2\left(d\theta^2 + \sin^2\theta d\phi^2\right),
+ds^2= -dt^2\left(1-2M/r\right) +\frac{dr^2}{1-2M/r} + r^2\left(d\theta^2 + \sin^2\theta d\phi^2\right),
 \end{equation}
 
-\noindent where units in which~$c=1$ are used, has been known for
+\noindent where units in which~$G=c=1$ are used, has been known for
 over a century~\cite{schwarzschild1916}.  Here I consider the spatial
-component~$ds^2= dr^2/\left(1-2GM/r\right) + r^2d\phi^2$, confined to
-equatorial points~($\theta=\pi/2$); we understand~$2GM\leq r<\infty$
+component~$ds^2= dr^2/\left(1-2M/r\right) + r^2d\phi^2$, confined to
+equatorial points~($\theta=\pi/2$); we understand~$2M\leq r<\infty$
 throughout.  A circular string at constant~$r$ will, trivially, have
@@
 proper length $L\left(r_1,r_2\right)$ of a string stretching
-from~$r=r_1\geq 2GM$ to~$r_2\geq r_1$ at constant $\theta,\phi$:
+from~$r=r_1\geq 2M$ to~$r_2\geq r_1$ at constant $\theta,\phi$:
 
@@
 \begin{equation}\label{radial_string_length}
-  \int_{r=r_1}^{r_2}\frac{dr}{\sqrt{1-2GM/r}}=
+  \int_{r=r_1}^{r_2}\frac{dr}{\sqrt{1-2M/r}}=
   \left.
-  \sqrt{r(r-2GM)} +\frac{1}{2}\log\left(
-  \frac{\sqrt{r}+\sqrt{r-2GM}}{\sqrt{r}-\sqrt{r-2GM}}\right)
+  \sqrt{r(r-2M)} +\frac{1}{2}\log\left(
+  \frac{\sqrt{r}+\sqrt{r-2M}}{\sqrt{r}-\sqrt{r-2M}}\right)
   \right|_{r_1}^{r_2}
@@
 Observe that integration may be performed with a lower limit
-of~$r_1=2GM$ with no difficulty, even though the radial component of the
+of~$r_1=2M$ with no difficulty, even though the radial component of the
 Schwarzschild metric approaches infinity there.
@@
 some peculiar features.  Firstly, consider the proper length from
-$r=2GM$ to $r=2GM(1+w)$ where $w>0, \left|w\right|\ll 1$.  Then
+$r=2M$ to $r=2M(1+w)$ where $w>0, \left|w\right|\ll 1$.  Then
 
@@
    \ell =
-   \int_{r=2GM}^{r=2GM(1+w)}ds
-   =2GM\left(2\sqrt{w} + \frac{1}{3}w^{3/2} + O(w^{5/2})\right).
+   \int_{r=2M}^{r=2M(1+w)}ds
+   =2M\left(2\sqrt{w} + \frac{1}{3}w^{3/2} + O(w^{5/2})\right).
  \end{equation}
 
-This would indicate that, close to the Schwarzschild radius $r=2GM$,
+This would indicate that, close to the Schwarzschild radius $r=2M$,
 large (but finite) amounts of string are required to connect two
@@
 \begin{equation}\label{asymptotic_ell}
-  \ell = 2GM\left(w +  \frac{1}{2}\log w + \frac{1}{2} + \log(2)  +  \frac{1}{8}w^{-1} + O(w^{-2})\right),
+  \ell = 2M\left(w +  \frac{1}{2}\log w + \frac{1}{2} + \log(2)  +  \frac{1}{8}w^{-1} + O(w^{-2})\right),
 \end{equation}
@@
 \noindent suggesting that there is ``extra radial space'', of length
-over $1.19\cdot 2GM$, sensible at large distances from the event
+over $1.19\cdot 2M$, sensible at large distances from the event
 horizon.  Radial strings have proper length longer than the
@@
 %  \int_{p_1}^{p_2}\sqrt{\frac{\left(r'\right)^2}{1-2M/r} + r^2}d\phi=
-  \int_{p_1}^{p_2}\left(\frac{\left(r'\right)^2}{1-2GM/r} + r^2\right)^\frac{1}{2}d\phi=
+  \int_{p_1}^{p_2}\left(\frac{\left(r'\right)^2}{1-2M/r} + r^2\right)^\frac{1}{2}d\phi=
   \int_{p_1}^{p_2}F\left(r,r'\right)d\phi
@@
   r''\left(\phi\right) =
-  (r-2GM) + \frac{(4r-6GM)\left(r'\right)^2}{2r\left(r-2GM\right)},\qquad r>2GM.
+  (r-2M) + \frac{(4r-6M)\left(r'\right)^2}{2r\left(r-2M\right)},\qquad r>2M.
 \end{equation}
@@
 equation~\ref{rdashdash} with the corresponding equation for null
-geodesics, which is usually given~\cite{wald} in terms of $u=2GM/r$ as
+geodesics, which is usually given~\cite{wald} in terms of $u=2M/r$ as
 $u''=3u^2/2-u$, but is equivalent
-to~$r''\left(\phi\right)=r-3GM+4GMr'/r$.
+to~$r''\left(\phi\right)=r-3M+4Mr'/r$.
 
@@
 integration.  There are two qualitatively different types of
-solutions: those with a tangential point ($r'=0,r>2GM$), and those
+solutions: those with a tangential point ($r'=0,r>2M$), and those
 without.  Consider first solutions with such a point: there exists a
-pair~$\left(r,\phi\right), r>2GM$ at which~$r'(\phi)=0$.  This point
+pair~$\left(r,\phi\right), r>2M$ at which~$r'(\phi)=0$.  This point
 must be unique, as~$r''(\phi)>0$, and be the closest approach of the
@@
   }{
-    r-\frac{\left(r'\right)^2}{2\left(r-2GM\right)}
+    r-\frac{\left(r'\right)^2}{2\left(r-2M\right)}
   }\qquad\mbox{along a taut string}
@@
 Considering~$r'(\phi)=0$ shows that if a string is tangential at a
-point with~$r_0>2GM$, then it has a radius of curvature of~$r_0^2/2GM$
+point with~$r_0>2M$, then it has a radius of curvature of~$r_0^2/2M$
 at that point.  Any circular path, including a circular orbit
-at~$r=r_0$, has a radius of curvature of $r_0<r_0^2/2GM$, showing that
+at~$r=r_0$, has a radius of curvature of $r_0<r_0^2/2M$, showing that
 circular orbits cannot follow a taut string at closest approach.  Also
-observe that null geodesics have a radius of curvature of $r_0^2/3GM <
-r_0^2/2GM$, showing that light cannot follow a taut string at closest
+observe that null geodesics have a radius of curvature of $r_0^2/3M <
+r_0^2/2M$, showing that light cannot follow a taut string at closest
 approach either.  The string has a point of inflection
-if~$\left(r'\right)^2=2r(r-2GM)$.
+if~$\left(r'\right)^2=2r(r-2M)$.
 
@@
 
-If, conversely, $r'(\phi)\neq 0$ whenever~$r>2GM$, then the string has
+If, conversely, $r'(\phi)\neq 0$ whenever~$r>2M$, then the string has
 only one free end; the other is tangential to the event horizon (in
@@
 except in the degenerate case of constant~$\phi$ when the string is
-radial for~$r>2GM$.
+radial for~$r>2M$.
 
@@
 for ordinary differential equations~\cite{soetart2010}.  Non-radial
-solutions fall into two distinct classes: those with $r>2GM$ at all
-points, and those which attain $r=2GM$, in which case the string is
+solutions fall into two distinct classes: those with $r>2M$ at all
+points, and those which attain $r=2M$, in which case the string is
 tangential to the event horizon.
 
-We first consider strings for which $r>2GM$ everywhere (that is,
+We first consider strings for which $r>2M$ everywhere (that is,
 completely exterior strings).
@@
 approach is greater than a critical value of about~$\simeq 1.0761\cdot
-2GM$, the string does not self-intersect; if the closest approach is
+2M$, the string does not self-intersect; if the closest approach is
 less than this value, then the string crosses itself at~$\phi=\pi/2$.
-As discussed above, if $\left(r'\right)^2=2r(r-2GM)$, the string has a
+As discussed above, if $\left(r'\right)^2=2r(r-2M)$, the string has a
 point of inflection at which point the radius of curvature becomes
@@
 Figure \ref{strings_r_equals_2}, for example, shows strings passing
-through~$(4GM,0)$.  Figure~\ref{light_r_equals_2} shows light paths for
+through~$(4M,0)$.  Figure~\ref{light_r_equals_2} shows light paths for
 comparison.
 
-The cases where the string attains $r=2GM$ are tangent to the event
+The cases where the string attains $r=2M$ are tangent to the event
 horizon (or, exceptionally, purely radial).
@@
 event horizon at~$\phi=0$ and Figure~\ref{one_free_end_r_equals_2}
-shows tangential strings passing through $(4GM,0)$.
+shows tangential strings passing through $(4M,0)$.
 
@@
   approach to the event horizon occurs at $\phi=0$ (3 o'clock) and
-  $r/2GM\in\left(1,1.0761\right)$; strings are symmetrical about $\phi=0$
+  $r/2M\in\left(1,1.0761\right)$; strings are symmetrical about $\phi=0$
   but for clarity only $\phi\geq 0$ is shown}
@@
 \includegraphics{angle_at_r_equals_2.pdf} % created by "angle_at_r_equals_2.R"
-\caption{Strings passing through $(4GM,0)$}
+\caption{Strings passing through $(4M,0)$}
 \label{strings_r_equals_2}
@@
 \includegraphics{light_start_at_r_equals_2.pdf} % created by "angle_start_at_r_equals_2.R"
-\caption{Null geodesics passing through $(4GM,0)$, shown for $0\leq\phi\leq 2\pi$}
+\caption{Null geodesics passing through $(4M,0)$, shown for $0\leq\phi\leq 2\pi$}
 \label{light_r_equals_2}
@@
 \includegraphics[width=170mm]{one_free_end_r_equals_2.pdf}
-\caption{Strings tangential to the event horizon passing through $(4GM,0)$}
+\caption{Strings tangential to the event horizon passing through $(4M,0)$}
 \label{one_free_end_r_equals_2}

### 2019-06-07 string_grg.tex
initial draft using the style for <General Relativity and Gravitation>

--- 2019-05-28 string_jpam.tex
@@
-\documentclass[prb,preprint]{revtex4-1} 
-% The line above defines the type of LaTeX document.
-% Note that AJP uses the same style as Phys. Rev. B (prb).
-
-% The % character begins a comment, which continues to the end of the line.
-
-\usepackage{amsmath}  % needed for \tfrac, \bmatrix, etc.
-\usepackage{amsfonts} % needed for bold Greek, Fraktur, and blackboard bold
-\usepackage{graphicx} % needed for figures
-
+%%%%%%%%%%%%%%%%%%%%%%% file template.tex %%%%%%%%%%%%%%%%%%%%%%%%%
+%
+% This is a general template file for the LaTeX package SVJour3
+% for Springer journals.          Springer Heidelberg 2010/09/16
+%
+% Copy it to a new file with a new name and use it as the basis
+% for your article. Delete % signs as needed.
+%
+% This template includes a few options for different layouts and
+% content for various journals. Please consult a previous issue of
+% your journal as needed.
+%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%
+% First comes an example EPS file -- just ignore it and
+% proceed on the \documentclass line
+% your LaTeX will extract the file if required
+\begin{filecontents*}{example.eps}
+%!PS-Adobe-3.0 EPSF-3.0
+%%BoundingBox: 19 19 221 221
+%%CreationDate: Mon Sep 29 1997
+%%Creator: programmed by hand (JK)
+%%EndComments
+gsave
+newpath
+  20 20 moveto
+  20 220 lineto
+  220 220 lineto
+  220 20 lineto
+closepath
+2 setlinewidth
+gsave
+  .4 setgray fill
+grestore
+stroke
+grestore
+\end{filecontents*}
+%
+\RequirePackage{fix-cm}
+%
+%\documentclass{svjour3}                     % onecolumn (standard format)
+%\documentclass[smallcondensed]{svjour3}     % onecolumn (ditto)
+\documentclass[smallextended]{svjour3}       % onecolumn (second format)
+%\documentclass[twocolumn]{svjour3}          % twocolumn
+%
+\smartqed  % flush right qed marks, e.g. at end of proof
+%
+\usepackage{graphicx}
+\usepackage{amssymb}
+%
+% \usepackage{mathptmx}      % use Times fonts if available on your TeX system
+%
+% insert here the call for the packages your document requires
+%\usepackage{latexsym}
+% etc.
+%
+% please place your own definitions here and don't use \def but
+% \newcommand{}{}
+%
+% Insert the name of "your journal" with
+% \journalname{myjournal}
+%
 \begin{document}
 
-% Be sure to use the \title, \author, \affiliation, and \abstract macros
-% to format your title page.  Don't use lower-level macros to  manually
-% adjust the fonts and centering.
+\title{Light inextensible strings under tension in the Schwarzschild geometry}%\thanks{Grants or other notes
+%about the article that should go on the front page should be
+%placed here. General acknowledgments should be placed at the end of the article.}
 
-\title{Light inextensible strings under tension in the Schwarzschild geometry}
-% In a long title you can use \\ to force a line break at a certain location.
+%\subtitle{Do you have a subtitle?\\ If so, write it here}
+
+%\titlerunning{Short form of title}        % if too long for running head
 
 \author{Robin K. S. Hankin}
-\email{hankin.robin@gmail.com} % optional
-\altaffiliation[permanent address: ]{55 Wellesley Street East,
-Auckland, New Zealand}
-% If there were a second author at the same address, we would put another 
-% \author{} statement here.  Don't combine multiple authors in a single
-% \author statement.
-\affiliation{School of Engineering, Computer and Mathematical Sciences, Auckland University of Technology}
-% Please provide a full mailing address here.
 
-% See the REVTeX documentation for more examples of author and affiliation lists.
+%\authorrunning{Short form of author list} % if too long for running head
 
-\date{\today}
+\institute{Robin K. S. Hankin \at
+  Auckland University of Technology\\
+  Wellesley Street\\
+  Auckland, New Zealand\\
+              Tel.: +64(9)5219999\\
+              \email{hankin.robin@gmail.com}           %  \\
+}
+
+\date{Received: date / Accepted: date}
+% The correct dates will be entered by the editor
+
+\maketitle
 
 \begin{abstract}
+
   Here I analyze the behaviour of stationary, light inextensible strings
@@
 suggests a possible way of manipulating a black hole.
+\keywords{Black holes\and inextensible string\and minimal-length path}
+% \PACS{PACS code1 \and PACS code2 \and more}
+% \subclass{MSC code1 \and MSC code2 \and more}
 \end{abstract}
-% AJP requires an abstract for all regular article submissions.
-% Abstracts are optional for submissions to the "Notes and Discussions" section.
 
-\maketitle % title page is now complete
+\section{Introduction}
+\label{intro}
 
@@
 expressions for the strength of gravity at the event horizon of a
-black hole gives $L\lesssim TM/\left(\mu c^2\right)$.  For the string
-to behave as though light on lengthscales comparable with the
-Schwarzschild radius would require $T\gtrsim\mu c^2$.  Exceeding this
-criterion would violate the dominant energy condition (and in this
-context we note that transverse waves would propagate at
+black hole with mass $M$ gives $L\lesssim TM/\left(\mu c^4\right)$.
+For the string to behave as though light on lengthscales comparable
+with the Schwarzschild radius would require $T\gtrsim\mu c^2$.
+Exceeding this criterion would violate the dominant energy condition
+(and in this context we note that transverse waves would propagate at
 speed~$\sqrt{T/\mu}\sim c$).  Nevertheless, the idealization of a
@@
 in the vicinity of a black hole is well
-known~\citep{abramowicz1992,abramowicz1993}, and the analysis of taut
+known~\cite{abramowicz1992,abramowicz1993}, and the analysis of taut
 strings furnishes another everyday window into general relativity.
-Another possibility, presented by \citet{damour1978} and discussed by
-\citet{thorne1986}, is to pass electric current through such taut
-strings, an option not pursued here.
+Another possibility, presented by discussed by
+Thorne~\cite{thorne1986}, is to pass electric current through such
+taut strings, an option not pursued here.
 
@@
 %%Vancouver style references.
-\bibliographystyle{model3-num-names}
-\bibliography{stringrefs}
 
@@
 
-Calculations are performed using the R programming language~\citep{rcore2018};
+Calculations are performed using the R programming language~\cite{rcore2018};
 source code for the figures is available at
@@
 \\
-\url{https://github.com/RobinHankin/string.git}
+{\tt https://github.com/RobinHankin/string.git}
 
@@
 
+ \section*{Conflict of interest}
+ The author declares that he has no conflict of interest.
+
+% BibTeX users please use one of
+%\bibliographystyle{spbasic}      % basic style, author-year citations
+\bibliographystyle{spmpsci}      % mathematics and physical sciences
+%\bibliographystyle{spphys}       % APS-like style for physics
+%\bibliography{}   % name your BibTeX data base
+
+\bibliography{stringrefs}
+
 \end{document} 
+% end of file template.tex

### 2019-06-07 string_grg.tex
minor typos

@@
 strings furnishes another everyday window into general relativity.
-Another possibility, presented by discussed by
-Thorne~\cite{thorne1986}, is to pass electric current through such
-taut strings, an option not pursued here.
+Another possibility, discussed by Thorne~\cite{thorne1986}, is to pass
+electric current through such taut strings, an option not pursued
+here.
 
@@
 component~$ds^2= dr^2/\left(1-2M/r\right) + r^2d\phi^2$, confined to
-equatorial points~($\theta=\pi/2$); we understand~$2M\leq r<\infty$
+equatorial points~($\theta=\pi/2$); we understand~$2M\leq r$
 throughout.  A circular string at constant~$r$ will, trivially, have

### 2019-06-16 trajectory_mercury.R
should subtract pi for better axis labelling

@@ -6,7 +6,7 @@
 
 library("deSolve")
 source("usefulfuncs.R")
-source("mercury.R")
+source("mercury.R")  # orbital parameters for Mercury in geometrized units
      
 LVmod <- function(Time, State, Pars) {
     with(as.list(c(State, Pars)), {
@@ -53,4 +53,4 @@ phi <- out$phi[change]
 phi <- (phi) %% (2*pi)
 
 
-plot(tee,phi)
+plot(tee,phi-pi)

### 2019-06-16 usefulgeodesicfuncs.R
unified interface, now use trajectory() for plotting orbits

@@ -1,0 +1,53 @@
+library("deSolve")   # ode()
+source("usefulfuncs.R")
+
+`geodesic` <- function(Time, State, Pars) {
+    with(as.list(c(State, Pars)), {
+      ##    V <- (r-2*m)*(r^2+h^2)/r^3  (with GR term)
+      if(GR){
+        dVdr <- -2*h^2/r^3 + 2*m/r^2 + 6*m*h^2/r^4 # inefficient idiom
+        } else {
+          dVdr <- -2*h^2/r^3 + 2*m/r^2
+        }
+      rdotdot <- -dVdr/2
+      phidot <- h/r^2
+      return(list(c(rdot,rdotdot,phidot)))
+    })
+}
+
+`trajectory` <- function(r_start, m, h, tau=NULL, GR=TRUE){
+  if(is.null(tau)){tau<- seq_len(3000)} # works nicely with trajectory(30, 1/2, 3)
+
+  yini <- c(
+      r    = r_start,
+      rdot = 0,
+      phi  = 0
+  )
+  
+  bh <- as.data.frame(ode(y=yini, times=tau, func=geodesic, c(m=m, h=h, GR=GR), rtol=1e-6))
+  
+  r   <- bh[,2]
+  phi   <- bh[,4]
+
+  xy <- cbind(x=r*cos(phi),y=r*sin(phi))
+  return(cbind(bh,xy))
+}
+
+## Example use-cases
+jj <- trajectory(30, 1/2, 3) # m=1/2 -> Schwarzschild radius = 1
+plot(jj$x,jj$y, asp=1, type='l', axes=FALSE, ylim=c(-30,30), xlab='', ylab='')
+polargrid(r=10*(1:3))
+event_horizon()
+
+
+jj <- trajectory(30, 1/2, 3, GR=FALSE)  # Newtonian approximation
+plot(jj$x,jj$y, asp=1, type='l', axes=FALSE, ylim=c(-30,30), xlab='', ylab='')
+polargrid(r=10*(1:3))
+event_horizon()
+
+
+source("mercury.R")
+jj <- trajectory(merc_perihelion, m=d, h=h, tau=seq(from=0,to=15e17,len=1e6))
+plot(jj$x,jj$y, asp=1, type='l', axes=FALSE, xlab='', ylab='')
+event_horizon()
+

### 2019-06-16 trajectory.R
trajectory functions using consistent usefulgeodesicfunctions.R

@@ -1,39 +1,20 @@
 ## Plots a trajectory of an (infinitesimal mass) object orbiting a
-## black hole, showing non-closed orbit.  Works standalone but is
-## intended to be called by maker.R which creates a pdf file.
+## black hole, showing the effect of the relativistic correction,
+## specifically GR's prediction of a non-closed orbit.
+
+## Compare the result of trajectory_newtonian.R which shows the same
+## simulation but under the Newtonian approximation.
+
+## Works standalone but is intended to be called by maker.R which
+## creates a pdf file.
 
 ## Notation follows Rindler, p239
 
-library("deSolve")
+
 source("usefulfuncs.R")
+source("usefulgeodesicfuncs.R")
      
-LVmod <- function(Time, State, Pars) {
-    with(as.list(c(State, Pars)), {
-    #    V <- (r-2*m)*(r^2+h^2)/r^3
-        dVdr <- -2*h^2/r^3 + 2*m/r^2 + 6*m*h^2/r^4 # inefficient idiom
-        rdotdot <- -dVdr/2
-        phidot <- h/r^2
-        return(list(c(rdot,rdotdot,phidot)))
-    })
-}
-
-## parameters, mass m and angular momentum h:
-pars  <- c(m=1/2, h=3)    # m=1/2, so r_schwarz=1
-
-## Initial conditions, rdot=0 means periastron:
-yini  <- c(r=30,rdot=0,phi=0)
-
-## times means *proper* times:
-times <- seq(from=0,to=3100,by=3)
-
-out   <- ode(yini, times, LVmod, pars)
-jj <- cbind(out[,2]*cos(out[,4]),out[,2]*sin(out[,4]))
-
-## Plot commands start:
-plot(jj, asp=1, type='l',
-     axes=FALSE, xlab='',ylab='',
-     main="Timelike geodesics in the Schwarzschild geometry")
-
-polargrid(r=(1:6)*5)
-text((1:6)*5,-1,(1:6)*5)
-event_horizon(fill=TRUE)
+jj <- trajectory(30, 1/2, 3) # m=1/2 -> Schwarzschild radius = 1
+plot(jj$x,jj$y, asp=1, type='l', axes=FALSE, xlab='', ylab='',main='Geodesics in the Schwarzschild geometry')
+polargrid(r=10*(1:3))
+event_horizon()

### 2019-06-16 trajectory_newtonian.R

@@ -1,0 +1,21 @@
+## Plots a trajectory of an (infinitesimal mass) object orbiting a
+## black hole, under the Newtonian approximation of slow speeds and
+## large distances compared with the Schwarzschild radius.  We see a
+## closed elliptical orbit as per Keplerian dynamics.
+
+## Compare the result of trajectory.R which shows the same simulation
+## but without making the Newtonian approximation.
+
+## Works standalone but is intended to be called by maker.R which
+## creates a pdf file.
+
+## Notation follows Rindler, p239
+
+
+source("usefulfuncs.R")
+source("usefulgeodesicfuncs.R")
+     
+jj <- trajectory(30, 1/2, 3, GR=FALSE) # m=1/2 -> Schwarzschild radius = 1
+plot(jj$x,jj$y, asp=1, type='l', axes=FALSE, xlab='', ylab='',main='Newtonian orbit',sub='Three successive orbits plotted, note closed path')
+polargrid(r=10*(1:3))
+event_horizon()

### 2019-06-16 trajectory_mercury.R
rationalize trajectory()

@@ -4,40 +4,14 @@
 
 ## Notation follows Rindler, p239
 
-library("deSolve")
-source("usefulfuncs.R")
+source("usefulgeodesicfuncs.R")
 source("mercury.R")  # orbital parameters for Mercury in geometrized units
      
-LVmod <- function(Time, State, Pars) {
-    with(as.list(c(State, Pars)), {
-    #    V <- (r-2*m)*(r^2+h^2)/r^3
-        dVdr <- -2*h^2/r^3 + 2*m/r^2 + 6*m*h^2/r^4 # inefficient idiom
-        rdotdot <- -dVdr/2
-        phidot <- h/r^2
-        return(list(c(rdot,rdotdot,phidot)))
-    })
-}
+out <- trajectory(merc_perihelion, m=d, h=h, tau=seq(from=0,to=35e17,len=5e6))
+plot(out$x,out$y, asp=1, type='l', axes=FALSE, xlab='', ylab='')
+event_horizon()
+dev.new()
 
-## parameters, mass m and angular momentum h:
-pars  <- c(m=d, h=h)    # m=1/2, so r_schwarz=1
-
-## Initial conditions, rdot=0 means periastron:
-yini  <- c(r=merc_perihelion,rdot=0,phi=0)
-
-## times means *proper* times:
-times <- seq(from=0,to=15e17,len=5000000)
-
-out   <- ode(yini, times, LVmod, pars)
-jj <- cbind(out[,2]*cos(out[,4]),out[,2]*sin(out[,4]))
-
-## Plot commands start:
-plot(jj, asp=1, type='l',
-     axes=FALSE, xlab='',ylab='',
-     main="Timelike geodesics in the Schwarzschild geometry")
-
-#polargrid(r=(1:6)*5)
-#text((1:6)*5,-1,(1:6)*5)
-event_horizon(fill=TRUE)
 
 out <- as.data.frame(out)
 
@@ -45,7 +19,6 @@ jj1 <- c(NA,out$rdot)
 jj2 <- c(out$rdot,NA)
 
 change <- (jj1>0) & (jj2<0)
-
 
 tee    <- out$time[change]
 

### 2019-06-16 usefulgeodesicfuncs.R
do not actually execute the example use-cases

@@ -34,6 +34,7 @@ source("usefulfuncs.R")
 }
 
 ## Example use-cases
+if(FALSE){
 jj <- trajectory(30, 1/2, 3) # m=1/2 -> Schwarzschild radius = 1
 plot(jj$x,jj$y, asp=1, type='l', axes=FALSE, ylim=c(-30,30), xlab='', ylab='')
 polargrid(r=10*(1:3))
@@ -51,3 +52,4 @@ jj <- trajectory(merc_perihelion, m=d, h=h, tau=seq(from=0,to=15e17,len=1e6))
 plot(jj$x,jj$y, asp=1, type='l', axes=FALSE, xlab='', ylab='')
 event_horizon()
 
+}

### 2019-06-17 usefulgeodesicfuncs.R
new SI argument to trajectory()

@@ -15,7 +15,7 @@ source("usefulfuncs.R")
     })
 }
 
-`trajectory` <- function(r_start, m, h, tau=NULL, GR=TRUE){
+`trajectory` <- function(r_start, m, h, tau=NULL, GR=TRUE, SI=TRUE){
   if(is.null(tau)){tau<- seq_len(3000)} # works nicely with trajectory(30, 1/2, 3)
 
   yini <- c(
@@ -25,10 +25,14 @@ source("usefulfuncs.R")
   )
   
   bh <- as.data.frame(ode(y=yini, times=tau, func=geodesic, c(m=m, h=h, GR=GR), rtol=1e-6))
-  
-  r   <- bh[,2]
-  phi   <- bh[,4]
 
+  if(SI){
+    bh[,1] <- bh[,1]/sol  # sol from mercury.R
+    bh[,3] <- bh[,3]*sol
+  }
+
+  r   <- bh[,2]
+  phi <- bh[,4]
   xy <- cbind(x=r*cos(phi),y=r*sin(phi))
   return(cbind(bh,xy))
 }

### 2019-06-17 mercury.R
more docs

@@ -1,42 +1,53 @@
-G <-  6.674e-11  # big G in SI
-M  <- 1.9885e30 # mass of Sun (kg)
-M_merc <- 3.3011e23 # mass of Mercury
-sol <- 299792458    # speed of light
+## This file creates some variables to be used by trajectory_mercury.R
+## to analyse Mercury's orbit.
 
-d <- G*M/sol^2  # mass of sun
-r_s <- 2*d  # Schwarzschild radius of Sun (=2M)
-ecc_wikipedia <- 0.205630 # eccentricity of Mercury's orbit as given by wikiped
-per <- 87.969 * 3600*24
-	
-merc_perihelion   <- 46001200e3  # meters
-merc_aphelion <- 69816900e3
 
+G <-  6.674e-11     # big G in SI
+M  <- 1.9885e30     # mass of Sun (kg)
+M_merc <- 3.3011e23 # mass of Mercury (kg)
+sol <- 299792458    # speed of light (m/s)
+
+d <- G*M/sol^2  # mass of sun in meters
+r_s <- 2*d      # Schwarzschild radius of Sun (=2M)
+ecc_wiki <- 0.205630 # eccentricity of Mercury's orbit from wikipedia
+per <- 87.969 * 3600*24   # period of Mercury's orbit (s)
+merc_perihelion <- 46001200e3  # meters
+merc_aphelion   <- 69816900e3
+
+
+omega <- 2*pi/per         # mean angular velocity of Mercury, radian/sec
+
+## Now calculate ellipse parameters from observational data:
 a <- (merc_perihelion + merc_aphelion)/2
 cee <- a-merc_perihelion
 b <- sqrt(a^2-cee^2)
-ecc <- sqrt(1-(b/a)^2)  # should match ecc above
+ecc <- sqrt(1-(b/a)^2)  # should match ecc_wiki above
 ell <- (a^2-cee^2)/a  # semi-latus rectum
 
 mu <- G*(M + M_merc)
-# equation 19 says v = sqrt(mu/ell) * (1+e*cos(theta))
+
+# equation 19 of wikipedia says v = sqrt(mu/ell) * (1+e*cos(theta))
 vmin <- sqrt(mu/ell)*(1-ecc)
 vmax <- sqrt(mu/ell)*(1+ecc)
 
 # Now we need to express the specific angular momentum in relativistic
-#  units in which G=c=1:
-
+# units in which G=c=1:
    
 specific_angular_momentum_SI <- merc_perihelion * vmax
 
 # consistency check:
[…]

### 2019-06-17 usefulgeodesicfuncs.R
slight change to default plus bugfix on sol

@@ -16,7 +16,7 @@ source("usefulfuncs.R")
 }
 
 `trajectory` <- function(r_start, m, h, tau=NULL, GR=TRUE, SI=TRUE){
-  if(is.null(tau)){tau<- seq_len(3000)} # works nicely with trajectory(30, 1/2, 3)
+  if(is.null(tau)){tau<- seq_len(3100)} # works nicely with trajectory(30, 1/2, 3)
 
   yini <- c(
       r    = r_start,
@@ -27,8 +27,8 @@ source("usefulfuncs.R")
   bh <- as.data.frame(ode(y=yini, times=tau, func=geodesic, c(m=m, h=h, GR=GR), rtol=1e-6))
 
   if(SI){
-    bh[,1] <- bh[,1]/sol  # sol from mercury.R
-    bh[,3] <- bh[,3]*sol
+    bh[,1] <- bh[,1]/299792458  # sol defined in mercury.R
+    bh[,3] <- bh[,3]*299792458
   }
 
   r   <- bh[,2]

### 2019-06-17 trajectory_mercury.R
more explicit calculation of perihelion advance

@@ -7,11 +7,10 @@
 source("usefulgeodesicfuncs.R")
 source("mercury.R")  # orbital parameters for Mercury in geometrized units
      
-out <- trajectory(merc_perihelion, m=d, h=h, tau=seq(from=0,to=35e17,len=5e6))
+out <- trajectory(merc_perihelion, m=d, h=h, tau=seq(from=0,to=35e16,len=1e7))
+
 plot(out$x,out$y, asp=1, type='l', axes=FALSE, xlab='', ylab='')
 event_horizon()
-dev.new()
-
 
 out <- as.data.frame(out)
 
@@ -25,5 +24,25 @@ tee    <- out$time[change]
 phi <- out$phi[change]
 phi <- (phi) %% (2*pi)
 
-
+dev.new()
 plot(tee,phi-pi)
+
+
+## Now calculate perihelion advance in seconds of arc per century:
+conversion <-
+  c(
+      seconds_per_minute = 60,
+      minutes_per_hour   = 60,
+      hours_per_day      = 24,
+      days_per_year      = 365.25,
+      years_per_century  = 100,
+      degrees_per_radian = 180/pi,
+      minutes_per_degree = 60,
+      seconds_per_minute = 60
+  )
+
+lm(phi~tee)$coefficients[2] * prod(conversion)
+
+
+
+## Compare with Einstein's value of 44

### 2019-06-17 perihelion_advance_of_mercury.Rmd
nice markdown documnet

---
title: "Perihelion advance of Mercury"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
source("mercury.R")
source("usefulfuncs.R")
source("usefulgeodesicfuncs.R")
```

## Summary

The perihelion advance of Mercury is a well-known test of general relativity.  Einstein calculated that Mercury's perihelion would advance by 44 seconds of arc per century due to general relativistic effects; here I reproduce that analysis using numerical reasoning.

## Introduction

The orbit of Mercury has long been known to advance at a rate of about 6400'' [seconds of arc] per century.  Various classical phenomena, such as the effects of Venus, and the oblateness of the sun, account for all but 44 seconds of arc and this remaining amount puzzled astronomers for many years. According to Einstien and others, the general theory of relativity predicts an additional precession of almost exactly 44 seconds per century.  In this document I will calculate the precessional rate using numerical methods.

## Closed and non-closed orbits.

Consider the following R idiom, in which ```trajectory()``` is used to directly simulate a geodesic in Schwarzschild space:

```{r}
jj <- trajectory(30, 1/2, 3) # m=1/2 -> Schwarzschild radius = 1
plot(jj$x,jj$y, asp=1, type='l', axes=FALSE, xlab='', ylab='',main='Geodesics in the Schwarzschild geometry')
polargrid(r=10*(1:3))
event_horizon()
```

In the above, we see a typical orbit in Schwarzschild space and a precession of around 40 degrees per orbit.  As a consistency check we can perform the same calculation but with Newtonian approximations:

```{r}
jj <- trajectory(30, 1/2, 3, GR=FALSE) # GR=FALSE means use Newtonian approximation
plot(jj$x,jj$y, asp=1, type='l', axes=FALSE, xlab='', ylab='',main='Newtonian approximation')
polargrid(r=10*(1:3))
event_horizon()
```

and above we see that the orbit is accurately closed, as expected.

## Mercury

First we specify some orbital parameters for the orbit of Mercury, mostly taken from wikipedia:
```{r}
G <-  6.674e-11     # big G in SI
M  <- 1.9885e30     # mass of Sun (kg)
M_merc <- 3.3011e23 # mass of Mercury (kg)
sol <- 299792458    # speed of light (m/s)
merc_perihelion <- 46001200e3 
merc_aphelion   <- 69816900e3
mu <- G*(M + M_merc)
```

Now we calculate some features of the orbit in geometrized units:
```{r} 
d <- G*M/sol^2  # mass of sun in meters
r_s <- 2*d      # Schwarzschild radius of Sun (=2M)
```

And some futher features using classical geometry:
```{r}
a <- (merc_perihelion + merc_aphelion)/2
cee <- a-merc_perihelion
b <- sqrt(a^2-cee^2)
ecc <- sqrt(1-(b/a)^2)  # should match ecc_wiki above
ell <- (a^2-cee^2)/a  # semi-latus rectum
```

Now calculate maximum and minimum orbital speeds:

```{r}
vmin <- sqrt(mu/ell)*(1-ecc)
vmax <- sqrt(mu/ell)*(1+ecc)
```

And we need to specify the orbital angular mometum in different units:

```{r}   
specific_angular_momentum_SI <- merc_perihelion * vmax
h <- specific_angular_momentum_SI/sol #geometrized units
```

With all these we can simulate the orbit:
```{r}
 out <- trajectory(merc_perihelion, m=d, h=h, tau=seq(from=0,to=35e15,len=5e5))
 plot(out$x,out$y,type='l',asp=1,axes=FALSE,xlab='',ylab='')
 event_horizon()
 max(out$phi/(2*pi))
```

In the above diagram we see over 15.4 orbits of Mercury, exhibiting no visible deviation from an exactly closed orbital path.  To calculate the precession we need to simulate the motion for a longer time:

```{r}
out <- trajectory(merc_perihelion, m=d, h=h, tau=seq(from=0,to=35e16,len=1e7))
```

We need to isolate the change points (actually calculating aphelion rather than perihelion for convencence):

```{r}
jj1 <- c(NA,out$rdot)
jj2 <- c(out$rdot,NA)
change <- (jj1>0) & (jj2<0)

tee    <- out$time[change]

phi <- out$phi[change]
phi <- (phi) %% (2*pi)

plot(tee,phi-pi)
```

There is a lot of noise on this caused by the discrete nature of the simulation results provided by ```ode()```.  To estimate the precession we can use linear regression:

```{r}
prec <- lm(phi~tee)$coefficients[2]
prec
```

And to convert to astronomical units we use
```{r}
conversion <-
  c(
      seconds_per_minute = 60,
      minutes_per_hour   = 60,
      hours_per_day      = 24,
      days_per_year      = 365.25,
      years_per_century  = 100,
      degrees_per_radian = 180/pi,
      minutes_per_degree = 60,
      seconds_per_minute = 60
  )

```

giving

```{r}
prec*prod(conversion)
```

closely agreeing with Einstein's original prediction of 44.  A more careful analysis would account for the nonzero mass of Mercury.

### 2019-06-17 mercury.R
unit conversions for arcseconds per century as per Einstein's predictions

@@ -51,3 +51,43 @@ if(FALSE){  # do not run; takes a long time
   plot(out$x,out$y,type='l',asp=1)
   ## this is done better in trajectory_mercury.R
 }
+
+
+conversion1 <- prod(c(  # used in the Rmd file
+    seconds_per_minute = 60, 
+    minutes_per_hour   = 60,
+    hours_per_day      = 24,
+    days_per_year      = 365.25,
+    years_per_century  = 100,
+    degrees_per_radian = 180/pi,
+    minutes_per_degree = 60,
+    seconds_per_minute = 60
+))
+
+seconds_per_orbit <- prod(c(
+    seconds_per_minute = 60,
+    minutes_per_hour   = 60,
+    hours_per_day      = 24,
+    days_per_orbit     = 87.969
+))
+
+seconds_per_century <- prod(c(
+    seconds_per_minute = 60,
+    minutes_per_hour   = 60,
+    hours_per_day      = 24,
+    days_per_year      = 365.25,
+    years_per_century  = 100
+))
+
+orbits_per_century <- seconds_per_century/seconds_per_orbit
+arcseconds_per_radian <- prod(
+    c(
+        degrees_per_radian = 180/pi,
+        minutes_per_degree = 60,
+        seconds_per_minute = 60
+        ))
+
+
+precession <- 24*pi^3*a^2/(per^2*sol^2*(1-ecc^2))  # radians/orbit
+arcseconds_per_orbit <- precession*arcseconds_per_radian
+arcseconds_per_century <- arcseconds_per_orbit * orbits_per_century

### 2019-06-17 perihelion_advance_of_mercury.Rmd
change correction from 44 to 43

@@
 The perihelion advance of Mercury is a well-known test of general relativity.
-Einstein calculated that Mercury's perihelion would advance by 44 seconds of arc per century due to general relativistic effects; here I reproduce that analysis using numerical reasoning.
+Einstein calculated that Mercury's perihelion would advance by 43 seconds of arc per century due to general relativistic effects; here I reproduce that analysis using numerical reasoning.
 
@@
 The orbit of Mercury has long been known to advance at a rate of about 6400'' [seconds of arc] per century.
-Various classical phenomena, such as the effects of Venus, and the oblateness of the sun, account for all but 44 seconds of arc and this remaining amount puzzled astronomers for many years.
-According to Einstien and others, the general theory of relativity predicts an additional precession of almost exactly 44 seconds per century.
+Various classical phenomena, such as the effects of Venus, and the oblateness of the sun, account for all but 43 seconds of arc and this remaining amount puzzled astronomers for many years.
+According to Einstien and others, the general theory of relativity predicts an additional precession of almost exactly 43 seconds per century.
 In this document I will calculate the precessional rate using numerical methods.
@@
 
-closely agreeing with Einstein's original prediction of 44.  A more careful analysis would account for the nonzero mass of Mercury.
+closely agreeing with Einstein's original prediction of 43.  A more careful analysis would account for the nonzero mass of Mercury.

### 2019-06-17 perihelion_advance_of_mercury.Rmd
more work

@@
 
-The orbit of Mercury has long been known to advance at a rate of about 6400'' [seconds of arc] per century.
-Various classical phenomena, such as the effects of Venus, and the oblateness of the sun, account for all but 43 seconds of arc and this remaining amount puzzled astronomers for many years.
-According to Einstien and others, the general theory of relativity predicts an additional precession of almost exactly 43 seconds per century.
-In this document I will calculate the precessional rate using numerical methods.
+The orbit of Mercury has long been known to advance at a rate of about
+575'' [seconds of arc] per century.  Various classical phenomena, such
+as the effects of Venus, and the oblateness of the sun, account for
+all but 43'' and this remaining amount puzzled astronomers for many
+years.  According to Einstien and others, the general theory of
+relativity predicts an additional precession of almost exactly 43
+seconds per century, thus closing the gap between expectation and
+observation.  In this document I will calculate the precessional rate
+using numerical methods.
 
@@
 
-closely agreeing with Einstein's original prediction of 43.  A more careful analysis would account for the nonzero mass of Mercury.
+closely agreeing with Einstein's original prediction of 43''.  A more
+careful analysis would account for the nonzero mass of Mercury.

### 2019-06-17 perihelion_advance_of_mercury.Rmd
theoretical analysis

@@
 
-In the above diagram we see over 15.4 orbits of Mercury, exhibiting no visible deviation from an exactly closed orbital path.
-To calculate the precession we need to simulate the motion for a longer time:
+In the above diagram we see over 15.4 orbits of Mercury, exhibiting no
+visible deviation from an exactly closed orbital path.  To calculate
+the precession we need to simulate the motion for a longer time:
 
@@
 
-We need to isolate the change points (actually calculating aphelion rather than perihelion for convencence):
+We need to isolate the change points (actually calculating aphelion
+rather than perihelion for convenience):
 
@@
   )
-
 ```
@@
 careful analysis would account for the nonzero mass of Mercury.
+
+##  Theoretical analysis
+
+In current notation, a first-order analysis gives
+
+\[
+\frac{24\pi^3a^2}{t^2c^2\left(1-e^2\right)}
+\]
+
+for the precession.  In R idiom:
+
+```{r}
+precession <- 24*pi^3*a^2/(per^2*sol^2*(1-ecc^2))  # radians/orbit
+precession
+```
+
+To convert this figure which is given in radians per orbit) to
+arcseconds per century, we need to multiply by a sequence of factors:
+
+```{r}
+seconds_per_orbit <- prod(c(
+    seconds_per_minute = 60,
+    minutes_per_hour   = 60,
+    hours_per_day      = 24,
+    days_per_orbit     = 87.969
+))
+
+seconds_per_century <- prod(c(
+    seconds_per_minute = 60,
+    minutes_per_hour   = 60,
+    hours_per_day      = 24,
+    days_per_year      = 365.25,
+    years_per_century  = 100
+))
+
+orbits_per_century <- seconds_per_century/seconds_per_orbit
+
+arcseconds_per_radian <- prod(
+    c(
+        degrees_per_radian = 180/pi,
+        minutes_per_degree = 60,
+        seconds_per_minute = 60
+   	     ))
+
+arcseconds_per_orbit <- precession*arcseconds_per_radian
+arcseconds_per_century <- arcseconds_per_orbit * orbits_per_century
+arcseconds_per_century
+```
+
+Compare with the value of about 43.37 obtained from numerical
+integration, showing close agreement.

### 2019-06-17 perihelion_advance_of_mercury.Rmd
typos and clarification

@@
 all but 43'' and this remaining amount puzzled astronomers for many
-years.  According to Einstien and others, the general theory of
+years.  According to Einstein and others, the general theory of
 relativity predicts an additional precession of almost exactly 43
@@
 
-Consider the following R idiom, in which ```trajectory()``` is used to directly simulate a geodesic in Schwarzschild space:
+Consider the following R idiom, in which ```trajectory()``` is used to
+directly simulate a geodesic in Schwarzschild space:
 
 ```{r}
-jj <- trajectory(30, 1/2, 3) # m=1/2 -> Schwarzschild radius = 1
+# Simulate a trajectory starting at r=30, mass of sun=1/2 so Schwarzschild radius=1:
+jj <- trajectory(r_start=30, m=1/2, h=3) # h=3 gives angular momentum
 plot(jj$x,jj$y, asp=1, type='l', axes=FALSE, xlab='', ylab='',main='Geodesics in the Schwarzschild geometry')
 polargrid(r=10*(1:3))
-event_horizon()
+event_horizon()  # show the event horizon, to scale
 ```
 
-In the above, we see a typical orbit in Schwarzschild space and a precession of around 40 degrees per orbit.
-As a consistency check we can perform the same calculation but with Newtonian approximations:
+(NB ```trajectory()``` takes geometrized units).  In the above, we see
+a typical trajectory in Schwarzschild space with a precession of
+around 40 degrees per orbit.  As a consistency check we can perform
+the same calculation but with Newtonian approximations:
 
@@
 
-and above we see that the orbit is accurately closed, as expected.
+and above we see that the orbit is accurately closed, as expected
+(three orbits are simulated, the lines overlap).
 
@@
 
-First we specify some orbital parameters for the orbit of Mercury, mostly taken from wikipedia:
+First we specify some orbital parameters for the orbit of Mercury, mostly taken from Wikipedia:
 ```{r}
@@
 
-And some futher features using classical geometry:
+And some further features using classical geometry:
 ```{r}
@@
 b <- sqrt(a^2-cee^2)
-ecc <- sqrt(1-(b/a)^2)  # should match ecc_wiki above
+ecc <- sqrt(1-(b/a)^2)  # should match eccentricity given by Wikipedia
 ell <- (a^2-cee^2)/a  # semi-latus rectum
@@
  out <- trajectory(merc_perihelion, m=d, h=h, tau=seq(from=0,to=35e15,len=5e5))
- plot(out$x,out$y,type='l',asp=1,axes=FALSE,xlab='',ylab='')
+ plot(out$x,out$y,type='l',asp=1,axes=FALSE,xlab='',ylab='',main='Orbit of Mercury')
  event_horizon()
@@
 In the above diagram we see over 15.4 orbits of Mercury, exhibiting no
-visible deviation from an exactly closed orbital path.  To calculate
-the precession we need to simulate the motion for a longer time:
+visible deviation from an exactly closed orbital path.  Nevertheless,
+the orbit is not closed as Mercury has a tiny amount of orbital
+precession.  To calculate this precession we need to simulate the
+motion for a longer time:
 
@@
 
-plot(tee,phi-pi)
+plot(tee,phi-pi,main='orbital angle at aphelion',xlab='time (seconds)',ylab='angle of aphelion')
 ```
 
-There is a lot of noise on this caused by the discrete nature of the simulation results provided by ```ode()```.  To estimate the precession we can use linear regression:
+The above diagram shows a clear trend, somewhat obscured by a vertical
+spread due to the discrete nature of the simulation.  To estimate the
+underlying trend, which is the general relativistic precession of
+Mercury, we can use linear regression:
 
@@
 
-To convert this figure which is given in radians per orbit) to
-arcseconds per century, we need to multiply by a sequence of factors:
+To convert this figure (which is given in radians per orbit) to the
+more usual units of arcseconds per century, we need to multiply by a
+sequence of factors:
 

### 2019-06-19 light_start_at_radius_of_sun.R
tidyup and dramatic pause

@@ -6,7 +6,14 @@
 
 ## More documentation is given in light_starts_at_r_equals_2.R
 
-source("usefullightfuncs.R") # defines lightpoints()
+cat(paste("Einstein's prediction: ",4*G*M/sol^2/radius_of_sun*180/pi*60^2,"\n"))
+cat(paste("Eddington's measurement: ",1.75,"\n"))
+## calculated value printed at the end
+cat("calculating...\n")
+Sys.sleep(10)  # dramatic pause (the actual calculation is almost instantaneous)
+
+
+source("usefullightfuncs.R") # defines nullgeodesic()
 
 G <- 6.67408e-11  # Gravitational constant (SI)
 M <- 1.989e30     # mass of sun (SI)
@@ -15,7 +22,7 @@ radius_of_sun <- 695510e3 # (m)
 r_schwarz <- 2*G*M/sol^2 # Schwarzschild radius of sun (m)
 r_start  <- radius_of_sun/r_schwarz # starting radius for light ray
 
-initialangletotangent <- 0  # tangential
+initialangletotangent <- 0  # tangential ray
  
 small <- 0.1
 
@@ -32,6 +39,27 @@ light <-
 
 jj <- tail(light,2)
 
-deflection <- -2*as.vector(diff(jj[,1]) / diff(jj[,2])) # radians (NB double)
+angle <- as.vector(diff(jj[,1]) / diff(jj[,2]))  # radians
 
-print(deflection*180/pi*60^2)  # arcseconds; compare Einstein's value of 1.75
+deflection <- -2*angle
+
+## In the above, note multiplication by [minus] two because the
+## astronomical measurement---which we are trying to predict---is of
+## total deflection of distant starlight.  The total deflection is
+## comprised of deflection in the ingoing leg *plus* deflection on the
+## outgoing leg.  Function nullgeodesic() gives deflection on the
+## outgoing leg, as the light starts at (r=radius_of_sun, phi=0) [that
+## is, at the 3 o'clock position, at the radius of the sun, moving
+## parallel to the y-axis].  Variable 'angle' is the angle that the
+## light beam makes from the y-axis, at a large distance [one day I'll
+## get round to deriving this properly] from the sun.  Ingoing
+## deflection and outgoing deflection are equal by symmetry, and we
+## need to add the two deflections, which is why we have a factor of 2
+## here.
+
+
+
+arcseconds <- deflection*180/pi*60^2
+
+
+cat(paste("direct numerical integration:", arcseconds, "\n"))

### 2019-06-19 light_start_at_radius_of_sun.R
bugfix

@@ -6,14 +6,8 @@
 
 ## More documentation is given in light_starts_at_r_equals_2.R
 
-cat(paste("Einstein's prediction: ",4*G*M/sol^2/radius_of_sun*180/pi*60^2,"\n"))
-cat(paste("Eddington's measurement: ",1.75,"\n"))
-## calculated value printed at the end
-cat("calculating...\n")
-Sys.sleep(10)  # dramatic pause (the actual calculation is almost instantaneous)
-
-
 source("usefullightfuncs.R") # defines nullgeodesic()
+
 
 G <- 6.67408e-11  # Gravitational constant (SI)
 M <- 1.989e30     # mass of sun (SI)
@@ -21,6 +15,16 @@ sol <- 299792458  # speed of light
 radius_of_sun <- 695510e3 # (m)
 r_schwarz <- 2*G*M/sol^2 # Schwarzschild radius of sun (m)
 r_start  <- radius_of_sun/r_schwarz # starting radius for light ray
+
+
+cat(paste("Einstein's prediction: ",4*G*M/sol^2/radius_of_sun*180/pi*60^2,"\n"))
+cat(paste("Eddington's measurement: ",1.75,"\n"))
+## calculated value printed at the end
+cat("calculating...\n")
+Sys.sleep(10)  # dramatic pause (the actual calculation is almost instantaneous)
+
+
+
 
 initialangletotangent <- 0  # tangential ray
  

### 2019-06-20 perihelion_advance_of_mercury.Rmd
docs
consistency check with GR=FALSE orbit

@@
 the orbit is not closed as Mercury has a tiny amount of orbital
-precession.  To calculate this precession we need to simulate the
-motion for a longer time:
+precession.  Note that the Schwarzschild radius would be microsopic at
+this scale (the size of the central disk is dominated by the width of
+the plotting line used for the event horizon).  To calculate this
+precession we need to simulate the motion for a longer time:
 
@@
 
-We need to isolate the change points (actually calculating aphelion
-rather than perihelion for convenience):
+We need to isolate the change points, here found isolating iterations
+$i$ such that ```r[i-1]<r[i]``` and r[i]>r[i+1]```:
 
@@
 
-plot(tee,phi-pi,main='orbital angle at aphelion',xlab='time (seconds)',ylab='angle of aphelion')
+plot(tee,phi-pi,main='orbital angle of Mercury at perihelion',xlab='time (seconds)',ylab='angle of perihelion')
 ```
@@
 careful analysis would account for the nonzero mass of Mercury.
+
+By way of a consistency check, we can plot a similar diagram but with
+the classical approximation:
+
+```{r}
+out <- trajectory(merc_perihelion, m=d, h=h, tau=seq(from=0,to=35e16,len=1e7),GR=FALSE)
+jj1 <- c(NA,out$rdot)
+jj2 <- c(out$rdot,NA)
+change <- (jj1>0) & (jj2<0)
+tee    <- out$time[change]
+phi <- out$phi[change]
+phi <- (phi) %% (2*pi)
+plot(tee,phi-pi,main='orbital angle at of Mercury at perihelion, classical case',xlab='time (seconds)',ylab='angle of perihelion')
+```
+
+In the above plot, note how there is no trend, consistent with the
+Newtonian case of zero perihelion advance.
 

### 2019-07-07 perihelion_advance_of_mercury.Rmd
slight clarification

@@
  out <- trajectory(merc_perihelion, m=d, h=h, tau=seq(from=0,to=35e15,len=5e5))
- plot(out$x,out$y,type='l',asp=1,axes=FALSE,xlab='',ylab='',main='Orbit of Mercury')
- event_horizon()
+ plot(out$x,out$y,type='l',asp=1,axes=FALSE,xlab='',ylab='',main='Orbit of Mercury, Sun not to scale')
+ points(0,0,pch=21,bg='yellow',col='orange') # Schwarzschild radius invisible at this scale
  max(out$phi/(2*pi))

### 2019-07-07 trajectory_mercury.R
clarification of inline docs

@@ -4,7 +4,7 @@
 
 ## Notation follows Rindler, p239
 
-source("usefulgeodesicfuncs.R")
+source("usefulgeodesicfuncs.R")  # defines trajectory()
 source("mercury.R")  # orbital parameters for Mercury in geometrized units
      
 out <- trajectory(merc_perihelion, m=d, h=h, tau=seq(from=0,to=35e16,len=1e7))

### 2019-07-07 usefulgeodesicfuncs.R

@@ -1,3 +1,5 @@
+# Some useful geodesic functions; used by (eg) trajectory_mercury.R
+
 library("deSolve")   # ode()
 source("usefulfuncs.R")
 

### 2019-07-07 maker.R
rename file to a less misleading name

@@ -43,7 +43,7 @@ pdf(file="light_closest_approach.pdf")
 source("light_closest_approach.R")
 dev.off()
 
-pdf(file="trajectory.pdf")
+pdf(file="illustrative_relativistic_trajectory.pdf")
 source("trajectory.R")
 dev.off()
 

### 2019-07-08 shapiro.R
Shapiro delay work

@@ -1,0 +1,37 @@
+# Numerical verification of the Shapiro delay experiment
+
+source("usefulnullgeodesicfuncs.R")
+
+st <- function(x){(x-min(x))/(max(x)-min(x))}
+
+r0 <- r_sun/d # grazing incidence of the sun in Schwarzschild radii
+b_radial <- r0/sqrt(1-1/r0)
+
+small <- 1e-6  # needed to prevent numerical problems
+
+d_earth <- seq(from=r_earth/r_s * 0.999, to=r_earth/r_s * 1.001,len=100)
+d_venus <- seq(from=r_venus/r_s * 0.999, to=r_venus/r_s * 1.001,len=100)
+
+
+out_earth <- nullgeodesic(r_start=r0,b=b_radial*(1-small),tau=c(0,d_earth))
+out_venus <- nullgeodesic(r_start=r0,b=b_radial*(1-small),tau=c(0,d_venus))
+
+g_earth <- approxfun(out_earth$r,out_earth$r)
+g_venus <- approxfun(out_venus$r,out_venus$r)
+
+
+transit_time_relativistic  <-   # transit time in seconds
+(g_earth(r_earth)+g_venus(r_venus))*r_s/sol
+
+  
+transit_time_classical <- (r_earth + r_venus)/sol # seconds
+
+plot(out_earth$x,out_earth$y,type='o',col='red',cex=st(out_earth$time),asp=1)
+event_horizon()
+
+
+
+shapiro_delay <- transit_time_relativistic - transit_time_classical
+# compare observation of 200 microseconds (!)
+
+

### 2019-07-11 string_grg.tex
multiply defined labels
more work

@@
 \section{Introduction}
-\label{intro}
 
-\section{Introduction} % Section titles are automatically converted to all-caps.
-% Section numbering is automatic.
-
-\label{intro}
 Light inextensible strings are a stalwart feature of elementary
@@
 (proper) unit length~$\mu$, then a string will behave as though light
-on lengthscales $L\lesssim T/\left(\mu g\right)$.  Using Newtonian
-expressions for the strength of gravity at the event horizon of a
-black hole with mass $M$ gives $L\lesssim TM/\left(\mu c^4\right)$.
-For the string to behave as though light on lengthscales comparable
-with the Schwarzschild radius would require $T\gtrsim\mu c^2$.
-Exceeding this criterion would violate the dominant energy condition
-(and in this context we note that transverse waves would propagate at
-speed~$\sqrt{T/\mu}\sim c$).  Nevertheless, the idealization of a
-string as light and inextensible is an interesting and informative
-limit: its analysis interprets the concept ``straight line'' and
-suggests a possible way of manipulating a black hole.  Peculiar
-counterintuitive behaviour such as inward-directed centrifugal force
-in the vicinity of a black hole is well
+on lengthscales $L\lesssim T/\left(\mu g\right)$.  Using
+$c^4/\left(4GM\right)$ for the strength of gravity at the event
+horizon of a black hole with mass $M$ gives $L\lesssim TM/\left(\mu
+c^4\right)$.  For the string to behave as though light on lengthscales
+comparable with the Schwarzschild radius would require $T\gtrsim\mu
+c^2$.  Exceeding this criterion would violate the dominant energy
+condition (and in this context we note that transverse waves would
+propagate at speed~$\sqrt{T/\mu}\sim c$).  Nevertheless, the
+idealization of a string as light and inextensible is an interesting
+and informative limit: Rindler~\cite{rindler}, for example, considers
+lowering a heavy object to the event horizon on the end of such a
+string to illustrate energetic arguments in GR.
+
+Analysis of non-radial strings interprets the concept ``straight
+line'' and suggests a possible way of manipulating a black hole.
+Peculiar counterintuitive behaviour such as inward-directed
+centrifugal force in the vicinity of a black hole is well
 known~\cite{abramowicz1992,abramowicz1993}, and the analysis of taut

### 2019-08-26 string_physics_teacher.tex
try "The Physics Teacher".  This needs jss.bst

--- 2019-07-11 string_grg.tex
@@
-%%%%%%%%%%%%%%%%%%%%%%% file template.tex %%%%%%%%%%%%%%%%%%%%%%%%%
-%
-% This is a general template file for the LaTeX package SVJour3
-% for Springer journals.          Springer Heidelberg 2010/09/16
-%
-% Copy it to a new file with a new name and use it as the basis
-% for your article. Delete % signs as needed.
-%
-% This template includes a few options for different layouts and
-% content for various journals. Please consult a previous issue of
-% your journal as needed.
-%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%
-% First comes an example EPS file -- just ignore it and
-% proceed on the \documentclass line
-% your LaTeX will extract the file if required
-\begin{filecontents*}{example.eps}
-%!PS-Adobe-3.0 EPSF-3.0
-%%BoundingBox: 19 19 221 221
-%%CreationDate: Mon Sep 29 1997
-%%Creator: programmed by hand (JK)
-%%EndComments
-gsave
-newpath
-  20 20 moveto
-  20 220 lineto
-  220 220 lineto
-  220 20 lineto
-closepath
-2 setlinewidth
-gsave
-  .4 setgray fill
-grestore
-stroke
-grestore
-\end{filecontents*}
-%
-\RequirePackage{fix-cm}
-%
-%\documentclass{svjour3}                     % onecolumn (standard format)
-%\documentclass[smallcondensed]{svjour3}     % onecolumn (ditto)
-\documentclass[smallextended]{svjour3}       % onecolumn (second format)
-%\documentclass[twocolumn]{svjour3}          % twocolumn
-%
-\smartqed  % flush right qed marks, e.g. at end of proof
-%
-\usepackage{graphicx}
+\documentclass[review]{elsarticle}
+
+\usepackage{lineno,hyperref}
+\usepackage{amsmath}
 \usepackage{amssymb}
-%
-% \usepackage{mathptmx}      % use Times fonts if available on your TeX system
-%
-% insert here the call for the packages your document requires
-%\usepackage{latexsym}
-% etc.
-%
-% please place your own definitions here and don't use \def but
-% \newcommand{}{}
-%
-% Insert the name of "your journal" with
-% \journalname{myjournal}
-%
+\usepackage{booktabs}
+\usepackage{units}
+\usepackage{xcolor}
+\modulolinenumbers[5]
+
+\journal{The Physics Teacher}
+
+\bibliographystyle{jss}
+
 \begin{document}
-
-\title{Light inextensible strings under tension in the Schwarzschild geometry}%\thanks{Grants or other notes
-%about the article that should go on the front page should be
-%placed here. General acknowledgments should be placed at the end of the article.}
-
-%\subtitle{Do you have a subtitle?\\ If so, write it here}
-
-%\titlerunning{Short form of title}        % if too long for running head
-
-\author{Robin K. S. Hankin}
-
-%\authorrunning{Short form of author list} % if too long for running head
-
-\institute{Robin K. S. Hankin \at
-  Auckland University of Technology\\
-  Wellesley Street\\
-  Auckland, New Zealand\\
-              Tel.: +64(9)5219999\\
-              \email{hankin.robin@gmail.com}           %  \\
-}
-
-\date{Received: date / Accepted: date}
-% The correct dates will be entered by the editor
-
-\maketitle
+\begin{frontmatter}
+\title{Light inextensible strings under tension in the Schwarzschild geometry}
 
@@
 suggests a possible way of manipulating a black hole.
-\keywords{Black holes\and inextensible string\and minimal-length path}
+
 % \PACS{PACS code1 \and PACS code2 \and more}
@@
 
+\begin{keyword}
+Black holes\sep light inextensible string\sep minimal-length path
+\end{keyword}
+
+\end{frontmatter}
+
+\linenumbers
 \section{Introduction}
 
-Light inextensible strings are a stalwart feature of elementary
+Light inextensible strin
+gs are a stalwart feature of elementary
 physics.  A light string is not subject to gravity, and being under
@@
 
-% BibTeX users please use one of
-%\bibliographystyle{spbasic}      % basic style, author-year citations
-\bibliographystyle{spmpsci}      % mathematics and physical sciences
-%\bibliographystyle{spphys}       % APS-like style for physics
-%\bibliography{}   % name your BibTeX data base
+\subsection*{Supplementary material}
 
-\bibliography{stringrefs}
+R code for all statistical analysis given here is available at
+
+{\tt https://github.com/RobinHankin/hyper2}.
+
+\section*{References}
+
+\bibliography{stringrefs.bib}
 
 \end{document}
-% end of file template.tex

### 2019-08-26 string_physics_teacher.tex
initial work, emphasising the educational nature of the results

@@
 
-Light inextensible strin
-gs are a stalwart feature of elementary
-physics.  A light string is not subject to gravity, and being under
-tension means that its configuration is an extremum of proper length.
-Of course, no real string is perfectly weightless or inextensible.
-With tension~$T$, gravitational acceleration $g$ and (proper) mass per
+General relativity is a topic of long-standing fascination for
+undergraduate physics students [CITATION NEEDED].  Einstein's concept
+of ``curved'' spacetime is an enduring metaphor that attracts large
+numbers of undergraduates into courses such as AUT's frontiers of
+astronomy courses.
+
+Students are weary of being told that ``space is warped'' and ``not
+even light can escape a black hole''.  Students typically stumble on
+the concept of warped spacetime on the grounds that torsion-free
+metrics are not embedded IN ANYTHING.
+
+Here I show how numerical methods can be used to interpret the concept
+of ``straight line'' in the vicinity of a black hole.  As
+\cite{overduin2018} point out, demonstrations of relativistic concepts
+using everyday materials (in their case gravitational waves visualised
+by oscillations in spandex) can exert a galvanising effect on
+students.
+
+Light inextensible strings are a stalwart feature of elementary
+physics.  It is perhaps not out of place to point out explicitly that
+a light string is not subject to gravity, and its being under tension
+means that its configuration is an extremum of proper length.  Of
+course, no real string is perfectly weightless or inextensible.  With
+tension~$T$, gravitational acceleration $g$ and (proper) mass per
 (proper) unit length~$\mu$, then a string will behave as though light

### 2019-08-27 string_physics_teacher.tex
more work for a teaching audience

--- 2019-08-26 string_physics_teacher.tex
@@
 
-  Here I analyze the behaviour of stationary, light inextensible strings
-under tension in the vicinity of a nonrotating black hole.  Such
-strings minimize their proper length, given by integrating the spatial
-component of the Schwarzschild metric along the string.  The path is
-calculated by numerical solution to the Euler-Lagrange equations.  If
-the string is entirely outside the event horizon, its closest approach
-to the singularity is tangential; if the distance of closest
-approach is less than about $1.0761\cdot 2M$, the
-string self-intersects.  If the string attains $r=2M$, then it
-will in general be tangential to the event horizon.  Light taut
-strings interpret the concept ``straight line'' and their behaviour
-suggests a possible way of manipulating a black hole.
+Light inextensible string under tension is a stalwart feature of
+elementary physics.  Here I show how considering such string in the
+vicinity of a black hole, together with computer algebra systems, can
+generate insight into the Schwarzschild geometry that is accessible to
+undergraduates.  Light taut strings minimize their proper length,
+given by integrating the spatial component of the Schwarzschild metric
+along the string.  The path itself is given by straightforward
+numerical solution to the Euler-Lagrange equations.  If the string is
+entirely outside the event horizon, its closest approach to the
+singularity is tangential.  At this point the string is {\em visibly}
+curved, surely a memorable and informative insight.  The geometry of
+the Schwarzschild metric induces some peculiar nonlocal behaviour: if
+the distance of closest approach is less than about $1.0761\cdot 2M$,
+the string self-intersects, even though it is everywhere under
+tension.  Light taut strings furnish a third interpretation of the
+concept ``straight line'', the other two being null geodesics and
+free-fall world lines.
 
@@
 
-Light inextensible strin
-gs are a stalwart feature of elementary
-physics.  A light string is not subject to gravity, and being under
-tension means that its configuration is an extremum of proper length.
+General relativity is a topic of long-standing fascination for
+undergraduate physics students [CITATION NEEDED].  Einstein's concept
+of ``curved'' spacetime is an enduring metaphor that attracts large
+numbers of undergraduates into introductory general relativity.
+
+Here I show how numerical methods, including computer algebra systems,
+can be used to interpret the concept of ``straight line'' in the
+vicinity of a black hole.  The problem is would be suitable for an
+undergraduate homework problem~\cite{ajpeditor}.
+
+A light string's being under tension means that its path will be an
+extremum of proper length: the calculus of variations proves that the
+path is straight in the sense of zero curvature.  In terrestrial
+gravity, standard $25\times 10^{-6}\mbox{kg}\cdot\mbox{m}^{-1}$ cotton
+sewing thread at its breaking strain of about $40\,\mbox{N}$ would
+have a sag of~$\sim 6\times 10^{-6}\mbox{m}$ over a $1\,\mbox{m}$
+horizontal scale: barely perceptible.  How does such a quotidian
+observation translate into the vicinity of a black hole?  Would the
+``warping of space'' result in {\em visibly} curved thread?
+
 Of course, no real string is perfectly weightless or inextensible.
@@
 
-Analysis of non-radial strings interprets the concept ``straight
-line'' and suggests a possible way of manipulating a black hole.
-Peculiar counterintuitive behaviour such as inward-directed
-centrifugal force in the vicinity of a black hole is well
-known~\cite{abramowicz1992,abramowicz1993}, and the analysis of taut
-strings furnishes another everyday window into general relativity.
-Another possibility, discussed by Thorne~\cite{thorne1986}, is to pass
-electric current through such taut strings, an option not pursued
-here.
+Dynamic definitions of straight line also run into difficulties:
+counterintuitive behaviour such as inward-directed centrifugal force
+in the vicinity of a black hole is well
+known~\cite{abramowicz1992,abramowicz1993}.
 
@@
 
-Elementary methods show that the proper length $\ell$ of radial strings has
-some peculiar features.  Firstly, consider the proper length from
-$r=2M$ to $r=2M(1+w)$ where $w>0, \left|w\right|\ll 1$.  Then
+Elementary methods show that the proper length $\ell$ of radial
+strings has some non-Euclidean features.  Firstly, consider the proper
+length from $r=2M$ to $r=2M(1+w)$ where $w>0, \left|w\right|\ll 1$.
+Then
 
@@
 horizon.  Radial strings have proper length longer than the
-Schwarzschild radial coordinate difference $w$.
+Schwarzschild radial coordinate difference $w$, a nice illustration of
+the failure of Euclidean geometry.
 
@@
 $u''=3u^2/2-u$, but is equivalent
-to~$r''\left(\phi\right)=r-3M+4Mr'/r$.
+to~$r''\left(\phi\right)=r-3M+2M\left(r'\right)^2/r$.  This shows that
+light, being sensitive to the time component of the metric, behaves
+differently from string, which is not.
 
@@
 at~$r=r_0$, has a radius of curvature of $r_0<r_0^2/2M$, showing that
-circular orbits cannot follow a taut string at closest approach.  Also
-observe that null geodesics have a radius of curvature of $r_0^2/3M <
-r_0^2/2M$, showing that light cannot follow a taut string at closest
-approach either.  The string has a point of inflection
-if~$\left(r'\right)^2=2r(r-2M)$.
+circular orbits cannot follow a taut string at closest approach (as in
+the classical case).  Also observe that null geodesics have a radius
+of curvature of $r_0^2/3M < r_0^2/2M$, showing that light cannot
+follow a taut string at closest approach either.  The string has a
+point of inflection if~$\left(r'\right)^2=2r(r-2M)$.
 

### 2019-08-27 string_physics_teacher.tex
more work

@@
 General relativity is a topic of long-standing fascination for
-undergraduate physics students [CITATION NEEDED].  Einstein's concept
-of ``curved'' spacetime is an enduring metaphor that attracts large
-numbers of undergraduates into introductory general relativity.
-
-Here I show how numerical methods, including computer algebra systems,
-can be used to interpret the concept of ``straight line'' in the
-vicinity of a black hole.  The problem is would be suitable for an
-undergraduate homework problem~\cite{ajpeditor}.
+undergraduate physics students~\cite{christensen2012}.  Here I show
+how numerical methods, including computer algebra systems, can be used
+to interpret the concept of ``straight line'' in the vicinity of a
+black hole.  The analysis would be suitable for an undergraduate
+homework problem~\cite{romano2019}.
 
@@
 path is straight in the sense of zero curvature.  In terrestrial
-gravity, standard $25\times 10^{-6}\mbox{kg}\cdot\mbox{m}^{-1}$ cotton
-sewing thread at its breaking strain of about $40\,\mbox{N}$ would
-have a sag of~$\sim 6\times 10^{-6}\mbox{m}$ over a $1\,\mbox{m}$
-horizontal scale: barely perceptible.  How does such a quotidian
-observation translate into the vicinity of a black hole?  Would the
-``warping of space'' result in {\em visibly} curved thread?
+gravity, standard $25\times 10^{-6}\unit{kg}\cdot\meter^{-1}$ cotton
+sewing thread at its breaking strain of about $40\unit{N}$ would have
+a sag of~$\sim 6\times 10^{-6}\meter$ over a $1\meter$ horizontal
+scale: barely perceptible.  How does such a quotidian observation
+translate into the vicinity of a black hole?  Would the ``warping of
+space'' result in {\em visibly} curved thread?
 
@@
 only one free end; the other is tangential to the event horizon (in
-the sense that ${\displaystyle \lim_{r\longrightarrow 1^+}r'=0}$),
+the sense that ${\displaystyle \lim_{r\longrightarrow 2M^+}r'=0}$),
 except in the degenerate case of constant~$\phi$ when the string is

### 2019-08-27 string_physics_teacher.tex
use \geqslant instead of \geq; much classier

@@
 component~$ds^2= dr^2/\left(1-2M/r\right) + r^2d\phi^2$, confined to
-equatorial points~($\theta=\pi/2$); we understand~$2M\leq r$
+equatorial points~($\theta=\pi/2$); we understand~$r\geqslant 2M$
 throughout.  A circular string at constant~$r$ will, trivially, have
@@
 
-Considering first the radial component we may calculate the
-proper length $L\left(r_1,r_2\right)$ of a string stretching
-from~$r=r_1\geq 2M$ to~$r_2\geq r_1$ at constant $\theta,\phi$:
+Considering first the radial component we may calculate the proper
+length $L\left(r_1,r_2\right)$ of a string stretching
+from~$r=r_1\geqslant 2M$ to~$r_2\geqslant r_1$ at constant
+$\theta,\phi$:
 
@@
 \includegraphics[width=170mm]{closest_approach_selfintersecting.pdf}
-\caption{Self-intersecting strings shown for $\phi\geq 0$.  Closest
+\caption{Self-intersecting strings shown for $\phi\geqslant 0$.  Closest
   approach to the event horizon occurs at $\phi=0$ (3 o'clock) and
   $r/2M\in\left(1,1.0761\right)$; strings are symmetrical about $\phi=0$
-  but for clarity only $\phi\geq 0$ is shown}
+  but for clarity only $\phi\geqslant< 0$ is shown}
 \label{closest_approach_self_intersecting}
@@
 \includegraphics{light_start_at_r_equals_2.pdf} % created by "angle_start_at_r_equals_2.R"
-\caption{Null geodesics passing through $(4M,0)$, shown for $0\leq\phi\leq 2\pi$}
+\caption{Null geodesics passing through $(4M,0)$, shown for
+  $0\leqslant\phi\leqslant 2\pi$}
 \label{light_r_equals_2}

### 2019-08-28 string_physics_teacher.tex
work

@@
 
-Dynamic definitions of straight line also run into difficulties:
-counterintuitive behaviour such as inward-directed centrifugal force
-in the vicinity of a black hole is well
-known~\cite{abramowicz1992,abramowicz1993}.
+\section{The Schwarzschild metric}
 
-The Schwarzschild metric
+The vacuum Einstein field equations for a static, spherically
+symmetric mass are solved by the Schwarzschild
+metric~\cite{schwarzschild1916}.  In units in which~$G=c=1$, this is:
 
 \begin{equation}\label{schwarzschild}
-ds^2= -dt^2\left(1-2M/r\right) +\frac{dr^2}{1-2M/r} + r^2\left(d\theta^2 + \sin^2\theta d\phi^2\right),
+ds^2= -dt^2\left(1-2M/r\right) +\frac{dr^2}{1-2M/r} + r^2\left(d\theta^2 + \sin^2\theta d\phi^2\right).
 \end{equation}
 
-\noindent where units in which~$G=c=1$ are used, has been known for
-over a century~\cite{schwarzschild1916}.  Here I consider the spatial
-component~$ds^2= dr^2/\left(1-2M/r\right) + r^2d\phi^2$, confined to
-equatorial points~($\theta=\pi/2$); we understand~$r\geqslant 2M$
-throughout.  A circular string at constant~$r$ will, trivially, have
-proper length~$2\pi r$.
+\noindent Here I consider the spatial component~$ds^2=
+dr^2/\left(1-2M/r\right) + r^2d\phi^2$, confined to equatorial
+points~($\theta=\pi/2$); we understand~$r\geqslant 2M$ throughout.  A
+circular string at constant~$r$ will, trivially, have proper
+length~$2\pi r$.
 
-Considering first the radial component we may calculate the proper
-length $L\left(r_1,r_2\right)$ of a string stretching
-from~$r=r_1\geqslant 2M$ to~$r_2\geqslant r_1$ at constant
-$\theta,\phi$:
+Considering first the radial component we may calculate the
+proper length $L\left(r_1,r_2\right)$ of a string stretching
+from~$r=r_1\geqslant 2M$ to~$r_2\geqslant r_1$ at constant $\theta,\phi$:
 
@@
 
-This would indicate that, close to the Schwarzschild radius $r=2M$,
-large (but finite) amounts of string are required to connect two
-points with slightly differing radial coordinates.  Passing to the
-case $w\gg 1$ we have an asymptotic expansion that begins
+(compare the flat space string length of $2Mw$).  This would indicate
+ that, close to the Schwarzschild radius $r=2M$, large (but finite)
+ amounts of string are required to connect two points with slightly
+ differing radial coordinates.  Passing to the case $w\gg 1$ we have
+ an asymptotic expansion that begins
 
@@
 
-\noindent suggesting that there is ``extra radial space'', of length
-over $1.19\cdot 2M$, sensible at large distances from the event
-horizon.  Radial strings have proper length longer than the
-Schwarzschild radial coordinate difference $w$, a nice illustration of
+\noindent suggesting that there is ``extra space'', sensible at large
+distances from the event horizon, with radial strings being
+arbitrarily longer than flat space would allow: a nice illustration of
 the failure of Euclidean geometry.
@@
 $\frac{d}{d\phi}\frac{\partial F}{\partial r'}-\frac{\partial
-  F}{\partial r}=0$ give, after simplification, an expression for the
-second derivative of~$r$:
+  F}{\partial r}=0$ give, after simplification\footnote{The Mathematica
+  idiom would be:
 
-% \begin{equation}\label{rdashdash_properunits}
-%   r''\left(\phi\right) =
-%   (r-1) + \frac{(4r-3)\left(r'\right)^2}{2r\left(r-1\right)},\qquad r>1.
-% \end{equation}
+\noindent
+{\tt
+Needs["VariationalMethods`"]\\
+p = EulerEquations[Sqrt[y'[x]\^{}2/(1-1/y[x]) + y[x]\^{}2],y[x],x] // FullSimplify\\
+ss = p[[1]]//Numerator\\
+DSolve[ss==0,y[x],x]
+}}, an expression for the second derivative of~$r$:
+
 \begin{equation}\label{rdashdash}
@@
 of curvature of $r_0^2/3M < r_0^2/2M$, showing that light cannot
-follow a taut string at closest approach either.  The string has a
-point of inflection if~$\left(r'\right)^2=2r(r-2M)$.
+follow a taut string at closest approach either: tangential string
+looks curved to a local observer.
 
@@
 
-Here I solve equation~\ref{rdashdash} using the \verb+deSolve+ package
-for ordinary differential equations~\cite{soetart2010}.  Non-radial
-solutions fall into two distinct classes: those with $r>2M$ at all
-points, and those which attain $r=2M$, in which case the string is
-tangential to the event horizon.
+Here I solve equation~\ref{rdashdash} numerically, using the
+\verb+deSolve+ package for ordinary differential
+equations~\cite{soetart2010}.  Non-radial solutions fall into two
+distinct classes: those with $r>2M$ at all points, and those which
+attain $r=2M$, in which case the string is tangential to the event
+horizon.
 
@@
 less than this value, then the string crosses itself at~$\phi=\pi/2$.
-As discussed above, if $\left(r'\right)^2=2r(r-2M)$, the string has a
-point of inflection at which point the radius of curvature becomes
-infinite.  Figure~\ref{curvature_switch} shows non-self intersecting
-strings rotated so that the point of inflection occurs at~$\phi=0$; we
-see that each string has a finite length in which it is convex to the
-origin and two semi-infinite stretches in which the curvature is
-negative.
+
+Equation~\ref{roc} shows that the string has a point of inflection
+if~$\left(r'\right)^2=2r(r-2M)$ at which point the radius of curvature
+becomes infinite.  Figure~\ref{curvature_switch} shows non-self
+intersecting strings rotated so that the point of inflection occurs
+at~$\phi=0$; we see that each string has a finite length in which it
+is convex to the origin and two semi-infinite stretches in which the
+curvature is negative.
 
@@
 relativity, others being null geodesics and world lines of massive
-objects.  Comparing a world line with that of nearby taut strings thus
+objects.  Comparing light paths with that of nearby taut strings thus
 furnishes another insight into path curvature, complementing the
@@
   $r/2M\in\left(1,1.0761\right)$; strings are symmetrical about $\phi=0$
-  but for clarity only $\phi\geqslant< 0$ is shown}
+  but for clarity only $\phi\geqslant 0$ is shown}
 \label{closest_approach_self_intersecting}

### 2019-08-28 string_physics_teacher.tex
more work

@@
 point with~$r_0>2M$, then it has a radius of curvature of~$r_0^2/2M$
-at that point.  Any circular path, including a circular orbit
-at~$r=r_0$, has a radius of curvature of $r_0<r_0^2/2M$, showing that
-circular orbits cannot follow a taut string at closest approach (as in
-the classical case).  Also observe that null geodesics have a radius
-of curvature of $r_0^2/3M < r_0^2/2M$, showing that light cannot
-follow a taut string at closest approach either: tangential string
-looks curved to a local observer.
+there.  Tangential null geodesics have a radius of curvature of
+$r_0^2/3M$, strictly less than that of a string.  This shows that a
+taut string would {\em look} curved to a local observer, having a
+(visual) radius of curvature of $r_0^2/6M$.  At the photon sphere of a
+solar mass black hole, such a string would appear to have a radius of
+about $2\unit{km}$, a readily understandable result.
 

### 2019-08-28 string_physics_teacher.tex
further work on radius, there is no such thing as "the" radius of curvature, only comparison with other curved paths (such as null geodesics)

@@
 Considering~$r'(\phi)=0$ shows that if a string is tangential at a
-point with~$r_0>2M$, then it has a radius of curvature of~$r_0^2/2M$
-there.  Tangential null geodesics have a radius of curvature of
-$r_0^2/3M$, strictly less than that of a string.  This shows that a
-taut string would {\em look} curved to a local observer, having a
-(visual) radius of curvature of $r_0^2/6M$.  At the photon sphere of a
-solar mass black hole, such a string would appear to have a radius of
-about $2\unit{km}$, a readily understandable result.
+point with~$r_0>2M$, then it obeys, to first order, $r''=r-2M$, while
+tangential null geodesics follow $r''=r-3M$.  This shows that a taut
+string would {\em look} curved to a local observer, having a (visual)
+radius of curvature of $r_0^2/M\sqrt{1-2M/r_0}$ (the extra term comes
+from the fact that appearance is driven by local $ds$, not
+Schwarzschild $r$).  At the photon sphere $r_0=3M$, such a string
+would appear to have a radius of~$\sqrt{27}M$.  For a solar mass black
+hole this would be about $15\unit{km}$, a readily understandable
+result.
 
@@
 \includegraphics[width=170mm]{light_closest_approach.pdf}
+
 \caption{Null geodesics in the Schwarzschild geometry, tangential

### 2019-08-28 string_physics_teacher.tex
remove erroneous graph

@@
 
-Equation~\ref{roc} shows that the string has a point of inflection
-if~$\left(r'\right)^2=2r(r-2M)$ at which point the radius of curvature
-becomes infinite.  Figure~\ref{curvature_switch} shows non-self
-intersecting strings rotated so that the point of inflection occurs
-at~$\phi=0$; we see that each string has a finite length in which it
-is convex to the origin and two semi-infinite stretches in which the
-curvature is negative.
-
 There are other ways to parametrize completely exterior strings;
@@
 \centering
-\includegraphics{curvature_switch.pdf} % created by "radius_of_curvature_switch.R"
-\caption{Strings arranged by point at which the radius of curvature
-  switches, which occurs at~$\phi=0$.  Colour changes from black to
-  blue at points of infinite~$R$; black indicates convex to the origin
-  and blue indicates concave regions}
-\label{curvature_switch}
-\end{figure}
-
-\begin{figure}[p]
-\centering
 \includegraphics{angle_at_r_equals_2.pdf} % created by "angle_at_r_equals_2.R"

### 2019-08-28 string_physics_teacher.tex
more terse

@@
 p = EulerEquations[Sqrt[y'[x]\^{}2/(1-1/y[x]) + y[x]\^{}2],y[x],x] // FullSimplify\\
-ss = p[[1]]//Numerator\\
-DSolve[ss==0,y[x],x]
+DSolve[p[[1]]//Numerator == 0, y[x],x]
 }}, an expression for the second derivative of~$r$:

### 2019-08-28 string_physics_teacher.tex
simplify now we are using stupid units
another insight

@@
   r''\left(\phi\right) =
-  (r-2M) + \frac{(4r-6M)\left(r'\right)^2}{2r\left(r-2M\right)},\qquad r>2M.
+  (r-2M) + \frac{(2r-3M)\left(r'\right)^2}{r\left(r-2M\right)},\qquad r>2M.
 \end{equation}
@@
 $u''=3u^2/2-u$, but is equivalent
-to~$r''\left(\phi\right)=r-3M+2M\left(r'\right)^2/r$.  This shows that
+to~$r''\left(\phi\right)=r-3M+2\left(r'\right)^2/r$.  This shows that
 light, being sensitive to the time component of the metric, behaves
-differently from string, which is not.
+differently from string, which is not (at this point it might be worth
+``reminding'' students that a straight line in flat space,
+$r=r_0\sec\phi$, obeys $r''=r+2\left(r'\right)^2/r$).
 

### 2019-08-28 string_physics_teacher.tex
another insight

@@
 \noindent which would give~$R$ as a function of~$\left(r,r'\right)$.
-Considering~$r'(\phi)=0$ shows that if a string is tangential at a
-point with~$r_0>2M$, then it obeys, to first order, $r''=r-2M$, while
-tangential null geodesics follow $r''=r-3M$.  This shows that a taut
-string would {\em look} curved to a local observer, having a (visual)
-radius of curvature of $r_0^2/M\sqrt{1-2M/r_0}$ (the extra term comes
-from the fact that appearance is driven by local $ds$, not
-Schwarzschild $r$).  At the photon sphere $r_0=3M$, such a string
-would appear to have a radius of~$\sqrt{27}M$.  For a solar mass black
-hole this would be about $15\unit{km}$, a readily understandable
-result.
+Considering~$r'(\phi)=0$ shows that a tangential string at a point
+with~$r_0>2M$ obeys, to first order, $r''=r-2M$, while tangential null
+geodesics follow $r''=r-3M$.  This shows that a taut string would {\em
+  look} curved to a local observer, having a (visual) radius of
+curvature of $r_0^2/M\sqrt{1-2M/r_0}$ (the extra term comes from the
+fact that appearance is driven by local $ds$, not Schwarzschild $r$).
+At the photon sphere $r_0=3M$, such a string would appear to have a
+radius of~$\sqrt{27}M$.  For a solar mass black hole this would be
+about $15\unit{km}$, a readily understandable result.
 

### 2019-08-28 string_physics_teacher.tex
more terseification

@@
 under~$\phi\longrightarrow-\phi$, the string must be symmetrical about
-at this point.  The (signed) radius of curvature~$R$ is readily
-evaluated:
-
-% \begin{equation}\label{roc_properunits}
-%   R = \frac{
-%     \left(r^2 + \left(r'\right)^2\right)^\frac{3}{2}
-%   }{
-% %    \left|
-%     r^2 + 2\left(r'\right)^2-rr''
-% %    \right|
-%   }
-%   =
-%   \frac{
-%     \left(r^2 + \left(r'\right)^2\right)^\frac{3}{2}
-%   }{
-%     r-\frac{\left(r'\right)^2}{2\left(r-1\right)}
-%   }\qquad\mbox{along a taut string}
-% \end{equation}
-
-\begin{equation}\label{roc}
-  R = \frac{
-    \left(r^2 + \left(r'\right)^2\right)^\frac{3}{2}
-  }{
-%    \left|
-    r^2 + 2\left(r'\right)^2-rr''
-%    \right|
-  }
-  =
-  \frac{
-    \left(r^2 + \left(r'\right)^2\right)^\frac{3}{2}
-  }{
-    r-\frac{\left(r'\right)^2}{2\left(r-2M\right)}
-  }\qquad\mbox{along a taut string}
-\end{equation}
-
-\noindent which would give~$R$ as a function of~$\left(r,r'\right)$.
-Considering~$r'(\phi)=0$ shows that a tangential string at a point
-with~$r_0>2M$ obeys, to first order, $r''=r-2M$, while tangential null
-geodesics follow $r''=r-3M$.  This shows that a taut string would {\em
-  look} curved to a local observer, having a (visual) radius of
-curvature of $r_0^2/M\sqrt{1-2M/r_0}$ (the extra term comes from the
-fact that appearance is driven by local $ds$, not Schwarzschild $r$).
-At the photon sphere $r_0=3M$, such a string would appear to have a
-radius of~$\sqrt{27}M$.  For a solar mass black hole this would be
-about $15\unit{km}$, a readily understandable result.
-
-Compare the above analysis with a taut {\em heavy} string which
-adopts, at least locally, a catenary configuration at closest
-approach~\cite{nguyen2007}.  A heavy string is manifestly convex to
-the singularity in the sense that $R<0$ if $\phi'=0$, in contrast to
-light strings which have $R>0$.
+at this point.  Considering~$r'(\phi)=0$ shows that a tangential
+string at a point with~$r_0>2M$ obeys, to first order, $r''=r-2M$,
+while tangential null geodesics follow $r''=r-3M$.  Thus a taut string
+would {\em look} curved to a local observer, having a (visual) radius
+of curvature of $r_0^2/M\sqrt{1-2M/r_0}$ (the extra term comes from
+the fact that appearance is driven by local $ds$, not Schwarzschild
+$r$).  At the photon sphere $r_0=3M$, such a string would appear to
+have a radius of~$\sqrt{27}M$.  For a solar mass black hole this would
+be about $15\unit{km}$, a readily understandable result.
 

### 2019-08-28 radius_of_curvature_switch.R
vandalise the script: fallacious reasoning

@@ -1,3 +1,12 @@
+} # This line vandalises the script, which is subject to a serious
+  # conceptual error.  The "radius of curvature" of a string (or
+  # indeed a null geodesic) is meaningless unless one is in flat
+  # space: it depends on the coordinate system used.  At least, that
+  # is my current thinking.
+
+
+
+
 ## This script runs standalone but is designed to be called from
 ## maker.R [cf Makefile], which creates the pdf file
 ## "curvature_switch.pdf"

### 2019-08-28 string_physics_teacher.tex
tersification; the mma code was misleading in any event

@@
 Needs["VariationalMethods`"]\\
-p = EulerEquations[Sqrt[y'[x]\^{}2/(1-1/y[x]) + y[x]\^{}2],y[x],x] // FullSimplify\\
-DSolve[p[[1]]//Numerator == 0, y[x],x]
-}}, an expression for the second derivative of~$r$:
+EulerEquations[Sqrt[y'[x]\^{}2/(1-1/y[x]) + y[x]\^{}2],y[x],x]}}, an expression for the second derivative of~$r$:
 

### 2019-08-28 string_physics_teacher.tex
slight rewording

@@
 A light string's being under tension means that its path will be an
-extremum of proper length: the calculus of variations proves that the
-path is straight in the sense of zero curvature.  In terrestrial
-gravity, standard $25\times 10^{-6}\unit{kg}\cdot\meter^{-1}$ cotton
-sewing thread at its breaking strain of about $40\unit{N}$ would have
-a sag of~$\sim 6\times 10^{-6}\meter$ over a $1\meter$ horizontal
-scale: barely perceptible.  How does such a quotidian observation
-translate into the vicinity of a black hole?  Would the ``warping of
-space'' result in {\em visibly} curved thread?
+extremum of proper length.  In flat space, the calculus of variations
+proves that the path is straight in the sense of zero curvature.  In
+terrestrial gravity, standard $25\times
+10^{-6}\unit{kg}\cdot\meter^{-1}$ cotton sewing thread at its breaking
+strain of about $40\unit{N}$ would have a sag of~$\sim 6\times
+10^{-6}\meter$ over a $1\meter$ horizontal scale: barely perceptible.
+How does such a quotidian observation translate into the vicinity of a
+black hole?  Would the ``warping of space'' result in {\em visibly}
+curved thread?
 
@@
 ``reminding'' students that a straight line in flat space,
-$r=r_0\sec\phi$, obeys $r''=r+2\left(r'\right)^2/r$).
+$r=r_0\sec\left(\phi-\phi_0\right)$, obeys
+$r''=r+2\left(r'\right)^2/r$).
 
@@
 
-Light inextensible strings under tension adopt non-trivial
-configurations including self-intersecting paths.  The path of such
-strings interprets the concept ``straight line'' in general
-relativity, others being null geodesics and world lines of massive
-objects.  Comparing light paths with that of nearby taut strings thus
+The concept of ``straight line'' is problematic in general relativity.
+Although photon paths provide a definition of straightness, their
+world lines are sensitive to the time component of the metric.  Light
+taut strings respond only to the spatial components of the metric, and
+their configuration is arguably a more intuitive and helpful way to
+think about straightness.  Non-trivial configurations are adopted by
+such strings, including self-intersecting paths.  Taut strings are
+arguably as fundamental a definition of straighn
+
+Comparing light paths with that of nearby taut strings thus
 furnishes another insight into path curvature, complementing the
 analysis of Abramowicz~\cite{abramowicz1992}.
-
-In Figure~\ref{closest_approach_non_self_intersecting}, one's
-classical intuition, and d'Alembert's principle, would suggest that
-the black hole should accelerate to the left.  Note that the strings
-are already ``straight'' in sense that they are minimal length; so
-altering the tension would not change their path.  However, it is not
-clear by what mechanism the required force would be transmitted: the
-situation is steady, so no gravitational waves are produced, and the
-string is inextensible so stores no energy.  Further work might
-include analysis of taut strings near a spinning black hole.
-
-%%Vancouver style references.
 

### 2019-08-28 string_physics_teacher.tex
work

@@
 
-General relativity is a topic of long-standing fascination for
-undergraduate physics students~\cite{christensen2012}.  Here I show
-how numerical methods, including computer algebra systems, can be used
-to interpret the concept of ``straight line'' in the vicinity of a
-black hole.  The analysis would be suitable for an undergraduate
-homework problem~\cite{romano2019}.
+General relativity is a topic of long-standing fascination for physics
+undergraduates~\cite{christensen2012}.  Students appreciate everyday
+illustrations of the concepts involved, as recent educational work on
+gravitational wave visualisations attests~\cite{overduin2018}.  Here I
+show how considering light inextensible string can be used to
+interpret the concept of ``straight line'' in the vicinity of a black
+hole.  The analysis presented here would be suitable for an
+undergraduate homework problem~\cite{romano2019}.
 
@@
 
-Further, because the system is symmetrical
+Further, because the system is invariant
 under~$\phi\longrightarrow-\phi$, the string must be symmetrical about
@@
 
-Here I solve equation~\ref{rdashdash} numerically, using the
-\verb+deSolve+ package for ordinary differential
-equations~\cite{soetart2010}.  Non-radial solutions fall into two
-distinct classes: those with $r>2M$ at all points, and those which
-attain $r=2M$, in which case the string is tangential to the event
-horizon.
+Here I solve equation~\ref{rdashdash} numerically, using R programming
+language~\cite{rcore2019} and the \verb+deSolve+ package for ordinary
+differential equations~\cite{soetart2010}.  Non-radial solutions fall
+into two distinct classes: those with $r>2M$ at all points, and those
+which attain $r=2M$, in which case the string is tangential to the
+event horizon.
 
@@
 their configuration is arguably a more intuitive and helpful way to
-think about straightness.  Non-trivial configurations are adopted by
-such strings, including self-intersecting paths.  Taut strings are
-arguably as fundamental a definition of straighn
-
-Comparing light paths with that of nearby taut strings thus
-furnishes another insight into path curvature, complementing the
-analysis of Abramowicz~\cite{abramowicz1992}.
+think about straightness.  Using numerical methods to compute string
+configuration is suitable for an undergraduate homework problem.
 

### 2019-08-28 string_physics_teacher.tex
make the width = the line width, not a mm value

@@
 \centering
-\includegraphics[width=170mm]{flamm_string.pdf}
+\includegraphics[width=\linewidth]{flamm_string.pdf}
 \caption{Perspective view of Flamm's paraboloid with superimposed minimal-length
@@
 \centering
-\includegraphics[width=170mm]{closest_approach_nonselfintersecting.pdf}
+\includegraphics[width=\linewidth]{closest_approach_nonselfintersecting.pdf}
 \caption{Non-self intersecting strings arranged by distance~$d$ of
@@
 \centering
-\includegraphics[width=170mm]{closest_approach_selfintersecting.pdf}
+\includegraphics[width=\linewidth]{closest_approach_selfintersecting.pdf}
 \caption{Self-intersecting strings shown for $\phi\geqslant 0$.  Closest
@@
 \centering
-\includegraphics[width=170mm]{light_closest_approach.pdf}
+\includegraphics[width=\linewidth]{light_closest_approach.pdf}
 
@@
 \centering
-\includegraphics{angle_at_r_equals_2.pdf} % created by "angle_at_r_equals_2.R"
+\includegraphics[width=\linewidth]{angle_at_r_equals_2.pdf} % created by "angle_at_r_equals_2.R"
 \caption{Strings passing through $(4M,0)$}
@@
 \centering
-\includegraphics{light_start_at_r_equals_2.pdf} % created by "angle_start_at_r_equals_2.R"
+\includegraphics[width=\linewidth]{light_start_at_r_equals_2.pdf} % created by "angle_start_at_r_equals_2.R"
 \caption{Null geodesics passing through $(4M,0)$, shown for
@@
 \centering
-\includegraphics[width=170mm]{one_free_end_fixed_EH_intersection.pdf}
+\includegraphics[width=\linewidth]{one_free_end_fixed_EH_intersection.pdf}
 \caption{Strings tangential to event horizon at $(1,0)$}
@@
 \centering
-\includegraphics[width=170mm]{one_free_end_r_equals_2.pdf}
+\includegraphics[width=\linewidth]{one_free_end_r_equals_2.pdf}
 \caption{Strings tangential to the event horizon passing through $(4M,0)$}

### 2019-08-28 string_physics_teacher.tex
work on the AIP style manual

@@
 vicinity of a black hole, together with computer algebra systems, can
-generate insight into the Schwarzschild geometry that is accessible to
-undergraduates.  Light taut strings minimize their proper length,
-given by integrating the spatial component of the Schwarzschild metric
-along the string.  The path itself is given by straightforward
-numerical solution to the Euler-Lagrange equations.  If the string is
-entirely outside the event horizon, its closest approach to the
-singularity is tangential.  At this point the string is {\em visibly}
-curved, surely a memorable and informative insight.  The geometry of
-the Schwarzschild metric induces some peculiar nonlocal behaviour: if
-the distance of closest approach is less than about $1.0761\cdot 2M$,
-the string self-intersects, even though it is everywhere under
-tension.  Light taut strings furnish a third interpretation of the
-concept ``straight line'', the other two being null geodesics and
-free-fall world lines.
+generate insight into the Schwarzschild geometry that is suitable for
+an undergraduate homework problem.  Light taut strings minimize their
+proper length, given by integrating the spatial component of the
+Schwarzschild metric along the string.  The path itself is given by
+straightforward numerical solution to the Euler-Lagrange equations.
+If the string is entirely outside the event horizon, its closest
+approach to the singularity is tangential.  At this point the string
+is {\em visibly} curved, surely a memorable and informative insight.
+The geometry of the Schwarzschild metric induces some peculiar
+nonlocal behaviour: if the distance of closest approach is less than
+about $1.0761\cdot 2M$, the string self-intersects, even though it is
+everywhere under tension.  Light taut strings furnish a third
+interpretation of the concept ``straight line'', the other two being
+null geodesics and free-fall world lines.
 
@@
 
-\noindent where dashes denote differentiation with respect to~$\phi$.  The
-Euler-Lagrange equations for this system,
+\noindent where dashes denote differentiation with respect to~$\phi$.
+The Euler-Lagrange equations for this system,
 $\frac{d}{d\phi}\frac{\partial F}{\partial r'}-\frac{\partial
-  F}{\partial r}=0$ give, after simplification\footnote{The Mathematica
-  idiom would be:
-
-\noindent
-{\tt
-Needs["VariationalMethods`"]\\
-EulerEquations[Sqrt[y'[x]\^{}2/(1-1/y[x]) + y[x]\^{}2],y[x],x]}}, an expression for the second derivative of~$r$:
+  F}{\partial r}=0$ give an expression for the second derivative
+of~$r$ (Mathematica idiom would be {\tt EulerEquations[Sqrt[y[x]\^{}2
+      + y'[x]\^{}2/(1-1/y[x])],y[x],x]}):
 
@@
 
-\noindent 
-Solutions to equation~\ref{rdashdash} correspond to a taut string
-confined to Flamm's paraboloid~\cite{flamm1916}; see
+\noindent Solutions to equation~\ref{rdashdash} correspond to a taut
+string confined to Flamm's paraboloid~\cite{flamm1916}; see
 Figure~\ref{flamm}.  It is interesting to compare
@@
 There does not appear to be a simple analytical solution to
-equation~\ref{rdashdash} but nevertheless we may make several
-observations.  Equation~\ref{rdashdash} is a second-order nonlinear
-ordinary differential equation and thus has two constants of
-integration.  There are two qualitatively different types of
-solutions: those with a tangential point ($r'=0,r>2M$), and those
-without.  Consider first solutions with such a point: there exists a
-pair~$\left(r,\phi\right), r>2M$ at which~$r'(\phi)=0$.  This point
-must be unique, as~$r''(\phi)>0$, and be the closest approach of the
-string to the event horizon.
+equation~\ref{rdashdash} but it is a straightforward second order ODE,
+amenable to numerical solution.  There are two qualitatively different
+types of solutions: those with a tangential point ($r'=0,r>2M$), and
+those without.  Consider first solutions with such a point: there
+exists a pair~$\left(r,\phi\right), r>2M$ at which~$r'(\phi)=0$.  This
+point must be unique, as~$r''(\phi)>0$, and be the closest approach of
+the string to the event horizon.
 
@@
 
-Here I solve equation~\ref{rdashdash} numerically, using R programming
-language~\cite{rcore2019} and the \verb+deSolve+ package for ordinary
-differential equations~\cite{soetart2010}.  Non-radial solutions fall
-into two distinct classes: those with $r>2M$ at all points, and those
-which attain $r=2M$, in which case the string is tangential to the
-event horizon.
-
-We first consider strings for which $r>2M$ everywhere (that is,
-completely exterior strings).
+Here I solve equation~\ref{rdashdash} numerically, using the R
+programming language~\cite{rcore2019} and the \verb+deSolve+ package
+for ordinary differential equations~\cite{soetart2010}.  First we
+consider strings for which $r>2M$ everywhere (that is, completely
+exterior strings).
 Figure~\ref{closest_approach_non_self_intersecting} shows a sequence
@@
 through~$(4M,0)$.  Figure~\ref{light_r_equals_2} shows light paths for
-comparison.
-
-The cases where the string attains $r=2M$ are tangent to the event
-horizon (or, exceptionally, purely radial).
+comparison.  The cases where the string attains $r=2M$ are tangent to
+the event horizon (or, exceptionally, purely radial).
 Figure~\ref{fixed_EH_intersection} shows strings tangential to the
@@
 
-\section*{Supplementary Material}
+\section*{References}
+\bibliography{stringrefs.bib}
 
-Calculations are performed using the R programming language~\cite{rcore2018};
-source code for the figures is available at
-\\
-\\
-{\tt https://github.com/RobinHankin/string.git}
+\clearpage
 
@@
 \includegraphics[width=\linewidth]{closest_approach_nonselfintersecting.pdf}
-\caption{Non-self intersecting strings arranged by distance~$d$ of
-  closest approach to the event horizon, occurring tangentially
-  at~$\phi=0$}
+\caption{Light inextensible strings under tension close to a black
+  hole.  Here we see non-self intersecting strings arranged by
+  increasing distance of closest approach to the event horizon,
+  occurring tangentially at~$\phi=0$}
 \label{closest_approach_non_self_intersecting}
@@
 \includegraphics[width=\linewidth]{closest_approach_selfintersecting.pdf}
-\caption{Self-intersecting strings shown for $\phi\geqslant 0$.  Closest
-  approach to the event horizon occurs at $\phi=0$ (3 o'clock) and
-  $r/2M\in\left(1,1.0761\right)$; strings are symmetrical about $\phi=0$
-  but for clarity only $\phi\geqslant 0$ is shown}
+\caption{Light inextensible strings under tension close to a black
+  hole.  Here self-intersecting strings are shown for $\phi\geqslant
+  0$.  Closest approach to the event horizon occurs at $\phi=0$ (3
+  o'clock) and $r/2M\in\left(1,1.0761\right)$; strings are symmetrical
+  about $\phi=0$ but for clarity only $\phi\geqslant 0$ is shown}
 \label{closest_approach_self_intersecting}
@@
 \includegraphics[width=\linewidth]{light_closest_approach.pdf}
-
 \caption{Null geodesics in the Schwarzschild geometry, tangential
   at~$\phi=0$.  Note the differences between these curves and the taut
-  strings shown elsewhere: unlike taut strings, null geodesics have
-  positive curvature everywhere, may cross the event horizon inwards,
-  and are never tangential to the event horizon}
+  strings shown elsewhere: unlike taut strings, null geodesices may
+  cross the event horizon inwards, and are never tangential to the
+  event horizon}
 \label{light_closest_approach}
@@
 \includegraphics[width=\linewidth]{angle_at_r_equals_2.pdf} % created by "angle_at_r_equals_2.R"
-\caption{Strings passing through $(4M,0)$}
+\caption{Light inextensible strings under tension close to a black
+  hole.  Here strings passing through $(4M,0)$ at different angles are
+  shown}
 \label{strings_r_equals_2}
@@
 \includegraphics[width=\linewidth]{light_start_at_r_equals_2.pdf} % created by "angle_start_at_r_equals_2.R"
-\caption{Null geodesics passing through $(4M,0)$, shown for
-  $0\leqslant\phi\leqslant 2\pi$}
+\caption{Null geodesics passing through $(4M,0)$ at different angles,
+  shown for $0\leqslant\phi\leqslant 2\pi$}
 \label{light_r_equals_2}
@@
 
- \section*{Conflict of interest}
- The author declares that he has no conflict of interest.
-
-\subsection*{Supplementary material}
-
-R code for all statistical analysis given here is available at
-
-{\tt https://github.com/RobinHankin/hyper2}.
-
-\section*{References}
-
-\bibliography{stringrefs.bib}
-
 \end{document}

### 2019-08-28 string_physics_teacher.tex
remove the tangential-to-the-event-horizon stuff

@@
 equation~\ref{rdashdash} but it is a straightforward second order ODE,
-amenable to numerical solution.  There are two qualitatively different
-types of solutions: those with a tangential point ($r'=0,r>2M$), and
-those without.  Consider first solutions with such a point: there
-exists a pair~$\left(r,\phi\right), r>2M$ at which~$r'(\phi)=0$.  This
-point must be unique, as~$r''(\phi)>0$, and be the closest approach of
-the string to the event horizon.
-
-Further, because the system is invariant
-under~$\phi\longrightarrow-\phi$, the string must be symmetrical about
-at this point.  Considering~$r'(\phi)=0$ shows that a tangential
-string at a point with~$r_0>2M$ obeys, to first order, $r''=r-2M$,
-while tangential null geodesics follow $r''=r-3M$.  Thus a taut string
-would {\em look} curved to a local observer, having a (visual) radius
-of curvature of $r_0^2/M\sqrt{1-2M/r_0}$ (the extra term comes from
-the fact that appearance is driven by local $ds$, not Schwarzschild
-$r$).  At the photon sphere $r_0=3M$, such a string would appear to
-have a radius of~$\sqrt{27}M$.  For a solar mass black hole this would
-be about $15\unit{km}$, a readily understandable result.
-
-If, conversely, $r'(\phi)\neq 0$ whenever~$r>2M$, then the string has
-only one free end; the other is tangential to the event horizon (in
-the sense that ${\displaystyle \lim_{r\longrightarrow 2M^+}r'=0}$),
-except in the degenerate case of constant~$\phi$ when the string is
-radial for~$r>2M$.
+amenable to numerical solution.  Here we consider only strings
+exterior to the event horizon, $r>2M$, and for these we have a
+well-defined point of closest approach to the event horizon.  Further,
+because the system is invariant under~$\phi\longrightarrow-\phi$, the
+string must be symmetrical about at this point.
+Considering~$r'(\phi)=0$ shows that a tangential string at a point
+with~$r_0>2M$ obeys, to first order, $r''=r-2M$, while tangential null
+geodesics follow $r''=r-3M$.  Thus a taut string would {\em look}
+curved to a local observer, having a (visual) radius of curvature of
+$r_0^2/M\sqrt{1-2M/r_0}$ (the extra term comes from the fact that
+appearance is driven by local $ds$, not Schwarzschild $r$).  At the
+photon sphere $r_0=3M$, such a string would appear to have a radius
+of~$\sqrt{27}M$.  For a solar mass black hole this would be about
+$15\unit{km}$, a readily understandable result.
 
@@
 through~$(4M,0)$.  Figure~\ref{light_r_equals_2} shows light paths for
-comparison.  The cases where the string attains $r=2M$ are tangent to
-the event horizon (or, exceptionally, purely radial).
-Figure~\ref{fixed_EH_intersection} shows strings tangential to the
-event horizon at~$\phi=0$ and Figure~\ref{one_free_end_r_equals_2}
-shows tangential strings passing through $(4M,0)$.
+comparison.
 
@@
 
-\begin{figure}[p] % created by "one_free_end_fixed_EH_intersection.R"
-\centering
-\includegraphics[width=\linewidth]{one_free_end_fixed_EH_intersection.pdf}
-\caption{Strings tangential to event horizon at $(1,0)$}
-\label{fixed_EH_intersection}
-\end{figure}
-
-\begin{figure}[p] % created by "one_free_end_r_equals_2.R"
-\centering
-\includegraphics[width=\linewidth]{one_free_end_r_equals_2.pdf}
-\caption{Strings tangential to the event horizon passing through $(4M,0)$}
-\label{one_free_end_r_equals_2}
-\end{figure}
-
 \end{document}

### 2019-08-29 string_physics_teacher.tex
clarification

@@
 There does not appear to be a simple analytical solution to
-equation~\ref{rdashdash} but it is a straightforward second order ODE,
-amenable to numerical solution.  Here we consider only strings
-exterior to the event horizon, $r>2M$, and for these we have a
-well-defined point of closest approach to the event horizon.  Further,
-because the system is invariant under~$\phi\longrightarrow-\phi$, the
-string must be symmetrical about at this point.
-Considering~$r'(\phi)=0$ shows that a tangential string at a point
-with~$r_0>2M$ obeys, to first order, $r''=r-2M$, while tangential null
-geodesics follow $r''=r-3M$.  Thus a taut string would {\em look}
-curved to a local observer, having a (visual) radius of curvature of
+equation~\ref{rdashdash}; but it is a straightforward second order
+ODE, amenable to numerical solution and no harder than the equivalent
+for null geodesics, a standard homework problem.  Here we consider
+only strings exterior to the event horizon, $r>2M$, which have a
+well-defined point of closest approach to the singularity.  Because
+the system is invariant under~$\phi\longrightarrow-\phi$, the string
+must be symmetrical about this point.  If $\phi'$ is small, then a
+string follows $r''=r-2M$, while tangential null geodesics follow
+$r''=r-3M$.  Thus a taut string would {\em look} curved to a local
+observer, having a (visual) radius of curvature of
 $r_0^2/M\sqrt{1-2M/r_0}$ (the extra term comes from the fact that
@@
 
-\section*{References}
 \bibliography{stringrefs.bib}

### 2019-08-29 string_physics_teacher.tex
typos

@@
 elementary physics.  Here I show how considering such string in the
-vicinity of a black hole, together with computer algebra systems, can
-generate insight into the Schwarzschild geometry that is suitable for
+vicinity of a black hole, with the help of computer algebra systems,
+can generate insight into the Schwarzschild geometry in the context of
 an undergraduate homework problem.  Light taut strings minimize their
@@
 is {\em visibly} curved, surely a memorable and informative insight.
-The geometry of the Schwarzschild metric induces some peculiar
-nonlocal behaviour: if the distance of closest approach is less than
+The geometry of the Schwarzschild metric induces some interesting
+nonlocal phenomena: if the distance of closest approach is less than
 about $1.0761\cdot 2M$, the string self-intersects, even though it is

### 2020-01-03 string_european_journal_of_physics.tex
start work on changing style to EJP

--- 2019-08-29 string_physics_teacher.tex
@@
+\pdfminorversion=4
+
 \documentclass[review]{elsarticle}
@@
 
-\journal{The Physics Teacher}
+\journal{European Journal of Physics}
 
-\bibliographystyle{jss}
+\bibliographystyle{unsrt}
 
@@
 \title{Light inextensible strings under tension in the Schwarzschild geometry}
+\author{Robin K S Hankin}
+\address{Auckland University of Technology}
+\ead{hankin.robin@gmail.com}
 
@@
 can generate insight into the Schwarzschild geometry in the context of
-an undergraduate homework problem.  Light taut strings minimize their
+a variational calculus problem.  Light taut strings minimize their
 proper length, given by integrating the spatial component of the
@@
 
-General relativity is a topic of long-standing fascination for physics
-undergraduates~\cite{christensen2012}.  Students appreciate everyday
-illustrations of the concepts involved, as recent educational work on
-gravitational wave visualisations attests~\cite{overduin2018}.  Here I
-show how considering light inextensible string can be used to
-interpret the concept of ``straight line'' in the vicinity of a black
-hole.  The analysis presented here would be suitable for an
-undergraduate homework problem~\cite{romano2019}.
+Light inextensible string under tension is a stalwart feature of
+elementary physics.  Here I consider how light inextensible string
+behaves in the vicinity of a black hole; and use this to interpret the
+concept of ``straight line'' in curved spacetime.
 
@@
 terrestrial gravity, standard $25\times
-10^{-6}\unit{kg}\cdot\meter^{-1}$ cotton sewing thread at its breaking
+10^{-6}\unit{kg}\cdot\unit{m}^{-1}$ cotton sewing thread at its breaking
 strain of about $40\unit{N}$ would have a sag of~$\sim 6\times
-10^{-6}\meter$ over a $1\meter$ horizontal scale: barely perceptible.
+10^{-6}\unit{m}$ over a $1\unit{m}$ horizontal scale: barely perceptible.
 How does such a quotidian observation translate into the vicinity of a
@@
    \int_{r=2M}^{r=2M(1+w)}ds
-   =2M\left(2\sqrt{w} + \frac{1}{3}w^{3/2} + O(w^{5/2})\right).
+   =2M\left(2\sqrt{w} + \frac{1}{3}w^{3/2} + O(w^{5/2})\right)
  \end{equation}
@@
 light, being sensitive to the time component of the metric, behaves
-differently from string, which is not (at this point it might be worth
-``reminding'' students that a straight line in flat space,
-$r=r_0\sec\left(\phi-\phi_0\right)$, obeys
+differently from string, which is not (recall that a straight line in
+flat space, $r=r_0\sec\left(\phi-\phi_0\right)$, obeys
 $r''=r+2\left(r'\right)^2/r$).
@@
 ODE, amenable to numerical solution and no harder than the equivalent
-for null geodesics, a standard homework problem.  Here we consider
+for null geodesics.  Here we consider
 only strings exterior to the event horizon, $r>2M$, which have a

### 2020-01-03 string_european_journal_of_physics.tex
use Roman d for differentials as per the style guide

@@
 \end{frontmatter}
+\newcommand{\rmd}{\mathrm{d}}
 
@@
 \begin{equation}\label{schwarzschild}
-ds^2= -dt^2\left(1-2M/r\right) +\frac{dr^2}{1-2M/r} + r^2\left(d\theta^2 + \sin^2\theta d\phi^2\right).
+\rmd s^2= -\rmd t^2\left(1-2M/r\right) +\frac{\rmd r^2}{1-2M/r} + r^2\left(\rmd\theta^2 + \sin^2\theta\rmd\phi^2\right).
 \end{equation}
 
-\noindent Here I consider the spatial component~$ds^2=
-dr^2/\left(1-2M/r\right) + r^2d\phi^2$, confined to equatorial
+\noindent Here I consider the spatial component~$\rmd s^2=
+\rmd r^2/\left(1-2M/r\right) + r^2\rmd\phi^2$, confined to equatorial
 points~($\theta=\pi/2$); we understand~$r\geqslant 2M$ throughout.  A
@@
 \begin{equation}\label{radial_string_length}
-  \int_{r=r_1}^{r_2}\frac{dr}{\sqrt{1-2M/r}}=
+  \int_{r=r_1}^{r_2}\frac{\rmd r}{\sqrt{1-2M/r}}=
   \left.
@@
    \ell =
-   \int_{r=2M}^{r=2M(1+w)}ds
+   \int_{r=2M}^{r=2M(1+w)}\rmd s
    =2M\left(2\sqrt{w} + \frac{1}{3}w^{3/2} + O(w^{5/2})\right)
@@
 %  \int_{p_1}^{p_2}\sqrt{\frac{\left(r'\right)^2}{1-2M/r} + r^2}d\phi=
-  \int_{p_1}^{p_2}\left(\frac{\left(r'\right)^2}{1-2M/r} + r^2\right)^\frac{1}{2}d\phi=
-  \int_{p_1}^{p_2}F\left(r,r'\right)d\phi
+  \int_{p_1}^{p_2}\left(\frac{\left(r'\right)^2}{1-2M/r} + r^2\right)^\frac{1}{2}\rmd\phi=
+  \int_{p_1}^{p_2}F\left(r,r'\right)\rmd\phi
 \end{equation}
@@
 The Euler-Lagrange equations for this system,
-$\frac{d}{d\phi}\frac{\partial F}{\partial r'}-\frac{\partial
+$\frac{\rmd}{\rmd\phi}\frac{\partial F}{\partial r'}-\frac{\partial
   F}{\partial r}=0$ give an expression for the second derivative

### 2020-01-03 string_european_journal_of_physics.tex
use \Or for order O() as per the style guide
do not use amsmath as per the style guide
left align equations as per the style guide
log -> ln as per the style guide
order of brackets {[()]} as per the style guide
noindent after an equation
clarification of notation

@@
 
-\documentclass[review]{elsarticle}
+\documentclass[review,fleqn]{elsarticle}
 
 \usepackage{lineno,hyperref}
-\usepackage{amsmath}
 \usepackage{amssymb}
@@
 \newcommand{\rmd}{\mathrm{d}}
+\newcommand{\Or}{\mathord{\mathrm{O}}}
 
@@
 %   \left.
-%   \sqrt{r(r-1)} +\frac{1}{2}\log\left(
+%   \sqrt{r(r-1)} +\frac{1}{2}\ln\left(
 %   \frac{\sqrt{r}+\sqrt{r-1}}{\sqrt{r}-\sqrt{r-1}}\right)
@@
   \left.
-  \sqrt{r(r-2M)} +\frac{1}{2}\log\left(
+  \sqrt{r(r-2M)} +\frac{1}{2}\ln\left(
   \frac{\sqrt{r}+\sqrt{r-2M}}{\sqrt{r}-\sqrt{r-2M}}\right)
@@
    \int_{r=2M}^{r=2M(1+w)}\rmd s
-   =2M\left(2\sqrt{w} + \frac{1}{3}w^{3/2} + O(w^{5/2})\right)
+   =2M\left[2\sqrt{w} + \frac{1}{3}w^{3/2} + \Or(w^{5/2})\right]
  \end{equation}
 
-(compare the flat space string length of $2Mw$).  This would indicate
- that, close to the Schwarzschild radius $r=2M$, large (but finite)
- amounts of string are required to connect two points with slightly
- differing radial coordinates.  Passing to the case $w\gg 1$ we have
- an asymptotic expansion that begins
+\noindent (compare the flat space string length of $2Mw$).  This would
+indicate that, close to the Schwarzschild radius $r=2M$, large (but
+finite) amounts of string are required to connect two points with
+slightly differing radial coordinates.  Passing to the case $w\gg 1$
+we have an asymptotic expansion that begins
 
 % \begin{equation}\label{asymptotic_ell_properunits}
-%   \ell = w +  \frac{1}{2}\log w + \frac{1}{2} + \log(2)  +  \frac{1}{8}w^{-1} + O(w^{-2}),
+%   \ell = w +  \frac{1}{2}\ln w + \frac{1}{2} + \ln(2)  +  \frac{1}{8}w^{-1} + O(w^{-2}),
 % \end{equation}
@@
 \begin{equation}\label{asymptotic_ell}
-  \ell = 2M\left(w +  \frac{1}{2}\log w + \frac{1}{2} + \log(2)  +  \frac{1}{8}w^{-1} + O(w^{-2})\right),
+  \ell = 2M\left[w +  \frac{1}{2}\ln w + \frac{1}{2} + \ln(2)  +  \frac{1}{8}w^{-1} + \Or(w^{-2})\right],
 \end{equation}
@@
 \caption{Null geodesics passing through $(4M,0)$ at different angles,
-  shown for $0\leqslant\phi\leqslant 2\pi$}
+  shown for $0\leqslant\phi\leqslant 2\pi$.  Notation as for figure~\ref{light_closest_approach}}
 \label{light_r_equals_2}

### 2020-04-10 string_physics_teacher_revision.tex
rephrase for TPT, issue #20

@@
 Of course, no real string is perfectly weightless or inextensible.
-With tension~$T$, gravitational acceleration $g$ and (proper) mass per
-(proper) unit length~$\mu$, then a string will behave as though light
-on lengthscales $L\lesssim T/\left(\mu g\right)$.  Using
-$c^4/\left(4GM\right)$ for the strength of gravity at the event
-horizon of a black hole with mass $M$ gives $L\lesssim TM/\left(\mu
-c^4\right)$.  For the string to behave as though light on lengthscales
-comparable with the Schwarzschild radius would require $T\gtrsim\mu
-c^2$.  Exceeding this criterion would violate the dominant energy
-condition (and in this context we note that transverse waves would
-propagate at speed~$\sqrt{T/\mu}\sim c$).  Nevertheless, the
-idealization of a string as light and inextensible is an interesting
-and informative limit: Rindler~\cite{rindler}, for example, considers
-lowering a heavy object to the event horizon on the end of such a
-string to illustrate energetic arguments in GR.
+Students could use dimensional analysis to assess the accuracy of such
+approximations.  With tension~$T$, gravitational acceleration $g$ and
+(proper) mass per (proper) unit length~$\mu$, then a string will
+behave as though light on lengthscales $L\lesssim T/\left(\mu
+g\right)$.  Using $c^4/\left(4GM\right)$ for the strength of gravity
+at the event horizon of a black hole with mass $M$ gives $L\lesssim
+TM/\left(\mu c^4\right)$.  For the string to behave as though light on
+lengthscales comparable with the Schwarzschild radius would require
+$T\gtrsim\mu c^2$.  Exceeding this criterion would violate the
+dominant energy condition (and in this context we might remind
+students that transverse waves would propagate at
+speed~$\sqrt{T/\mu}\sim c$).  Nevertheless, the idealization of a
+string as light and inextensible is an interesting and instructive
+limit: Rindler~\cite{rindler}, for example, considers lowering a heavy
+object to the event horizon on the end of such a string to illustrate
+energetic arguments in GR.
 
@@
 
-\noindent Here I consider the spatial component~$ds^2=
-dr^2/\left(1-2M/r\right) + r^2d\phi^2$, confined to equatorial
-points~($\theta=\pi/2$); we understand~$r\geqslant 2M$ throughout.  A
-circular string at constant~$r$ will, trivially, have proper
-length~$2\pi r$.
+Many undergraduate courses will present the Schwarzschild metric {\em
+  ex nihilo}, in which case its pedagogical value would be to
+illustrate non-Newtonian features of GR such as time dilation.  The
+most striking of these are temporal: suppose pulses of light are
+emitted from a distant source [such as a pulsar] with interval $t_P$;
+the coefficient of $dt^2$ implies that a student, stationary at
+distance $r>2M$ from the singularity will perceive these pulses to
+have an interval of~$t_P/\sqrt{1-2M/r} < t_P$ [it might be worth
+  emphasising that coordinate ``$t$'' coincides with Lorentzian time
+  for an observer at infinity].  Further, if our student emits pulses
+at (local) interval $t_L$, then the distant observer would see these
+pulses with an interval of $t_L\sqrt{1-2M/r}$, and this approaches
+infinity as $r\longrightarrow 2M$ and the distant observer sees the
+student ``freezing''.  Nevertheless, the free-fall time from the Event
+Horizon to the singularity is finite.
 
-Considering first the radial component we may calculate the
-proper length $L\left(r_1,r_2\right)$ of a string stretching
-from~$r=r_1\geqslant 2M$ to~$r_2\geqslant r_1$ at constant $\theta,\phi$:
+We also see a failure of classical geometry: the coordinate $r$ has
+the defining feature that the circumference of a circle of radius $r$
+has circumference $2\pi r$; but radial distances are not Euclidean.
+To investigate this phenomenon, we consider the spatial
+component~$ds^2= dr^2/\left(1-2M/r\right) + r^2d\phi^2$, confined to
+equatorial points~($\theta=\pi/2$); we understand~$r\geqslant 2M$
+throughout.
 
-% \begin{equation}\label{radial_string_length_properunits}
-%   \int_{r=r_1}^{r_2}\frac{dr}{\sqrt{1-1/r}}=
-%   \left.
-%   \sqrt{r(r-1)} +\frac{1}{2}\log\left(
-%   \frac{\sqrt{r}+\sqrt{r-1}}{\sqrt{r}-\sqrt{r-1}}\right)
-%   \right|_{r_1}^{r_2}
-%   \end{equation}
+Students might need to be reminded of the concepts of proper time and
+proper length: these are quantities measured {\em locally}.  In
+particular, we note that proper length of an object is the length as
+measured by a local observer at rest with respect to the object.  The
+proper length of a string is the length measured by an inchworm as it
+crawls along the string.  A circular string at constant~$r$ will,
+trivially, have proper length~$2\pi r$.  Considering first the radial
+component we may calculate the proper length $L\left(r_1,r_2\right)$
+of a string stretching from~$r=r_1\geqslant 2M$ to~$r_2\geqslant r_1$
+at constant $\theta,\phi$:
 
@@
 
-%  \begin{equation}\label{ell_properunits}
-%    \ell=\int_{r=1}^{r=1+w}ds
-%    =2\sqrt{w} + \frac{1}{3}w^{3/2} + O(w^{5/2})
-%  \end{equation}
-
  \begin{equation}\label{ell}
@@
 
-% \begin{equation}\label{asymptotic_ell_properunits}
-%   \ell = w +  \frac{1}{2}\log w + \frac{1}{2} + \log(2)  +  \frac{1}{8}w^{-1} + O(w^{-2}),
-% \end{equation}
-% 
 \begin{equation}\label{asymptotic_ell}
@@
 
+\section{Three types of straight line}
+
+There are at least three types of ``straight line'' to consider when
+studying general relativity.  In the context of
+equation~\ref{schwarzschild} we have:
+
+\begin{enumerate}
+\item null geodesics: paths for which $ds^2=0$ at all points.  Light
+  travels on null geodesics and is arguably the ``straightest'' of
+  straight lines.
+\item timelike geodesics: paths for which $ds^2>0$ everywhere, and for
+  which $\int\sqrt{ds^2}$ is a minimum.  These are the paths that
+  objects in freefall follow.
+\item taut string: paths for which $dt=0$, thereby considering the
+  configuration of a string.  If the proper length
+  $\int\sqrt{dr^2+r^2d\theta^2}$ is a minimum, the string is taut.
+\end{enumerate}
+
+The path of a taut string seems to be a natural and intuitive
+interpretation of straight line, with immediate everyday
+interpretation, yet has been neglected in the pedagogical literature.
+Nontrivial taut strings require analysis of nonradial paths, which I
+consider below.
+
 \section{Nonradial string}
@@
 terms of~$r=r\left(\phi\right)$ gives us a path length of
-
-% \begin{equation}
-% %  \int_{p_1}^{p_2}\sqrt{\frac{\left(r'\right)^2}{1-2M/r} + r^2}d\phi=
-%   \int_{p_1}^{p_2}\left(\frac{\left(r'\right)^2}{1-1/r} + r^2\right)^\frac{1}{2}d\phi=
-%   \int_{p_1}^{p_2}F\left(r,r'\right)d\phi
-% \end{equation}
 
@@
 string confined to Flamm's paraboloid~\cite{flamm1916}; see
-Figure~\ref{flamm}.  It is interesting to compare
-equation~\ref{rdashdash} with the corresponding equation for null
-geodesics, which is usually given~\cite{wald} in terms of $u=2M/r$ as
-$u''=3u^2/2-u$, but is equivalent
-to~$r''\left(\phi\right)=r-3M+2\left(r'\right)^2/r$.  This shows that
-light, being sensitive to the time component of the metric, behaves
-differently from string, which is not (at this point it might be worth
-``reminding'' students that a straight line in flat space,
+Figure~\ref{flamm}.  At this point it might be worth ``reminding''
+students that a straight line in flat space,
 $r=r_0\sec\left(\phi-\phi_0\right)$, obeys
-$r''=r+2\left(r'\right)^2/r$).
+$r''=r+2\left(r'\right)^2/r$, and that equation~\ref{rdashdash}
+reduces to this in the limit $r/2M\longrightarrow\infty$.
+
+It is interesting to compare equation~\ref{rdashdash} with the
+corresponding equation for null geodesics, which is usually
+given~\cite{wald} in terms of $u=2M/r$ as $u''=3u^2/2-u$, but is
+equivalent to~$r''\left(\phi\right)=r-3M+2\left(r'\right)^2/r$.  This
+shows that light, being sensitive to the time component of the metric,
+behaves differently from string, which is not.
+
+Here we consider only strings exterior to the event horizon, $r>2M$,
+which have a well-defined point of closest approach to the
+singularity.  Because the system is invariant
+under~$\phi\longrightarrow-\phi$, the string must be symmetrical about
+this point.  If $\phi'$ is small, then a string follows $r''=r-2M$,
+while tangential null geodesics follow $r''=r-3M$.  Thus a taut string
+would {\em look} curved to a local observer, having a (visual) radius
+of curvature of $r_0^2/M\sqrt{1-2M/r_0}$ (the extra term comes from
+the fact that appearance is driven by local $ds$, not Schwarzschild
+$r$).  At the photon sphere $r_0=3M$, such a string would appear to
+have a radius of~$\sqrt{27}M$.  For a solar mass black hole this would
+be about $15.4\unit{km}$, a readily understandable result.  Of course,
+real cotton thread would break here, but the deviation of tangential
+light from a taut string's path would be about $0.13\unit{mm}$,
+certainly detectable by the naked eye.
+\section{Numerical results}
 
 There does not appear to be a simple analytical solution to
-equation~\ref{rdashdash}; but it is a straightforward second order
-ODE, amenable to numerical solution and no harder than the equivalent
-for null geodesics, a standard homework problem.  Here we consider
-only strings exterior to the event horizon, $r>2M$, which have a
-well-defined point of closest approach to the singularity.  Because
-the system is invariant under~$\phi\longrightarrow-\phi$, the string
-must be symmetrical about this point.  If $\phi'$ is small, then a
-string follows $r''=r-2M$, while tangential null geodesics follow
-$r''=r-3M$.  Thus a taut string would {\em look} curved to a local
-observer, having a (visual) radius of curvature of
-$r_0^2/M\sqrt{1-2M/r_0}$ (the extra term comes from the fact that
-appearance is driven by local $ds$, not Schwarzschild $r$).  At the
-photon sphere $r_0=3M$, such a string would appear to have a radius
-of~$\sqrt{27}M$.  For a solar mass black hole this would be about
-$15\unit{km}$, a readily understandable result.
+equation~\ref{rdashdash} and numerical methods are needed.  In an
+educational context, we might use Euler's method to illustrate the
+concepts and find $\left(r,r'\right)$---that is, a point in a
+two-dimensional phase space---as a function of independent variable
+$\phi$.  We specify initial conditions $(r_0,r'_0)$ at $\phi_0$ and
+repeatedly increment $\phi$ by a small amount $\delta\phi$:
 
-\section{Numerical results}
+\begin{eqnarray}
+\phi&\longrightarrow&\phi+\delta\phi\nonumber\\
+r   &\longrightarrow&r+r'\delta\phi\nonumber\\
+r'  &\longrightarrow&r' + r''\delta\phi\qquad\mbox{where $r''$ is from equation~\ref{rdashdash}.}
+\end{eqnarray}
 
-Here I solve equation~\ref{rdashdash} numerically, using the R
-programming language~\cite{rcore2019} and the \verb+deSolve+ package
-for ordinary differential equations~\cite{soetart2010}.  First we
-consider strings for which $r>2M$ everywhere (that is, completely
-exterior strings).
+Of course, Euler's method is numerically unstable and inefficient;
+here use the R programming language~\cite{rcore2019} and the
+\verb+deSolve+ package for ordinary differential
+equations~\cite{soetart2010} using the standard \verb+lsoda+ solver.
+The equations typically imply
+$\displaystyle\lim_{\phi\longrightarrow\pm\phi_C}=\infty$ for some
+critical value $\phi_C$ of $\phi$ [observe $\phi_C=\pi/2$ for flat
+  space] and this makes numerical analysis challenging.
+
+First we consider strings for which $r>2M$ everywhere (that is,
+completely exterior strings).
 Figure~\ref{closest_approach_non_self_intersecting} shows a sequence
@@
 closest approach occurs at~$\phi=0$.
-Figure~\ref{closest_approach_self_intersecting} shows the
-corresponding diagram for self-interesecting strings
-(Figure~\ref{light_closest_approach} shows equivalent null geodesics
-for comparison).  A bisection technique reveals that if the closest
-approach is greater than a critical value of about~$\simeq 1.0761\cdot
-2M$, the string does not self-intersect; if the closest approach is
-less than this value, then the string crosses itself at~$\phi=\pi/2$.
+
+Further analysis reveals something unexpected: for some values of
+initial conditions, the string self-intersects even though it is under
+tension everywhere and cannot support any nonzero bending moment.
+Figure~\ref{closest_approach_self_intersecting} shows some
+self-interesecting strings (Figure~\ref{light_closest_approach} shows
+equivalent null geodesics for comparison).  A bisection technique
+reveals that if the closest approach is greater than a critical value
+of about~$\simeq 1.0761\cdot 2M$, the string does not self-intersect;
+if the closest approach is less than this value, then the string
+crosses itself at~$\phi=\pi/2$.
 
@@
 think about straightness.  Using numerical methods to compute string
-configuration is suitable for an undergraduate homework problem.
+configuration is suitable for an undergraduate homework problem; the
+difficult part is specifying sensible initial conditions.
+
+The code producing the figures is available at github,
+\url{http://github.com/RobinHankin/string}.
 
@@
 
-\begin{figure}[p] % created by "closest_approach2.R"
+\begin{figure}[p] % created by "closest_approach3.R"
 \centering
-\includegraphics[width=\linewidth]{closest_approach_selfintersecting.pdf}
+\includegraphics[width=\linewidth]{closest_approach_selfintersecting_onegreen.pdf}
+
 \caption{Light inextensible strings under tension close to a black
-  hole.  Here self-intersecting strings are shown for $\phi\geqslant
-  0$.  Closest approach to the event horizon occurs at $\phi=0$ (3
-  o'clock) and $r/2M\in\left(1,1.0761\right)$; strings are symmetrical
-  about $\phi=0$ but for clarity only $\phi\geqslant 0$ is shown}
+  hole.  Here, self-intersecting strings are shown for $\phi\geqslant
+  0$ for clarity, but a single string is shown in green for
+  $-3.69\leqslant\phi\leqslant 3.69$, making the self-intersection
+  evident (all strings in the diagram, whether blue or green,
+  self-intersect at $\phi=\pm\pi$, even if the point of intersection
+  is far to the right).  Closest approach to the event horizon occurs
+  at $\phi=0$ (3 o'clock) and $r/2M\in\left(1,1.0761\right)$; all
+  strings are symmetrical about $\phi=0$}
 \label{closest_approach_self_intersecting}

### 2020-04-10 closest_approach3.R

@@ -1,0 +1,115 @@
+## Identical to closest_approach2.R but plots a single string looping
+## round the black hole in green, showing the self-intersection.
+
+## This script runs standalone but is designed to be called from
+## maker.R [cf Makefile], which creates the pdf file
+## "closest_approach2.pdf"
+
+source("usefulfuncs.R") # defines stringpoints() and dseq()
+
+
+domask <- TRUE
+
+
+## plot setup:
+jj <- 6
+par(xpd=TRUE)
+plot(NULL,asp=1,xlim=c(-jj,jj),ylim=c(-jj,jj),type='l',axes=F,xlab='',ylab='')
+userainbow <- FALSE   # colour of strings (FALSE = all strings blue)
+## setup ends
+
+
+dist <- dseq(from=(1.00012), to=1.076109317,len=70,power=3)
+
+cutoffmatrix <- matrix(c(
+    1.0000, pi + 3.30 -3.3 ,
+    1.00011, pi + 4.6 ,
+    1.00012, pi + 4.4,
+    1.00013, pi + 4.3,
+    1.00014, pi + 4.3,
+    1.00015, pi + 4.2,
+    1.00016, pi + 4.1,
+    1.00017, pi + 4.1  ,
+    1.00020, pi + 4.06  ,
+    1.00023, pi + 3.95  ,
+    1.00025, pi + 3.90 ,
+    1.00027, pi + 3.80 ,
+    1.00030, pi + 3.70  ,
+    1.00035, pi + 3.60 ,
+    1.00040, pi + 3.60 ,
+    1.00045, pi + 3.40 ,
+    1.0005, pi + 3.345 ,
+    1.0006, pi + 3.20  ,
+    1.0007, pi + 3.10  ,
+    1.0008, pi + 3.00 ,
+    1.0009, pi + 3.00  ,
+    1.0010, pi + 2.90  ,
+    1.0011, pi + 2.85  ,
+    1.0012, pi + 2.75  ,
+    1.0014, pi + 2.68  ,
+    1.0016, pi + 2.56  ,
+    1.0018, pi + 2.47  ,
+    1.0020, pi + 2.45  ,
+    1.0022, pi + 2.32  ,
+    1.0025, pi + 2.30  ,
+    1.0028, pi + 2.20  ,
+    1.0030, pi + 2.15  ,
+    1.0035, pi + 2.05  ,
+    1.0040, pi + 1.95  ,
[…]

### 2020-04-10 maker.R

@@ -6,6 +6,10 @@ pdf(file="closest_approach_selfintersecting.pdf")
 source("closest_approach2.R")
 dev.off()
 
+pdf(file="closest_approach_selfintersecting_onegreen.pdf")
+source("closest_approach3.R")
+dev.off()
+
 pdf(file="angle_at_r_equals_2.pdf")
 source("angle_at_r_equals_2.R")
 dev.off()

### 2020-09-25 perihelion_advance_of_mercury.Rmd
replace triple backtics with single

@@
 
-Consider the following R idiom, in which ```trajectory()``` is used to
+Consider the following R idiom, in which `trajectory()` is used to
 directly simulate a geodesic in Schwarzschild space:
@@
 
-(NB ```trajectory()``` takes geometrized units).  In the above, we see
+(NB `trajectory()` takes geometrized units).  In the above, we see
 a typical trajectory in Schwarzschild space with a precession of
@@
 We need to isolate the change points, here found isolating iterations
-$i$ such that ```r[i-1]<r[i]``` and r[i]>r[i+1]```:
+$i$ such that `r[i-1]<r[i]` and `r[i]>r[i+1]`:
 

### 2020-09-25 perihelion_advance_of_mercury.Rmd
label each chunk and use cache

@@
 
-```{r}
+```{r,label=geodesicsimulatorGR}
 # Simulate a trajectory starting at r=30, mass of sun=1/2 so Schwarzschild radius=1:
@@
 
-```{r}
+```{r,label=geodesicsimulatorclassical}
 jj <- trajectory(30, 1/2, 3, GR=FALSE) # GR=FALSE means use Newtonian approximation
@@
 First we specify some orbital parameters for the orbit of Mercury, mostly taken from Wikipedia:
-```{r}
+
+```{r,label=mercuryorbitalparameters}
 G <-  6.674e-11     # big G in SI
@@
 Now we calculate some features of the orbit in geometrized units:
-```{r} 
+
+```{r,label=geometrizedunits} 
 d <- G*M/sol^2  # mass of sun in meters
@@
 And some further features using classical geometry:
-```{r}
+
+```{r,label=furtherfeaturesclassical}
 a <- (merc_perihelion + merc_aphelion)/2
@@
 
-```{r}
+```{r,label=maxandminorbitalperiods}
 vmin <- sqrt(mu/ell)*(1-ecc)
@@
 
-```{r}   
+```{r,label=orbitalangularmomentumSI}   
 specific_angular_momentum_SI <- merc_perihelion * vmax
@@
 With all these we can simulate the orbit:
-```{r}
+```{r,label=orbitsimulatorincludingGR,cache=TRUE}
  out <- trajectory(merc_perihelion, m=d, h=h, tau=seq(from=0,to=35e15,len=5e5))
@@
 
-```{r}
+```{r,label=orbitsimulatorGRlongtime,cache=TRUE}
 out <- trajectory(merc_perihelion, m=d, h=h, tau=seq(from=0,to=35e16,len=1e7))
@@
 
-```{r}
+```{r,label=orbitalangleatperihelion}
 jj1 <- c(NA,out$rdot)
@@
 
-```{r}
+```{r,label=calculateprecession}
 prec <- lm(phi~tee)$coefficients[2]
@@
 And to convert to astronomical units we use
-```{r}
+
+```{r,label=converttoastronomialunits}
 conversion <-
@@
 
-```{r}
+```{r,label=resultofprecessionconverted}
 prec*prod(conversion)
@@
 
-```{r}
+```{r,precessionwithclassicaldynamics,cache=TRUE}
 out <- trajectory(merc_perihelion, m=d, h=h, tau=seq(from=0,to=35e16,len=1e7),GR=FALSE)
@@
 
-```{r}
+```{r,label=theoreticalprecessioncalculator}
 precession <- 24*pi^3*a^2/(per^2*sol^2*(1-ecc^2))  # radians/orbit
@@
 
-```{r}
+```{r,label=converttohumancentricunits}
 seconds_per_orbit <- prod(c(

### 2020-09-25 perihelion_advance_of_mercury.Rmd
use cache for time consuming calculations

@@
 
-```{r,label=orbitsimulatorGRlongtime,cache=TRUE}
+```{r,label=orbitsimulatorGRlongtime,cache=TRUE}  # takes ~5 mins to run
 out <- trajectory(merc_perihelion, m=d, h=h, tau=seq(from=0,to=35e16,len=1e7))
@@
 
-```{r,precessionwithclassicaldynamics,cache=TRUE}
+```{r,precessionwithclassicaldynamics,cache=TRUE}  # takes ~ 5 mins to run
 out <- trajectory(merc_perihelion, m=d, h=h, tau=seq(from=0,to=35e16,len=1e7),GR=FALSE)

### 2020-10-07 perihelion_advance_of_mercury.Rmd
more work, mostly clarifying motivation for the methods used

--- 2020-09-25 perihelion_advance_of_mercury.Rmd
@@
 and above we see that the orbit is accurately closed, as expected
-(three orbits are simulated, the lines overlap).
+(three orbits are simulated, the lines overlap).  However, a more
+exacting test would be to plot the same lines but focussing in on the
+aphelion point $(30,0)$:
+
+```{r,label=closeupofaphelion}
+dx <- 0.01
+dy <- 0.5
+plot(jj$x,jj$y,xlim=c(29.99,30.001),ylim=c(-dy,dy),type='b',xlab='x',ylab='y')
+```
+
+In the above we see an accurately closed orbit (which simulates more
+than three full revolutions round the sun).
 
@@
 ```{r,label=orbitsimulatorGRlongtime,cache=TRUE}
-out <- trajectory(merc_perihelion, m=d, h=h, tau=seq(from=0,to=35e16,len=1e7))
+out <- trajectory(merc_perihelion,m=d,h=h,tau=seq(from=0,to=35e16,len=1e7)) # takes ~5 mins
+tail(out)
 ```
 
-We need to isolate the change points, here found isolating iterations
-$i$ such that `r[i-1]<r[i]` and `r[i]>r[i+1]`:
+From the above we see that the trajectory has been simulated up to
+about $965$ radians or about $\frac{965}{2\pi}\simeq 153$ revolutions
+round the sun.  Plotting the entire trajectory is computationally
+expensive, but we can plot the simulated trajectory close to aphelion
+as for the Newtonian case:
+
+```{r, label-plotmercuryorbitGR}
+maxx <- 46001200027
+n <- 650000
+s <- seq_len(n)
+plot(out$x[s],out$y[s],xlim=c(maxx-500,maxx+100),ylim=c(-5e6,5e6),type='o',xlab="x",ylab="y",pch=c(16,rep(1,n-1)))
+```
+
+In the above plot, note the differing axis scales.  The start point is
+shown as a solid circle, subseqent points shown with open circles.
+Lines connect consecutive points in the simulation.  It is difficult
+to see evidence for precession in this plot because the points
+correspond to regular snapshots of Mercury's position in successive
+orbits, and the time evolution is not shown.  To proceed we need to
+find the aphelion (or an approximation to it) somehow, and show that
+the aphelion systematically increases with time.  Given the numerics,
+we find points on the simulated trajectory which are a local maximum
+distance from the sun.  This is straightfoward numerically, we
+identify iterations $i$ (which have radius `r[i]`) satisfying both
+`r[i-1]<r[i]` and `r[i]>r[i+1]`:
 
@@
 out <- trajectory(merc_perihelion, m=d, h=h, tau=seq(from=0,to=35e16,len=1e7),GR=FALSE)
+# trajectory() takes ~ 5 mins to run
 jj1 <- c(NA,out$rdot)

### 2020-10-07 perihelion_advance_of_mercury.Rmd
typo

@@
 round the sun.  Plotting the entire trajectory is computationally
-expensive, but we can plot the simulated trajectory close to aphelion
-as for the Newtonian case:
+expensive, but we can plot part of the simulated trajectory close to
+aphelion as for the Newtonian case:
 

### 2020-10-07 perihelion_advance_of_mercury.Rmd
identify perihelion instead of aphelion

@@
 correspond to regular snapshots of Mercury's position in successive
-orbits, and the time evolution is not shown.  To proceed we need to
-find the aphelion (or an approximation to it) somehow, and show that
-the aphelion systematically increases with time.  Given the numerics,
-we find points on the simulated trajectory which are a local maximum
-distance from the sun.  This is straightfoward numerically, we
-identify iterations $i$ (which have radius `r[i]`) satisfying both
-`r[i-1]<r[i]` and `r[i]>r[i+1]`:
+orbits, and the time evolution is not shown.
+
+To proceed we need to find the perihelion (or an approximation to it)
+somehow, and show that the estimated perihelion systematically
+increases with time.  Given the numerics, we find points on the
+simulated trajectory which are a local minimum distance from the sun.
+This is straightfoward numerically; we identify iterations $i$ (which
+have radius `r[i]`) satisfying both `r[i-1]<r[i]` and `r[i]>r[i+1]`:
 
@@
 jj2 <- c(out$rdot,NA)
-change <- (jj1>0) & (jj2<0)
+change <- (jj1<0) & (jj2>0)
 
@@
 
-plot(tee,phi-pi,main='orbital angle of Mercury at perihelion',xlab='time (seconds)',ylab='angle of perihelion')
+plot(tee,phi,main='orbital angle of Mercury at perihelion',xlab='time (seconds)',ylab='angle of perihelion')
 ```

### 2020-10-07 perihelion_advance_of_mercury.Rmd
add OLS analysis to Newtonian case

@@
 ```{r,label=calculateprecession}
+fit <- lm(phi~tee)
+summary(fit)
 prec <- lm(phi~tee)$coefficients[2]
@@
 
-closely agreeing with Einstein's original prediction of 43''.  A more
-careful analysis would account for the nonzero mass of Mercury.
-
-By way of a consistency check, we can plot a similar diagram but with
-the classical approximation:
+The estimate of closely agreeing with Einstein's original prediction
+of 43'' (note that the standard error in `fit` of about
+`r round(summary(fit)$coefficients[2,2] * prod(conversion),2)`'' is a
+severe overestimate as this ignores the serial correlation in the
+residuals); a more careful analysis would account for the nonzero mass
+of Mercury.  By way of a consistency check, we can plot a similar
+diagram but with the classical approximation:
 
@@
 phi <- (phi) %% (2*pi)
+newtfit <- summary(lm(phi~tee))
 plot(tee,phi-pi,main='orbital angle at of Mercury at perihelion, classical case',xlab='time (seconds)',ylab='angle of perihelion')
@@
 
-In the above plot, note how there is no trend, consistent with the
-Newtonian case of zero perihelion advance.
+In the above plot, note how there is no trend [the OLS estimate of the
+slope is $`r round(prod(conversion)*newtfit$coefficients[2,1],2)` \pm
+`r round(prod(conversion)*newtfit$coefficients[2,2],2)`$, again in
+units of arcseconds per century], consistent with the Newtonian case
+of zero perihelion advance.
 
@@
 
-Compare with the value of about 43.37 obtained from numerical
-integration, showing close agreement.
+Compare with the value of about `r round(prec*prod(conversion),2)`
+obtained above from numerical integration, showing close agreement.

### 2020-10-08 perihelion_advance_of_mercury.Rmd
typo
typos

@@
 the orbit is not closed as Mercury has a tiny amount of orbital
-precession.  Note that the Schwarzschild radius would be microsopic at
+precession.  Note that the Schwarzschild radius would be microscopic at
 this scale (the size of the central disk is dominated by the width of
@@
 In the above plot, note the differing axis scales.  The start point is
-shown as a solid circle, subseqent points shown with open circles.
+shown as a solid circle, subsequent points shown with open circles.
 Lines connect consecutive points in the simulation.  It is difficult
@@
 simulated trajectory which are a local minimum distance from the sun.
-This is straightfoward numerically; we identify iterations $i$ (which
+This is straightforward numerically; we identify iterations $i$ (which
 have radius `r[i]`) satisfying both `r[i-1]<r[i]` and `r[i]>r[i+1]`:

### 2020-10-19 string_physics_teacher_revision.tex
further work for #20

@@
 \begin{frontmatter}
-\title{Light inextensible strings under tension in the Schwarzschild geometry}
+\title{Light inextensible strings (threads) under tension in the
+  Schwarzschild geometry: a new interpretation for ``straight line''
+  in general relativity}
 
@@
 
-Light inextensible string under tension is a stalwart feature of
-elementary physics.  Here I show how considering such string in the
+Light inextensible string (thread) under tension is a stalwart feature
+of elementary physics.  Here I show how considering such string in the
 vicinity of a black hole, with the help of computer algebra systems,
 can generate insight into the Schwarzschild geometry in the context of
-an undergraduate homework problem.  Light taut strings minimize their
-proper length, given by integrating the spatial component of the
+an undergraduate homework problem, specifically by re-interpreting the
+concept ``straight line''.  Light taut strings minimize their proper
+length, given by integrating the spatial component of the
 Schwarzschild metric along the string.  The path itself is given by
@@
 symmetric mass are solved by the Schwarzschild
-metric~\cite{schwarzschild1916}.  In units in which~$G=c=1$, this is:
+metric~\cite{schwarzschild1916}.  In geometrized units, that is, a
+system in which~$G=c=1$, the Schwarzschild metric is:
 
@@
 has circumference $2\pi r$; but radial distances are not Euclidean.
-To investigate this phenomenon, we consider the spatial
-component~$ds^2= dr^2/\left(1-2M/r\right) + r^2d\phi^2$, confined to
-equatorial points~($\theta=\pi/2$); we understand~$r\geqslant 2M$
-throughout.
+This is frequently expressed by saying that space is ``warped'', a
+concept that has proved troublesome for generations of students.
+Considering taut strings surely furnishes students with a useful
+conceptual tool to understand the non-Euclidean geometry near a black
+hole.  To investigate this phenomenon, we consider the spatial
+component~$ds^2= dr^2/\left(1-2M/r\right) + r^2d\phi^2$ of the
+Schwarzschild metric, confined to equatorial points~($\theta=\pi/2$);
+we understand~$r\geqslant 2M$ throughout.
 
@@
 crawls along the string.  A circular string at constant~$r$ will,
-trivially, have proper length~$2\pi r$.  Considering first the radial
-component we may calculate the proper length $L\left(r_1,r_2\right)$
-of a string stretching from~$r=r_1\geqslant 2M$ to~$r_2\geqslant r_1$
-at constant $\theta,\phi$:
+trivially, have proper length~$2\pi r$ [this follows from considering
+  the Schwarzschild metric at constant $t,r,\theta$; the string will
+  then have proper (inchworm) length $\int_{\phi=0}^{2\pi}ds=2\pi r$].
+Considering now the radial component we may calculate the proper
+length $L\left(r_1,r_2\right)$ of a string stretching
+from~$r=r_1\geqslant 2M$ to~$r_2\geqslant r_1$ at constant
+$\theta,\phi$:
 
@@
 
-Observe that integration may be performed with a lower limit
-of~$r_1=2M$ with no difficulty, even though the radial component of the
-Schwarzschild metric approaches infinity there.
+(students might need to be told explicitly that the right hand side of
+equation~\ref{radial_string_length} not being equal to its Euclidean
+value of $r_2-r_1$ is a direct illustration of the non-flatness of
+space near a black hole).  Observe that the integration may be
+performed with a lower limit of~$r_1=2M$ with no difficulty, even
+though the radial component of the Schwarzschild metric approaches
+infinity there.
 
-Elementary methods show that the proper length $\ell$ of radial
-strings has some non-Euclidean features.  Firstly, consider the proper
-length from $r=2M$ to $r=2M(1+w)$ where $w>0, \left|w\right|\ll 1$.
-Then
+One criticism of this analysis from a pedagogical perspective is that
+the algebra is involved and nonintuitive.  However, we may apply
+inchworm thinking via calculus and asymptotic series to show
+simplified expressions for the proper length $\ell$ of radial strings.
+Firstly, consider the proper length from $r=2M$ to $r=2M(1+w)$ where
+$w>0, \left|w\right|\ll 1$.  Then
 
@@
   configuration of a string.  If the proper length
-  $\int\sqrt{dr^2+r^2d\theta^2}$ is a minimum, the string is taut.
+  $\int\sqrt{dr^2+r^2d\theta^2}$ is a minimum, the string is taut in
+  the sense that its inchworm-measured length is a minimum.
 \end{enumerate}
@@
 string confined to Flamm's paraboloid~\cite{flamm1916}; see
-Figure~\ref{flamm}.  At this point it might be worth ``reminding''
+Figure~\ref{flamm}.  At this point it might be worth reminding
 students that a straight line in flat space,
@@
 
-Of course, Euler's method is numerically unstable and inefficient;
-here use the R programming language~\cite{rcore2019} and the
+Of course, Euler's method is numerically unstable and inefficient.
+Here we use the R programming language~\cite{rcore2019} and the
 \verb+deSolve+ package for ordinary differential
@@
 The equations typically imply
-$\displaystyle\lim_{\phi\longrightarrow\pm\phi_C}=\infty$ for some
-critical value $\phi_C$ of $\phi$ [observe $\phi_C=\pi/2$ for flat
-  space] and this makes numerical analysis challenging.
+$\displaystyle\lim_{\phi\longrightarrow\pm\phi_C} r(\phi)=\infty$ for
+some critical value $\phi_C$ of $\phi$ [observe $\phi_C=\pi/2$ for
+  flat space] and this makes numerical analysis challenging.
 
@@
 initial conditions, the string self-intersects even though it is under
-tension everywhere and cannot support any nonzero bending moment.
+tension everywhere and cannot support any nonzero bending moment.  One
+might imagine our inchworm confidently declaring that the string was
+everywhere ``straight'' in the sense of being a minimal proper length
+path, and finding to his astonishment that he has traversed a closed
+loop.
+
 Figure~\ref{closest_approach_self_intersecting} shows some
@@
 their configuration is arguably a more intuitive and helpful way to
-think about straightness.  Using numerical methods to compute string
-configuration is suitable for an undergraduate homework problem; the
-difficult part is specifying sensible initial conditions.
+think about straightness.  The device of using an inchworm to measure
+proper distance along a non-extensible string serves to aid
+understanding of the non-Euclidean nature of space near a black hole.
 
-The code producing the figures is available at github,
+Using numerical methods to compute string configuration is suitable
+for an undergraduate homework problem; the difficult part is
+specifying sensible initial conditions.  The code producing the
+figures is available at github,
 \url{http://github.com/RobinHankin/string}.
@@
 \includegraphics[width=\linewidth]{flamm_string.pdf}
-\caption{Perspective view of Flamm's paraboloid with superimposed minimal-length
-  path corresponding to a taut, light string}
+\caption{Perspective view of Flamm's paraboloid with superimposed
+  minimal-length path corresponding to a taut, light string.  Proper
+  length along this string corresponds to the path integral in
+  $\mathbb{R}^3$.  Students might not be familiar with the fact that
+  the vertical axis has no physical interpretation: there is no
+  meaningful sense in which the spatial component of the Schwarzschild
+  metric is a submanifold of a higher-dimensional Euclidean space}
 \label{flamm}

### 2020-10-21 string_physics_teacher_revision.tex
Physics Teacher house style

@@
 
-\begin{abstract}
-
-Light inextensible string (thread) under tension is a stalwart feature
-of elementary physics.  Here I show how considering such string in the
-vicinity of a black hole, with the help of computer algebra systems,
-can generate insight into the Schwarzschild geometry in the context of
-an undergraduate homework problem, specifically by re-interpreting the
-concept ``straight line''.  Light taut strings minimize their proper
-length, given by integrating the spatial component of the
-Schwarzschild metric along the string.  The path itself is given by
-straightforward numerical solution to the Euler-Lagrange equations.
-If the string is entirely outside the event horizon, its closest
-approach to the singularity is tangential.  At this point the string
-is {\em visibly} curved, surely a memorable and informative insight.
-The geometry of the Schwarzschild metric induces some interesting
-nonlocal phenomena: if the distance of closest approach is less than
-about $1.0761\cdot 2M$, the string self-intersects, even though it is
-everywhere under tension.  Light taut strings furnish a third
-interpretation of the concept ``straight line'', the other two being
-null geodesics and free-fall world lines.
-
 % \PACS{PACS code1 \and PACS code2 \and more}
 % \subclass{MSC code1 \and MSC code2 \and more}
-\end{abstract}
 
@@
 \section{Introduction}
+
+Light inextensible string (thread) under tension is a stalwart feature
+of elementary physics.  Here I show how considering such string in the
+vicinity of a black hole, with the help of computer algebra systems,
+can generate insight into the Schwarzschild geometry in the context of
+an undergraduate homework problem, specifically by re-interpreting the
+concept ``straight line''.  Other definitions of ``straight'' might
+include null geodesic paths and worldlines of objects in free fall;
+light taut string furnishes a third distinct interpretation.  The path
+itself is given by straightforward numerical solution to the
+Euler-Lagrange equations.  At closest approach to the event horizon,
+the string is tangential; at this point the string is {\em visibly}
+curved, surely a memorable and informative insight.  The geometry of
+the Schwarzschild metric induces some interesting nonlocal phenomena:
+if the distance of closest approach is less than about $1.0761\cdot
+2M$, the string self-intersects, even though it is everywhere under
+tension.
 

### 2020-10-21 string_physics_teacher_revision.tex
capitalization

@@
 \begin{enumerate}
-\item null geodesics: paths for which $ds^2=0$ at all points.  Light
+\item Null geodesics: paths for which $ds^2=0$ at all points.  Light
   travels on null geodesics and is arguably the ``straightest'' of
   straight lines.
-\item timelike geodesics: paths for which $ds^2>0$ everywhere, and for
+\item Timelike geodesics: paths for which $ds^2>0$ everywhere, and for
   which $\int\sqrt{ds^2}$ is a minimum.  These are the paths that
   objects in freefall follow.
-\item taut string: paths for which $dt=0$, thereby considering the
+\item Taut string: paths for which $dt=0$, thereby considering the
   configuration of a string.  If the proper length

### 2020-10-21 string_physics_teacher_revision.tex
house style for references
Make \sim a unary operator not a binary operator

@@
 \usepackage{xcolor}
+\usepackage[superscript,biblabel]{cite}
+
 \modulolinenumbers[5]
@@
 
-\bibliographystyle{jss}
+\bibliographystyle{unsrt}
 
@@
 10^{-6}\unit{kg}\cdot\unit{m}^{-1}$ cotton sewing thread at its breaking
-strain of about $40\unit{N}$ would have a sag of~$\sim 6\times
+strain of about $40\unit{N}$ would have a sag of~${\sim} 6\times
 10^{-6}\unit{m}$ over a $1\unit{m}$ horizontal scale: barely perceptible.

### 2020-10-22 string_physics_educator.tex
manuscript for The Physics Educator

--- 2019-08-29 string_physics_teacher.tex
@@
-\documentclass[review]{elsarticle}
+% Authors may submit in single column format.
+% Once your paper got accepted, the publisher will reformat it according to TPE house style (two column).
+%\documentclass[wsdraft]{ws-tpe}
 
-\usepackage{lineno,hyperref}
-\usepackage{amsmath}
-\usepackage{amssymb}
-\usepackage{booktabs}
+\documentclass{ws-tpe}
+\usepackage{multicol}
 \usepackage{units}
-\usepackage{xcolor}
-\modulolinenumbers[5]
+\usepackage[superscript]{cite}
+\begin{document}
 
-\journal{The Physics Teacher}
+\markboth{Robin K. S. Hankin}{Light inextensible strings (thread) in the Schwarzschild metric}
 
-\bibliographystyle{jss}
-
-\begin{document}
-\begin{frontmatter}
 \title{Light inextensible strings under tension in the Schwarzschild geometry}
 
+\author{Robin K. S. Hankin}
+
+\address{School of Engineering, Computing, and Mathematical Sciences, Auckland Iniversity of Technology, 55 Wellesley Street East\\
+Auckland 1142, New Zealand
+\email{hankin.robin@gmail.com}}
+
+\maketitle
+
+\begin{history}
+\received{Day Month Year}
+\revised{Day Month Year}
+\end{history}
+
 \begin{abstract}
-
 Light inextensible string under tension is a stalwart feature of
@@
 null geodesics and free-fall world lines.
-
-% \PACS{PACS code1 \and PACS code2 \and more}
-% \subclass{MSC code1 \and MSC code2 \and more}
 \end{abstract}
 
-\begin{keyword}
-Black holes\sep light inextensible string\sep minimal-length path
-\end{keyword}
+\keywords{Black holes; Schwarzschild metric; light inextensible
+  string; minimal-length path}
 
-\end{frontmatter}
-
-\linenumbers
 \section{Introduction}
@@
 terrestrial gravity, standard $25\times
-10^{-6}\unit{kg}\cdot\meter^{-1}$ cotton sewing thread at its breaking
-strain of about $40\unit{N}$ would have a sag of~$\sim 6\times
-10^{-6}\meter$ over a $1\meter$ horizontal scale: barely perceptible.
-How does such a quotidian observation translate into the vicinity of a
-black hole?  Would the ``warping of space'' result in {\em visibly}
-curved thread?
+10^{-6}\unit{kg}\cdot\unit{m}^{-1}$ cotton sewing thread at its
+breaking strain of about $40\unit{N}$ would have a sag of~${\sim}
+6\times 10^{-6}\unit{m}$ over a $1\unit{m}$ horizontal scale: barely
+perceptible.  How does such a quotidian observation translate into the
+vicinity of a black hole?  Would the ``warping of space'' result in
+{\em visibly} curved thread?
 
@@
 
+\begin{figurehere} % Created by "flamm.R"
+\centering
+\includegraphics[width=\linewidth]{flamm_string.pdf}
+\caption{Perspective view of Flamm's paraboloid with superimposed minimal-length
+  path corresponding to a taut, light string}
+\label{flamm}
+\end{figurehere}
+
 There does not appear to be a simple analytical solution to
@@
 
+\begin{figurehere} % Created by "closest_approach.R"
+  \centerline{
+    \includegraphics[width=\linewidth]{closest_approach_nonselfintersecting.pdf}
+  }
+  \caption{Light inextensible strings under tension close to a black
+    hole.  Here we see non-self intersecting strings arranged by
+    increasing distance of closest approach to the event horizon,
+    occurring tangentially at~$\phi=0$}
+  \label{closest_approach_non_self_intersecting}
+\end{figurehere}
+
+\begin{figurehere} % created by "closest_approach2.R"
+  \centerline{
+    \includegraphics[width=\linewidth]{closest_approach_selfintersecting.pdf}
+  }
+  \caption{Light inextensible strings under tension close to a black
+    hole.  Here self-intersecting strings are shown for $\phi\geqslant
+    0$.  Closest approach to the event horizon occurs at $\phi=0$ (3
+    o'clock) and $r/2M\in\left(1,1.0761\right)$; strings are symmetrical
+    about $\phi=0$ but for clarity only $\phi\geqslant 0$ is shown}
+  \label{closest_approach_self_intersecting}
+\end{figurehere}
+
+\begin{figurehere} % Created by "light_closest_approach.R"
+  \centerline{
+    \includegraphics[width=\linewidth]{light_closest_approach.pdf}
+  }
+  \caption{Null geodesics in the Schwarzschild geometry, tangential
+    at~$\phi=0$.  Note the differences between these curves and the taut
+    strings shown elsewhere: unlike taut strings, null geodesices may
+    cross the event horizon inwards, and are never tangential to the
+    event horizon}
+  \label{light_closest_approach}
+\end{figurehere}
+
 There are other ways to parametrize completely exterior strings;
@@
 comparison.
+
+\begin{figurehere}
+  \centerline{
+    \includegraphics[width=\linewidth]{angle_at_r_equals_2.pdf} % created by "angle_at_r_equals_2.R"
+  }
+  \caption{Light inextensible strings under tension close to a black
+    hole.  Here strings passing through $(4M,0)$ at different angles are
+    shown}
+  \label{strings_r_equals_2}
+\end{figurehere}
+
+\begin{figurehere}
+  \centerline{
+    \includegraphics[width=\linewidth]{light_start_at_r_equals_2.pdf} % created by "angle_start_at_r_equals_2.R"
+  }
+  \caption{Null geodesics passing through $(4M,0)$ at different angles,
+    shown for $0\leqslant\phi\leqslant 2\pi$}
+  \label{light_r_equals_2}
+\end{figurehere}
 
@@
 
+\bibliographystyle{unsrt}
 \bibliography{stringrefs.bib}
-
-\clearpage
-
-\section*{Figures}
-
-\begin{figure}[h!] % Created by "flamm.R"
-\centering
-\includegraphics[width=\linewidth]{flamm_string.pdf}
-\caption{Perspective view of Flamm's paraboloid with superimposed minimal-length
-  path corresponding to a taut, light string}
-\label{flamm}
-\end{figure}
-
-\begin{figure}[p] % Created by "closest_approach.R"
-\centering
-\includegraphics[width=\linewidth]{closest_approach_nonselfintersecting.pdf}
-\caption{Light inextensible strings under tension close to a black
-  hole.  Here we see non-self intersecting strings arranged by
-  increasing distance of closest approach to the event horizon,
-  occurring tangentially at~$\phi=0$}
-\label{closest_approach_non_self_intersecting}
-\end{figure}
-
-\begin{figure}[p] % created by "closest_approach2.R"
-\centering
-\includegraphics[width=\linewidth]{closest_approach_selfintersecting.pdf}
-\caption{Light inextensible strings under tension close to a black
-  hole.  Here self-intersecting strings are shown for $\phi\geqslant
-  0$.  Closest approach to the event horizon occurs at $\phi=0$ (3
-  o'clock) and $r/2M\in\left(1,1.0761\right)$; strings are symmetrical
-  about $\phi=0$ but for clarity only $\phi\geqslant 0$ is shown}
-\label{closest_approach_self_intersecting}
-\end{figure}
-
-\begin{figure}[p] % Created by "light_closest_approach.R"
-\centering
-\includegraphics[width=\linewidth]{light_closest_approach.pdf}
-\caption{Null geodesics in the Schwarzschild geometry, tangential
-  at~$\phi=0$.  Note the differences between these curves and the taut
-  strings shown elsewhere: unlike taut strings, null geodesices may
-  cross the event horizon inwards, and are never tangential to the
-  event horizon}
-\label{light_closest_approach}
-\end{figure}
-
-\begin{figure}[p]
-\centering
-\includegraphics[width=\linewidth]{angle_at_r_equals_2.pdf} % created by "angle_at_r_equals_2.R"
-\caption{Light inextensible strings under tension close to a black
-  hole.  Here strings passing through $(4M,0)$ at different angles are
-  shown}
-\label{strings_r_equals_2}
-\end{figure}
-
-\begin{figure}[p]
-\centering
-\includegraphics[width=\linewidth]{light_start_at_r_equals_2.pdf} % created by "angle_start_at_r_equals_2.R"
-\caption{Null geodesics passing through $(4M,0)$ at different angles,
-  shown for $0\leqslant\phi\leqslant 2\pi$}
-\label{light_r_equals_2}
-\end{figure}
-
 \end{document}

### 2021-01-25 string_physics_educator.tex
mention thread to avoid confusion with quantum string theory
typo

@@
 
-\title{Light inextensible strings under tension in the Schwarzschild geometry}
+\title{Light inextensible strings (thread) under tension in the
+  Schwarzschild geometry}
 
@@
 
-\address{School of Engineering, Computing, and Mathematical Sciences, Auckland Iniversity of Technology, 55 Wellesley Street East\\
+\address{School of Engineering, Computing, and Mathematical Sciences, Auckland University of Technology, 55 Wellesley Street East\\
 Auckland 1142, New Zealand

### 2021-01-25 physics_educator_rebuttal.txt
issue #22

Reviewers' comments:

Reviewer 1: This paper studies geodesics on the equatorial plane of
the spatial sections (constant time slices) of the Schwarzschild
spacetime. The geodesic equation in this case reduces to a second
order non-linear ODE, given by Eq. (6) in the paper, and it is solved
numerically. A few geodesics are plotted in Fig. 1, 2, 3, and 5.

The author's idea is to use the path of light taut strings to
visualize "straight lines in the vicinity of a black hole". The paper
claims that light taut strings will follow paths that minimize proper
length, so the paths are geodesics in spatial sections of the
Schwarzschild spacetime in the second paragraph in the introduction
section and also in the discussion section. However, this major claim
was made without any justification. Justification for this major claim
needs to be given.

The paper has imprecise usage of GR terminology. A few statements are
unclear. Examples include:

1.  "straight line" in the vicinity of black hole... (Introduction/paragraph 1)
I believe it refers to straight line in spatial sections of black hole.

2. large (but finite) amounts of string are required to... (second line after Eq. 3)
why large amounts are needed since w is small?

3. with radial strings being arbitrarily longer than flat space would
allow... (below Eq. 4) what is the meaning of this statement?

4. corresponding equation for null geodesics... (second line below Eq. 6)
I believe the author refers to the orbit equation of null geodesics confined to the equatorial plane.

5. phi' should be r' (page 4 line 5)

6. Local observer (page 4 line 7)
I believe the author means static observer in Schwarzschild coordinates.

7. Radius of curvature (page 4 line 7) Since the equatorial plane is
now curved, what radius of curvature is the author talking about?
Clarification is needed.

8. A radius of \sqrt{27M} (page 4 line 9) What radius is this?

9. Although photon paths provide a definition of
straightness... (discussion section line 2) In GR, worldlines of
photon are null geodesics. The paper does not distinguish geodesics of
spatial sections and geodesics of spacetime.

### 2021-01-25 physics_educator_rebuttal.txt
whitespace and line breaks
start rebuttal

@@
-Reviewers' comments:
+Rebuttal follows.  Reviewers' comments preceded by a chevron (">") and
+my response included inline.
 
-Reviewer 1: This paper studies geodesics on the equatorial plane of
-the spatial sections (constant time slices) of the Schwarzschild
-spacetime. The geodesic equation in this case reduces to a second
-order non-linear ODE, given by Eq. (6) in the paper, and it is solved
-numerically. A few geodesics are plotted in Fig. 1, 2, 3, and 5.
+Overall response: the comments are prefectly reasonable and I have
+addressed them individually.  The manuscript is much stengthened as a
+result and I commend it to you.
 
-The author's idea is to use the path of light taut strings to
-visualize "straight lines in the vicinity of a black hole". The paper
-claims that light taut strings will follow paths that minimize proper
-length, so the paths are geodesics in spatial sections of the
-Schwarzschild spacetime in the second paragraph in the introduction
-section and also in the discussion section. However, this major claim
-was made without any justification. Justification for this major claim
-needs to be given.
+> Reviewers' comments:
+>  
+> Reviewer 1: This paper studies geodesics on the equatorial plane of
+> the spatial sections (constant time slices) of the Schwarzschild
+> spacetime. The geodesic equation in this case reduces to a second
+> order non-linear ODE, given by Eq. (6) in the paper, and it is solved
+> numerically. A few geodesics are plotted in Fig. 1, 2, 3, and 5.
+>  
+> The author's idea is to use the path of light taut strings to
+> visualize "straight lines in the vicinity of a black hole". The paper
+> claims that light taut strings will follow paths that minimize proper
+> length, so the paths are geodesics in spatial sections of the
+> Schwarzschild spacetime in the second paragraph in the introduction
+> section and also in the discussion section. However, this major claim
+> was made without any justification. Justification for this major claim
+> needs to be given.
+>
 
-The paper has imprecise usage of GR terminology. A few statements are
-unclear. Examples include:
+On the one hand, this is a perfectly reasonable request: the claim is
+indeed a major one, and one that had no explicit justification in the
+manuscript.  But on the other, it requires no justification as it is
+true by defnition.  Didactically, one might conceptualise a string as
+being comprised of many tiny links, each of which contributes some
+small length to the string, and each of which is sufficiently small
+for Euclidean geometry to describe its length.  Each link is subject
+to a well-defined local tension with respect to its neighbours.
 
-1.  "straight line" in the vicinity of black hole... (Introduction/paragraph 1)
-I believe it refers to straight line in spatial sections of black hole.
+We may then *count* the links and use this as a proxy for proper
+length, demonstrating the consistency of the concept of proper length;
+and of course small links oblige one to use infinitesimal calculus.
 
-2. large (but finite) amounts of string are required to... (second line after Eq. 3)
-why large amounts are needed since w is small?
+Imagine two strings, S1 and S2, which both smoothly join point p to
+point q (perhaps in a Schwarzschild space), and their paths are close
+in the sense of the Calculus of Variations.  Further, imagine that S1
+has a proper length slightly exceeding that of S2.  If we then impose
+a tension on the strings, we see that the S1 is unstable with respect
+to S2, because S2 has fewer links.  Thus equilibrium paths of light
+inextensible strings are at least local minima of length.
 
-3. with radial strings being arbitrarily longer than flat space would
-allow... (below Eq. 4) what is the meaning of this statement?
+Much of the above reasoning is neccarily tautological; it is perfectly
+reasonable to *define* an "inextensible light string under tension" as
+one that adopts a minimal [or at least extremal] proper-length path
+between its end points, AND I HAVE INCLUDED A SHORT DISCUSSION IN THE
+MANUSCRIPT.
 
-4. corresponding equation for null geodesics... (second line below Eq. 6)
-I believe the author refers to the orbit equation of null geodesics confined to the equatorial plane.
+> The paper has imprecise usage of GR terminology. A few statements are
+> unclear. Examples include:
+>  
+> 1.  "straight line" in the vicinity of black
+> hole... (Introduction/paragraph 1) I believe it refers to straight
+> line in spatial sections of black hole.
+>
 
-5. phi' should be r' (page 4 line 5)
+The original sentence was:
 
-6. Local observer (page 4 line 7)
-I believe the author means static observer in Schwarzschild coordinates.
+:: The sentence was: Here I show how considering light inextensible 
+:: string can be used to interpret the concept of "straight line" in the
+:: vicinity of a black hole.
 
-7. Radius of curvature (page 4 line 7) Since the equatorial plane is
-now curved, what radius of curvature is the author talking about?
-Clarification is needed.
+It might be rephrased: "...interpret the concept of "straight line" in
+spatial sections of a black hole", but I believe this misses the
+point.  My thinking was that the concept of "straight line" in the
+abstract, is to be re-interpreted as the path taken by a light
+inextensible string.
 
-8. A radius of \sqrt{27M} (page 4 line 9) What radius is this?
+I HAVE NOW REPHRASED THE MANUSCRIPT, THE NEW SENTENCE IS
 
-9. Although photon paths provide a definition of
-straightness... (discussion section line 2) In GR, worldlines of
-photon are null geodesics. The paper does not distinguish geodesics of
-spatial sections and geodesics of spacetime.
+:: Here I interpret a straight line as the path adopted by a light
+inexensible string passing between two points.
+
+NOT YET FINIALISED THIS.
+
+> 2. large (but finite) amounts of string are required to... (second
+> line after Eq. 3) why large amounts are needed since w is small?
+
+That is an extremely good observation.  The word "large" is
+inappropriate here, I think.  Asymptotic expansions can be confusing
+and what I meant to say was that the ratio of proper string length to
+Euclidean string length approached infinity. But the word "large" is
+not right here, and I have rephrased accordingly.
+
+    NOT YET FINALISED THIS
+
+>  
+> 3. with radial strings being arbitrarily longer than flat space would
+> allow... (below Eq. 4) what is the meaning of this statement?
+>
+
+Euclidean geometry predicts that l=2Mw; the intent of that sentence
+was to point out that we Euclidean predictions are arbirtarily
+incorrect as we move away from the black hole.  I have rephrased.
+
+    NOT YET FINALISED THIS
+
+> 4. corresponding equation for null geodesics... (second line below
+> Eq. 6) I believe the author refers to the orbit equation of null
+> geodesics confined to the equatorial plane.
+
+I have rephrased accordingly.
+
+    NOT YET FINALISED THIS
+
+>  
+> 5. phi' should be r' (page 4 line 5)
+>
+
+Manuscript now corrected.
+
+    NOT YET FINALISED THIS
+
+> 6. Local observer (page 4 line 7) I believe the author means static
+> observer in Schwarzschild coordinates.
+
+I realise that I have been misusing the word "local" to mean "static"
+and have now corrected the manuscript.
+
+    NOT YET FINALISED THIS
+
+>  
+> 7. Radius of curvature (page 4 line 7) Since the equatorial plane is
+> now curved, what radius of curvature is the author talking about?
+> Clarification is needed.
+
+On re-reading, the original manuscript was a little cryptic.  Now rephrased:
+
+    NOT YET FINALISED THIS
+
+>  
+> 8. A radius of \sqrt{27M} (page 4 line 9) What radius is this?
+
+this is the visual radius, any confusion cleared up by point 7 correction above.
+
+>  
+> 9. Although photon paths provide a definition of
+> straightness... (discussion section line 2) In GR, worldlines of
+> photon are null geodesics. The paper does not distinguish geodesics of
+> spatial sections and geodesics of spacetime.
+>
+
+I don't think the concept of "spatial sections" is particularly
+helpful, even though I do present a diagram of Flamm's paraboloid
+[which is obligatory I would say].  A spatial section S is a
+three-dimensional subset of a spacetime in which we require the
+squared interval between any two events of S to be spacelike.  It is
+easy to give an example of a spatial section in flat Minkowski space
+(t,x,y,z) as (t0,x,y,z) for some fixed t0.  Such constructions are
+somewhat more difficult in Schwarzschild space but we might use
+Schwarzschild coordinates (t,r,theta,phi) and then specify
+S2=(t0,r,theta,phi).  But for r1,r2<2M and fixed [theta,phi] we find
+that the interval between (t0,r1,theta,phi) and (t0,r2,theta,phi) is
+timelike; thus S2 is not a spatial section in the sense defined above.
+But for r>2M it is a perfectly acceptable spatial section, and this is
+illustrated in Figure 1.
+
+When considering static strings it is easier and more convenient to
+use standard Schwarzschild coordinates (t,r,theta,phi) and then simply
+ignore the t coordinate as it plays no part in calculating proper
+length of a string (of course, we need to be careful to specify that
+we are outside the event horizon).  If the string is static, we may
+imagine two static observers tied to the string; in calculating the
+amount of proper distance between the two observers (string length) we
+simply ignore the t coordinate and integrate the spatial component of
+the Schwarzschild metric.  This process is distinct from the concept
+of spatial section in which we condition on t=t0.
+
+The world lines of freely-falling objects follow non-null geodesics
+and photons follow null geodesics.  Light inextensible string follows
+neither of these paths.  We might consider a string in terms of the
+world lines of its links, and then require that the proper [spatial]
+distance between the links be constant.  The

### 2021-01-26 physics_educator_rebuttal.txt
fixes #23

@@
 
-Much of the above reasoning is neccarily tautological; it is perfectly
-reasonable to *define* an "inextensible light string under tension" as
-one that adopts a minimal [or at least extremal] proper-length path
-between its end points, AND I HAVE INCLUDED A SHORT DISCUSSION IN THE
-MANUSCRIPT.
+However, much of the above reasoning is necessarily tautological; it
+is perfectly reasonable to *define* an "inextensible light string
+under tension" as one that adopts a minimal [or at least extremal]
+proper-length path between its end points, and i have included a short
+discussion in the manuscript.
 

### 2021-01-26 string_physics_educator_revision2.tex
fixes #23

--- 2021-01-25 string_physics_educator.tex
@@
 A light string's being under tension means that its path will be an
-extremum of proper length.  In flat space, the calculus of variations
+extremum of proper length: non-extremal paths between two fixed points
+cannot be in stable equilibrium as the tension acts to shorten the
+path; stable paths must therefore be of extremal proper-length.
+In flat space, the calculus of variations
 proves that the path is straight in the sense of zero curvature.  In

### 2021-01-26 physics_educator_rebuttal.txt
fixes #24

@@
 It might be rephrased: "...interpret the concept of "straight line" in
-spatial sections of a black hole", but I believe this misses the
-point.  My thinking was that the concept of "straight line" in the
-abstract, is to be re-interpreted as the path taken by a light
-inextensible string.
+spatial sections of a black hole", but I believe this form of words
+would be potentially misleading.  My thinking was that the concept of
+"straight line" ---in the abstract---is to be re-interpreted in the
+manuscript as the path taken by a light inextensible string.
 
-I HAVE NOW REPHRASED THE MANUSCRIPT, THE NEW SENTENCE IS
+It now reads "Here I interpret a straight line as the path adopted by
+a stationary, light inexensible string under tension passing between
+two points"
 
-:: Here I interpret a straight line as the path adopted by a light
-inexensible string passing between two points.
-
-NOT YET FINIALISED THIS.
+There is a little bit of repetition in the revised manuscript, but I
+think it is worth it for the extra clarity.
 

### 2021-01-26 string_physics_educator_revision2.tex
fixes #24

@@
 gravitational wave visualisations attests~\cite{overduin2018}.  Here I
-show how considering light inextensible string can be used to
-interpret the concept of ``straight line'' in the vicinity of a black
-hole.  The analysis presented here would be suitable for an
-undergraduate homework problem~\cite{romano2019}.
+interpret a straight line as the path adopted by a stationary, light
+inexensible string under tension passing between two points.  The
+analysis presented here would be suitable for an undergraduate
+homework problem~\cite{romano2019}.
 
-A light string's being under tension means that its path will be an
-extremum of proper length: non-extremal paths between two fixed points
-cannot be in stable equilibrium as the tension acts to shorten the
-path; stable paths must therefore be of extremal proper-length.
-In flat space, the calculus of variations
-proves that the path is straight in the sense of zero curvature.  In
-terrestrial gravity, standard $25\times
-10^{-6}\unit{kg}\cdot\unit{m}^{-1}$ cotton sewing thread at its
-breaking strain of about $40\unit{N}$ would have a sag of~${\sim}
+Considering only stationary spacetimes, we define the proper length of
+a string to be the result of integrating the spatial part of the
+metric tensor along string's path between its endpoints.  A light
+string's being under tension means that its path will be an extremum
+of proper length: non-extremal paths between two fixed points cannot
+be in stable equilibrium as the tension acts to shorten the path;
+stable paths must therefore be of extremal proper-length.  In flat
+space, the calculus of variations proves that the path is straight in
+the sense of zero curvature.  In terrestrial gravity, standard
+$25\times 10^{-6}\unit{kg}\cdot\unit{m}^{-1}$ cotton sewing thread at
+its breaking strain of about $40\unit{N}$ would have a sag of~${\sim}
 6\times 10^{-6}\unit{m}$ over a $1\unit{m}$ horizontal scale: barely

### 2021-01-26 physics_educator_rebuttal.txt
fixes #25

@@
 That is an extremely good observation.  The word "large" is
-inappropriate here, I think.  Asymptotic expansions can be confusing
-and what I meant to say was that the ratio of proper string length to
-Euclidean string length approached infinity. But the word "large" is
-not right here, and I have rephrased accordingly.
+inappropriate and misleading here, I think.  Asymptotic expansions can
+be confusing and what I meant to say was that the ratio of proper
+string length to Euclidean string length (that is, length as estimated
+by the Euclidean formula (1+r)-1 where 'r' is circumferential radius)
+approaches infinity.  But the word "large" is not right here, and I
+have rephrased accordingly.
 
-    NOT YET FINALISED THIS
+I have also realised that this equation and the preceding one for
+small w can be made much more comprehensible by giving an example, and
+I have included two examples using solar-mass black holes.
 

### 2021-01-26 string_physics_educator_revision2.tex
fixes #25

@@
 (compare the flat space string length of $2Mw$).  This would indicate
- that, close to the Schwarzschild radius $r=2M$, large (but finite)
- amounts of string are required to connect two points with slightly
- differing radial coordinates.  Passing to the case $w\gg 1$ we have
- an asymptotic expansion that begins
+ that, close to the Schwarzschild radius $r=2M$, more string is
+ required to connect two points with slightly differing radial
+ coordinates.  Taking a solar-mass black hole and $w=10^{-3}$ as an
+ example, the proper length computed in flat space would be
+ $10^{-2}\cdot 2M\simeq 3\,\mathrm{m}$---compare $\sim
+ 187\,\mathrm{m}$ from equation~\ref{ell}.  Passing to the case $w\gg
+ 1$ we have an asymptotic expansion that begins
 
@@
 \begin{equation}\label{asymptotic_ell}
-  \ell = 2M\left(w +  \frac{1}{2}\log w + \frac{1}{2} + \log(2)  +  \frac{1}{8}w^{-1} + O(w^{-2})\right),
+  \ell = 2M\left(w +  \frac{1}{2}\log w + \frac{1}{2} + \log(2)  +  \frac{1}{8}w^{-1} + O(w^{-2})\right).
 \end{equation}
 
-\noindent suggesting that there is ``extra space'', sensible at large
-distances from the event horizon, with radial strings being
-arbitrarily longer than flat space would allow: a nice illustration of
-the failure of Euclidean geometry.
+\noindent The discrepancy between the flat-space result of $2Mw$ and
+equation~\ref{asymptotic_ell} suggests that, even far from the event
+horizon, radial strings are arbitrarily longer than flat space would
+imply: a nice illustration of the failure of Euclidean geometry.
+Again taking a solar mass black hole but with $w=2\times 10^7$
+(corresponding to the orbital radius of Mercury) the difference
+between the classical result and equation~\ref{asymptotic_ell} is
+about $28\,\mathrm{km}$ of extra string.  However, the prediction is
+unlikely to be the basis of a fourth test of general relativity---the
+others being Shapiro time delay, bending of starlight, and the
+perihelion advance of Mercury~\cite{einstein1916}---as the Sun is not
+inside its own Schwarzschild radius.
 

### 2021-01-26 string_physics_educator_revision2.tex
fixes #27

@@
 Figure~\ref{flamm}.  It is interesting to compare
-equation~\ref{rdashdash} with the corresponding equation for null
-geodesics, which is usually given~\cite{wald} in terms of $u=2M/r$ as
-$u''=3u^2/2-u$, but is equivalent
+equation~\ref{rdashdash} with the corresponding equation for
+equatorial null geodesics, which is usually given~\cite{wald} in terms
+of $u=2M/r$ as $u''=3u^2/2-u$, but is equivalent
 to~$r''\left(\phi\right)=r-3M+2\left(r'\right)^2/r$.  This shows that

### 2021-01-26 string_physics_educator_revision2.tex
fixes #28
fixes #29

@@
 the system is invariant under~$\phi\longrightarrow-\phi$, the string
-must be symmetrical about this point.  If $\phi'$ is small, then a
+must be symmetrical about this point.  If $\r'$ is small, then a
 string follows $r''=r-2M$, while tangential null geodesics follow
-$r''=r-3M$.  Thus a taut string would {\em look} curved to a local
-observer, having a (visual) radius of curvature of
+$r''=r-3M$.  Thus a taut string would {\em look} curved to a static
+observer in Schwarzschild coordinates, having a (visual) radius of
+curvature of
 $r_0^2/M\sqrt{1-2M/r_0}$ (the extra term comes from the fact that

### 2021-01-26 string_physics_educator_revision2.tex
typo
start work on issue #30

@@
 the system is invariant under~$\phi\longrightarrow-\phi$, the string
-must be symmetrical about this point.  If $\r'$ is small, then a
-string follows $r''=r-2M$, while tangential null geodesics follow
-$r''=r-3M$.  Thus a taut string would {\em look} curved to a static
+must be symmetrical about this point.  If $r'$ is small, then a string
+follows $r''=r-2M$, while tangential null geodesics follow $r''=r-3M$.
+Working to first order, and writing $d$ for the difference in
+Schwarzschild $r$ between string and null geodesic, we see that
+$d''=M$ and with appropriate boundary conditions
+$d=\frac{Mx^2}{2r_0^2}$, where $x=r_0\theta$ is an appropriate
+circumferential coordinate.  However, to a static local observer tied
+to the string would
+$d_\mathrm{local}=Mx^2/\left(2r_0^2\sqrt{1-2M/r_0}\right)$,
+
+$d_\mathrm{local}=d/\sqrt{1-2M/r_0}$,
+
+Thus a taut string would {\em look} curved to a static
 observer in Schwarzschild coordinates, having a (visual) radius of

### 2021-01-26 physics_educator_rebuttal.txt
fixes #27
fixes #28
fixes #29
fixes #30 fixes #31

@@
 
-I have rephrased accordingly.
-
-    NOT YET FINALISED THIS
+corrected.
 
@@
 
-    NOT YET FINALISED THIS
-
 > 6. Local observer (page 4 line 7) I believe the author means static
@@
 
-    NOT YET FINALISED THIS
-
 >  
@@
 
-On re-reading, the original manuscript was a little cryptic.  Now rephrased:
-
-    NOT YET FINALISED THIS
+On re-reading, the original manuscript was a little cryptic.  I have
+added a detailed derivation of this result that should clarify the
+issue.  In essence, a light beam and a tangential string will diverge
+quadratically and this leads to the radius of curvature quoted.
 
@@
 
-this is the visual radius, any confusion cleared up by point 7 correction above.
+Derivation added as a result of point 7 above should clarify this
+issue.
 

### 2021-01-26 string_physics_educator_revision2.tex
fixes #30 fixes #31

--- 2021-01-26 string_physics_educator_revision2.tex
@@
 the system is invariant under~$\phi\longrightarrow-\phi$, the string
-must be symmetrical about this point.  If $\phi'$ is small, then a
-string follows $r''=r-2M$, while tangential null geodesics follow
-$r''=r-3M$.  Thus a taut string would {\em look} curved to a local
-observer, having a (visual) radius of curvature of
-$r_0^2/M\sqrt{1-2M/r_0}$ (the extra term comes from the fact that
-appearance is driven by local $ds$, not Schwarzschild $r$).  At the
-photon sphere $r_0=3M$, such a string would appear to have a radius
-of~$\sqrt{27}M$.  For a solar mass black hole this would be about
-$15\unit{km}$, a readily understandable result.
+must be symmetrical about this point.  If $r'$ is small, then a string
+follows $r''=r-2M$, while tangential null geodesics follow $r''=r-3M$.
+Working to first order, and writing~$d$ for the difference in
+Schwarzschild $r$ between string and null geodesic, we see that
+$d''=M$.  Transforming coordinates to $(x,y) = (r\phi,r-r_0)$ and with
+boundary conditions corresponding to a string and null geodesic
+tangential at the $(x,y)$ origin, we see that
+$d=\frac{M}{2}(x/r_0)^2$.  However, a static local observer tied to
+the string would measure
+$d_\mathrm{local}=d/\sqrt{1-2M/r_0}$---applying
+equation~\ref{schwarzschild}---and obtain
+$d_\mathrm{local}=\frac{M}{2}(x/r_0)^2\left(1-2M/r_0\right)^{-1/2}$
+which would correspond to a radius of curvature
+of~$r_0\cdot\frac{r_0}{M}\cdot\left(1-2M/r_0\right)^{-1/2}$.  Thus a
+taut string would {\em look} curved to a static observer in
+Schwarzschild coordinates, in the sense that a tangential light beam
+and taut string would diverge quadratically.  At the photon sphere
+$r_0=3M$, such a string would appear to have a radius of~$\sqrt{27}M$.
+For a solar mass black hole this would be about $15\unit{km}$, a
+readily understandable result.
 

### 2021-01-27 physics_educator_rebuttal.txt
fixes #32

@@
 
-I don't think the concept of "spatial sections" is particularly
-helpful, even though I do present a diagram of Flamm's paraboloid
-[which is obligatory I would say].  A spatial section S is a
-three-dimensional subset of a spacetime in which we require the
-squared interval between any two events of S to be spacelike.  It is
-easy to give an example of a spatial section in flat Minkowski space
-(t,x,y,z) as (t0,x,y,z) for some fixed t0.  Such constructions are
-somewhat more difficult in Schwarzschild space but we might use
-Schwarzschild coordinates (t,r,theta,phi) and then specify
-S2=(t0,r,theta,phi).  But for r1,r2<2M and fixed [theta,phi] we find
-that the interval between (t0,r1,theta,phi) and (t0,r2,theta,phi) is
-timelike; thus S2 is not a spatial section in the sense defined above.
-But for r>2M it is a perfectly acceptable spatial section, and this is
-illustrated in Figure 1.
+I would argue that emphasising the concept of "spatial sections" might
+be somewhat counterproductive, even though I admittedly present a
+diagram of Flamm's paraboloid [which is obligatory here, I would say].
+A spatial section S is a 3D subset of a 4D spacetime in which we
+require the squared interval between any two distinct events of S to
+be spacelike.  It is easy to give an example of a spatial section in
+flat Minkowski space: with standard coordinates (t,x,y,z) we consider
+events (t0,x,y,z) for some fixed t0.  Such constructions are somewhat
+more difficult in Schwarzschild space but we might use Schwarzschild
+coordinates (t,r,theta,phi) and then specify S2=(t0,r,theta,phi).  The
+Schwarzschild metric then shows that S2 is a spatial section, provided
+that r>2M, because the Schwarzschild metric (equation 1 in the
+manuscript) is diagonal, and the nonvanishing components are strictly
+positive.
 
@@
 use standard Schwarzschild coordinates (t,r,theta,phi) and then simply
-ignore the t coordinate as it plays no part in calculating proper
-length of a string (of course, we need to be careful to specify that
-we are outside the event horizon).  If the string is static, we may
-imagine two static observers tied to the string; in calculating the
-amount of proper distance between the two observers (string length) we
-simply ignore the t coordinate and integrate the spatial component of
-the Schwarzschild metric.  This process is distinct from the concept
-of spatial section in which we condition on t=t0.
+ignore the 't' coordinate as it plays no part in calculating proper
+length of a string (being careful to specify that we are outside the
+event horizon).  If the string is static, we may imagine two static
+observers tied to the string; in calculating the amount of proper
+distance between the two observers (string length) we simply ignore
+the 't' coordinate and integrate the spatial component of the
+Schwarzschild metric.  This process is distinct from the concept of
+spatial section in which we condition on t=t0.
 
@@
 and photons follow null geodesics.  Light inextensible string follows
-neither of these paths.  We might consider a string in terms of the
-world lines of its links, and then require that the proper [spatial]
-distance between the links be constant.  The
+neither of these paths.  I think the revised manuscript makes this
+clear

### 2021-01-27 physics_educator_rebuttal.txt
slight rewording

@@
 
-Overall response: the comments are prefectly reasonable and I have
+Overall response: the comments are perfectly reasonable and I have
 addressed them individually.  The manuscript is much stengthened as a
@@
 indeed a major one, and one that had no explicit justification in the
-manuscript.  But on the other, it requires no justification as it is
-true by defnition.  Didactically, one might conceptualise a string as
-being comprised of many tiny links, each of which contributes some
-small length to the string, and each of which is sufficiently small
-for Euclidean geometry to describe its length.  Each link is subject
-to a well-defined local tension with respect to its neighbours.
+manuscript.  But I could argue that it requires no justification: one
+could say that the claim is true by definition.  Didactically, one
+might conceptualise a string as being comprised of many tiny links,
+each of which contributes some small length to the string, and each of
+which is sufficiently small for Euclidean geometry to describe its
+length.  Each link is subject to a well-defined local tension with
+respect to its neighbours.
 
@@
 
-Imagine two strings, S1 and S2, which both smoothly join point p to
-point q (perhaps in a Schwarzschild space), and their paths are close
-in the sense of the Calculus of Variations.  Further, imagine that S1
-has a proper length slightly exceeding that of S2.  If we then impose
-a tension on the strings, we see that the S1 is unstable with respect
-to S2, because S2 has fewer links.  Thus equilibrium paths of light
-inextensible strings are at least local minima of length.
-
 However, much of the above reasoning is necessarily tautological; it
@@
 under tension" as one that adopts a minimal [or at least extremal]
-proper-length path between its end points, and i have included a short
+proper-length path between its end points.  I have included a short
 discussion in the manuscript.
@@
 
-:: The sentence was: Here I show how considering light inextensible 
+:: Here I show how considering light inextensible 
 :: string can be used to interpret the concept of "straight line" in the
@@
 would be potentially misleading.  My thinking was that the concept of
-"straight line" ---in the abstract---is to be re-interpreted in the
-manuscript as the path taken by a light inextensible string.
+"straight line" ---in the abstract sense of how straightness itself is
+defined---is to be re-interpreted in the manuscript as the path taken
+by a light inextensible string.
 
@@
 was to point out that we Euclidean predictions are arbirtarily
-incorrect as we move away from the black hole.  I have rephrased.
-
-    NOT YET FINALISED THIS
+incorrect as we move away from the black hole.  I have rephrased and
+included a numerical example which should appeal to students of GR.
 
@@
 
-corrected.
+corrected, equatorial restriction now mentioned explicitly
 
@@
 issue.  In essence, a light beam and a tangential string will diverge
-quadratically and this leads to the radius of curvature quoted.
+quadratically and this leads to the radius of curvature quoted.  The
+manuscript is much clearer than before.
 
@@
 Derivation added as a result of point 7 above should clarify this
-issue.
+issue.  This is the curvature of the string that a static observer,
+who is tied to the string, would measure if he considered a beam of
+light to be "straight".
 
@@
 and photons follow null geodesics.  Light inextensible string follows
-neither of these paths.  I think the revised manuscript makes this
-clear
+neither of these paths.  I hope that the revised manuscript makes this
+clear, and the various diagrams of strings and null geodesics should
+emphasise the difference between the two.

### 2021-02-24 tpe_edreview_rebuttal.txt
start addressing issue #22

Review of "Light inextensible strings (thread) under tension in the
Schwarzschild geometry" submitted to The Physics Educator

I want to begin by congratulating the author on a very clever
exercise.  If the route of a non-stretchable string under tension
would take the path of least distance, asking about the behavior of a
tight string in a Schwarzschild geometry, and comparing that path to
that of a light ray, offers an interesting question.  One might
initially suppose the two paths would be the same, or approximately
so.  The author rightly observes that General Relativity is a "topic
of long-standing fascination for physics undergraduates." True, but I
am unsure whether this particular exercise will offer more
understanding or more confusion for the intended undergraduates,
because the distinction between a stretched string and a null metrics
may be blurred (see point 11 below).

I would point out that at least one calculation presented in the paper
can found in existing introductory General Relativity textbooks.
Specifically, the meaning of the r-coordinate, and the difference
between Delta r and the distance between two concentric shells of
different r-coordinates (eq. 2) is thoroughly discussed in Taylor &
Wheeler's book Exploring Black Holes: Introduction to General
Relativity (Addison-Wesley-Longman, 2000), Ch. 2.  While the
calculation of Eq. (2) should be included in this paper for
completeness and done before going on to other calculations, Taylor &
Wheeler's black holes book is a standard pedagogical reference for
such calculations, and should be included among the references.
Taylor & Wheeler's book is very likely the intended undergraduate
student's first encounter with General Relativity.

The paper raises a few questions in this reviewer's mind, and in
elsewhere it offers opportunities for suggestions.

1.  Question.  First paragraph in "Introduction": The third sentence
"Here I interpret straight line as the path adopted by a stationary,
light inextensible string under tension passing between two points."
Is this to be understood as identical to the definition of "geodesic"?
If that is the assumption, can it be justified, because when one
thinks of geodesic in General Relativity, one has "free-fall" in mind,
and the stretched string, being "stationary," would not be in
free-fall in the global coordinate system.

2.  Suggestion about the line that begins on p. 1 and continues to
p. 2, about "lengthscales $L\leq T/(\mu g)$: Saying it this way may
present a sticking point for the undergraduate audience towards whom
this paper in intended.  The logic behind this inequality would show
greater clarity if one says something like "the weight of the string
does not exceed the tension, so that $L\mu g<T$".  From this the
stated inequality about lengthscales follows easily.

3.  Suggestion.  Stating "$c^4/4GM$ for the strength of gravity at the
event horizon" may leave an undergraduate audience baffled.  By
starting from where they are, with $g=GM/r^2$, then setting r equal to
the Schwarzschild radius, the stated result follows immediately,
without mystery.

4.  Question. Lines 3-4, page 2.  Can the author explain what is meant
by "the dominant energy condition"?

5.  Suggestion.  2nd and 3rd line below eq. (2): the statement "A
circular string at constant r will, trivially, have proper length
$2\pi r$."  Stating the relation this way muddies the logic and the
distinction between flat and Schwarzschild space.  The r-coordinate of
a spherical shell centered on a point mass M is defined operationally
by measuring the circumference of the shell and dividing the
circumference's value by $2\pi$.  That must be understood before the
calculations of Eq. (2) can make sense.  As the sentence is presently
structured it may obscure the distinction between the r coordinate and
radial lengths because it invites the undergraduate reader to
interpret the "triviality" as a statement about plane Euclidean
geometry instead of the Schwarzschild geometry's definition of the
r-coordinate.

6.  Question.  There may be a typographical error in Eq. (2).
Shouldn't the coefficient of the logarithm be 2M instead of 1/2?  As
written the equation is not dimensionally consistent, since r and M
are measured in the same units.  Compare to Taylor & Wheeler's
Exploring Black Holes, p. 2-28, Eq. (27).

7.  Suggestion: It appears that the $L$ and the $l$ that appear
between Eq. (1) and Eq. (3) denote the same quantity.  Why use
different symbols?

8.  Suggestion: In the line above Eq. (3), if $w >> 0$ then one does
not need absolute value bars in $|w|<< 1$.

9.  Question.  2nd line below Eq. (4), why are radial strings
"arbitrarily" longer than flat space would imply?  It appears the
difference is a matter of calculation (or measurement), not arbitrary.
Please explain this choice of words.

10. Suggestion.  Line below Eq. (5), I think the word "dashes" should
be "primes".  This is a dash: --- and this is a prime: ', at least in
usage with which I am familiar.

11.  The paragraph below Eq. (6) is puzzling.  Here the path of the
tight string is being compared to an "equatorial null geodesic," which
I take to mean the path of a light beam moving along the equator.  If
we are considering a light beam, then shouldn't a term with $t$ in it
appear somewhere?  The puzzlement continues given the statement that
"light, being sensitive to the time component of the metric" whereas
the equations discussed in this paragraph say nothing (that I can see)
about the time component of the metric.  The proper time interval
would be zero, but the coordinate time interval $dt$ would, I think,
be nonzero.  What am I missing?

12.  In the passage below Eq. (6), I am having trouble making the
connection between $r''(\phi)=r-3M + 2(r')^2/r$ and the claim that the
solution is $r=r_0\sec(\phi-\phi_0)$.  What trig identity am I
missing?

13.  Can the author explain more fully what is meant by a "tangential
null geodesic" (several lines after Eq. 6)?  That terminology makes it
sound like the r-coordinate does not change, whence $r'$ would be
zero, but that appears to not be the case.  In this section, if $r'$
is small I see where the approximation $r''=r-2M$ comes from, but I am
having trouble seeing where $r''-r-3M$ arises for the alleged
tangential null geodesics.  What am I not seeing?

I will have to take the author's word for the numerical results.

This is an interesting paper.  I congratulate the author for thinking
of such an intriguing application of the Schwarzschild metric.
However, in this reviewer's opinion, before the paper can be well
fitted to its announced audience ("an undergraduate homework problem")
some points need to be clarified.

### 2021-02-24 tpe_edreview_rebuttal.txt
start work on #22

@@
-Review of "Light inextensible strings (thread) under tension in the
-Schwarzschild geometry" submitted to The Physics Educator
+Herewith my rebuttal to the review of "Light inextensible strings..."
+The referee's comments are preceded by a chevron (">") and my response
+is inline.  I have tinkered with linebreaks here and there in the
+interests of clarity, and have replaced some mathematical notation in
+the original review with latex equivalents.
 
-I want to begin by congratulating the author on a very clever
-exercise.  If the route of a non-stretchable string under tension
-would take the path of least distance, asking about the behavior of a
-tight string in a Schwarzschild geometry, and comparing that path to
-that of a light ray, offers an interesting question.  One might
-initially suppose the two paths would be the same, or approximately
-so.  The author rightly observes that General Relativity is a "topic
-of long-standing fascination for physics undergraduates." True, but I
-am unsure whether this particular exercise will offer more
-understanding or more confusion for the intended undergraduates,
-because the distinction between a stretched string and a null metrics
-may be blurred (see point 11 below).
+Short story: I agree with all the comments made and have modified the
+manuscript accordingly; changes appear in the manuscript using colour
+coding for ease of reference.  The document is stronger, more
+cohesive, more accurate, and hopefully more instructive; and I commend
+it to you.
 
-I would point out that at least one calculation presented in the paper
-can found in existing introductory General Relativity textbooks.
-Specifically, the meaning of the r-coordinate, and the difference
-between Delta r and the distance between two concentric shells of
-different r-coordinates (eq. 2) is thoroughly discussed in Taylor &
-Wheeler's book Exploring Black Holes: Introduction to General
-Relativity (Addison-Wesley-Longman, 2000), Ch. 2.  While the
-calculation of Eq. (2) should be included in this paper for
-completeness and done before going on to other calculations, Taylor &
-Wheeler's black holes book is a standard pedagogical reference for
-such calculations, and should be included among the references.
-Taylor & Wheeler's book is very likely the intended undergraduate
-student's first encounter with General Relativity.
-
-The paper raises a few questions in this reviewer's mind, and in
-elsewhere it offers opportunities for suggestions.
-
-1.  Question.  First paragraph in "Introduction": The third sentence
-"Here I interpret straight line as the path adopted by a stationary,
-light inextensible string under tension passing between two points."
-Is this to be understood as identical to the definition of "geodesic"?
-If that is the assumption, can it be justified, because when one
-thinks of geodesic in General Relativity, one has "free-fall" in mind,
-and the stretched string, being "stationary," would not be in
-free-fall in the global coordinate system.
-
-2.  Suggestion about the line that begins on p. 1 and continues to
-p. 2, about "lengthscales $L\leq T/(\mu g)$: Saying it this way may
-present a sticking point for the undergraduate audience towards whom
-this paper in intended.  The logic behind this inequality would show
-greater clarity if one says something like "the weight of the string
-does not exceed the tension, so that $L\mu g<T$".  From this the
-stated inequality about lengthscales follows easily.
-
-3.  Suggestion.  Stating "$c^4/4GM$ for the strength of gravity at the
-event horizon" may leave an undergraduate audience baffled.  By
-starting from where they are, with $g=GM/r^2$, then setting r equal to
-the Schwarzschild radius, the stated result follows immediately,
-without mystery.
-
-4.  Question. Lines 3-4, page 2.  Can the author explain what is meant
-by "the dominant energy condition"?
-
-5.  Suggestion.  2nd and 3rd line below eq. (2): the statement "A
-circular string at constant r will, trivially, have proper length
-$2\pi r$."  Stating the relation this way muddies the logic and the
-distinction between flat and Schwarzschild space.  The r-coordinate of
-a spherical shell centered on a point mass M is defined operationally
-by measuring the circumference of the shell and dividing the
-circumference's value by $2\pi$.  That must be understood before the
-calculations of Eq. (2) can make sense.  As the sentence is presently
-structured it may obscure the distinction between the r coordinate and
-radial lengths because it invites the undergraduate reader to
-interpret the "triviality" as a statement about plane Euclidean
-geometry instead of the Schwarzschild geometry's definition of the
-r-coordinate.
-
-6.  Question.  There may be a typographical error in Eq. (2).
-Shouldn't the coefficient of the logarithm be 2M instead of 1/2?  As
-written the equation is not dimensionally consistent, since r and M
-are measured in the same units.  Compare to Taylor & Wheeler's
-Exploring Black Holes, p. 2-28, Eq. (27).
-
-7.  Suggestion: It appears that the $L$ and the $l$ that appear
-between Eq. (1) and Eq. (3) denote the same quantity.  Why use
-different symbols?
-
-8.  Suggestion: In the line above Eq. (3), if $w >> 0$ then one does
-not need absolute value bars in $|w|<< 1$.
-
-9.  Question.  2nd line below Eq. (4), why are radial strings
-"arbitrarily" longer than flat space would imply?  It appears the
-difference is a matter of calculation (or measurement), not arbitrary.
-Please explain this choice of words.
-
-10. Suggestion.  Line below Eq. (5), I think the word "dashes" should
-be "primes".  This is a dash: --- and this is a prime: ', at least in
-usage with which I am familiar.
-
-11.  The paragraph below Eq. (6) is puzzling.  Here the path of the
-tight string is being compared to an "equatorial null geodesic," which
-I take to mean the path of a light beam moving along the equator.  If
-we are considering a light beam, then shouldn't a term with $t$ in it
-appear somewhere?  The puzzlement continues given the statement that
-"light, being sensitive to the time component of the metric" whereas
-the equations discussed in this paragraph say nothing (that I can see)
-about the time component of the metric.  The proper time interval
-would be zero, but the coordinate time interval $dt$ would, I think,
-be nonzero.  What am I missing?
-
-12.  In the passage below Eq. (6), I am having trouble making the
-connection between $r''(\phi)=r-3M + 2(r')^2/r$ and the claim that the
-solution is $r=r_0\sec(\phi-\phi_0)$.  What trig identity am I
-missing?
-
-13.  Can the author explain more fully what is meant by a "tangential
-null geodesic" (several lines after Eq. 6)?  That terminology makes it
-sound like the r-coordinate does not change, whence $r'$ would be
-zero, but that appears to not be the case.  In this section, if $r'$
-is small I see where the approximation $r''=r-2M$ comes from, but I am
-having trouble seeing where $r''-r-3M$ arises for the alleged
-tangential null geodesics.  What am I not seeing?
-
-I will have to take the author's word for the numerical results.
-
-This is an interesting paper.  I congratulate the author for thinking
-of such an intriguing application of the Schwarzschild metric.
-However, in this reviewer's opinion, before the paper can be well
-fitted to its announced audience ("an undergraduate homework problem")
-some points need to be clarified.
+> Review of "Light inextensible strings (thread) under tension in the
+> Schwarzschild geometry" submitted to The Physics Educator
+> 
+> I want to begin by congratulating the author on a very clever
+> exercise.  If the route of a non-stretchable string under tension
+> would take the path of least distance, asking about the behavior of a
+> tight string in a Schwarzschild geometry, and comparing that path to
+> that of a light ray, offers an interesting question.  One might
+> initially suppose the two paths would be the same, or approximately
+> so.  The author rightly observes that General Relativity is a "topic
+> of long-standing fascination for physics undergraduates." True, but I
+> am unsure whether this particular exercise will offer more
+> understanding or more confusion for the intended undergraduates,
+> because the distinction between a stretched string and a null metrics
+> may be blurred (see point 11 below).
+> 
+> I would point out that at least one calculation presented in the paper
+> can found in existing introductory General Relativity textbooks.
+> Specifically, the meaning of the r-coordinate, and the difference
+> between Delta r and the distance between two concentric shells of
+> different r-coordinates (eq. 2) is thoroughly discussed in Taylor &
+> Wheeler's book Exploring Black Holes: Introduction to General
+> Relativity (Addison-Wesley-Longman, 2000), Ch. 2.  While the
+> calculation of Eq. (2) should be included in this paper for
+> completeness and done before going on to other calculations, Taylor &
+> Wheeler's black holes book is a standard pedagogical reference for
+> such calculations, and should be included among the references.
+> Taylor & Wheeler's book is very likely the intended undergraduate
+> student's first encounter with General Relativity.
+> 
+> The paper raises a few questions in this reviewer's mind, and in
+> elsewhere it offers opportunities for suggestions.
+> 
+> 1.  Question.  First paragraph in "Introduction": The third sentence
+> "Here I interpret straight line as the path adopted by a stationary,
+> light inextensible string under tension passing between two points."
+> Is this to be understood as identical to the definition of "geodesic"?
+> If that is the assumption, can it be justified, because when one
+> thinks of geodesic in General Relativity, one has "free-fall" in mind,
+> and the stretched string, being "stationary," would not be in
+> free-fall in the global coordinate system.
+> 
+> 2.  Suggestion about the line that begins on p. 1 and continues to
+> p. 2, about "lengthscales $L\leq T/(\mu g)$: Saying it this way may
+> present a sticking point for the undergraduate audience towards whom
+> this paper in intended.  The logic behind this inequality would show
+> greater clarity if one says something like "the weight of the string
+> does not exceed the tension, so that $L\mu g<T$".  From this the
+> stated inequality about lengthscales follows easily.
+> 
+> 3.  Suggestion.  Stating "$c^4/4GM$ for the strength of gravity at the
+> event horizon" may leave an undergraduate audience baffled.  By
+> starting from where they are, with $g=GM/r^2$, then setting r equal to
+> the Schwarzschild radius, the stated result follows immediately,
+> without mystery.
+> 
+> 4.  Question. Lines 3-4, page 2.  Can the author explain what is meant
+> by "the dominant energy condition"?
+> 
+> 5.  Suggestion.  2nd and 3rd line below eq. (2): the statement "A
+> circular string at constant r will, trivially, have proper length
+> $2\pi r$."  Stating the relation this way muddies the logic and the
+> distinction between flat and Schwarzschild space.  The r-coordinate of
+> a spherical shell centered on a point mass M is defined operationally
+> by measuring the circumference of the shell and dividing the
+> circumference's value by $2\pi$.  That must be understood before the
+> calculations of Eq. (2) can make sense.  As the sentence is presently
+> structured it may obscure the distinction between the r coordinate and
+> radial lengths because it invites the undergraduate reader to
+> interpret the "triviality" as a statement about plane Euclidean
+> geometry instead of the Schwarzschild geometry's definition of the
+> r-coordinate.
+> 
+> 6.  Question.  There may be a typographical error in Eq. (2).
+> Shouldn't the coefficient of the logarithm be 2M instead of 1/2?  As
+> written the equation is not dimensionally consistent, since r and M
+> are measured in the same units.  Compare to Taylor & Wheeler's
+> Exploring Black Holes, p. 2-28, Eq. (27).
+> 
+> 7.  Suggestion: It appears that the $L$ and the $l$ that appear
+> between Eq. (1) and Eq. (3) denote the same quantity.  Why use
+> different symbols?
+> 
+> 8.  Suggestion: In the line above Eq. (3), if $w >> 0$ then one does
+> not need absolute value bars in $|w|<< 1$.
+> 
+> 9.  Question.  2nd line below Eq. (4), why are radial strings
+> "arbitrarily" longer than flat space would imply?  It appears the
+> difference is a matter of calculation (or measurement), not arbitrary.
+> Please explain this choice of words.
+> 
+> 10. Suggestion.  Line below Eq. (5), I think the word "dashes" should
+> be "primes".  This is a dash: --- and this is a prime: ', at least in
+> usage with which I am familiar.
+> 
+> 11.  The paragraph below Eq. (6) is puzzling.  Here the path of the
+> tight string is being compared to an "equatorial null geodesic," which
+> I take to mean the path of a light beam moving along the equator.  If
+> we are considering a light beam, then shouldn't a term with $t$ in it
+> appear somewhere?  The puzzlement continues given the statement that
+> "light, being sensitive to the time component of the metric" whereas
+> the equations discussed in this paragraph say nothing (that I can see)
+> about the time component of the metric.  The proper time interval
+> would be zero, but the coordinate time interval $dt$ would, I think,
+> be nonzero.  What am I missing?
+> 
+> 12.  In the passage below Eq. (6), I am having trouble making the
+> connection between $r''(\phi)=r-3M + 2(r')^2/r$ and the claim that the
+> solution is $r=r_0\sec(\phi-\phi_0)$.  What trig identity am I
+> missing?
+> 
+> 13.  Can the author explain more fully what is meant by a "tangential
+> null geodesic" (several lines after Eq. 6)?  That terminology makes it
+> sound like the r-coordinate does not change, whence $r'$ would be
+> zero, but that appears to not be the case.  In this section, if $r'$
+> is small I see where the approximation $r''=r-2M$ comes from, but I am
+> having trouble seeing where $r''-r-3M$ arises for the alleged
+> tangential null geodesics.  What am I not seeing?
+> 
+> I will have to take the author's word for the numerical results.
+> 
+> This is an interesting paper.  I congratulate the author for thinking
+> of such an intriguing application of the Schwarzschild metric.
+> However, in this reviewer's opinion, before the paper can be well
+> fitted to its announced audience ("an undergraduate homework problem")
+> some points need to be clarified.

### 2021-02-24 string_physics_educator_revision3.tex
start work on #22

--- 2021-01-26 string_physics_educator_revision2.tex
@@
 
-\address{School of Engineering, Computing, and Mathematical Sciences, Auckland University of Technology, 55 Wellesley Street East\\
-Auckland 1142, New Zealand
+\address{School of Engineering, Computing, and Mathematical Sciences\\Auckland University of Technology\\55 Wellesley Street East\\
+  Auckland CBD, Auckland 1010, New Zealand\\
 \email{hankin.robin@gmail.com}}
@@
 be in stable equilibrium as the tension acts to shorten the path;
-stable paths must therefore be of extremal proper-length.  In flat
+stable paths must therefore be of extremal proper length.  In flat
 space, the calculus of variations proves that the path is straight in
@@
 circular string at constant~$r$ will, trivially, have proper
-length~$2\pi r$.
-
-Considering first the radial component we may calculate the
-proper length $L\left(r_1,r_2\right)$ of a string stretching
-from~$r=r_1\geqslant 2M$ to~$r_2\geqslant r_1$ at constant $\theta,\phi$:
+length~$2\pi r$.  Considering first the radial component we may
+calculate the proper length $L\left(r_1,r_2\right)$ of a string
+stretching from~$r=r_1\geqslant 2M$ to~$r_2\geqslant r_1$ at constant
+$\theta,\phi$:
 
@@
   \frac{\sqrt{r}+\sqrt{r-2M}}{\sqrt{r}-\sqrt{r-2M}}\right)
-  \right|_{r_1}^{r_2}
+  \right|_{r_1}^{r_2}.
   \end{equation}
@@
 Observe that integration may be performed with a lower limit
-of~$r_1=2M$ with no difficulty, even though the radial component of the
-Schwarzschild metric approaches infinity there.
-
-Elementary methods show that the proper length $\ell$ of radial
-strings has some non-Euclidean features.  Firstly, consider the proper
-length from $r=2M$ to $r=2M(1+w)$ where $w>0, \left|w\right|\ll 1$.
-Then
+of~$r_1=2M$ with no difficulty, even though the radial component of
+the Schwarzschild metric approaches infinity there.  Elementary
+methods show that the proper length $\ell$ of radial strings has some
+non-Euclidean features.  Firstly, consider the proper length from
+$r=2M$ to $r=2M(1+w)$ where $w>0, \left|w\right|\ll 1$.  Then
 
@@
  required to connect two points with slightly differing radial
- coordinates.  Taking a solar-mass black hole and $w=10^{-3}$ as an
- example, the proper length computed in flat space would be
- $10^{-2}\cdot 2M\simeq 3\,\mathrm{m}$---compare $\sim
- 187\,\mathrm{m}$ from equation~\ref{ell}.  Passing to the case $w\gg
- 1$ we have an asymptotic expansion that begins
+ coordinates than would be expected classically.  Taking a solar-mass
+ black hole and $w=10^{-3}$ as an example, the proper length computed
+ in flat space would be $10^{-2}\cdot 2M\simeq 3\,\mathrm{m}$; compare
+ $\simeq 187\,\mathrm{m}$ from equation~\ref{ell}.  Passing to the
+ case $w\gg 1$ we have an asymptotic expansion that begins
 
@@
 Working to first order, and writing~$d$ for the difference in
-Schwarzschild $r$ between string and null geodesic, we see that
-$d''=M$.  Transforming coordinates to $(x,y) = (r\phi,r-r_0)$ and with
-boundary conditions corresponding to a string and null geodesic
+Schwarzschild~$r$ between string and null geodesic, we have $d''=M$.
+Transforming coordinates to $(x,y) = (r\phi,r-r_0)$ and with boundary
+conditions corresponding to both string and null geodesic being
 tangential at the $(x,y)$ origin, we see that

### 2021-02-24 tpe_edreview_rebuttal.txt
fixes #40

@@
 >
+
+corrected, now using $\ell$ consistently for proper length.
+
 > 8.  Suggestion: In the line above Eq. (3), if $w >> 0$ then one does

### 2021-02-24 string_physics_educator_revision3.tex
fixes #40

@@
 \usepackage{units}
+\usepackage{xcolor}
 \usepackage[superscript]{cite}
@@
 length~$2\pi r$.  Considering first the radial component we may
-calculate the proper length $L\left(r_1,r_2\right)$ of a string
+calculate the proper length $\ell\left(r_1,r_2\right)$ of a string
 stretching from~$r=r_1\geqslant 2M$ to~$r_2\geqslant r_1$ at constant

### 2021-02-25 tpe_edreview_rebuttal.txt
fixes #39

@@
 >
+
+Factor of M now added (actually this caused me some considerable
+confusion and uncertainty.  The equation presented in Taylor and
+Wheeler has a factor of 2M but it should be M---see the attached
+mathematica screenshot, plus the identity arctanh(x) =
+log((1+x)/(1-x))/2.
+
+This was just a typo in that particular line and has no effect on any
+other part of the manuscript.
+
 > 7.  Suggestion: It appears that the $L$ and the $l$ that appear

### 2021-02-25 string_physics_educator_revision3.tex
fixes #39
fixes #41
fixes #43

@@
   \left.
-  \sqrt{r(r-2M)} +\frac{1}{2}\log\left(
+  \sqrt{r(r-2M)} +M\log\left(
   \frac{\sqrt{r}+\sqrt{r-2M}}{\sqrt{r}-\sqrt{r-2M}}\right)
@@
 non-Euclidean features.  Firstly, consider the proper length from
-$r=2M$ to $r=2M(1+w)$ where $w>0, \left|w\right|\ll 1$.  Then
+$r=2M$ to $r=2M(1+w)$ where $w$ is small, $0<w\ll 1$.  Then
 
@@
 
-\noindent where dashes denote differentiation with respect to~$\phi$.
+\noindent where primes denote differentiation with respect to~$\phi$.
 The Euler-Lagrange equations for this system,
@@
 
-\begin{equation}\label{rdashdash}
+\begin{equation}\label{rprimeprime}
   r''\left(\phi\right) =
@@
 
-\noindent Solutions to equation~\ref{rdashdash} correspond to a taut
+\noindent Solutions to equation~\ref{rprimeprime} correspond to a taut
 string confined to Flamm's paraboloid~\cite{flamm1916}; see
 Figure~\ref{flamm}.  It is interesting to compare
-equation~\ref{rdashdash} with the corresponding equation for
+equation~\ref{rprimeprime} with the corresponding equation for
 equatorial null geodesics, which is usually given~\cite{wald} in terms
@@
 There does not appear to be a simple analytical solution to
-equation~\ref{rdashdash}; but it is a straightforward second order
+equation~\ref{rprimeprime}; but it is a straightforward second order
 ODE, amenable to numerical solution and no harder than the equivalent
@@
 
-Here I solve equation~\ref{rdashdash} numerically, using the R
+Here I solve equation~\ref{rprimeprime} numerically, using the R
 programming language~\cite{rcore2019} and the \verb+deSolve+ package
@@
 of non-self-intersecting numerical solutions to
-equation~\ref{rdashdash}, rotated so that the tangential point of
+equation~\ref{rprimeprime}, rotated so that the tangential point of
 closest approach occurs at~$\phi=0$.

### 2021-02-25 tpe_edreview_rebuttal.txt
mention "one-cm rulers laid end to end" in Taylor and Wheeler
fixes #41
fixes #43
fixes #33

@@
 > student's first encounter with General Relativity.
+
+I was unaware of the derivation in Taylor and Wheeler that the
+reviewer points me to.  I did not want to suggest that equation 2 was
+new, but was unable to find a derivation of it in the literature.  I
+have cited Taylor and Wheeler (but see my response to Q6 below).
+
 > 
@@
 mathematica screenshot, plus the identity arctanh(x) =
-log((1+x)/(1-x))/2.
+log((1+x)/(1-x))/2.  In the same section Taylor and Wheeler discuss
+one-centimeter long rulers laid end to end many times, a good metaphor
+for proper length.
 
@@
 >
+
+Fixed, now reads "consider the proper length from $r=2M$ to
+$r=2M(1+w)$ where $w$ is small, $0<w\ll 1$".
+
 > 9.  Question.  2nd line below Eq. (4), why are radial strings
@@
 > usage with which I am familiar.
+
+Fixed, notation now consistent
+
 > 

### 2021-02-25 string_physics_educator_revision3.tex
fixes #33

@@
 General relativity is a topic of long-standing fascination for physics
-undergraduates~\cite{christensen2012}.  Students appreciate everyday
+undergraduates~\cite{christensen2012,taylor2000}.  Students appreciate everyday
 illustrations of the concepts involved, as recent educational work on
@@
 circular string at constant~$r$ will, trivially, have proper
-length~$2\pi r$.  Considering first the radial component we may
+length~$2\pi r$.  Considering first the radial component we may follow Taylor and Wheeler \cite{taylor2000} and 
 calculate the proper length $\ell\left(r_1,r_2\right)$ of a string

### 2021-02-25 string_physics_educator_revision3.tex
tidyup
fixes #35

@@
 General relativity is a topic of long-standing fascination for physics
-undergraduates~\cite{christensen2012,taylor2000}.  Students appreciate everyday
-illustrations of the concepts involved, as recent educational work on
-gravitational wave visualisations attests~\cite{overduin2018}.  Here I
-interpret a straight line as the path adopted by a stationary, light
-inexensible string under tension passing between two points.  The
-analysis presented here would be suitable for an undergraduate
+undergraduates~\cite{christensen2012,taylor2000}.  Students appreciate
+everyday illustrations of the concepts involved, as recent educational
+work on gravitational wave visualisations attests~\cite{overduin2018}.
+Here I interpret a straight line as the path adopted by a stationary,
+light inexensible string under tension passing between two points.
+The analysis presented here would be suitable for an undergraduate
 homework problem~\cite{romano2019}.
@@
 (proper) unit length~$\mu$, then a string will behave as though light
-on lengthscales $L\lesssim T/\left(\mu g\right)$.  Using
+if the weight of the string does not exceed the tension, so that $L\mu g\lesssim T$.
+Using
 $c^4/\left(4GM\right)$ for the strength of gravity at the event
@@
 circular string at constant~$r$ will, trivially, have proper
-length~$2\pi r$.  Considering first the radial component we may follow Taylor and Wheeler \cite{taylor2000} and 
-calculate the proper length $\ell\left(r_1,r_2\right)$ of a string
-stretching from~$r=r_1\geqslant 2M$ to~$r_2\geqslant r_1$ at constant
-$\theta,\phi$:
+length~$2\pi r$.  Considering first the radial component we may follow
+Taylor and Wheeler \cite{taylor2000} and calculate the proper length
+$\ell\left(r_1,r_2\right)$ of a string stretching from~$r=r_1\geqslant
+2M$ to~$r_2\geqslant r_1$ at constant $\theta,\phi$:
 

### 2021-02-25 tpe_edreview_rebuttal.txt
start work on #42

@@
 >
+
+In mathematical usage (but not in general usage, at least according to
+the OED) there is a disctinction in mathematics between arbitrary
+arbitrarily.  One usually uses phrases like "arbitrarily large" to
+convey the idea that some object may become as large as one desires.
+
+In the manuscript the wording was "radial strings are arbitrarily
+longer than flat space would imply", which I think gets the intended
+meaning across.
+
+In this case we have the astonishing result that the discrepancy
+approaches infinity ("becomes arbitrarily large") as w approaches
+infinity.  I have not seen this result explictly before, yet am
+reluctant to claim that it is truly new: after all it is a simple
+consequence of a known result (equation 3).
+
+In any case the original wording was somewhat terse and I have
+rephrased.  The sentence now reads:
+
 > 10. Suggestion.  Line below Eq. (5), I think the word "dashes" should

### 2021-02-25 string_physics_educator_revision3.tex
fixes #36

--- 2021-02-25 string_physics_educator_revision3.tex
@@
 With tension~$T$, gravitational acceleration $g$ and (proper) mass per
-(proper) unit length~$\mu$, then a string will behave as though light
-on lengthscales $L\lesssim T/\left(\mu g\right)$.  Using
-$c^4/\left(4GM\right)$ for the strength of gravity at the event
-horizon of a black hole with mass $M$ gives $L\lesssim TM/\left(\mu
-c^4\right)$.  For the string to behave as though light on lengthscales
-comparable with the Schwarzschild radius would require $T\gtrsim\mu
-c^2$.  Exceeding this criterion would violate the dominant energy
-condition (and in this context we note that transverse waves would
-propagate at speed~$\sqrt{T/\mu}\sim c$).  Nevertheless, the
-idealization of a string as light and inextensible is an interesting
-and informative limit: Rindler~\cite{rindler}, for example, considers
-lowering a heavy object to the event horizon on the end of such a
-string to illustrate energetic arguments in GR.
+(proper) unit length~$\mu$, then a string of length $\ell$ will behave
+as though light if the weight of the string does not exceed the
+tension, so that $T\gtrsim\ell\mu g$.  At the Schwarzschild radius
+$2GM/c^2$ of a black hole with mass $M$, the Newtonian value
+$g=GM/r^2$ gives $T\gtrsim\ell\mu c^4/M$.  For the string to behave as
+though light on lengthscales comparable with the Schwarzschild radius
+(as in the simulations below), we would require $T\gtrsim\mu c^2$.
+Exceeding this criterion would violate the dominant energy condition
+(and in this context we note that transverse waves would propagate at
+speed~$\sqrt{T/\mu}\sim c$).  Nevertheless, the idealization of a
+string as light and inextensible is an interesting and informative
+limit: Rindler~\cite{rindler}, for example, considers lowering a heavy
+object to the event horizon on the end of such a string to illustrate
+energetic arguments in GR.
 

### 2021-02-25 tpe_edreview_rebuttal.txt
rebuttal to comment, issue #35
fixes #36
fixes #37

@@
 >
+
+This is a much better way of phrasing the argument and I have
+rephrased as suggested.
+
 > 3.  Suggestion.  Stating "$c^4/4GM$ for the strength of gravity at the
@@
 >
+
+I have totally rephrased that part of the manuscript using this
+suggestion and it is much clearer (I have also changed uppercase L for
+lowercase l as per suggestion number 7.
+
 > 4.  Question. Lines 3-4, page 2.  Can the author explain what is meant
 > by "the dominant energy condition"?
+
+On reflection the dominant energy condition is not really an
+appropriate or informative or useful or interesting concept for the
+intended readership.  The important thing is that tension greater $\mu
+c^2$ is unphysical and this context, the unphysicality is manifested
+by transverse waves travelling faster than light.
+
+In answer to the question, we may take the stress-energy tensor for a
+stretched string to be diagonal with elements $\mu,-T,0,0$.  Then the
+dominant energy condition is that, for any future-pointing causal
+vector, $-T^\mu_\nu V^\nu$ must also be a future-pointing causal
+vector.  We can see directly that if $T>\mu$ the DEC is violated.
+
+But this is a massive digression to the main thrust of the manuscript
+and any adequate pedagogical explanation would be an intrusive
+distraction.  It is like using a slegehammer to crack a nut.  I have
+removed any reference to the DEC on these grounds, and further
+emphasised the unphysicality of the wave speed being $>c$.
+
 > 

### 2021-02-25 tpe_edreview_rebuttal.txt
fixes #34

@@
 > free-fall in the global coordinate system.
+
+I think the sentence in question was technically correct but
+potentially misleading.  What I wanted to do was to make readers think
+about taut thread in the context of GR.  I didn't want to consider
+geodesics at this point in the introduction, or free-falling elevator
+frames.  I want to focus on taut string.
+
+[one of the conceptual difficulties in preparing the manuscript, as
+pointed out by a previous reviewer, is that the string is not in a
+freely-falling elevator frame; as the reviewer points out, such string
+does not follow a geodesic path in any meaningful sense].
+
+I have rephrased the sentence which now reads: "Here I consider the
+path adopted by a stationary, light inexensible string under tension
+passing between two points, and consider the extent to which the
+expression "straight line" would be applicable to such an object".
+
 > 

### 2021-02-25 string_physics_educator_revision3.tex
fixes #37
fixes #34

@@
 work on gravitational wave visualisations attests~\cite{overduin2018}.
-Here I interpret a straight line as the path adopted by a stationary,
-light inexensible string under tension passing between two points.
+Here I consider the path adopted by a stationary,
+light inexensible string under tension passing between two points, and consider the extent to which the expression  ``straight line'' would be applicable to such an object.
 The analysis presented here would be suitable for an undergraduate
@@
 (as in the simulations below), we would require $T\gtrsim\mu c^2$.
-Exceeding this criterion would violate the dominant energy condition
-(and in this context we note that transverse waves would propagate at
-speed~$\sqrt{T/\mu}\sim c$).  Nevertheless, the idealization of a
+Exceeding this criterion is unphysical:
+transverse waves propagate at
+speed~$\sqrt{T/\mu}$ which would exceed $c$.  Nevertheless, the idealization of a
 string as light and inextensible is an interesting and informative

### 2021-02-25 tpe_edreview_rebuttal.txt
the points are fixed (issue #34)

@@
 path adopted by a stationary, light inexensible string under tension
-passing between two points, and consider the extent to which the
+passing between two fixed points, and consider the extent to which the
 expression "straight line" would be applicable to such an object".

### 2021-02-25 string_physics_educator_revision3.tex
the points are fixed (issue #34)

@@
 Here I consider the path adopted by a stationary,
-light inexensible string under tension passing between two points, and consider the extent to which the expression  ``straight line'' would be applicable to such an object.
+light inexensible string under tension passing between two fixed points, and consider the extent to which the expression  ``straight line'' would be applicable to such an object.
 The analysis presented here would be suitable for an undergraduate

### 2021-02-26 tpe_edreview_rebuttal.txt
start work on issue #38

@@
 >
+
+The referee is absolutely correct here and raises an important point.
+When I said that the circumference was "trivially" $2\pi r$, what I
+meant was that integrating equation (1) from $phi=0$ to $2\pi$ at
+fixed $t,r,\theta$ was trivial.  This is a direct consequence of the
+*definition* of "r" [which is sometimes called the circumferential
+radial coordinate].  I remember struggling with this myself as an
+undergraduate (I wanted to know what the "real" radius was ...).
+Einstein himself said "it is not so easy to free oneself from the idea
+that coordinates must have an immediate metrical meaning", so I have
+good company.
+
+On reflection this point in the manuscript represents a good
+"teachable moment".
+
 > 6.  Question.  There may be a typographical error in Eq. (2).

### 2021-02-26 tpe_edreview_rebuttal.txt
fixes #38

--- 2021-02-25 tpe_edreview_rebuttal.txt
@@
 >
+
+The referee is absolutely correct here and raises an important point.
+When I said that the circumference was "trivially" $2\pi r$, what I
+meant was that integrating equation (1) from $phi=0$ to $2\pi$ at
+fixed $t,r,\theta$ was trivial.  The resulting proper string length is
+a direct consequence of the *definition* of "r" [which is sometimes
+called the circumferential radial coordinate].
+
+This comment has made me realise that the $2\pi r$ proper string
+length is actually the simplest proper length calculation possible in
+the Schwarzschild metric, and I have rephrased accordingly, and I hope
+it is more instructive.  I remember struggling with this myself as an
+undergraduate (I wanted to know what the "real" radius was ...).
+Einstein himself said "it is not so easy to free oneself from the idea
+that coordinates must have an immediate metrical meaning", so I have
+good company!
+
 > 6.  Question.  There may be a typographical error in Eq. (2).

### 2021-02-26 string_physics_educator_revision3.tex
fixes #38

@@
 dr^2/\left(1-2M/r\right) + r^2d\phi^2$, confined to equatorial
-points~($\theta=\pi/2$); we understand~$r\geqslant 2M$ throughout.  A
-circular string at constant~$r$ will, trivially, have proper
-length~$2\pi r$.  Considering first the radial component we may follow
-Taylor and Wheeler \cite{taylor2000} and calculate the proper length
+points~($\theta=\pi/2$); we understand~$r\geqslant 2M$ throughout.
+
+The simplest configuration for a string would be circular at
+constant~$r$ whose proper length would be given by integration of
+equation~\ref{schwarzschild} as~$\int_{\phi=0}^{2\pi}r\,d\phi=2\pi r$
+(this is why $r$ is sometimes called ``circumferential radius'').
+Note carefully that $r$ is not the distance {\em to} anything, so this
+result---which operates in Schwarzschild geometry and not plane
+Euclidean geometry---is less obvious than it might appear.
+Considering next a purely radial string we may follow Taylor and
+Wheeler \cite{taylor2000} and calculate the proper length
 $\ell\left(r_1,r_2\right)$ of a string stretching from~$r=r_1\geqslant

### 2021-02-26 tpe_edreview_rebuttal.txt
fixes #42

@@
 In mathematical usage (but not in general usage, at least according to
-the OED) there is a disctinction in mathematics between arbitrary
-arbitrarily.  One usually uses phrases like "arbitrarily large" to
-convey the idea that some object may become as large as one desires.
+the OED) there is a distinction between "arbitrary" and "arbitrarily".
+One usually uses phrases like "arbitrarily large" to convey the idea
+that some object may become as large as one desires.
 
@@
 
-In this case we have the astonishing result that the discrepancy
+In this case we have the startling result that the discrepancy
 approaches infinity ("becomes arbitrarily large") as w approaches
@@
 reluctant to claim that it is truly new: after all it is a simple
-consequence of a known result (equation 3).
+consequence of a known result (equation 3).  But it is certainly
+startling, to me at least.
 
 In any case the original wording was somewhat terse and I have
-rephrased.  The sentence now reads:
+rephrased.  The manuscript now explicitly states that the discrepancy
+in proper length between the Schwarzschild and Euclidean results
+approaches infinity as w approaches infinity.
 

### 2021-02-26 string_physics_educator_revision3.tex
fixes #42

@@
 
-\noindent The discrepancy between the flat-space result of $2Mw$ and
-equation~\ref{asymptotic_ell} suggests that, even far from the event
+\noindent The discrepancy $\Delta=\ell-2Mw$ between the flat-space result of $2Mw$ and
+equation~\ref{asymptotic_ell} has the startling feature that $\lim_{w\longrightarrow\infty}\Delta=\infty$ (because of the $\log w$ term).  This suggests that, even far from the event
 horizon, radial strings are arbitrarily longer than flat space would

### 2021-02-26 tpe_edreview_rebuttal.txt
fixes #44

@@
 >
+
+The equation given as $u''=3u^2/2-u$ gives the path of a beam of light
+through Schwarzschild space ideally in terms of $r$ as a function of
+angle, $r=r(\phi)$.  The idea is that a stationary observer has a
+laser pointer pointing in a particular direction, switches it on and
+leaves it on for a long time.  We then record all the points that the
+beam touches.  The path of the light is then simply $r=r(\phi)$, no
+time coordinate involved, as Figures 4 and 6 show.  I have changed the
+citation to Rindler, who gives a detailed derivation.
+
 > 12.  In the passage below Eq. (6), I am having trouble making the

### 2021-02-26 string_physics_educator_revision3.tex
fixes #44

@@
 equation~\ref{rprimeprime} with the corresponding equation for
-equatorial null geodesics, which is usually given~\cite{wald} in terms
-of $u=2M/r$ as $u''=3u^2/2-u$, but is equivalent
+equatorial null geodesics, which is usually given~\cite{rindler} in
+terms of $u=2M/r$ as $u''=3u^2/2-u$, but is equivalent
 to~$r''\left(\phi\right)=r-3M+2\left(r'\right)^2/r$.  This shows that

### 2021-02-27 tpe_edreview_rebuttal.txt
fixes #45

@@
 >
+
+There are two separate issues here.  Firstly, null geodesic paths in
+the Schwarzschild metric obey $u''=3u^2/2-u$ where $u=2M/r$.  This is
+equivalent to $r''(\phi) = r-3M + 2(r')^2/r$.  The second issue is
+straight lines in Euclidean flat space (in flat space the expression
+"straight lne" is unambiguous).  WLOG $\phi_0=0$ and if
+$r=r_0\sec\phi$ then $r'=-r_0\sin\phi\cdot\sec^2\phi$ and
+$r''=r_0(\sec\phi + 2\sin^2\phi\sec^3\phi$, from which $r''=2(r')^2/r$
+follows from cancellation.
+
+One can "go the other way"---that is, given ODE $r''=r+2(r')^2/r$, one
+can deriving $r=r_0\sec\phi$---by substituting $u=1/r$ and obtaining
+$u''+u=0$, whence $u=A\cos(\phi-\phi_0)$.
+
+I have rephrased slightly and numbered one of the equations so I can
+refer to it.
+
 > 13.  Can the author explain more fully what is meant by a "tangential
@@
 > tangential null geodesics.  What am I not seeing?
+
 > 

### 2021-02-27 string_physics_educator_revision3.tex
fixes #45

--- 2021-02-26 string_physics_educator_revision3.tex
@@
 equation~\ref{rprimeprime} with the corresponding equation for
-equatorial null geodesics, which is usually given~\cite{wald} in terms
-of $u=2M/r$ as $u''=3u^2/2-u$, but is equivalent
-to~$r''\left(\phi\right)=r-3M+2\left(r'\right)^2/r$.  This shows that
+equatorial null geodesics, which is usually given~\cite{rindler} in
+terms of $u=2M/r$ as $u''=3u^2/2-u$, but is equivalent
+to
+
+\begin{equation}\label{nullgeodesicODE}
+  r''\left(\phi\right)=r-3M+2\left(r'\right)^2/r.
+\end{equation}
+
+This shows that
 light, being sensitive to the time component of the metric, behaves
-differently from string, which is not (at this point it might be worth
-``reminding'' students that a straight line in flat space,
+differently from string, which is not.  At this point it might be worth
+observing that a straight line in flat space,
 $r=r_0\sec\left(\phi-\phi_0\right)$, obeys
-$r''=r+2\left(r'\right)^2/r$).
+$r''=r+2\left(r'\right)^2/r$: identical to equation~\ref{nullgeodesicODE} but without the $M$ term.
 

### 2021-02-27 tpe_edreview_rebuttal.txt
fixes #46

@@
 
+Everything can be resolved and clarified by omitting the word
+"tangential" which was misleading, and replacing "small" with
+"negligible".  The sentence now reads "If $r'$ is negligible, then a
+string follows $r''=r-2M$, while null geodesics follow $r''=r-3M$".
+
+The $r''=r-3M$ is just the (new) equation 7 -viz
+$r''\left(\phi\right)=r-3M+2\left(r'\right)^2/r$ but with $r'=0$.
+
 > 

### 2021-02-28 tpe_edreview_rebuttal.txt
fixes #47
further minor rewording

@@
 Short story: I agree with all the comments made and have modified the
-manuscript accordingly; changes appear in the manuscript using colour
-coding for ease of reference.  The document is stronger, more
-cohesive, more accurate, and hopefully more instructive; and I commend
-it to you.
+manuscript accordingly.  The document is stronger, more cohesive, more
+accurate, and hopefully more instructive.  I submit that the extra
+clarity and pedagogical value more than compensates for the slight
+extra length.
+
+I have changed the discussion slightly (emphasising the
+staticness of the metric considered) as a result of reflecting on the
+totality of the comments too.
 
@@
 The equation given as $u''=3u^2/2-u$ gives the path of a beam of light
-through Schwarzschild space ideally in terms of $r$ as a function of
+through Schwarzschild space in which $r$ is given as a function of
 angle, $r=r(\phi)$.  The idea is that a stationary observer has a
@@
 >
+
+Not any more! All the software used to create the diagrams is now
+publically available under version control at
+
+https://github.com/RobinHankin/string
+
+and I have added a citation to this repo.  The code is heavily
+documented.
+
 > This is an interesting paper.  I congratulate the author for thinking

### 2021-02-28 string_physics_educator_revision3.tex
fixes #46
fixes #47
further minor rewording

@@
 interpretation of the concept ``straight line'', the other two being
-null geodesics and free-fall world lines.
+null geodesics and free-fall world lines.  All the software used is
+available under the GPL~\cite{hankin2021}.
 \end{abstract}
@@
 The simplest configuration for a string would be circular at
-constant~$r$ whose proper length would be given by integration of
+constant~$r$ whose proper length is given by integration of
 equation~\ref{schwarzschild} as~$\int_{\phi=0}^{2\pi}r\,d\phi=2\pi r$
@@
 the system is invariant under~$\phi\longrightarrow-\phi$, the string
-must be symmetrical about this point.  If $r'$ is small, then a string
-follows $r''=r-2M$, while tangential null geodesics follow $r''=r-3M$.
+must be symmetrical about this point.  If $r'$ is negligible, then a string
+follows $r''=r-2M$, while null geodesics follow $r''=r-3M$.
 Working to first order, and writing~$d$ for the difference in
@@
 programming language~\cite{rcore2019} and the \verb+deSolve+ package
-for ordinary differential equations~\cite{soetart2010}.  First we
-consider strings for which $r>2M$ everywhere (that is, completely
-exterior strings).
-Figure~\ref{closest_approach_non_self_intersecting} shows a sequence
-of non-self-intersecting numerical solutions to
+for ordinary differential equations~\cite{soetart2010}.  We consider
+only strings for which $r>2M$ everywhere (that is, completely exterior
+strings).  Figure~\ref{closest_approach_non_self_intersecting} shows a
+sequence of non-self-intersecting numerical solutions to
 equation~\ref{rprimeprime}, rotated so that the tangential point of
@@
     at~$\phi=0$.  Note the differences between these curves and the taut
-    strings shown elsewhere: unlike taut strings, null geodesices may
+    strings shown elsewhere: unlike taut strings, null geodesics may
     cross the event horizon inwards, and are never tangential to the
@@
 world lines are sensitive to the time component of the metric.  Light
-taut strings respond only to the spatial components of the metric, and
-their configuration is arguably a more intuitive and helpful way to
-think about straightness.  Using numerical methods to compute string
-configuration is suitable for an undergraduate homework problem.
+taut strings respond only to the spatial components of a static
+metric, and comparing their configuration to null geodesics is a
+helpful way to think about straightness.  Using numerical methods to
+compute string configuration is suitable for an undergraduate homework
+problem.
 

### 2021-02-28 tpe_edreview_rebuttal.txt
transcript from sage to support factor of M rather than 2M as in T&W

@@
 
-Factor of M now added (actually this caused me some considerable
+Factor of M now added...actually this caused me some considerable
 confusion and uncertainty.  The equation presented in Taylor and
-Wheeler has a factor of 2M but it should be M---see the attached
-mathematica screenshot, plus the identity arctanh(x) =
-log((1+x)/(1-x))/2.  In the same section Taylor and Wheeler discuss
-one-centimeter long rulers laid end to end many times, a good metaphor
-for proper length.
+Wheeler has a factor of 2M but it should be M, for which I offer the
+following sage transcript:
+
+sage: var('r M r1 r2') 
+(r, M, r1, r2)
+sage: assume(r>0,r1>0,r2>0, r1>2*M, r2>2*M, r2-r1>0, M>0) 
+sage: integrate(1/sqrt(1-2*M/r),r) 
+M*log(sqrt(-2*M/r + 1) + 1) - M*log(sqrt(-2*M/r + 1) - 1) + r*sqrt(-2*M/r + 1)
+
+In the same section Taylor and Wheeler discuss one-centimeter long
+rulers laid end to end many times, a good metaphor for proper length.
 

### 2021-02-28 tpe_edreview_rebuttal.txt
slight rewording

@@
 
-I have changed the discussion slightly (emphasising the
-staticness of the metric considered) as a result of reflecting on the
-totality of the comments too.
+I have changed the discussion section at the end very slightly
+(emphasising the staticness of the metric considered) as a result of
+reflecting on the totality of the comments too.
 
@@
 suggestion and it is much clearer (I have also changed uppercase L for
-lowercase l as per suggestion number 7.
+lowercase l as per suggestion number 7).
 
@@
 
-The equation given as $u''=3u^2/2-u$ gives the path of a beam of light
-through Schwarzschild space in which $r$ is given as a function of
-angle, $r=r(\phi)$.  The idea is that a stationary observer has a
-laser pointer pointing in a particular direction, switches it on and
-leaves it on for a long time.  We then record all the points that the
-beam touches.  The path of the light is then simply $r=r(\phi)$, no
-time coordinate involved, as Figures 4 and 6 show.  I have changed the
-citation to Rindler, who gives a detailed derivation.
+The equation given as $u''=3u^2/2-u$ [here $u=2M/r$] gives the path of
+a beam of light through Schwarzschild space in which $r$ is given as a
+function of angle, $r=r(\phi)$.  The idea is that a stationary
+observer has a laser pointer pointing in a particular direction,
+switches it on and leaves it on for a long time.  We then record all
+the points that the beam touches.  The path of the light is then
+simply $r=r(\phi)$, no time coordinate involved, as Figures 4 and 6
+show.  I have changed the citation to Rindler, who gives a detailed
+derivation.
 
@@
 straight lines in Euclidean flat space (in flat space the expression
-"straight lne" is unambiguous).  WLOG $\phi_0=0$ and if
+"straight line" is unambiguous).  WLOG $\phi_0=0$ and if
 $r=r_0\sec\phi$ then $r'=-r_0\sin\phi\cdot\sec^2\phi$ and
-$r''=r_0(\sec\phi + 2\sin^2\phi\sec^3\phi$, from which $r''=2(r')^2/r$
-follows from cancellation.
+$r''=r_0(\sec\phi + 2\sin^2\phi\sec^3\phi$, from which
+$r''=r+2(r')^2/r$ follows directly.
 

### 2021-03-11 tpe_edreview2_rebuttal.tex
start work on further rebuttal

\documentclass[12pt]{article}
\usepackage{xcolor}
\begin{document}

I appreciate the author responding to each of my previous questions
and suggestions. For the most part I am satisfied with the author’s
responses and subsequent modifications to the manuscript. While it is
not my purpose to annoy the author or delay an editorial decision on
publication, a small number of comments about minor points remain:

1. The discussion about circumference and $2\pi r$ may be, to the
undergraduate reader to whom the Schwarzschild metric is still quite
new, a source of confusion regarding the logic behind the metric. In
the revised manuscript we read of ``constant $r$ whose proper length
is given by integration," where the integral is

\[
\int_{0}^{2\pi} r\,d\phi = 2\pi r
\]

While the integration is certainly true, it this context it appears to
express a tautology because the $r$- coordinate is defined as the
circumference of a shell (centered on the origin) divided by $2\pi$.
The passage just quoted makes it sound (at least to me) like
``circumference = $2\pi r$ is a derived result, when in fact it the
logic goes the other way---the measured circumference provides the
definition of $r$ (see Taylor \& Wheeler, Black Holes, page 2-7, which
in T\&W pagination means page 7 in Ch. 2; T\&W is a popular
entry-level textbook for undergraduates interested in general
relativity).

2. The author's explanation registers with more clarity now in the
passage about string being ``arbitrarily" large.  At first reading
this sounds like the author says that the length of a tight string
fixed between two points may be arbitrary, notwithstanding the
distinction between ``arbitrary" and ``arbitrarily."  Here is what I
now understand the author to be saying about the value of l in flat
and in Schwarzschild spacetimes:

\[
\ell_\mathrm{flat} = 2Mw
\]

and, asymptotically,

\[
\ell_\mathrm{Sch}\sim 2Mw + \log w
\]

and thus their difference is

\[
\ell_\mathrm{Sch} - \ell_\mathrm{flat}\sim M\log w
\]

which goes to infinity logarithmically; in other words, for an
infinite string the Schwarzschild metric is a ``worse infinity" which
diverges faster than what occurs in flat spacetime.  So far, so good.
I guess what bothers me is the statement that ``even far from [emphasis
  added] the event horizon, radial strings are arbitrarily longer than
flat space would imply...".  So long as $w$ is large (however far from
the event horizon) but finite, the lengths in both metrics are precise
numbers, not arbitrary.  Only when $w$ goes to infinity does the
length in Schwarzschild spacetime become infinitely longer than the
length in flat spacetime, although though both are infinite.  I
suppose this passage would sit better with this reader by merely
stating that the asymptotic difference between $\ell_\mathrm{Sch}$ and
$\ell_\mathrm{flat}$ diverges logarithmically.  But perhaps this is
only a matter of taste.

3. I still must disagree with the author about the coefficient of the
logarithm term in Eq. (2).  The original manuscript put this
coefficient at $1/2$; I argued that it is $2M$ because of the
corresponding equation in T\&W (their Eq. [27] on p. 2-28).  I also
find disagreement with Eq. (2) on a sign in the denominator of the log
term.  Standing ready to be corrected, please allow me to show why
this reader thinks these discrepancies exist.  We are to evaluate

\[
\ell=\int_{r_1}^{r_2}\frac{dr}{\sqrt{1-2M/r}} =
\int_{r_1}^{r_2}\frac{\sqrt{r}\,dr}{\sqrt{r-2M}},
\]

Let $r=z^2$ so that $dr=2z\,dz$, and let $2M=a^2$.  Then

\[
\ell=2\int_{r_1}^{r_2} \frac{z^2\,dz}{\sqrt{z^2-a^2}}.
\]

 Using the integral table in the CRC handbook, integral No. 171,

\[
\int\frac{dx}{\sqrt{x^2\pm a^2}}=\frac{x}{2} \sqrt{z^2\pm a^2}
\mp \frac{a^2}{2}\log\left| z+\sqrt{x^2\pm a^2}\right|
\]

 and with the minus sign under the radical in the integrand, I obtain

\[
\ell = 2\left[
  \frac{z}{2}\sqrt{z^2-a^2} + \frac{a^2}{2}\log\left|z+\sqrt{z^2-a^2}\right|
  \right]_{z_1}^{z_2}
\]

\[
=
\sqrt{r_2(r_2-2M)}-\sqrt{r_1(r_1-2M)} + 2M\log\left|
\frac{\sqrt{r_2}+\sqrt{r_2-2M}}{\sqrt{r_1}+\sqrt{r_1-2M}}\right|
\]

In contrast, Eq (2) reads (points of disagreement in red):

\[
\sqrt{r_2(r_2-2M)}-\sqrt{r_1(r_1-2M)} 
+\color{red}M
\color{black}\log\left|
\frac{\sqrt{r_2}+\sqrt{r_2-2M}}{\sqrt{r_1}
  \color{red}-
  \color{black}\sqrt{r_1-2M}}\right|
\]

Other than these minor points, I have no further concerns about the
paper. I find the project very interesting, offering much to think
about. Having worked with undergraduate physics majors who wish to
learn some general relativity, once such students are made familiar
with the Schwarzschild metric, this exercise with a tight string
offers a useful tool. I leave it to the editors and authors to decide
what to do with these my latest suggestions regarding the revised
manuscript. I thank the author for taking my previous suggestions and
concerns seriously.

\end{document}

### 2021-03-12 tpe_edreview2_rebuttal.tex
work on circumferential radius

@@
 relativity).
+
+\textcolor{blue}{This is a perfectly reasonable comment and 
+  I have changed the manuscript.\\
+{\bf Old text:}
+ ...(this is why $r$ is sometimes called ``circumferential radius'').\\
+{\bf New text:} ...(this is the motivation for the definition of $r$;
+indeed, it is  sometimes called ``circumferential radius'' for this
+reason).
+}
 
@@
 
+\textcolor{blue}{ This is a subtle point which took me some
+  considerable time to grok.
+}
+
 3. I still must disagree with the author about the coefficient of the
@@
 
+{\color{blue} The RHS of the equation above is quoted correctly, but
+  the left hand side should be $\int\frac{x^2\,dx}{\sqrt{x^2\pm
+      a^2}}$, not $\int\frac{dx}{\sqrt{x^2\pm a^2}}$ as quoted [this
+    is CRC equation 164 in my edition].  }
+
  and with the minus sign under the radical in the integrand, I obtain
@@
 
+{\color{blue} The reviewer is correct here but I would observe that
+  the expression above (with or without the red alternative) is not
+  equation 2 of the manuscript.  The reviewer's expression actually
+  matches the RHS of Equation 2 in the manuscript which is
+
+  \[
+  \left.
+  \sqrt{r(r-2M)} +M\log\left(
+  \frac{\sqrt{r}+\sqrt{r-2M}}{\sqrt{r}-\sqrt{r-2M}}\right)
+  \right|_{r_1}^{r_2}
+  \]
+
+  \[
+  = 
+  \sqrt{r_2(r_2-2M)}-  \sqrt{r_1(r_1-2M)}
+ +M\log\left(
+ \frac{\sqrt{r_2}+\sqrt{r_2-2M}}{\sqrt{r_2}-\sqrt{r_2-2M}}
+ \frac{\sqrt{r_1}+\sqrt{r_1-2M}}{\sqrt{r_1}-\sqrt{r_1-2M}}
+ \right)
+  \]
+
+The match is not obvious: one needs to manipulate logarithms of
+negative numbers and keep track of two cut lines on the complex plane
+which manifest as (imaginary) constants of integration.  The apparent
+factor of 2 resolves itself because multiplying numerator and
+denominator by the appropriate conjugate surd results in the {\em
+  square} of the expression given by the referee.
+
+Arguably the referee's formulation is neater and more compact than
+that appearing in the manuscript but I would support my own version on
+the grounds that it is closer to mathematica's output (and indeed is
+much more amenable to asymptotic analysis).
+
+{\bf Summary: the reviewer's algebra was correct but our expressions
+  in fact agree.}
+}
+
 Other than these minor points, I have no further concerns about the

### 2021-03-12 string_physics_educator_revision4.tex
work on circumferential radius

--- 2021-02-28 string_physics_educator_revision3.tex
@@
 equation~\ref{schwarzschild} as~$\int_{\phi=0}^{2\pi}r\,d\phi=2\pi r$
-(this is why $r$ is sometimes called ``circumferential radius'').
-Note carefully that $r$ is not the distance {\em to} anything, so this
+(this is the motivation for the definition of $r$; indeed, it is
+sometimes called ``circumferential radius'' for this reason).  Note
+carefully that $r$ is not the distance {\em to} anything, so this
 result---which operates in Schwarzschild geometry and not plane

### 2021-03-12 tpe_edreview2_rebuttal.tex
work on another comment

@@
 
-\textcolor{blue}{ This is a subtle point which took me some
-  considerable time to grok.
+\textcolor{blue}{This is a subtle point which took me some
+  considerable time to understand.  I would express the problem with
+  the existing text as that the radial strings concerned stretch all
+  the way to the event horizon; yet the phrase ``far from the event
+  horizon'' kind of suggests that {\em all} the string is far from the
+  event horizon.\\ It is possible to get round this informally by
+  considering the proper length of a radial string from $2M(1+w_1)$ to
+  $2M(1+w_2)$ (where we understand $2M\leq w_1\leq w_2$), which is seen from
+  equation 4 to be $2M(w_1-w_2 + \log\left(\frac{w_2}{w_1}\right) +
+  O(w_1^{-1})$.\\ But on balance this is a rather unspectacular result
+  which would require a distractingly large amount of explanation.  On
+  balance I think it best to leave this out, resulting, I hope in
+  sharper, more focussed wording.
+\\ {\bf old text}
+The discrepancy\ldots has the startling feature that
+$\lim_{w\longrightarrow\infty}\Delta=\infty$ (because of the $\log w$
+term).  This suggests that, even far from the event horizon, radial
+strings are arbitrarily longer than flat space would imply: a nice
+illustration of the failure of Euclidean geometry.
+\\ {\bf new text}
+The discrepancy\ldots has the startling feature that
+$\lim_{w\longrightarrow\infty}\Delta=\infty$ (because of the $\log w$
+term): a nice illustration of the failure of Euclidean geometry.
 }

### 2021-03-12 string_physics_educator_revision4.tex
work on another comment

@@
 \noindent The discrepancy $\Delta=\ell-2Mw$ between the flat-space result of $2Mw$ and
-equation~\ref{asymptotic_ell} has the startling feature that $\lim_{w\longrightarrow\infty}\Delta=\infty$ (because of the $\log w$ term).  This suggests that, even far from the event
-horizon, radial strings are arbitrarily longer than flat space would
-imply: a nice illustration of the failure of Euclidean geometry.
+equation~\ref{asymptotic_ell} has the startling feature that $\lim_{w\longrightarrow\infty}\Delta=\infty$ (because of the $\log w$ term): a nice illustration of the failure of Euclidean geometry.
 Again taking a solar mass black hole but with $w=2\times 10^7$

### 2021-03-12 tpe_edreview2_rebuttal.tex
slight rewording

@@
   $2M(1+w_2)$ (where we understand $2M\leq w_1\leq w_2$), which is seen from
-  equation 4 to be $2M(w_1-w_2 + \log\left(\frac{w_2}{w_1}\right) +
-  O(w_1^{-1})$.\\ But on balance this is a rather unspectacular result
+  equation 4 to be $2M\left[w_1-w_2 + \log\left(\frac{w_2}{w_1}\right) +
+  O(w_1^{-1})\right]$.\\ But on balance this is a rather unspectacular result
   which would require a distractingly large amount of explanation.  On
@@
 
-{\color{blue} The reviewer is correct here but I would observe that
-  the expression above (with or without the red alternative) is not
-  equation 2 of the manuscript.  The reviewer's expression actually
-  matches the RHS of Equation 2 in the manuscript which is
+{\color{blue} The reviewer's expression above agrees with my own
+  calculations, but I would observe that the expression above (with or
+  without the red alternative) is {\em not} equation 2 of the
+  manuscript.  The reviewer's expression actually matches the RHS of
+  Equation 2 in the manuscript which is
 

### 2021-03-12 tpe_edreview2_rebuttal.tex
more rephrasing

@@
 
+{\color{blue} The reviewer's expression above is correct}
+
 In contrast, Eq (2) reads (points of disagreement in red):
@@
 
-{\color{blue} The reviewer's expression above agrees with my own
-  calculations, but I would observe that the expression above (with or
-  without the red alternative) is {\em not} equation 2 of the
-  manuscript.  The reviewer's expression actually matches the RHS of
-  Equation 2 in the manuscript which is
+{\color{blue} The expression above (with or without the red
+  alternative) is {\em not} equation 2 of the manuscript. Equation 2
+  in the manuscript is:
 
@@
   \frac{\sqrt{r_2}+\sqrt{r_2-2M}}{\sqrt{r_2}-\sqrt{r_2-2M}}
- \frac{\sqrt{r_1}+\sqrt{r_1-2M}}{\sqrt{r_1}-\sqrt{r_1-2M}}
+  \frac{\sqrt{r_1}-\sqrt{r_1-2M}}{\sqrt{r_1}+\sqrt{r_1-2M}}
   \right)
@@
 
-The match is not obvious: one needs to manipulate logarithms of
-negative numbers and keep track of two cut lines on the complex plane
-which manifest as (imaginary) constants of integration.  The apparent
-factor of 2 resolves itself because multiplying numerator and
+  \[
+  = 
+  \sqrt{r_2(r_2-2M)}-  \sqrt{r_1(r_1-2M)}
+  +M\log\left(
+  \frac{
+    \left(\sqrt{r_2}+\sqrt{r_2-2M}\right)^2
+  }{
+    \left(\sqrt{r_1}+\sqrt{r_1-2M}\right)^2
+  }
+  \right)
+  \]
+
+  \[
+  = 
+  \sqrt{r_2(r_2-2M)}-  \sqrt{r_1(r_1-2M)}
+  +2M\log\left(
+  \frac{\sqrt{r_2}+\sqrt{r_2-2M}}{\sqrt{r_1}+\sqrt{r_1-2M}}
+  \right)
+  \]
+
+We see that my original expression matches that of the reviewer.  The
+apparent factor of 2 resolves itself because multiplying numerator and
 denominator by the appropriate conjugate surd results in the {\em
@@
 
-{\bf Summary: the reviewer's algebra was correct but our expressions
-  in fact agree.}
-}
+{\bf Summary: both my, and the reviewer's, algebra was correct and our
+  expressions in fact agree.}  }
 

### 2021-03-14 tpe_edreview2_rebuttal.tex
description of file's format

@@
 \begin{document}
+
+\section*{Rebuttal to second editorial review: overall response}
+
+(Below, the reviewer's comments are in black, and my replies to the
+issues are in \textcolor{blue}{blue} [the reviewer used
+  \textcolor{red}{red} to highlight differences between our
+  calculations].  Short story: I have accommodated the review's minor
+comments with slight rewording; and resolved the algebraic discrepancy
+by proving that our two expressions are in fact equal).
+
+\section*{Detailed rebuttal}
 

### 2021-04-16 usefulnullgeodesicfuncs.R
typo

@@ -24,7 +24,7 @@ source("mercury.R")   # defines M=2e30, the mass of the sun, orbital
 `nullgeodesic` <- function(r_start, b, tau=NULL, SI=FALSE){
 
   if(is.null(tau)){
-        tau<- seq(from=0,to=100.1,len=100)
+        tau <- seq(from=0,to=100.1,len=100)
   } 
 
   yini <- c(

### 2021-04-16 shapiro.Rmd
start work on #21

---
title: "Shapiro delay"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
source("mercury.R")
source("usefulfuncs.R")
source("usefulgeodesicfuncs.R")
```

## Summary

The Shapiro delay is one of the classical tests of general relativity.
Here I attempt to reproduce that analysis using numerical reasoning
but fail because of numerical roundoff error.

## Introduction

Shapiro delay blah blah blah

First some setups:
```{r}

source("usefulnullgeodesicfuncs.R")

r0 <- r_sun/d # grazing incidence of the sun in Schwarzschild radii
r0
r_s
```

Thus the radius of the sun is about 471000 Schwarzschild radii.

```{r}
b_radial <- r0/sqrt(1-1/r0)

small <- 1e-6  # needed to prevent numerical problems

d_earth <-
seq(from=orbital_radius_earth/r_s * 0.999, to=orbital_radius_earth/r_s * 1.001,len=100)

d_venus <- 
seq(from=orbital_radius_venus/r_s * 0.999, to=orbital_radius_venus/r_s * 1.001,len=100)

d_earth <- c(1:10,d_earth)
```

Now calculate a null geodesic starting from the sun and ending at earth's orbit:

```{r}
out_earth <- nullgeodesic(r_start=r0,b=b_radial*(1-small),tau=c(0,d_earth))
head(out_earth)
tail(out_earth)
```

(function `nullgeodesic()` is a general-purpose solver for the path
light takes in the Schwarzschild metric; see file
`light_start_at_r_equals_2.R` for a more visually appealing simulation
than presented here, and `light_start_at_radius_of_sun.R` for a
discussion of the bending of starlight by the sun during an eclipse,
as (e.g.) measured by Eddington in 1919).  Above, see how the path
starts at $t=0$ at a location given by the radius of the sun (the `r`
column) and angle $\phi=0$ (the `phi` column).  We can perform the
same calculation but for Venus"

```{r}
out_venus <- nullgeodesic(r_start=r0,b=b_radial*(1-small),tau=c(0,d_venus))
```

The next step is to approximate the funtion

```{r}
f_earth <- approxfun(out_earth$r,out_earth$time)
f_venus <- approxfun(out_venus$r,out_venus$time)
```

So now `f_earth(t)` gives the time required to achieve a particular
radius.  Thus to calculate the transit time from Earth to Venus we
simply add the times to get to the Earth's radius to Venus's:

```{r}
transit_time_relativistic  <- 
(f_earth(orbital_radius_earth/r_s) + f_venus(orbital_radius_venus/r_s))*r_s/sol

transit_time_relativistic  # transit time in seconds
```

```{r}

## Now calculate the classical time, simply distance/speed:
transit_time_classical <- (orbital_radius_earth + orbital_radius_venus)/sol # seconds

## plot (rather dull):
st <- function(x){(x-min(x))/(max(x)-min(x))}
plot(out_earth$x,out_earth$y,type='o',col='red',cex=st(out_earth$time),asp=1)
event_horizon()

shapiro_delay <- transit_time_relativistic - transit_time_classical
# compare observation of 200 microseconds (!)

shapiro_delay <- transit_time_relativistic - transit_time_classical
cat("shapiro delay = ", shapiro_delay, "seconds")
```

### 2021-04-19 shapiro.Rmd
fixes #21

@@
 title: "Shapiro delay"
-output: html_document
+output:
+  pdf_document: default
+  html_document: default
 ---
@@
 The Shapiro delay is one of the classical tests of general relativity.
-Here I attempt to reproduce that analysis using numerical reasoning
-but fail because of numerical roundoff error.
+Here I use direct numerical simulation to
 
@@
 ```{r}
+source("usefulnullgeodesicfuncs.R")
+```
 
-source("usefulnullgeodesicfuncs.R")
+Schwarzschild radius of sun:
 
-r0 <- r_sun/d # grazing incidence of the sun in Schwarzschild radii
+```{r}
+r_s
+```
+
+```{r}
+r0 <- r_sun/r_s # grazing incidence of the sun in Schwarzschild radii
 r0
-r_s
 ```
@@
 ```{r}
+
 b_radial <- r0/sqrt(1-1/r0) 
-
-small <- 1e-6  # needed to prevent numerical problems
+r0
+b_radial
+small <- 1e-10 # needed to prevent numerical problems
 
@@
 column) and angle $\phi=0$ (the `phi` column).  We can perform the
-same calculation but for Venus"
+same calculation but for Venus:
 
@@
 
-The next step is to approximate the funtion
+The next step is to approximate the function
 
@@
 
+```{r}
+st <- function(x){(x-min(x))/(max(x)-min(x))}
+plot(out_earth$x,out_earth$y,type='o',col='red',cex=st(out_earth$time),asp=1)
+event_horizon()
+```
+
 So now `f_earth(t)` gives the time required to achieve a particular
-radius.  Thus to calculate the transit time from Earth to Venus we
-simply add the times to get to the Earth's radius to Venus's:
+radius.  Thus to calculate the transit time from Earth to Venus (NB:
+one way!) we simply add the times to get to the Earth's radius to
+Venus's:
 
@@
 
+Now we calculate the classical time.  The naive estimate would be
+
 ```{r}
+transit_time_classical <- (orbital_radius_earth + orbital_radius_venus)/sol # seconds
+```
 
-## Now calculate the classical time, simply distance/speed:
-transit_time_classical <- (orbital_radius_earth + orbital_radius_venus)/sol # seconds
+but this neglects the fact that the light path starts at grazing
+incidence (note that the relativistic path does not graze the sun but
+actually grazes radius $r_s+2MG/c^2$, but this is the same to first
+order).
 
-## plot (rather dull):
-st <- function(x){(x-min(x))/(max(x)-min(x))}
-plot(out_earth$x,out_earth$y,type='o',col='red',cex=st(out_earth$time),asp=1)
-event_horizon()
+![Diagram of light path from Earth to Venus via grazing incidence at sun](earth_venus_sun_eddington_shapiro.png){width=100px}
 
-shapiro_delay <- transit_time_relativistic - transit_time_classical
-# compare observation of 200 microseconds (!)
+```{r}
+distance_grazing_to_earth <- sqrt(orbital_radius_earth^2 - r_sun^2)
+distance_grazing_to_venus <- sqrt(orbital_radius_venus^2 - r_sun^2)
+transit_time_classical <- (distance_grazing_to_earth + distance_grazing_to_venus)/sol
+transit_time_classical
+```
 
-shapiro_delay <- transit_time_relativistic - transit_time_classical
-cat("shapiro delay = ", shapiro_delay, "seconds")
+Then the Shapiro delay is just the difference:
+
+```{r}
+shapiro_delay <- (transit_time_classical - transit_time_relativistic)*2
+shapiro_delay
 ```
+
+(we multiply by two because the radar had an outward and return leg,
+to Venus and back).  Compare this with the measured effect of 200
+microseconds:
+
+```{r}
+shapiro_delay/200e-6
+```

### 2021-04-19 shapiro.R

@@ -1,45 +1,0 @@
-## Numerical verification of the Shapiro delay experiment.
-##
-## Technically, this is a failed experiment: the predicted Shapiro
-## delay, between Earth and Venus, is zero (as printed at the bottom
-## of this file).  What baffles me is why the agreement is so precise,
-## given that the two methods of calculation are so utterly different.
-
-
-source("usefulnullgeodesicfuncs.R")
-
-st <- function(x){(x-min(x))/(max(x)-min(x))}
-
-r0 <- r_sun/d # grazing incidence of the sun in Schwarzschild radii
-b_radial <- r0/sqrt(1-1/r0)
-
-small <- 1e-6  # needed to prevent numerical problems
-
-d_earth <- seq(from=orbital_radius_earth/r_s * 0.999, to=orbital_radius_earth/r_s * 1.001,len=100)
-d_venus <- seq(from=orbital_radius_venus/r_s * 0.999, to=orbital_radius_venus/r_s * 1.001,len=100)
-
-d_earth <- c(1:10,d_earth)
-
-
-out_earth <- nullgeodesic(r_start=r0,b=b_radial*(1-small),tau=c(0,d_earth))
-out_venus <- nullgeodesic(r_start=r0,b=b_radial*(1-small),tau=c(0,d_venus))
-
-f_earth <- approxfun(out_earth$r,out_earth$r)
-f_venus <- approxfun(out_venus$r,out_venus$r)
-
-transit_time_relativistic  <-   # transit time in seconds
-(f_earth(orbital_radius_earth/r_s) + f_venus(orbital_radius_venus/r_s))*r_s/sol
-
-
-## Now calculate the classical time, simply distance/speed:
-transit_time_classical <- (orbital_radius_earth + orbital_radius_venus)/sol # seconds
-
-## plot (rather dull):
-plot(out_earth$x,out_earth$y,type='o',col='red',cex=st(out_earth$time),asp=1)
-event_horizon()
-
-shapiro_delay <- transit_time_relativistic - transit_time_classical
-# compare observation of 200 microseconds (!)
-
-shapiro_delay <- transit_time_relativistic - transit_time_classical
-cat("shapiro delay = ", shapiro_delay, "seconds")

### 2022-05-08 arxiv/templateArxiv.tex
start work on an arxiv submission

\documentclass{article}

\usepackage{PRIMEarxiv}

\usepackage[utf8]{inputenc} % allow utf-8 input
\usepackage[T1]{fontenc}    % use 8-bit T1 fonts
\usepackage{hyperref}       % hyperlinks
\usepackage{url}            % simple URL typesetting
\usepackage{booktabs}       % professional-quality tables
\usepackage{amsfonts}       % blackboard math symbols
\usepackage{nicefrac}       % compact symbols for 1/2, etc.
\usepackage{microtype}      % microtypography
\usepackage{lipsum}
\usepackage{fancyhdr}       % header
\usepackage{graphicx}       % graphics
\graphicspath{{media/}}     % organize your images and other figures under media/ folder

%Header
\pagestyle{fancy}
\thispagestyle{empty}
\rhead{ \textit{ }}

% Update your Headers here
\fancyhead[LO]{Proper length as a radial coordinate in the Schwarzschild metric}
% \fancyhead[RE]{Firstauthor and Secondauthor} % Firstauthor et al. if more than 2 - must use \documentclass[twoside]{article}

%% Title
\title{A template for Arxiv Style
%%%% Cite as
%%%% Update your official citation here when published 
\thanks{\textit{\underline{Citation}}: 
\textbf{Authors. Title. Pages.... DOI:000000/11111.}} 
}

\author{
  Robin K. S. Hankin \\
Auckland University of Technology \\
Auckland, New Zealand\\
  \texttt{hankin.robin@gmail.com} 
}

\begin{document}
\maketitle

\begin{abstract}
The Schwarzschild solution to the spherically symmetric vacuum
Einstein field equations is usually presented in terms of
Schwarzschild coordinates $(t,r,\phi,\theta)$.  However, we are free
to consider reparametrizations in terms of functions of the
Schwarzschild coordinates.  Here I consider the Schwarzschild solution
in terms of $(t,u_1\phi,\theta)$, where $u_1=u_1(r)$ is the
straight-line proper length from a point to the event horizon.  This
coordinate system has some interesting properties and affords new
physical insight to black holes.  I also consider related
parameterizations $u_2$ and $u_3$ which preserve proper area or volume
instead of proper length.
\end{abstract}

% keywords can be removed
\keywords{Black holes\and Schwarzschild metric\and proper distance}

\section{Introduction}

The Schwarzschild metric is usually written

\begin{equation}
  ds^2=-dt^2\left(1-\frac{2GM}{c^2r}\right)
  +\frac{dr^2}{\left(1-\frac{2GM}{c^2r}\right)} + r^2(\sin^2\theta
  d\phi^2 + d\theta^2)
\end{equation}

where $G$ is the gravitational constant, $c$ the speed of light, and
$M$ the mass of the black hole.  However, it is much better to use
units in which $c=1$ and $2MG/c^2=1$ [that is, the speed of light is
  unity and the Schwarzschild radius is unity].  For a solar mass
black hole ($2\times 10^{30}\,\mathrm{kg}$) we have the unit of
distance being about $3\,\mathrm{km}$ and the unit of time about
$10\,\mathrm{ms}$.  With such units the Schwarzschild metric becomes

\begin{equation}
  ds^2=-dt^2\left(1-1/r\right)
  +\frac{dr^2}{\left(1-1/r\right)} + r^2(\sin^2\theta d\phi^2 + d\theta^2)
\end{equation}

If we consider constant $\phi,\theta$ then we have

\label{sec:headings}

\begin{equation}
  ds^2=-dt^2\left(1-1/r\right)
  +\frac{dr^2}{\left(1-1/r\right)}.
\end{equation}

The proper distance from a point at Schwarzschild coordinate $r$ would be

\begin{equation}
  u_1(r)=
  \int_{r'=1}^r\frac{dr'}{\sqrt{1-1/r'}}=
  \sqrt{r(r-1)} + \frac{1}{2}\log\frac{\sqrt{r}+\sqrt{r-1}}{\sqrt{r}-\sqrt{r-1}}
  \end{equation}

Hankin \cite{hankin2021} shows that

\begin{equation}
  u_1(1+w)=2\sqrt{w} + \frac{1}{3}w^{3/2} + O(w^{7/2})
\end{equation}
for $w\ll 1$, and

\begin{equation}
  u_1(1+w)=w + \frac{1}{2}\log w + \frac{1}{2} +
  \log(2) + \frac{1}{8}w^{-1} + O(w^{-2})
\end{equation}

for $w\gg 1$.

\section{Conclusion}
Your conclusion here

%Bibliography
\bibliographystyle{unsrt}  
\bibliography{references}

\end{document}

### 2022-05-19 arxiv/template.tex
better arxiv template

--- 2022-05-08 arxiv/templateArxiv.tex
@@
 
-\usepackage{PRIMEarxiv}
+\usepackage{arxiv}
 
@@
 \usepackage{microtype}      % microtypography
-\usepackage{lipsum}
-\usepackage{fancyhdr}       % header
-\usepackage{graphicx}       % graphics
-\graphicspath{{media/}}     % organize your images and other figures under media/ folder
+\usepackage{cleveref}       % smart cross-referencing
+\usepackage{lipsum}         % Can be removed after putting your text content
+\usepackage{graphicx}
+\usepackage{natbib}
+\usepackage{doi}
+\title{Proper radial distance as a coordinate in the Schwarzschild metric}
 
-%Header
-\pagestyle{fancy}
-\thispagestyle{empty}
-\rhead{ \textit{ }}
+% Here you can change the date presented in the paper title
+%\date{September 9, 1985}
+% Or remove it
+%\date{}
 
-% Update your Headers here
-\fancyhead[LO]{Proper length as a radial coordinate in the Schwarzschild metric}
-% \fancyhead[RE]{Firstauthor and Secondauthor} % Firstauthor et al. if more than 2 - must use \documentclass[twoside]{article}
-
-%% Title
-\title{A template for Arxiv Style
-%%%% Cite as
-%%%% Update your official citation here when published 
-\thanks{\textit{\underline{Citation}}: 
-\textbf{Authors. Title. Pages.... DOI:000000/11111.}} 
+\author{ \href{https://orcid.org/0000-0000-0000-0000}{\includegraphics[scale=0.06]{orcid.pdf}\hspace{1mm}Robin K. S. Hankin}\thanks{Use footnote for providing further
+		information about author (webpage, alternative
+		address)---\emph{not} for acknowledging funding agencies.} \\
+              AUT University\\
+              Auckland, New Zealand\\
+	\texttt{hankin.robin@gmail.com} \\
 }
 
-\author{
-  Robin K. S. Hankin \\
-Auckland University of Technology \\
-Auckland, New Zealand\\
-  \texttt{hankin.robin@gmail.com} 
+% Uncomment to override  the `A preprint' in the header
+%\renewcommand{\headeright}{Technical Report}
+%\renewcommand{\undertitle}{Technical Report}
+\renewcommand{\shorttitle}{Proper radial distance}
+
+%%% Add PDF metadata to help others organize their library
+%%% Once the PDF is generated, you can check the metadata with
+%%% $ pdfinfo template.pdf
+\hypersetup{
+pdftitle={Proper radial distance as a coordinate in the Schwarzschild metric}
+pdfsubject={q-bio.NC, q-bio.QM},
+pdfauthor={Robin K. S.~Hankin}
+pdfkeywords={First keyword, Second keyword, More},
 }
@@
 % keywords can be removed
-\keywords{Black holes\and Schwarzschild metric\and proper distance}
+\keywords{First keyword \and Second keyword \and More}
 
 \section{Introduction}
-
 The Schwarzschild metric is usually written
@@
   ds^2=-dt^2\left(1-\frac{2GM}{c^2r}\right)
-  +\frac{dr^2}{\left(1-\frac{2GM}{c^2r}\right)} + r^2(\sin^2\theta
+<  +\frac{dr^2}{\left(1-\frac{2GM}{c^2r}\right)} + r^2(\sin^2\theta
   d\phi^2 + d\theta^2)
@@
 
-Hankin \cite{hankin2021} shows that
+\cite{hankin2021} shows that
+
+\citep{hankin2021} shows that
+
+\citet{hankin2021} shows that
+
+\cite{hankin2020}
 
@@
 
-\section{Conclusion}
-Your conclusion here
-
-%Bibliography
-\bibliographystyle{unsrt}  
+\bibliographystyle{unsrtnat}
 \bibliography{references}

### 2022-05-20 arxiv/template.tex
minor changes

@@
 
-\author{ \href{https://orcid.org/0000-0000-0000-0000}{\includegraphics[scale=0.06]{orcid.pdf}\hspace{1mm}Robin K. S. Hankin}\thanks{Use footnote for providing further
-		information about author (webpage, alternative
-		address)---\emph{not} for acknowledging funding agencies.} \\
+\author{ \href{https://orcid.org/0000-0001-5982-0415}{\includegraphics[scale=0.06]{orcid.pdf}\hspace{1mm}Robin K. S. Hankin}\thanks{
+\url{https://academics.aut.ac.nz/robin.hankin}}\\
               AUT University\\
@@
 pdfauthor={Robin K. S.~Hankin}
-pdfkeywords={First keyword, Second keyword, More},
+pdfkeywords={Black holes, Schwarzschild metric, proper distance,
 }

### 2022-05-20 light_start_at_r_equals_1.R
minor changes

@@ -6,7 +6,7 @@
 source("usefulfuncs.R") # defines polargrid() etc
 source("usefullightfuncs.R") # defines lightpoints()
 
-r_start <- 1 # starting radius for light ray
+r_start <- 0.8 # starting radius for light ray
 mask <- TRUE    # set to TRUE to mask the diagram to a circle; set to
                 # FALSE for debugging (masking is time-consuming)
 size_of_plot <- 3
@@ -112,8 +112,7 @@ cutoffmatrix  <- matrix(c(
 colnames(cutoffmatrix) <- c("cuts","vals")
 fmax <- approxfun(cutoffmatrix[,1],cutoffmatrix[,2])
 
-start_angle <- seq(from=-pi/2,to=pi/2,len=100)
-
+start_angle <- seq(from=-pi/2,to=-pi/8,len=30)
 start_angle <- start_angle[-c(1,length(start_angle))]
 
 ## Uncomment the following when refining cutoffmatrix:
@@ -124,13 +123,25 @@ for(initialangletotangent in start_angle){
 
   final_phi <- min(fmax(initialangletotangent),2*pi)
 #  print(c(initialangletotangent,final_phi))
-  xy1 <-
+  xyinc <- 
     nullgeodesic(
         r_start=r_start,  # r=3/2 is a circular (but unstable) orbit
         dubydphistart = tan(initialangletotangent)/r_start,
-        phi=seq(from=0,to=final_phi,len=100)
-    )
+        phi=seq(from=0,to=final_phi,len=100),include=T)
+    
   
+
+
+f <- function(r){
+A <- sqrt(r)
+B <- sqrt(r-1)
+A*B + log((A+B)/(A-B))/2
+
+}
+xyinc[,3] <- f(xyinc[,3])
+xy1 <- cbind(xyinc[,3]*cos(xyinc[,4]),xyinc[,3]*sin(xyinc[,4]))
+
+
   points(xy1,col='red',type='l')
 }
 
@@ -151,5 +162,5 @@ if(mask){
   }
 }
 
-event_horizon(fill=FALSE)
-points(1,0,pch=16)
+
+points(f(1),0,pch=16)

### 2022-05-20 light_start_at_u_equals_2_stringcoords.R

@@ -12,7 +12,8 @@ func <- u1  # can be u2 or u3 (or u0, I guess).  Used to transform the
             # radial coordinate, for light paths and also the source
             # point.
 
-r_start <- 2 # starting radius for light ray
+r_start <- u1_inv(2) # starting radius for light ray, starts at u=2, but the
+                     # calcs are in terms of schwarzschild r
 domask <- TRUE    # set to FALSE to see entire geodesic
 size_of_plot <- 3
 
@@ -72,45 +73,45 @@ cutoffmatrix  <- matrix(c(
     0.0, pi/2 + 0.4,
     0.1, pi/2   + 0.9,
     0.15, pi/2  + 1.56,
-    0.2, pi/2   + pi/2,
-    0.22, pi/2  + 2.3,
-    0.23, pi/2  + 2.3,
-    0.26, pi/2  + 2.5,
-    0.28, pi/2  + pi,
-    0.285, pi/2  + 3.6,
-    0.29, pi/2  + 4.6,
-    0.295, pi/2  + 3*pi/2,
-    0.30, pi/2   + 3*pi/2,
-    0.301, pi/2 + 5,
-    0.302, pi/2 + 6,
-    0.305, pi/2 + 6,
-    0.31, pi/2 + 6,
-    0.32, pi/2 + 6,
-    0.325, pi/2 + 4.5,
-    0.33, pi/2 + 4.4,
-    0.335, pi/2 + 4.3,
-    0.34, pi/2 + 4.2,
-    0.35, pi/2 + 4,
-    0.36, pi/2 + 4,
-    0.37, pi/2 + 3.7,
-    0.38, pi/2 + 3.5,
-    0.4, pi/2 + 3.4,
-    0.43, pi/2 + 3.2,
-    0.45, pi/2 + 2.9,
-    0.5, pi/2 + 2.8,
-    0.55, pi/2 + 2.4,
-    0.6, pi/2 + 2.3,
-    0.7, pi/2 + 2.1,
-    0.8, pi/2 + 1.7,
-    0.9, pi/2 + 1.5,
-    1.0, pi/2 +1.2,
-    1.1, pi/2 +1.1,
-    1.15, pi/2 +1.03,
-    1.2, pi/2 +0.9,
-    1.25, pi/2 +0.8,
-    1.3, pi/2 +0.6,
[…]

### 2022-05-20 usefullightfuncs.R

@@ -43,7 +43,9 @@ library("deSolve")   # ode()
   r <- 1/bh[,2]  # r=1/u is the dependent variable
   ## NB: bh[,3] is du/phi
   
+
   xy <- cbind(r*cos(phi),r*sin(phi))
+
   if(include){
       xy <- cbind(xy,r,phi)
   }

### 2022-06-16 closest_approach3_stringcoords.R
bugfix

@@ -109,7 +109,8 @@ o <- stringpoints(y_start=extra ,initial_string_angle = 0,theta=tseq,give=TRUE)
 u <- u1(o[,2])
 theta <- o[,1]
 xy <- cbind(x=u*cos(theta),y=u*sin(theta))
-points(rbind(xy,jj),type='l',col="green",lwd=3)
+
+points(xy,type='l',col="green",lwd=3)
 xy[,2] <- -xy[,2]
 points(rbind(xy,jj),type='l',col="green",lwd=3)
 ## mask strings too far from the black hole:

### 2022-06-16 shapiro.Rmd
tiny changes

@@
 The Shapiro delay is one of the classical tests of general relativity.
-Here I use direct numerical simulation to
+In this unfinished document I use direct numerical simulation to
+verify the predictions of GR.
 
@@
 
-![Diagram of light path from Earth to Venus via grazing incidence at sun](earth_venus_sun_eddington_shapiro.png){width=100px}
+![Diagram of light path from Earth to Venus via grazing incidence at sun](earth_venus_sun_eddington_shapiro.png){width=700px}
 

### 2022-06-16 closest_approach3_stringcoords.R
say where stringfuncs.R lives

@@ -1,12 +1,13 @@
-## Identical to closest_approach2.R but plots a single string looping
-## round the black hole in green, showing the self-intersection.
+## Shows paths of light inextensible string near a black hole but
+## using string coordinates.
 
 ## This script runs standalone but is designed to be called from
 ## maker.R [cf Makefile], which creates the pdf file
-## "closest_approach2.pdf"
+## "closest_approach3_stringcoords.pdf"
 
 source("usefulfuncs.R") # defines stringpoints() and dseq()
-source("stringfuncs.R") # u1() and u2()  see schwarzschild repo
+source("stringfuncs.R") # defines u1() and u2() ; file stringfuncs.R
+                        # lives in the schwarzschild repo under R/
 
 domask <- TRUE
 

### 2022-06-16 arxiv/template.tex
typos

@@
 %\renewcommand{\undertitle}{Technical Report}
-\renewcommand{\shorttitle}{Proper radial distance}
+%\renewcommand{\shorttitle}{Proper radial distance}
 
@@
 \hypersetup{
-pdftitle={Proper radial distance as a coordinate in the Schwarzschild metric}
+pdftitle={Proper radial distance as a coordinate in the Schwarzschild metric},
 pdfsubject={q-bio.NC, q-bio.QM},
-pdfauthor={Robin K. S.~Hankin}
-pdfkeywords={Black holes, Schwarzschild metric, proper distance,
+pdfauthor={Robin K. S.~Hankin},
+pdfkeywords={Black holes, Schwarzschild metric, proper distance}
 }
-
 \begin{document}

### 2022-06-16 arxiv/template.tex
include spacetime diagrams

@@
 \usepackage{arxiv}
-
 \usepackage[utf8]{inputenc} % allow utf-8 input
@@
 \usepackage{doi}
+\usepackage{caption}
+\usepackage{subcaption}
 \title{Proper radial distance as a coordinate in the Schwarzschild metric}
@@
   ds^2=-dt^2\left(1-\frac{2GM}{c^2r}\right)
-<  +\frac{dr^2}{\left(1-\frac{2GM}{c^2r}\right)} + r^2(\sin^2\theta
+  +\frac{dr^2}{\left(1-\frac{2GM}{c^2r}\right)} + r^2(\sin^2\theta
   d\phi^2 + d\theta^2)
@@
 
-\citep{hankin2021} shows that
-
-\citet{hankin2021} shows that
-
-\cite{hankin2020}
+% \citep{hankin2021} \citet{hankin2021}
 
@@
 
-for $w\gg 1$.
+for $w\gg 1$.  We can use this as a coordinate on the same footing as
+circumferential radius $r$.  The first thing to do would be to
+suppress $\theta,\phi$ coordinates and consider only $r,t$; see
+Figures~\ref{schwarzschild} and~\ref{eddington}.
+
+\begin{figure}%
+    \centering
+    \subfloat{{\includegraphics[width=7cm]{schwarzschild.pdf} }}%produced by the schwarzschild package
+    \subfloat{{\includegraphics[width=7cm]{schwarzschild_stringcoords.pdf} }}%
+    \caption{Spacetime near a black hole with $\theta,\phi$
+      suppressed.  Left, diagram in $r,t$ coordinates.  Right, same
+      spacetime but presented in $u_1,t$ coordinates: note absence of interior on right}
+    \label{schwarzschild}%
+\end{figure}
+
+\begin{figure}%
+    \centering
+    \subfloat{{\includegraphics[width=7cm]{eddington.pdf} }}%
+    \subfloat{{\includegraphics[width=7cm]{eddington_stringcoords.pdf} }}%
+    \caption{Spacetime near a black hole with $\theta,\phi$ suppressed
+      using Eddington-Finkelstein coordinates.  Left, diagram in $r,t$
+      coordinates.  Right, same spacetime but presented in $u_1,t$
+      coordinates: note absence of interior on right}
+    \label{eddington}%
+\end{figure}
+
+\begin{figure}%
+    \centering
+    \subfloat{{\includegraphics[width=7cm]{gullstrand.pdf} }}%
+    \subfloat{{\includegraphics[width=7cm]{gullstrand_stringcoords.pdf} }}%
+    \caption{Spacetime near a black hole with $\theta,\phi$ suppressed
+      using Gullstrand coordinates.  Left, diagram in $r,t$
+      coordinates.  Right, same spacetime but presented in $u_1,t$
+      coordinates: note absence of interior on right}
+    \label{gullstrand}%
+\end{figure}
 

### 2022-06-16 light_closest_approach_stringcoords.R
new files for null geodesics in string coordinates

@@ -1,0 +1,120 @@
+## Null geodesics in the Schwarzschild geometry.  Same as
+## light_closest_approach.R but plots in string coordinates.
+
+## This script runs standalone but is designed to be called from
+## maker.R [cf Makefile], which creates the pdf file
+## "light_closest_approach_stringcoords.pdf"
+
+## Here, the light starts tangentially from phi=0.
+## It is like closest_approach.R and closest_approach2.R but for light
+## beams rather than string.  Compare light_start_at_r_equals_2.R,
+## which has light beams starting from (2,0) going in all directions.
+
+
+## This file uses the same ODE engine and idiom as
+## light_start_at_r_equals_2.R (henceforth "lsar2"), but differs in
+## detail, particularly cutoffmatrix.
+
+## ODE is d^2u/d phi^2 = 3u^2/2-u  (where u=1/r)
+
+source("usefulfuncs.R") # defines polargrid() etc
+source("usefullightfuncs.R") # defines lightpoints()
+source("stringfuncs.R") # u1() and u2() etc; lives in the schwarzschild repo
+
+domask <- TRUE    # set to FALSE to see entire geodesic (TRUE is time-consuming but needed for publication quality)
+size_of_plot <- 3.5
+
+## plot setup:
+par(xpd=TRUE)
+
+jj <- c(-size_of_plot, size_of_plot)
+plot(NULL,asp=1,xlim=jj,ylim=jj,type='l',axes=FALSE,xlab='',ylab='')
+
+cutoffmatrix  <- matrix(c(
+    0,0.001 ,
+    0.01,0.2,
+    0.02,0.3,
+    0.05,0.4,
+    0.07,0.5,
+    0.08,0.5,
+    0.1,0.7,
+    0.2,1.0,
+    0.5,1.8,
+    0.7,2.2,
+    0.8,2.5,
+    0.9,2.8,
+
+    1,3,
+    1.1, pi+0.2,
+    
+    1.2,pi + 0.6,
+    1.3,pi + 1.1,
+
+    1.35,pi + 1.4,
+    1.40,pi + pi/2+0.3,
+    1.41,pi + pi/2+0.3,
+    1.42,pi+2.2,
+    1.43,pi+2.3,
+    1.44,pi+2.4,
+    1.47,pi+pi          ,
[…]

### 2022-06-16 light_start_at_r_equals_2_stringcoords.R

@@ -1,0 +1,159 @@
+## Null geodesics starting at r=2 but with different starting angles.
+
+## ODE is d^2u/d phi^2 = 3u^2/2-u  (where u=1/r)
+
+source("usefulfuncs.R") # defines polargrid() etc
+source("usefullightfuncs.R") # defines lightpoints()
+source("stringfuncs.R") # u1() and u2() etc; lives in the schwarzschild repo
+
+r_start <- 2 # starting radius for light ray
+domask <- TRUE    # set to FALSE to see entire geodesic
+size_of_plot <- 3
+
+## plot setup:
+par(xpd=TRUE)
+
+jj <- c(-size_of_plot, size_of_plot)
+plot(NULL,asp=1,xlim=jj,ylim=jj,type='l',axes=FALSE,xlab='',ylab='')
+
+
+## The chief numerical difficulty here is defining the maximum angle
+## that the path (here a light ray, elsewhere a taut string) can
+## attain.  We are starting at (r=2,phi=0) and indexing by angle made
+## by the string to the tangent at that point.  We need to determine
+## what the maximum value of phi is.  For example, consider flat space
+## and a light ray starting tangentially at (2,0).  Then the maximum
+## value of phi would be pi/2.  In curved Schwarzschild space, this
+## angle is more than pi/2.
+
+
+## Variable 'cutoffmatrix' has two columns.  The first gives the angle
+## of the string at its start point (r=r_start, theta=0) to the
+## tangent of the circle of radius r_start at that point.  Negative
+## values mean outward-going strings and positive values mean
+## inward-going strings.  +pi/2 corresponds to a radially inward
+## string and -pi/2 corresponds to a radially outward string.  0
+## corresponds to a string that is tangential.  The second column
+## gives the maximum value of phi to integrate to.  Make this too big,
[…]

### 2022-06-16 light_closest_approach_stringcoords.R
mods to make pdf files

@@ -80,7 +80,6 @@ cutoffmatrix  <- matrix(c(
 colnames(cutoffmatrix) <- c("cuts","vals")
 fmax <- approxfun(cutoffmatrix[,1],cutoffmatrix[,2])
 
-
 r_start <- seq(from=1.1,to=4,len=47)
 
 for(r in r_start){

### 2022-06-16 light_start_at_r_equals_2.R

@@ -134,7 +134,8 @@ for(initialangletotangent in start_angle){
     nullgeodesic(
         r_start=r_start,  # r=3/2 is a circular (but unstable) orbit
         dubydphistart = tan(initialangletotangent)/r_start,
-        phi=seq(from=0,to=final_phi,len=100)
+        phi=seq(from=0,to=final_phi,len=100),
+        include=TRUE
     )
   
   points(xy1,col='red',type='l')

### 2022-06-16 maker.R

@@ -44,10 +44,16 @@ source("light_closest_approach.R")
 dev.off()
 
 
+pdf(file="light_closest_approach_stringcoords.pdf")
+source("light_closest_approach_stringcoords.R")
+dev.off()
+
+
 pdf(file="illustrative_relativistic_trajectory.pdf")
 source("illustrative_relativistic_trajectory.R")
 dev.off()
 
-
-
+pdf(file="light_start_at_r_equals_2_stringcoords.pdf")
+source("light_start_at_r_equals_2_stringcoords.R")
+dev.off()
 

### 2022-06-16 closest_approach3.R
slight mods remove event horizon

@@ -6,7 +6,7 @@
 ## "closest_approach2.pdf"
 
 source("usefulfuncs.R") # defines stringpoints() and dseq()
-
+source("stringfuncs.R") # u1() and u2()  see schwarzschild repo
 
 domask <- TRUE
 
@@ -93,7 +93,10 @@ for(i in seq_along(dist)){
   }
 
   tseq <- seq(from=0,to=f(dist[i]),len=100)
-  xy <- stringpoints(y_start=dist[i],initial_string_angle = 0,theta=tseq)
+  o <- stringpoints(y_start=dist[i],initial_string_angle = 0,theta=tseq,give=TRUE)
+  u <- u1(o[,2])
+  theta <- o[,1]
+  xy <- cbind(x=u*cos(theta),y=u*sin(theta))
   points(xy,type='l',col=stringcol,lwd=2)
   
 #  xy <- stringpoints(y_start=dist[i],initial_string_angle = 0, theta=-tseq)
@@ -112,4 +115,4 @@ points(rbind(xy,jj),type='l',col="green",lwd=3)
 if(domask){mask(7)}
 
 polargrid(rlab=6.5)
-event_horizon()
+

### 2022-06-16 arxiv/template.tex
include string

@@
 
+\begin{figure}%
+    \centering
+    \subfloat{{\includegraphics[width=7cm]{light_start_at_r_equals_2.pdf} }}%
+    \subfloat{{\includegraphics[width=7cm]{light_start_at_r_equals_2_stringcoords.pdf} }}
+    \caption{Null geodesics near a black hole.  Left, diagram using $r,\phi$
+      coordinates.  Right, same spacetime but presented in $u_1,\phi$
+      coordinates: note absence of interior on right}
+    \label{light_2}%
+\end{figure}
+
+\begin{figure}%
+    \centering
+    \subfloat{{\includegraphics[width=7cm]{light_closest_approach.pdf} }}%
+    \subfloat{{\includegraphics[width=7cm]{light_closest_approach_stringcoords.pdf} }}
+    \caption{Null geodesics near a black hole.  Left, diagram using $r,\phi$
+      coordinates.  Right, same spacetime but presented in $u_1,\phi$
+      coordinates: note absence of interior on right}
+    \label{light_closest_2}%
+\end{figure}
+
+\begin{figure}%
+    \centering
+    \subfloat{{\includegraphics[width=7cm]{closest_approach_nonselfintersecting.pdf} }}%
+    \subfloat{{\includegraphics[width=7cm]{closest_approach_nonselfintersecting_stringcoords.pdf} }}
+    \caption{Light inextensible string near a black hole,
+      non-self-intersecting case.  Left, diagram using $r,\phi$
+      coordinates.  Right, same spacetime but presented in $u_1,\phi$
+      coordinates: note absence of interior on right}
+    \label{closest_approach_nonself}%
+\end{figure}
+
+\begin{figure}%
+    \centering
+    \subfloat{{\includegraphics[width=7cm]{closest_approach_selfintersecting.pdf} }}%
+    \subfloat{{\includegraphics[width=7cm]{closest_approach_selfintersecting_stringcoords.pdf} }}
+    \caption{Light inextensible string near a black hole, self intersecting case.  Left, diagram using $r,\phi$
+      coordinates.  Right, same spacetime but presented in $u_1,\phi$
+      coordinates: note absence of interior on right}
+    \label{closest_approach_self}%
+\end{figure}
+
 \bibliographystyle{unsrtnat}

### 2022-06-16 maker.R

@@ -2,12 +2,24 @@ pdf(file="closest_approach_nonselfintersecting.pdf")
 source("closest_approach.R")
 dev.off()
 
+pdf(file="closest_approach_nonselfintersecting_stringcoords.pdf")
+source("closest_approach_stringcoords.R")
+dev.off()
+
 pdf(file="closest_approach_selfintersecting.pdf")
 source("closest_approach2.R")
 dev.off()
 
+pdf(file="closest_approach_selfintersecting_stringcoords.pdf")
+source("closest_approach2_stringcoords.R")
+dev.off()
+
 pdf(file="closest_approach_selfintersecting_onegreen.pdf")
 source("closest_approach3.R")
+dev.off()
+
+pdf(file="closest_approach_selfintersecting_onegreen_strincoords.pdf")
+source("closest_approach3_stringcoords.R")
 dev.off()
 
 pdf(file="angle_at_r_equals_2.pdf")

### 2022-06-18 closest_approach2_stringcoords.R
add singularity

@@ -110,3 +110,4 @@ if(domask){mask(7)}
 
 polargrid(rlab=6.5)
 
+points(0,0)

### 2022-06-18 closest_approach_stringcoords.R

@@ -78,3 +78,4 @@ if(domask){mask(7)}
 
 polargrid(rlab=6.5)
 
+points(0,0)

### 2022-06-18 arxiv/template.tex
u2 and u3

@@
 
-for $w\gg 1$.  We can use this as a coordinate on the same footing as
-circumferential radius $r$.  The first thing to do would be to
-suppress $\theta,\phi$ coordinates and consider only $r,t$; see
-Figures~\ref{schwarzschild} and~\ref{eddington}.
+for $w\gg 1$.
+
+\subsection{Equal areas}
+
+We might be interested in preserving areas.  In this case we would use
+a coordinate $u_2$, defined by
+
+\begin{equation}
+  \frac{u_2^2}{2}=\int\frac{r\,dr}{\sqrt{1-1/r}}
+\end{equation}
+
+giving $u_2=\sqrt{I-K}$ where $K$ is a constant of integration and
+
+\begin{equation}
+    I = r^2\sqrt{1-{1}/{r}}\left(1+\frac{3}{2r}\right)
+ + \frac{3}{8}\log\left(\frac{\sqrt{r}+\sqrt{r-1}}{\sqrt{r}-\sqrt{r-1}}\right)
+\end{equation}
+
+Similarly for volume, $u_3=(I-K)^{1/3}$ with
+
+\begin{equation}
+    I = r^3\sqrt{1-{1}/{r}}\left(1+\frac{5}{4r} + \frac{15}{8r^2}\right)
++ \frac{5}{16}\log\left(\frac{\sqrt{r}+\sqrt{r-1}}{\sqrt{r}-\sqrt{r-1}}\right)
+\end{equation}
+\section{Results}
+
+We can use $u_1$ as a coordinate on the same footing as
+circumferential radius $r$.  First we suppress $\theta,\phi$
+coordinates and consider only $r,t$; see Figures~\ref{schwarzschild},
+\ref{eddington}, and ~\ref{gullstrand}.
 
@@
 
+It is also possible to plot null geodesics for light in an equatorial
+plane: figures~\ref{light_2} and \ref{light_closest_2}.  Light
+inextensible string is plotted, again for the equatorial case, in
+figures~\ref{closest_approach_nonself} and
+\ref{closest_approach_self}, for the nonself-intersecting and self
+intersecting cases respectively.
+
 \begin{figure}%
@@
     \subfloat{{\includegraphics[width=7cm]{closest_approach_selfintersecting_stringcoords.pdf} }}
-    \caption{Light inextensible string near a black hole, self intersecting case.  Left, diagram using $r,\phi$
-      coordinates.  Right, same spacetime but presented in $u_1,\phi$
-      coordinates: note absence of interior on right}
+    \caption{Light inextensible string near a black hole, self
+      intersecting case.  Left, diagram using $r,\phi$ coordinates.
+      Right, same spacetime but presented in $u_1,\phi$ coordinates:
+      note absence of interior on right}
     \label{closest_approach_self}%
 \end{figure}
+
+\section{Conclusions}
 

### 2022-06-19 arxiv/template.tex
start conclusions

@@
 
+We have presented physical diagrams of a spherically symmetric black
+hole using proper radial coordinate $u_1$ instead of circumferential
+radius $r$.  One issue is the fact that $u_1$ coordinates do not cover
+the interior of the event horizon $r<1$.  This gives the various null
+geodesic and light inextensible string diagrams a different appearance
+in $u_1$ coordinates.
+
 \bibliographystyle{unsrtnat}
-\bibliography{references}
+\bibliography{../stringrefs}
 

### 2022-06-19 arxiv/template.tex
slight rewording

@@
 Einstein field equations is usually presented in terms of
-Schwarzschild coordinates $(t,r,\phi,\theta)$.  However, we are free
-to consider reparametrizations in terms of functions of the
-Schwarzschild coordinates.  Here I consider the Schwarzschild solution
-in terms of $(t,u_1\phi,\theta)$, where $u_1=u_1(r)$ is the
-straight-line proper length from a point to the event horizon.  This
-coordinate system has some interesting properties and affords new
-physical insight to black holes.  I also consider related
+Schwarzschild coordinates $(t,r,\phi,\theta)$, where $r$ is the
+circumferential radius.  However, we are free to consider
+reparametrizations in terms of functions of the Schwarzschild
+coordinates.  Here I consider the Schwarzschild solution in terms of
+$(t,u_1,\phi,\theta)$, where $u_1=u_1(r)$ is the straight-line proper
+length from a point to the event horizon.  This coordinate system has
+some interesting properties and affords new physical insight to black
+holes.  I present some diagrams using proper length radial coordinate
+instead of circumferential radius $r$.  I also consider related
 parameterizations $u_2$ and $u_3$ which preserve proper area or volume
@@
 % keywords can be removed
-\keywords{First keyword \and Second keyword \and More}
+\keywords{Black hole\and Proper distance \and Proper radial coordinate\and Schwarzschild metric}
 
 \section{Introduction}
-The Schwarzschild metric is usually written
+The Schwarzschild metric \citep{schwarzschild1916} is usually written
 

### 2022-06-19 arxiv/template.tex
refer to JOSE

@@
 
-The proper distance from a point at Schwarzschild coordinate $r$ would be
+The proper distance from a point at Schwarzschild coordinate $r$ would
+be
 
@@
 We can use $u_1$ as a coordinate on the same footing as
-circumferential radius $r$.  First we suppress $\theta,\phi$
-coordinates and consider only $r,t$; see Figures~\ref{schwarzschild},
-\ref{eddington}, and ~\ref{gullstrand}.
+circumferential radius $r$ and present a number of diagrams taken
+from~\citep{hankin2020}.  Below, each figure has two parts: on the
+left, we see the standard figure using circumferential radius $r$, and
+on the right we see the same figure but with the horizontal axis
+showing $u_1$.  First we suppress $\theta,\phi$ coordinates and
+consider only $r,t$; see Figures~\ref{schwarzschild}, \ref{eddington},
+and ~\ref{gullstrand}.
 

### 2023-05-31 mercury_and_venus.Rmd
lightweight analysis of Mercury and Venus orbital radii

---
title: "Non-Euclidean space as revealed by proper distances from Mercury to Venus"
author: "Robin Hankin"
date: "2023-05-30"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Here I consider the difference between two ways to assess the distance
between Mercury and Venus.  We use units in which the Schwarzschild
radius is unity, converting to SI when needed.  Astronomically, the
correct values are

```{r}
M <- 1.989e30 # mass of Sun in kg
G <- 6.6743e-11 # G in SI
sol <- 299792458 # speed of light
(r <- 2*M*G/sol^2) # Schwarzschild radius of Sun in m
(mercury <- 57909050e3/r) # semimajor axis
(venus <- 108208000e3/r)
```

For simplicity, I will first consider Mercury and Venus to have an
orbital radius of a notional $2\times 10^7$ and $4\times 10^7$
respectively.  The Euclidean distance between the two is just $4\times
10^7-2\times 10^7=2\times 10^7$ but the proper distance is given by

$$
\int_{r=2\times 10^7}^{4\times 10^7}\frac{dr}{\sqrt{1-1/r}}
$$

Mathematically this is easy.  Mathematica gives

```
In[27]:= Assuming[x>0,Integrate[1/Sqrt[1-1/r],{r,2*10^7,4*10^7}]]                                                                                                                     
Out[27]= -2000 (Sqrt[99999995] - Sqrt[399999990]) - ArcSinh[Sqrt[19999999]] + ArcSinh[Sqrt[39999999]]
```

Numerically we see close agreement between Mathematica and R:

```
In[28]:= p = Assuming[x>0,Integrate[1/Sqrt[1-1/r],{r,2*10^7,4*10^7}]]                        
In[29]:= (p - 2*10^7)//N
Out[29]= 0.346574
```

```{r}
(v <- integrate(function(r){1/sqrt(1-1/r)},2e7,4e7))
v$value - 2e7
```

The discrepancy is in Schwarzschild radii.

```{r}
(v$value-2e7)*r
```

This gives us the discrepancy in meters.  We see a proper distance
between the two planets of about a kilometer in excess of the
Euclidean distance calculated on the basis of circumferential radius
and flat space.  It's not that much harder to use the astronomically
accurate values instead:

```{r}
diff <- (integrate(function(r){1/sqrt(1-1/r)},mercury,venus)$value - (venus-mercury))
diff   # discrepancy in Schwarzschild radius
diff*r  # in meters
```

Asymptotically we have

$$
\int_{r=w}^{wx}\frac{dr}{\sqrt{1-1/r}}=w(x-1)+\frac{\log(x)}{2} -w^{-1}\frac{3(1-x)}{8x}+{\mathcal O}\left(w^{-2}\right)
$$
(writing $w$ for Mercury's orbital radius).  Noting that $x\sim 1$, the first two terms are about

```{r}
x <- venus/mercury
log(x)/2
3*(1-x)/(8*mercury*x)
```

showing that the zeroth order term is sufficient here, and also close
agreement with the numerical integration using astronomical data given
above.

### 2023-11-28 mercury_and_venus.Rmd
underbrace

@@
 $$
-\int_{r=w}^{wx}\frac{dr}{\sqrt{1-1/r}}=w(x-1)+\frac{\log(x)}{2} -w^{-1}\frac{3(1-x)}{8x}+{\mathcal O}\left(w^{-2}\right)
+\int_{r=w}^{wx}\frac{dr}{\sqrt{1-1/r}}=\underbrace{w(x-1)}_{\mbox{Euclid}}+\frac{\log(x)}{2} -w^{-1}\frac{3(1-x)}{8x}+{\mathcal O}\left(w^{-2}\right)
 $$

### 2024-10-31 weighted_string.R
functionalisation

@@ -8,11 +8,11 @@ rdashdash <- function(r,rdash,lambda,M=1/2){
     )
 }
 
-string <- function(lambda){
+string <- function(r0,lambda=0,angles=seq(from = 0, to = 0.5, by = 0.01)){
 
     parameters <- c(lambda = lambda, M = 1/2)
-    state <- c(r = 2, rdash = 0)
-    times <- seq(from = 0, to = 0.5, by = 0.01)
+    state <- c(r = r0, rdash = 0)
+    times <- angles
 
     Lorenz <- function(t, state, parameters){
         with(as.list(c(state, parameters)), {
@@ -31,8 +31,11 @@ string <- function(lambda){
     return(cbind(r*cos(phi),r*sin(phi)))
 }
 
-plot(c(-3,3),c(-3,3),type="n",asp=1)
+size <- 5
+plot(c(-size,size),c(-size,size),type="n",asp=1)
 th <- seq(from=0,to=2*pi,len=100)
 points(cos(th),sin(th),type='l')  # event horizon
-points(string(0)) 
-
+points(string(r0=1.4,lambda= 0,angles=seq(0,2,0.01)),type='l')
+for(l in seq(from=-0.1,to=-1.4,by=-0.1)){
+    points(string(r0=1.4,lambda=l,angles=seq(0,pi/2,0.01)),type='l')
+}

### 2024-10-31 weighted_string.Rmd
use Rmd format here

---
title: "Heavy inextensible strings in the Schwarzschild geometry"
author: "Robin Hankin"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library("deSolve")
```

Now define the  second derivative from Euler's equation:

```{r definerdashdash}
rdashdash <- function(r,rdash,lambda,M=1/2){
    x <- sqrt(1-2*M/r)
    return(
    (M*(4+lambda*(4-3*x))*r^2 + (-1+lambda*(-1+x))*r^3 + M*(3 + lambda*(3-2*x))*rdash^2 +
     2*r*(M^2*(-2+lambda*(-2+x)) + (-1+lambda*(-1+x))*rdash^2))/((-1+lambda*(-1+x))*r*(-2*M+r))
    )
}
```

```{r useode}
string <- function(r0,lambda=0,angles=seq(from = 0, to = 0.5, by = 0.01)){

    parameters <- c(lambda = lambda, M = 1/2)
    state <- c(r = r0, rdash = 0)
    times <- angles

    Lorenz <- function(t, state, parameters){
        with(as.list(c(state, parameters)), {
            dr <- rdash
            drdash <- rdashdash(r, rdash, lambda, M)
            list(c(dr, drdash))
        })
    }

    out <- ode(y = state, times = times, func = Lorenz, parms = parameters)

    phi <- out[,1]
    r <- out[,2]

    return(cbind(r*cos(phi),r*sin(phi)))
}
```

```{r plotstrings}

size <- 5
plot(c(-size,size),c(-size,size),type="n",asp=1)
th <- seq(from=0,to=2*pi,len=100)
points(cos(th),sin(th),type='l')  # event horizon
points(string(r0=1.4,lambda= 0,angles=seq(0,2,0.01)),type='l')
for(l in seq(from=-0.1,to=-1.4,by=-0.1)){
    points(string(r0=1.4,lambda=l,angles=seq(0,pi/2,0.01)),type='l')
}
```

### 2024-10-31 weighted_string.Rmd
start documenting

@@
 library("deSolve")
+```
+
+The light inextensible string functional was
+
+$$
+\int_{p_1}^{p_2}\left(\frac{\left(r'\right)^2}{1-2M/r} + r^2\right)^\frac{1}{2}d\phi
+$$
+
+But now we use a Lagrange multiplier:
+
+$$
+\int_{p_1}^{p_2}\left[\left(\frac{\left(r'\right)^2}{1-2M/r} + r^2\right)^\frac{1}{2}
++\lambda\left(\frac{\left(r'\right)^2}{1-2M/r} + r^2\right)^\frac{1}{2}
+\left(1+\lambda(1-\sqrt{1-2M/r})\right)
+\right]d\phi
+$$
+
+that is, miminizing the potential energy subject to a (proper) length
+constraint.  Simplifying:
+
+$$
+\int_{p_1}^{p_2}\left(\frac{\left(r'\right)^2}{1-2M/r} +
+r^2\right)^\frac{1}{2} \left(1+\lambda(1-\sqrt{1-2M/r})\right) d\phi
+$$
+
+```
+EulerEquations[ 
+  Sqrt[r[\[Phi]]^2 + 
+     r'[\[Phi]]^2/(1 - 2*M/r[\[Phi]])]*(1 + \[Lambda]*(1 - 
+        Sqrt[1 - 2*M/r[\[Phi]]])), r[\[Phi]], \[Phi]] // FullSimplify
 ```

### 2024-10-31 weighted_string.Rmd
use approxfun

@@
 +\lambda\left(\frac{\left(r'\right)^2}{1-2M/r} + r^2\right)^\frac{1}{2}
-\left(1+\lambda(1-\sqrt{1-2M/r})\right)
+\left(1-\sqrt{1-2M/r}\right)
 \right]d\phi
@@
 
-size <- 5
-plot(c(-size,size),c(-size,size),type="n",asp=1)
+lambda <- seq(from = -0.05, to = -1.7, by = -0.1)
+f <- approxfun(
+    c(0,-0.1,-0.2,-0.5, -0.7, -1,  -1.4,     -1.6,-1.7),
+    c(2.4,2.4, 2.1,1.5,   1.5, 1.1,  0.63, 0.3,0)
+)
+
+plot(lambda,f(lambda))
+
+size <- 2
+par(pty="s")
+plot(NA,type="n",asp=1,xlim=c(-size,size),ylim=c(-size,size))
 th <- seq(from=0,to=2*pi,len=100)
 points(cos(th),sin(th),type='l')  # event horizon
-points(string(r0=1.4,lambda= 0,angles=seq(0,2,0.01)),type='l')
-for(l in seq(from=-0.1,to=-1.4,by=-0.1)){
-    points(string(r0=1.4,lambda=l,angles=seq(0,pi/2,0.01)),type='l')
+
+for(l in lambda){
+    points(string(r0=1.2,lambda=l,angles=seq(0,f(l),0.01)),type='l')
 }
+
+segments(1.2,0,4,0)
+docolour <- function(lambda,...){
+    points(string(r0=1.2,lambda=lambda,angles=seq(0,f(lambda),0.01)), type='l', lwd=3, ...)
+}
+docolour(-0.05,col='blue')
+docolour(-0.7,col='green')
+docolour(-1.4,col='red')
+points(1.2,0,pch=16)
+
 ```

### 2024-10-31 weighted_string.Rmd
more work

@@
 
-```{r plotstrings}
+```{r plotstringsonepointtwo}
 
+dist <- 1.2
 lambda <- seq(from = -0.05, to = -1.7, by = -0.1)
@@
 for(l in lambda){
-    points(string(r0=1.2,lambda=l,angles=seq(0,f(l),0.01)),type='l')
+    points(string(r0 = dist, lambda = l, angles = seq(0,f(l),0.01)), type = 'l')
 }
 
-segments(1.2,0,4,0)
+segments(dist,0,4,0)
 docolour <- function(lambda,...){
-    points(string(r0=1.2,lambda=lambda,angles=seq(0,f(lambda),0.01)), type='l', lwd=3, ...)
+    points(string(r0 = dist, lambda = lambda, angles=seq(0,f(lambda),0.01)), type='l', lwd=3, ...)
 }
-docolour(-0.05,col='blue')
-docolour(-0.7,col='green')
-docolour(-1.4,col='red')
+#docolour(-0.05,col='blue')
+#docolour(-0.7,col='green')
+#docolour(-1.4,col='red')
 points(1.2,0,pch=16)
+```
 
+```{r plotstringsonepointzerofour}
+
+dist <- 1.04
+lambda <- seq(from = 0, to = -1, by = -0.1)
+
+f <- approxfun(
+    c(0,-0.1,-0.2,-0.5, -0.7, -1,  -1.4,     -1.6,-1.7),
+    c(2.4,2.4, 2.1,1.5,   1.5, 0.5,  0.63, 0.3,0)
+)
+
+plot(lambda,f(lambda))
+
+size <- 2
+par(pty="s")
+plot(NA,type="n",asp=1,xlim=c(-size,size),ylim=c(-size,size))
+th <- seq(from=0,to=2*pi,len=100)
+points(cos(th),sin(th),type='l')  # event horizon
+
+for(l in lambda){
+    points(string(r0 = dist, lambda = l, angles = seq(0,f(l),0.01)), type = 'l')
+}
+
+segments(dist,0,4,0)
+docolour <- function(lambda,...){
+    points(string(r0 = dist, lambda = lambda, angles=seq(0,f(lambda),0.01)), type='l', lwd=3, ...)
+}
+docolour(0,col='red')
+docolour(-1,col='blue')
+points(dist,0,pch=16)
 ```

### 2024-11-01 weighted_string.Rmd
more work

@@
 
+Mathematica can deal with this.  The idiom is:
+
 ```
+Needs["VariationalMethods`"]
 EulerEquations[ 
@@
 
-Now define the  second derivative from Euler's equation:
+The above comments give a second-order ODE, which is analytically
+intractable.  However, we can proceeed by
+
+```
+Solve[%, r''[phi]] // FullSimplify
+```
+
+which gives an expression for $r''(\phi)$.  In R idiom, this is:
 
@@
     return(
-    (M*(4+lambda*(4-3*x))*r^2 + (-1+lambda*(-1+x))*r^3 + M*(3 + lambda*(3-2*x))*rdash^2 +
-     2*r*(M^2*(-2+lambda*(-2+x)) + (-1+lambda*(-1+x))*rdash^2))/((-1+lambda*(-1+x))*r*(-2*M+r))
+    (
+        M*(4+lambda*(4-3*x))*r^2 +
+        (-1+lambda*(-1+x))*r^3 +
+        M*(3 + lambda*(3-2*x))*rdash^2 +
+        2*r*(M^2*(-2+lambda*(-2+x)) + (-1+lambda*(-1+x))*rdash^2)
+    ) / (
+        (-1+lambda*(-1+x))*r*(-2*M+r)
+    )
     )
@@
     c(0,-0.1,-0.2,-0.5, -0.7, -1,  -1.4,     -1.6,-1.7),
-    c(2.4,2.4, 2.1,1.5,   1.5, 0.5,  0.63, 0.3,0)
+    c(2.4,2.4, 2.1,1.5,   1.5, 0.9,  0.63, 0.3,0)
 )

### 2024-11-01 weighted_string.Rmd
longer integration
tighter limits

@@
 dist <- 1.04
-lambda <- seq(from = 0, to = -1, by = -0.1)
+lambda <- seq(from = 0, to = -1.15, by = -0.1)
 
 f <- approxfun(
-    c(0,-0.1,-0.2,-0.5, -0.7, -1,  -1.4,     -1.6,-1.7),
-    c(2.4,2.4, 2.1,1.5,   1.5, 0.9,  0.63, 0.3,0)
+    c(0,-0.1,-0.2,-0.5, -0.7, -1,  -1.15, -1.1, -1.4,     -1.6,-1.7),
+    c(3.4,2.9, 2.7,1.9,   1.5, 0.9,  0.4,  0.6,   0.63,    0.3    ,0)
     )
@@
 docolour(-1,col='blue')
+docolour(-1.15,col='green')
 points(dist,0,pch=16)

### 2024-11-05 weighted_string.Rmd
mathematica screenshot

@@
 
-which gives an expression for $r''(\phi)$.  In R idiom, this is:
+which gives an expression for $r''(\phi)$.
+
+![Mathematica screenshot](mathematica_screenshot.png)
+
+In R idiom, this is:
 
@@
 rdashdash <- function(r,rdash,lambda,M=1/2){
-    x <- sqrt(1-2*M/r)
+    x <- sqrt(1 - 2*M/r)
     return(
     (
-        M*(4+lambda*(4-3*x))*r^2 +
-        (-1+lambda*(-1+x))*r^3 +
-        M*(3 + lambda*(3-2*x))*rdash^2 +
-        2*r*(M^2*(-2+lambda*(-2+x)) + (-1+lambda*(-1+x))*rdash^2)
+        M*(4 + lambda*(4 - 3*x))*r^2 +
+        (-1 + lambda*(-1 + x))*r^3 +
+        M*(3 + lambda*(3 - 2*x))*rdash^2 +
+        2*r*(M^2*(-2 + lambda*(-2 + x)) + (-1 + lambda*(-1 + x))*rdash^2)
     ) / (
-        (-1+lambda*(-1+x))*r*(-2*M+r)
+        (-1 + lambda*(-1 + x))*r*(-2*M + r)
     )
